### 2021-03-30 tex/astropy-paper-3.tex
initial check-in

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%  Style notes:
%
%   -
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[modern]{aastex631}
\usepackage{xspace}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{ulem}

% To allow putting figures in a subdir
\graphicspath{{figures/}}

\submitjournal{ApJ}

\shorttitle{Astropy Project III}
\shortauthors{The Astropy Collaboration}

\input{preamble.tex}

% Links to documentation: subpackages
\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}

% Links to documentation: classes
\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}

\begin{document}

\draft{\today}

\title{The Astropy Project: xx and status of the v5.0 core package}

\correspondingauthor{Astropy Coordination Committee}
\email{coordinators@astropy.org}

\author{Astropy Collaboration}
\noaffiliation
{\let\thefootnote\relax\footnote{{The author list has three parts: the authors that made significant contributions to the writing of the paper in order of contribution, the four members of the \astropy Project coordination committee, and contributors to the \astropy Project in alphabetical order. The position in the author list does not correspond to contributions to the \astropy Project as a whole. A more complete list of contributors to the core package can be found in the \href{https://github.com/astropy/astropy/graphs/contributors}{package repository}, and at the \href{http://www.astropy.org/team.html}{\astropy team webpage}.}}}

\input{author.tex}

\begin{abstract}

    To be written...

\end{abstract}

\keywords{%
    Astrophysics - Instrumentation and Methods for Astrophysics
    ---
    methods: data analysis
    ---
    methods: miscellaneous
}

\section{Introduction} \label{sec:intro}

To be written...

\acknowledgments

We would like to thank the members of the community that have contributed to
\astropy, that have opened issues and provided feedback, and have supported the
project in a number of different ways.

The \astropy community is supported by and makes use
of a number of organizations and services outside the traditional
academic community. We thank Google for financing and organizing the
Google Summer of Code (GSoC) program, that has funded severals
students per year to work on \astropy related projects over the
summer. These students often turn into long-term contributors. We also
thank NumFOCUS and the Python Software Foundation for financial
support. Within the academic community, we thank
institutions that make it possible that astronomers and other developers on
their staff can contribute their time to the development of
\astropy projects.  We would like acknowledge the support of the
Space Telescope Science Institute, Harvard–Smithsonian Center for Astrophysics,
and the South African Astronomical Observatory.

Furthermore, the \astropy packages would not exist in their current form without
a number of web services for code hosting, continuous integration, and
documentation; in particular, \astropy heavily relies on GitHub, Travis CI,
Appveyor, CircleCI, and Read the Docs.

\astropypkg interfaces with the SIMBAD database, operated at CDS, Strasbourg,
France. It also makes use of the ERFA library \citep{erfa}, which in turn
derives from the IAU SOFA Collection\footnote{\url{http://www.iausofa.org}}
developed by the International Astronomical Union Standards of Fundamental
Astronomy \citep{sofa}.

\software{\package{astropy} (\citealt{astropy}),
          \package{numpy} (\citealt{numpy}),
          \package{scipy} (\citealt{scipy}),
          \package{matplotlib} (\citealt{matplotlib}),
          \package{Cython} (\citealt{cython}),
          \package{SOFA} (\citealt{sofa}),
          \package{ERFA} (\citealt{erfa})
          }

\bibliographystyle{aasjournal}
\bibliography{bibliography}

% \appendix

% \section{List of Affiliated Packages}

% \begin{longrotatetable}
%     \begin{deluxetable*}{cccp{3in}c}
%     \tablecaption{Registry of affiliated packages.}
%     \label{tab:registry}
%     \tablehead{
%         \colhead{Package Name} &
%         \colhead{Stable} &
%         \colhead{PyPI Name} &
%         \colhead{Maintainer} &
%         \colhead{Citation}
%       }
%       \startdata
%         \input{registry.tex}
%       \enddata
%   \end{deluxetable*}
% \end{longrotatetable}

\end{document}

### 2021-11-04 src/ms.tex
remove old latex build stuff and start switch to snakemake

--- 2021-03-30 tex/astropy-paper-3.tex
@@
 \noaffiliation
-{\let\thefootnote\relax\footnote{{The author list has three parts: the authors that made significant contributions to the writing of the paper in order of contribution, the four members of the \astropy Project coordination committee, and contributors to the \astropy Project in alphabetical order.
-The position in the author list does not correspond to contributions to the \astropy Project as a whole.
-A more complete list of contributors to the core package can be found in the \href{https://github.com/astropy/astropy/graphs/contributors}{package repository}, and at the \href{http://www.astropy.org/team.html}{\astropy team webpage}.}}}
+{\let\thefootnote\relax\footnote{{The author list has two parts: the authors that made significant contributions to the writing of the paper in order of contribution, followed by contributors to the \astropy Project in alphabetical order.
+\textbf{The position in the author list does not correspond to contributions to the \astropy Project as a whole.} A more complete list of contributors to the core package can be found in the \href{https://github.com/astropy/astropy/graphs/contributors}{package repository}, and at the \href{http://www.astropy.org/team.html}{\astropy team webpage}.}}}
 
-\input{author.tex}
+% \input{author.tex}
 
@@
 
-\acknowledgments
+\section{Placeholder} \label{sec:placeholder}
+
+\begin{figure}
+    \begin{centering}
+        \includegraphics{figures/contributor-summary.pdf}
+        \caption{Placeholder figure!}
+        \label{fig:contributor-summary}
+    \end{centering}
+\end{figure}
+
+\begin{acknowledgments}
 
@@
 Astronomy \citep{sofa}.
+
+\end{acknowledgments}
 

### 2021-11-16 src/ms.tex
add outline of paper from google doc

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%  Style notes:
 %
-%   -
+%   Please read the CONTRIBUTING.md file in this repository for notes about
+%   style and latex conventions!
 %
@@
 
-\title{The Astropy Project: xx and status of the v5.0 core package}
+\title{The Astropy Project: xx}
 
@@
 
-\section{Placeholder} \label{sec:placeholder}
+\section{Major Updates in the Astropy Project} \label{sec:project-updates}
+
+\subsection{Project governance} \label{sec:project-governance}
+
+New CoCo and election overview.
+
+\subsection{Contributor base} \label{sec:project-contributors}
+
+Overview and statistics of contributors. Changes since v2.0.
 
@@
 \end{figure}
+
+\subsection{Funding} \label{sec:project-funding}
+
+New funding sources.
+
+\section{Major Updates to the Astropy Core Package} \label{sec:core-updates}
+
+\subsection{New Long-term Support (LTS) Version: v5.0} \label{sec:core-v50}
+
+\subsection{Highlighted Feature Development} \label{sec:core-features}
+
+Timeseries, Box Least Squares
+
+TODO: go through What's New pages and identify a few big additions / changes.
+
+\section{Learn Astropy} \label{sec:learn}
+
+Current status and scope
+
+Vision for the future (review what we said in v2.0 paper)
+
+User forums and engagement with user base
+
+\section{Supporting the Ecosystem of Astronomical Python Software}
+\label{sec:ecosystem}
+
+\subsection{Community-oriented infrastructure}
+
+Infrastructure packages that exist to support and help others maintain their package infrastructure.
+
+\subsection{Affiliated packages}
+
+Updates!
+
+\subsection{Connections with observatories}
+
+JWST (motivated by excitement around the launch?)
+
+Gemini
+
+\section{Future Plans for the Astropy Project} \label{sec:future}
+
+\subsection{Roadmap}
+
+\subsection{Challenges}
+
+Attracting new contributors when the code has become quite complex,
+Contributor to maintainer mentoring,
+Long-term / sustained funding for maintaining infrastructure,
+...
 

### 2021-11-30 src/ms.tex
add new figure

@@
 
-To be written...
+Over the last decade, \python has grown enormously in popularity and has become
+a dominant programming language in the astronomical community.
+
+\begin{figure}
+    \begin{centering}
+        \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
+        \caption{
+            Yearly full-text mentions of programming languages (indicated in the
+            figure legend) in refereed publications in the astronomical
+            literature database in the Astrophysics Data System (ADS;
+            \citealt{ads}).
+            \python has rapidly become the dominant programming language
+            mentioned in refereed articles over the last 10 years.
+        }
+        \label{fig:python-mentions}
+    \end{centering}
+\end{figure}
 
@@
 
-Timeseries, Box Least Squares
-
-TODO: go through What's New pages and identify a few big additions / changes.
+\begin{itemize}
+    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+    \item TODO: something about WCS, SpectralCoord?
+\end{itemize}
 
@@
 
-Updates!
+New packages and major updates!
 

### 2021-11-30 src/figures/python-mentions.py

@@ -1,0 +1,136 @@
+"""
+TODO: describe
+"""
+
+# Standard library
+import datetime
+import pathlib
+
+# Third-party
+import ads
+import astropy.table as at
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+
+
+def get_papers(query, rows_per_page=100, max_pages=100):
+    """
+    Parameters
+    ----------
+    query : str
+    rows_per_page : int (optional)
+    max_pages : int (optional)
+
+    Returns
+    -------
+    search_query : `ads.SearchQuery`
+    papers_df : `pandas.DataFrame`
+    """
+    q = ads.SearchQuery(q=query,
+                        sort="date",
+                        max_pages=max_pages,
+                        rows=rows_per_page,
+                        fl=["identifier", "title", "author", "year",
+                            "pubdate", "pub", "citation_count"])
+
+    all_dicts = []
+    for paper in q:
+        # Get arxiv ID
+        aid = [":".join(t.split(":")[1:]) for t in paper.identifier
+               if t.startswith("arXiv:")]
+
+        all_dicts.append(dict(
+            authors='; '.join(paper.author) if paper.author else '',
+            year=int(paper.year) if paper.year is not None else -1,
+            pubdate=paper.pubdate if paper.pubdate is not None else '',
+            title=paper.title[0] if paper.title is not None else '',
+            pub=paper.pub if paper.pub is not None else '',
+            arxiv=aid[0] if len(aid) else '',
+            citations=(paper.citation_count
+                       if paper.citation_count is not None else 0)
+        ))
+
+    papers = at.Table(sorted(all_dicts, key=lambda x: x['pubdate'],
+                             reverse=True))
+
+    # Convert list of papers to pandas DataFrame
+    df = papers.to_pandas()
+    df['pubdate'] = np.array([
+        datetime.datetime(
+            int(x.split('-')[0]),  # year
+            max(1, int(x.split('-')[1])),  # month
+            1  # first day of the month
+        )
+        for x in df['pubdate']
+    ])
+    df.index = df['pubdate']
+    df = df.sort_index()
+
+    return q, df
+
+
+def get_dfs(cache_path):
+    cache_path = pathlib.Path(cache_path)
+    queries = {
+        'Python': (
+            'full:"python" property:"refereed" collection:"astronomy"'),
+        'IDL': (
[…]

### 2021-12-01 src/ms.tex
some rough words in the intro

@@
 
-Over the last decade, \python has grown enormously in popularity and has become
-a dominant programming language in the astronomical community.
+Over the last decade, \python has grown enormously in popularity to become a
+dominant programming language in the astronomical (and broader) scientific
+community.
+For example, Figure~\ref{fig:python-mentions} shows the number of yearly
+full-text mentions of \python as compared to a few other programming languages
+in refereed articles in the astronomical literature, demonstrating its nearly
+exponential growth in popularity.
+The rapid adoption of \python by astronomy researchers, students, observatories,
+and technical staff combined with an associated increase in awareness and
+interest about open-source software tools is leading a paradigm shift in the way
+research is done, data is analyzed, and results are shared.
+
+TODO:  https://physicsworld.com/a/standing-on-the-shoulders-of-programmers/
+
+A major reason that Python has been adopted in the sciences is the huge amount
+of community-oriented, open-source software tools that are available to
+anyone...TODO.
+TODO: Numpy, scipy, matplotlib, scientific software ecosystem.
+In astronomy, one of the first large, community-oriented, open-source \python packages ... was Astropy ... (note: yt)
+TODO: words about astropy core package.
+TODO: The core package is now largely stable and provides core foundational tools.
+TODO: Astropy has expanded to become the Astropy Project: Broader scope than core package. Core package, Ecosystem of tools, Community of users, developers, maintainers
+
+Transition of needs: From new features, to community development, sustainability, ...
+- pipeline between User, Participant, Code contributor, Maintainer, Coordinator.
+
+Funding received:
+- Moore
+- NASA
 

### 2021-12-01 src/figures/python-mentions.py

@@ -102,7 +102,7 @@ def get_dfs(cache_path):
 def plot_yearly_mentions(paper_dfs):
     fig, ax = plt.subplots(figsize=(8, 6), constrained_layout=True)
 
-    for name, df in paper_dfs.items():
+    for i, (name, df) in enumerate(paper_dfs.items()):
         g = df.groupby(by=[df.index.year]).count()
         group_dates = np.array([datetime.date(x, 1, 1) for x in g.index])
         mentions = g['pubdate.1'].values
@@ -115,7 +115,7 @@ def plot_yearly_mentions(paper_dfs):
         ax.plot(group_dates,
                 mentions,
                 marker='', drawstyle='steps-mid',
-                lw=2, label=name)
+                lw=2, label=name, zorder=-i)
 
     ax.set_xlim(datetime.date(1991, 8, 1),
                 datetime.date(2021, 6, 1))  # 2021.5

### 2021-12-01 src/ms.tex
intro first bits

@@
 
+The \python programming language is a high-level, interpreted (versus compiled)
+programming language that has become an industry standard across many
+computational domains, technological sectors, and fields of research.
+Despite claims to the contrary \citep{Portegies-Zwart:2020}, \python enables
+scalable, time- and energy-efficient code execution \citep[e.g.,][]{Augier:2021}
+with a focus on code readability, ease of use, and interoperability with other
+languages.
 Over the last decade, \python has grown enormously in popularity to become a
-dominant programming language in the astronomical (and broader) scientific
-community.
+dominant programming language in the astronomical and broader scientific
+communities.
 For example, Figure~\ref{fig:python-mentions} shows the number of yearly
@@
 and technical staff combined with an associated increase in awareness and
-interest about open-source software tools is leading a paradigm shift in the way
-research is done, data is analyzed, and results are shared.
+interest about open-source software tools is contributing to a paradigm shift in
+the way research is done, data is analyzed, and results are shared in astronomy
+and beyond.
+
+One of the factors that has led to its rapid ascent in popularity in scientific
+contexts has been significant volunteer-driven effort behind developing
+community-oriented open-source software tools and fostering communities of users
+and developers that have grown around these packages.
+TODO: Numpy, scipy, matplotlib, scientific software ecosystem.
+The astropy core package ... one of the first large, community-oriented, open-source \python packages ... was Astropy ... (note: yt)
+TODO: words about astropy core package.
+TODO: The core package is now largely stable and provides core foundational tools.
+
+TODO: Astropy has expanded to become the Astropy Project: Broader scope than core package. Core package, Ecosystem of tools, Community of users, developers, maintainers
 
 TODO:  https://physicsworld.com/a/standing-on-the-shoulders-of-programmers/
-
-A major reason that Python has been adopted in the sciences is the huge amount
-of community-oriented, open-source software tools that are available to
-anyone...TODO.
-TODO: Numpy, scipy, matplotlib, scientific software ecosystem.
-In astronomy, one of the first large, community-oriented, open-source \python packages ... was Astropy ... (note: yt)
-TODO: words about astropy core package.
-TODO: The core package is now largely stable and provides core foundational tools.
-TODO: Astropy has expanded to become the Astropy Project: Broader scope than core package. Core package, Ecosystem of tools, Community of users, developers, maintainers
 

### 2021-12-01 src/ms.tex
more intro words

@@
 One of the factors that has led to its rapid ascent in popularity in scientific
-contexts has been significant volunteer-driven effort behind developing
+contexts has been the significant, volunteer-driven effort behind developing
 community-oriented open-source software tools and fostering communities of users
-and developers that have grown around these packages.
-TODO: Numpy, scipy, matplotlib, scientific software ecosystem.
-The astropy core package ... one of the first large, community-oriented, open-source \python packages ... was Astropy ... (note: yt)
-TODO: words about astropy core package.
-TODO: The core package is now largely stable and provides core foundational tools.
+and developers that have grown around these efforts.
+Today, a broad and feature-diverse ``ecosystem'' of packages exists in the
+\python scientific computing landscape: Roughly ordered from general-use to
+domain-specific, this landscape now includes packages that provide core
+numerical analysis functionality like \package{numpy} \citep{todo} and
+\package{scipy} \citep{todo}, to visualization frameworks like
+\package{matplotlib} \citep{todo}, to machine learning and data analysis
+packages like \package{tensorflow} \citep{todo}, \package{pymc3} \citep{todo},
+and \package{emcee} \citep{todo}, to specialized, domain-specific packages like
+\package{yt} \citep{todo}, \package{plasmapy} \citep{todo}, \package{sunpy}
+\citep{todo}, \package{Biopython} \citep{todo}, and \package{sympy}
+\citep{sympy} (to name a few in each category).
+The \astropypkg \citep{astropy:2013, astropy:2018} core package began as an
+effort to consolidate the development of commonly-used functionality needed to
+perform astronomical research into a community-developed \python package.
 
-TODO: Astropy has expanded to become the Astropy Project: Broader scope than core package. Core package, Ecosystem of tools, Community of users, developers, maintainers
+The \astropypkg core package was one of the first large, open-source \python
+packages developed for astronomy and provides, among other things, software
+functionality for reading and writing astronomy-specific data formats (e.g.,
+FITS), transforming and representing astronomical coordinates, and representing
+and propagating physical units in code.
+An early description of the core functionality in \astropypkg can be found in
+the first Astropy paper \citep{astropy:2013} or in detail in the core package
+documentation.\footnote{\url{https://docs.astropy.org/}}
+The \astropypkg core package is now largely stable in that the software
+interface does not change without sufficient and significant motivation, and the
+addition of new features into the core package has slowed significantly as
+compared to the first years of its development.
+This is largely driven by the fact that the name Astropy has grown in scope to
+become ``the Astropy Project'': A name used to refer to the union of the
+\astropypkg core package, the ecosystem of astronomy-specific software tools
+that are interoperable with \astropypkg, \emph{and} the community of users,
+developers, and maintainers that participate in Astropy efforts.
 
-TODO:  https://physicsworld.com/a/standing-on-the-shoulders-of-programmers/
+% TODO:  https://physicsworld.com/a/standing-on-the-shoulders-of-programmers/
 

### 2021-12-02 src/ms.tex
title attempt
some minor tweaks

@@
 
-\title{The Astropy Project: xx}
+\title{The Astropy Project: \\
+       Sustaining and Growing a Community-oriented Open-source Project and
+       the Latest Major Release (v5.0) of the Core Package}
 
@@
 
-The \python programming language is a high-level, interpreted (versus compiled)
-programming language that has become an industry standard across many
+The \python programming language is a high-level, interpreted (as opposed to
+compiled) programming language that has become an industry standard across many
 computational domains, technological sectors, and fields of research.
@@
 numerical analysis functionality like \package{numpy} \citep{todo} and
-\package{scipy} \citep{todo}, to visualization frameworks like
-\package{matplotlib} \citep{todo}, to machine learning and data analysis
-packages like \package{tensorflow} \citep{todo}, \package{pymc3} \citep{todo},
-and \package{emcee} \citep{todo}, to specialized, domain-specific packages like
-\package{yt} \citep{todo}, \package{plasmapy} \citep{todo}, \package{sunpy}
-\citep{todo}, \package{Biopython} \citep{todo}, and \package{sympy}
-\citep{sympy} (to name a few in each category).
-The \astropypkg \citep{astropy:2013, astropy:2018} core package began as an
-effort to consolidate the development of commonly-used functionality needed to
-perform astronomical research into a community-developed \python package.
+\package{scipy} \citep{todo}, visualization frameworks like \package{matplotlib}
+\citep{todo}, machine learning and data analysis packages like
+\package{tensorflow} \citep{todo}, \package{pymc3} \citep{todo}, and
+\package{emcee} \citep{todo}, domain-specific libraries like \package{yt}
+\citep{todo}, \package{plasmapy} \citep{todo}, \package{sunpy} \citep{todo},
+\package{Biopython} \citep{todo}, and \package{sympy} \citep{sympy} (to name a
+few in each category).
+The \astropypkg \citep{astropy:2013, astropy:2018} core package began in this
+vein, as an effort to consolidate the development of commonly-used functionality
+needed to perform astronomical research into a community-developed \python
+package.
 

### 2021-12-02 src/ms.tex
intro end

@@
 compared to the first years of its development.
-This is largely driven by the fact that the name Astropy has grown in scope to
-become ``the Astropy Project'': A name used to refer to the union of the
+This is largely driven by the fact that the core package now represents just one
+piece of the broader astronomy \python context, and much new feature development
+is now happening in more specialized packages that are expanding the
+capabilities of the Astropy ecosystem by building on top of the foundations laid
+by the \astropypkg core package.
+Because of this natural expansion, the name Astropy has grown in scope beyond a
+single \python library to become ``the Astropy Project.''
+
+The Astropy Project is a community effort that represents the union of the
 \astropypkg core package, the ecosystem of astronomy-specific software tools
-that are interoperable with \astropypkg, \emph{and} the community of users,
-developers, and maintainers that participate in Astropy efforts.
+that are interoperable with \astropypkg (Astropy Affiliated Packages),
+\emph{and} the community of users, developers, and maintainers that participate
+in Astropy efforts.
+However, there is no institution responsible for managing the Astropy Project,
+for funding or maintaining its development, or sustaining it into the future:
+The Project is maintained and coordinated largely by volunteers.
+While new Astropy-affiliated packages are being developed that expand upon the
+core functionality in the \astropypkg package, representing a natural expansion
+of the Astropy Project ecosystem, the needs of and challenges faced by the
+Project are evolving.
+In particular, the transition from focusing our energy on development and
+maintenance of a single core package, to instead sustaining the core package and
+fostering the development of the community and its expansion has been a key
+issue faced by the Astropy Project in the last several years.
+
+In this Article, we briefly describe recent key updates in the \astropypkg core
+package since the last Astropy paper (``Paper II''; \citealt{astropy:2018}),
+major updates in the governance, contributor base, and funding of the Project,
+and discuss some of the future plans and challenges faced by the Astropy
+Project.
 
@@
 
-Transition of needs: From new features, to community development, sustainability, ...
-- pipeline between User, Participant, Code contributor, Maintainer, Coordinator.
+% Transition of needs: From new features, to community development, sustainability, ...
+% - pipeline between User, Participant, Code contributor, Maintainer, Coordinator.
 
-Funding received:
-- Moore
-- NASA
+% Funding received:
+% - Moore
+% - NASA
 

### 2021-12-02 src/figures/python-mentions.py
if running on a pull request, no ads token will be found, so make a blank plot

@@ -117,6 +117,14 @@ def plot_yearly_mentions(paper_dfs):
                 marker='', drawstyle='steps-mid',
                 lw=2, label=name, zorder=-i)
 
+    if len(paper_dfs) == 0:
+        # ADS key not found - pull request?
+        ax.set_title(
+            "ADS key not found! If this is a pull request, that is normal: "
+            "ADS queries are not performed.",
+            fontsize=14
+        )
+
     ax.set_xlim(datetime.date(1991, 8, 1),
                 datetime.date(2021, 6, 1))  # 2021.5
 
@@ -132,5 +140,9 @@ if __name__ == "__main__":
     cache_path = pathlib.Path('cache/prog-lang-fulltext')
     cache_path.mkdir(exist_ok=True, parents=True)
 
-    paper_dfs = get_dfs(cache_path)
+    try:
+        paper_dfs = get_dfs(cache_path)
+    except ads.exceptions.APIResponseError:
+        paper_dfs = {}
+
     plot_yearly_mentions(paper_dfs)

### 2021-12-02 src/ms.tex
reorg
move comments down
fix references

@@
 domain-specific, this landscape now includes packages that provide core
-numerical analysis functionality like \package{numpy} \citep{todo} and
-\package{scipy} \citep{todo}, visualization frameworks like \package{matplotlib}
-\citep{todo}, machine learning and data analysis packages like
-\package{tensorflow} \citep{todo}, \package{pymc3} \citep{todo}, and
-\package{emcee} \citep{todo}, domain-specific libraries like \package{yt}
-\citep{todo}, \package{plasmapy} \citep{todo}, \package{sunpy} \citep{todo},
-\package{Biopython} \citep{todo}, and \package{sympy} \citep{sympy} (to name a
-few in each category).
+numerical analysis functionality like \package{numpy} \citep{numpy:nature} and
+\package{scipy} \citep{scipy}, visualization frameworks like
+\package{matplotlib} \citep{matplotlib}, machine learning and data analysis
+packages like \package{tensorflow} \citep{tensorflow}, \package{pymc3}
+\citep{Salvatier:2016}, and \package{emcee} \citep{emcee}, domain-specific
+libraries like \package{yt} \citep{yt:2011}, \package{plasmapy}
+\citep{plasmapy}, \package{sunpy} \citep{sunpy:apj}, \package{Biopython}
+\citep{biopython}, and \package{sympy} \citep{sympy} (to name a few in each
+category).
 The \astropypkg \citep{astropy:2013, astropy:2018} core package began in this
@@
 
-% TODO:  https://physicsworld.com/a/standing-on-the-shoulders-of-programmers/
-
-% Transition of needs: From new features, to community development, sustainability, ...
-% - pipeline between User, Participant, Code contributor, Maintainer, Coordinator.
-
-% Funding received:
-% - Moore
-% - NASA
-
 \begin{figure}
@@
 \end{figure}
+
+\section{Major Updates to the Astropy Core Package} \label{sec:core-updates}
+
+\subsection{New Long-term Support (LTS) Version: v5.0} \label{sec:core-v50}
+
+\subsection{Highlighted Feature Development} \label{sec:core-features}
+
+\begin{itemize}
+    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+    \item TODO: something about WCS, SpectralCoord?
+\end{itemize}
 
@@
 
+% Transition of needs: From new features, to community development, sustainability, ...
+% - pipeline between User, Participant, Code contributor, Maintainer, Coordinator.
+
 \subsection{Funding} \label{sec:project-funding}
@@
 New funding sources.
-
-\section{Major Updates to the Astropy Core Package} \label{sec:core-updates}
-
-\subsection{New Long-term Support (LTS) Version: v5.0} \label{sec:core-v50}
-
-\subsection{Highlighted Feature Development} \label{sec:core-features}
-
-\begin{itemize}
-    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-    \item TODO: something about WCS, SpectralCoord?
-\end{itemize}
-
-\section{Learn Astropy} \label{sec:learn}
-
-Current status and scope
-
-Vision for the future (review what we said in v2.0 paper)
-
-User forums and engagement with user base
+% Funding received:
+% - Moore
+% - NASA
 
@@
 \subsection{Roadmap}
+
+\section{Learn Astropy} \label{sec:learn}
+
+Current status and scope
+
+Vision for the future (review what we said in v2.0 paper)
+
+User forums and engagement with user base
 
@@
 
-\software{\package{astropy} (\citealt{astropy}),
-          \package{numpy} (\citealt{numpy}),
-          \package{scipy} (\citealt{scipy}),
-          \package{matplotlib} (\citealt{matplotlib}),
-          \package{Cython} (\citealt{cython}),
-          \package{SOFA} (\citealt{sofa}),
-          \package{ERFA} (\citealt{erfa})
+\software{\package{astropy} \citep{astropy:2013, astropy:2018},
+          \package{numpy} \citep{numpy:nature},
+          \package{scipy} \citep{scipy},
+          \package{matplotlib} \citep{matplotlib},
+          \package{Cython} \citep{cython}.
+        %   \package{SOFA} \citep{sofa},
+        %   \package{ERFA} (\citealt{erfa})
           }
@@
 \bibliographystyle{aasjournal}
-\bibliography{bibliography}
+\bibliography{refs}
 

### 2021-12-17 src/ms.tex
proposed section authors

@@
 \newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}
+
+\newcommand{\secauthor}[1]{{\color{blue}Author:~\textit{#1}}}
+\newcommand{\secunfilled}{{\color{red}Author:~\textit{Looking for volunteers!}}}
 
@@
 
+\secauthor{Tom Robitaille}
+
+Summarize a few notable things in v5.0 as compared to the last LTS release.
+
 \subsection{Highlighted Feature Development} \label{sec:core-features}
 
+\secauthor{Nathaniel Starkman, Marten van Kerkwijk}
+
+Summarize major pieces of development since the last paper (since v2.0). A few
+ideas (taken from What's New pages) below, but feel free to expand on or remove
+from this list:
 \begin{itemize}
@@
 
-New CoCo and election overview.
+\secauthor{Erik Tollerud}
+
+Briefly summarize new procedures and governance structure, new CoCo, election
+process overview, etc.
 
@@
 
-Overview and statistics of contributors. Changes since v2.0.
+\secunfilled
+
+Overview and statistics of contributors. Highlight changes since v2.0.
 
@@
 
+\subsection{Inclusion, Diversity, and Equity Programs} \label{sec:project-ide}
+
+\secauthor{Lía Corrales}
+
 \subsection{Funding} \label{sec:project-funding}
 
-New funding sources.
-% Funding received:
-% - Moore
-% - NASA
+\secauthor{Aarya Patil}
+
+Summarize funding sources (Moore, NASA) and amounts and what this has been used
+for.
 
@@
 
-Infrastructure packages that exist to support and help others maintain their package infrastructure.
+\secunfilled
+
+Infrastructure packages that exist to support and help others maintain their
+package infrastructure.
 
@@
 
-New packages and major updates!
+\secauthor{Matt Craig, Brett Morris}
 
-\subsection{Connections with observatories}
+Highlight a few new affiliated packages and major updates to existing ones.
+Include a big table of all affiliated packages and references (as in v2.0
+paper).
 
-JWST (motivated by excitement around the launch?)
+\subsection{Connections with data archives}
+
+\secauthor{Adam Ginsburg}
+
+Summary of Astroquery and engagement with archives. Cite astroquery paper.
+
+\subsection{Connections with Observatories and Missions}
+
+Brief summary of efforts in observatory- or mission-driven development that have
+contributed to Astropy, and vice versa.
+
+JWST
+\secunfilled
 
 Gemini
+\secunfilled
+
+Cherenkov Telescope Array
+\secauthor{Axel Donath, Max Noethe}
+
+Rubin Observatory
+\secunfilled
 
@@
 
+\secauthor{Clara Brasseur}
+
+Overview of roadmap and some highlights.
+
 \section{Learn Astropy} \label{sec:learn}
 
-Current status and scope
+\secauthor{Lía Corrales, David Shupe + Learn team}
 
-Vision for the future (review what we said in v2.0 paper)
+\begin{itemize}
+\item Current status and scope
+\item Vision for the future (review what we said in v2.0 paper)
+\item User forums and engagement with user base
+\item Workshops
+\end{itemize}
 
-User forums and engagement with user base
+\subsection{Current and Future Challenges}
 
-\subsection{Challenges}
+\secunfilled
 
-Attracting new contributors when the code has become quite complex,
-Contributor to maintainer mentoring,
-Long-term / sustained funding for maintaining infrastructure,
-...
+\begin{itemize}
+\item Attracting new contributors when the code has become quite complex,
+\item Contributor to maintainer mentoring,
+\item Long-term / sustained funding for maintaining infrastructure,
+\item ...
+\end{itemize}
 

### 2022-01-04 src/ms.tex
Add Leo to LIGO et al
Co-authored-by: Leo Singer <leo.p.singer@nasa.gov>
add CRT at the end of the document
Update ms.tex
remove example
latex en dash

@@
 JWST
-\secunfilled
+\secauthor{Larry Bradley}
 
@@
 
+LIGO/Virgo/KAGRA
+\secauthor{Leo Singer}
 \section{Future Plans for the Astropy Project} \label{sec:future}
@@
 \astropy projects.  We would like acknowledge the support of the
-Space Telescope Science Institute, Harvard–Smithsonian Center for Astrophysics,
+Space Telescope Science Institute, Harvard--Smithsonian Center for Astrophysics,
 and the South African Astronomical Observatory.

### 2022-01-06 src/ms.tex
add more context to community infrastructure section
Add section author for community-oriented infrastructure
First draft of abstract

@@
 \begin{abstract}
-
-    To be written...
-
+The \astropy Project supports and fosters the development of open-source and openly-developed
+\python packages that provide commonly-needed functionality to the astronomical
+community.
+A key element of the \astropy Project is the core package \astropypkg, which serves as the
+foundation for more specialized projects and packages.
+In this article, we summarize key features in the core package as of the recent major
+release, version 5.0, and provide major updates for the project.
+We then discuss supporting a broader ecosystem of inter-operable packages,
+including connections with several observatories and missions.
+We also revisit the future outlook of the \astropy Project and the current
+status of Learn Astropy. Last but not least, we contemplate the current
+and future challenges facing the project.
 \end{abstract}
@@
 \section{Introduction} \label{sec:intro}
+
+\secauthor{Adrian Price-Whelan}
 
@@
 
-\secunfilled
+\secauthor{Adrian Price-Whelan}
 
@@
 
-\secunfilled
+\secauthor{Nicholas Earl}
 
@@
 package infrastructure.
+
+This includes an overview of the sphinx-astropy tools, the pytest extensions
+maintained by astropy, and the package template.
 

### 2022-01-18 src/ms.tex
added governance section

--- 2022-01-04 src/ms.tex
@@
 
-Briefly summarize new procedures and governance structure, new CoCo, election
-process overview, etc.
+%Briefly summarize new procedures and governance structure, new CoCo, election
+%process overview, etc.
+
+As part of the process of developing Astropy into a long-term sustainable product, and to improve transparency and accountability, the Project agreed to write down and formalize our governance structure (partly supported by explicit funding for this purpose - see \S \ref{sec:project-funding}).
+At the 2019 coordination meeting input was gathered from participants on what governance structures existed in the associated Open Source Software communities, and what would fit well with the needs of Astropy.
+This led into a ``retreat'' planned for March 2020, but due to the COVID-19 pandemic, this became a serious of virtual meetings of the ``Astropy Governance Working Group''.
+This group drafted the APE0 document \citep{ape0}, which was then eventually ratified and implemented by the ``Astropy Governance Implementation Working Group'' in Fall 2021.
+While the process emphasized flexibility and the ability to adapt to changing circumstances, it is expected that this is the framework Astropy's governance will operate in for at least the medium-term future.
+
+The APE0 \citep{ape0} document lays out the principles of this governance structure, so we refer the reader to that document for a more thorough description.
+However here we highlight some key elements.
+While many of these principles were already de facto true or have been discussed organically (and have been discussed in earlier papers in this series), the APE0-based governance aimed to provide a single place where the community can agree as a starting point.
+With this in mind, it highlights the developer and user community of the Astropy Project as the ultimate sources of authority, as well as the core principle of ``do-ocracy'' that those who do work for the Project (be it coding, training, or other less concrete contributions) gain more influence on the outputs of the Project by virtue of their effort.
+APE0 adds, however, the concept of ``voting members'' - a self-governed part of that community who are entrusted to elect the Coordination Committee.
+While this committee has existed from the inception of the project, APE0 establishes a formal voting process for this committee, and explicitly outlines the rights and responsibilities of the Coordination Committee.
+This role is mainly to facilitate consensus and act as the decision maker when other mechanisms have failed.
+However, it also includes powers that either require central authority or secrets (e.g.
+passwords), but AOE0 also charges the Committee to devolve responsibilities and seek community input on these items as often as possible.
+
+The first coordinating committee election under these rules took place in fall 2021, electing a mix of prior existing and new coordination committee members, and was contested in the sense of more candidates than available slots.
+This suggests the process is already working to serve the long-term interests of the committee to both spread the coordination effort, and to ensure it is not dominated by the same people for as long as the Project continues.
+While other, more fine-grained governance improvements are planned for the future, it is clear the foundation is now in place.
 

### 2022-01-19 src/ms.tex

--- 2022-01-06 src/ms.tex
@@
 We then discuss supporting a broader ecosystem of inter-operable packages,
-including connections with several observatories and missions.
+including connections with several astronomical observatories and missions.
 We also revisit the future outlook of the \astropy Project and the current
-status of Learn Astropy. Last but not least, we contemplate the current
-and future challenges facing the project.
+status of Learn Astropy. 
+We conclude by raising and discussing the current and future challenges facing the project.
 \end{abstract}

### 2022-01-20 src/ms.tex
add initial learn text

@@
 
+\subsubsection{Current status and scope}
+
+The Astropy-ii paper discussed the {\it Learn Astropy} effort only in the
+context of online tutorials and guides. This vision has been realized as
+{\it Learn Astropy} has released a new version of this website, and has
+expanded its efforts
+in user engagement and in conducting workshops. The Learn team meets
+weekly to drive these issues forward.
+
 \begin{itemize}
-\item Current status and scope
-\item Vision for the future (review what we said in v2.0 paper)
-\item User forums and engagement with user base
-\item Workshops
+
+\item {\bf Learn Astropy website:} The site has a new search interface powered by
+Algolia which is used by major shopping sites. This work has been
+supported in part by the Dunlap grant. Tutorials have been added and
+a {\it ccdproc} Jupyter book has been completed.
+
+\item {\bf User forums:} In addition to the pre-existing Astropy mailing list,
+and the Python in Astronomy Facebook group, users have been invited to the
+Astropy Slack workspace. The Moore Foundation grant allows this Slack space to
+be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+open-source projets. There is some internal debate amongst Astropy developers
+on whether Astropy Slack should be reserved for developer discussions, with
+users redirected to another site. Accordingly, the Astropy Project has commissioned
+a Discourse site which is more open than the Facebook group and more user-oriented
+than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+developers are able to invoice as independent contractors the time they
+spend helping users on these forums.
+
+\item {\bf Workshops:} the Project has been conducting workshops at winter meetings
+of the American Astronomical Society since AAS 225 in January 2015. Up to the
+start of the coronavirus pandemic, these were full-day in-person workshops with
+as many as 90 participants and a dozen facilitators from the project. During the
+pandemic, these workshops were moved to an online format and split into
+basic and advanced sessions. Additionally, beginning with the AAS 238 online
+meeting, the workshops have been expanded to Summer AAS meetings. The Learn
+team finds that the workshop audience is best found as AAS meetings as opposed
+to more general Python meetings.
+
 \end{itemize}
+
+\subsubsection{Learn vision for the future}
+
+The Learn plan going forward is to continue to improve the tutorial website;
+to index third-party tutorials; to promote user engagement on the Discourse
+site; and to look for supportable opportunities to expand the reach of
+Astropy workshops beyond the American Astronomical Society meetings.
 

### 2022-01-21 src/ms.tex
First draft of astroquery component

--- 2022-01-18 src/ms.tex
@@
 \newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
+\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
+\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
+\newcommand{\astropyunits}{\astropysubpkg{units}}
+\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
+\newcommand{\astropyconstants}{\astropysubpkg{constants}}
+\newcommand{\astropytime}{\astropysubpkg{time}}
+\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
 
 % Links to documentation: classes
-\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
-\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
-\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
-\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}
+\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
+\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}
+
+\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
+\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
+\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
+\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
+GalacticLSR}{GalacticLSR}}
+\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
+\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
+\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
+\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
+\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
+\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
+\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
+\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
+\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
+\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
+\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
+\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
+
+% suggestion: shortcut reference to Paper II
+\newcommand{\paperii}{Paper II}
 
@@
 \begin{abstract}
-
-    To be written...
-
+The \astropy Project supports and fosters the development of open-source and openly-developed
+\python packages that provide commonly-needed functionality to the astronomical
+community.
+A key element of the \astropy Project is the core package \astropypkg, which serves as the
+foundation for more specialized projects and packages.
+In this article, we summarize key features in the core package as of the recent major
+release, version 5.0, and provide major updates for the project.
+We then discuss supporting a broader ecosystem of inter-operable packages,
+including connections with several astronomical observatories and missions.
+We also revisit the future outlook of the \astropy Project and the current
+status of Learn Astropy.
+We conclude by raising and discussing the current and future challenges facing the project.
 \end{abstract}
@@
     \begin{centering}
-        \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
+      % \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
         \caption{
@@
 
-Summarize a few notable things in v5.0 as compared to the last LTS release.
+Major versions of the core package -- that is to say versions which add and/or
+modify functionality -- are released approximately every six months, and are
+then maintained with releases that fix issues, until the next major version
+is released. However, every two years a major release is designated as a
+long-term support (LTS) release which continues to be maintained for up to two
+years \citep{ape2}. The motivation for LTS releases is to provide longer-term stable
+versions of astropy that users requiring a high level of stability can make
+use of if they do not always need the latest features (this could include
+for example telescope pipelines and so on). The 5.0 release of the core package was designated as LTS,
+and since it was released at the end of 2021, it will be maintained until the
+end of 2023.
 
@@
     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-    \item TODO: something about WCS, SpectralCoord?
+    \item APE 14 WCS
+    \item A new SpectralCoord class for representing and transforming spectral quantities
+    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
 \end{itemize}
@@
 
-As part of the process of developing Astropy into a long-term sustainable product, and to improve transparency and accountability, the Project agreed to write down and formalize our governance structure (partly supported by explicit funding for this purpose - see \S \ref{sec:project-funding}).
-At the 2019 coordination meeting input was gathered from participants on what governance structures existed in the associated Open Source Software communities, and what would fit well with the needs of Astropy.
-This led into a ``retreat'' planned for March 2020, but due to the COVID-19 pandemic, this became a serious of virtual meetings of the ``Astropy Governance Working Group''.
-This group drafted the APE0 document \citep{ape0}, which was then eventually ratified and implemented by the ``Astropy Governance Implementation Working Group'' in Fall 2021.
-While the process emphasized flexibility and the ability to adapt to changing circumstances, it is expected that this is the framework Astropy's governance will operate in for at least the medium-term future.
+As part of the process of developing Astropy into a long-term sustainable
+product, and to improve transparency and accountability, the Project agreed to
+write down and formalize our governance structure (partly supported by explicit
+funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
+Astropy Coordination Meeting, input was gathered from participants on what governance
+structures existed in the associated Open Source Software communities, and what
+would fit well with the needs of Astropy. This led into a ``retreat'' planned
+for March 2020, but due to the COVID-19 pandemic, this became a series of
+virtual meetings of the ``Astropy Governance Working Group''.  This group
+drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
+implemented by the ``Astropy Governance Implementation Working Group'' in Fall
+2021. While the process emphasized flexibility and the ability to adapt to
+changing circumstances, it is expected that this is the framework Astropy's
+governance will operate in for at least the medium-term future.
 
-The APE0 \citep{ape0} document lays out the principles of this governance structure, so we refer the reader to that document for a more thorough description.
-However here we highlight some key elements.
-While many of these principles were already de facto true or have been discussed organically (and have been discussed in earlier papers in this series), the APE0-based governance aimed to provide a single place where the community can agree as a starting point.
-With this in mind, it highlights the developer and user community of the Astropy Project as the ultimate sources of authority, as well as the core principle of ``do-ocracy'' that those who do work for the Project (be it coding, training, or other less concrete contributions) gain more influence on the outputs of the Project by virtue of their effort.
-APE0 adds, however, the concept of ``voting members'' - a self-governed part of that community who are entrusted to elect the Coordination Committee.
-While this committee has existed from the inception of the project, APE0 establishes a formal voting process for this committee, and explicitly outlines the rights and responsibilities of the Coordination Committee.
-This role is mainly to facilitate consensus and act as the decision maker when other mechanisms have failed.
-However, it also includes powers that either require central authority or secrets (e.g.
-passwords), but AOE0 also charges the Committee to devolve responsibilities and seek community input on these items as often as possible.
+The APE0 \citep{ape0} document lays out the principles of this governance
+structure, so we refer the reader to that document for a more thorough
+description.  However here we highlight some key elements. While many of these
+principles were already de facto true or have been discussed organically (and
+have been discussed in earlier papers in this series), the APE0-based governance
+aimed to provide a single place where the community can agree as a starting
+point. With this in mind, it highlights the developer and user community of the
+Astropy Project as the ultimate sources of authority, as well as the core
+principle of ``do-ocracy'' that those who do work for the Project (be it coding,
+training, or other less concrete contributions) gain more influence on the
+outputs of the Project by virtue of their effort.  APE0 adds, however, the
+concept of ``voting members'' - a self-governed part of that community who are
+entrusted to elect the Coordination Committee. While this committee has existed
+from the inception of the project, APE0 establishes a formal voting process for
+this committee, and explicitly outlines the rights and responsibilities of the
+Coordination Committee. This role is mainly to facilitate consensus and act as
+the decision maker when other mechanisms have failed.  However, it also includes
+powers that either require central authority or secrets (e.g., passwords), but
+APE0 also charges the Committee to devolve responsibilities and seek community
+input on these items as often as possible.
 
-The first coordinating committee election under these rules took place in fall 2021, electing a mix of prior existing and new coordination committee members, and was contested in the sense of more candidates than available slots.
-This suggests the process is already working to serve the long-term interests of the committee to both spread the coordination effort, and to ensure it is not dominated by the same people for as long as the Project continues.
-While other, more fine-grained governance improvements are planned for the future, it is clear the foundation is now in place.
+The first Coordination Committee election under these rules took place in Fall
+2021, electing a mix of prior existing and new coordination committee members,
+and was contested in the sense of more candidates than available slots. This
+suggests the process is already working to serve the long-term interests of the
+committee to both spread the coordination effort, and to ensure it is not
+dominated by the same people for as long as the Project continues. While other,
+more fine-grained governance improvements are planned for the future, it is
+clear the foundation is now in place.
 
@@
     \begin{centering}
-        \includegraphics{figures/contributor-summary.pdf}
+      % \includegraphics{figures/contributor-summary.pdf}
         \caption{Placeholder figure!}
@@
 \secauthor{Lía Corrales}
+
+With support from the Moore Foundation, the \astropy Project was
+allocated funding to support mentoring programs. In 2020 a call was
+made to submit proposals for IDE (Inclusion, Diversity, and
+Empowerment) initiatives for project-wide consideration on
+Github. This process was deemed the most ``open'' because it allowed
+for community wide feedback to focus and improve proposal
+initiatives. Two programs were selected via this process, described
+below.
+
+\textbf{Outreachy:} Describe the outreachy project
+
+\textbf{Women of Color Code (WoCCode)} is a peer-mentoring network for
+coders from traditionally marginalized groups, most notably women of
+color. Participants were invited to the WoCCode Slack space and
+encouraged to attend monthly webinars to share skills in the context
+of open source software libraries. Every other month, a guest speaker
+was invited to talk about their career path and share a skill. Program
+participants were solicited in the fall of 2020, yielding 73
+applications from 17 countries (48\% from the United States, 37\% from
+Africa, and 15\% from remaining continents). We selected participants
+who we identified as having a high potential for contributing to open
+source projects: intermediate to advanced programming skills with a
+vocal interest in contributing. Of the thirty applicants invited to
+the program, nineteen joined the community on Slack. Participants were
+organized into cohorts based on interest and each cohort was assigned
+one of three mentors that were also selected via open
+application. Mentors acted as a general resource to participants, gave
+one webinar, and organized a hack day. WoCCode also supported
+registration of two participants to attend the American Astronomical
+Society virtual summer 2021 meeting.
+
+Participants rated the impact of the program as very high. In the
+final webinar, participants reported a change in their perception of
+coding in general, for example, accepting coding as something they can
+do for fun. Participants reported generally feeling comfortable asking
+questions and interacting with a community where ``[e]veryone's
+thoughts are welcomed, no one is made to feel less important'' and
+where one can enjoy ``warm interactions with likeminded people.''
+WoCCode is continuing into 2022 by broadening participation in the
+Slack space and publicly advertising the guest webinar events.
+
+In addition to the supported mentoring initiatives, the \astropy
+project as a whole has taken steps to examine representation of
+marginalized groups within the project and search for avenues of
+improvement.
+
+% Much of this text was adapted from Robel Geda's report from NSBP and
+% SACNAS. Please include him on the author list.
+\textbf{\astropy representation at national diversity conferences:}
+Several members of the \astropy community attended virtual conferences
+of the National Society of Black Physicists (NSBP) and SACNAS
+(focusing on Hispanic and Native American scientists across all STEM
+fields). \astropy representatives noted that one underlying topic came
+up multiple times. A major barrier for persons who come from
+underrepresented communities is the lack of resources and expertise
+that are necessary to train students. Writing PEP 8 compliant software
+and understanding the Git and Github workflow is not part of the
+standard Physics and Astrophysics curriculum. Additionally, the larger
+Astronomnical community is still in the process of transitioning
+towards Python based tools, making it difficult for students not
+currently under advisement by some one with extensive Python expertise
+to get involved with the open source Python community. Projects such
+as Learn Astropy could have a profound impact by empowering
+underrepresented groups because it provides a free, searchable, and
+accessible introduction to Python tools for astrophysical
+research. Representatives also noted that offering training and
+teaching materials to PIs of Research Experience for Undergraduate
+(REU) programs throughout the United States could be helpful for
+encouraging advisors to teach students how to use Astropy, create
+their own libraries, and use version control on open source platforms
+like Github. Such materials could also be offered as a workshop at
+national diversity conferences themselves, as SACNAS solicits special
+session proposals each year.
+
+\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
+  initiative:} Describe more about this.
 
@@
 
-Infrastructure packages that exist to support and help others maintain their
-package infrastructure.
+The Astropy project supports the broader ecosystem by providing pre-configured
+infrastructure packages that the community can use to support and maintain
+their own software package infrastructure. These include tools to easily
+generate documentation and setup automated testing, as well as provide
+package scaffolding for new projects.
 
-This includes an overview of the sphinx-astropy tools, the pytest extensions
-maintained by astropy, and the package template.
+Sphinx is a common and useful tool for generating documentation for Python
+packages. The Astropy project maintains a default Sphinx configuration along
+with Astropy-specific extensions which can be easily added to community
+projects via the
+\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}} meta
+package. This tool provides a pre-configured Sphinx setup compatible with
+Astropy projects, which includes several extensions useful for generating API
+documentation
+(\href{https://github.com/astropy/sphinx-automodapi}{\texttt{sphinx-automodapi}}),
+allowing for Numpy docstring parsing
+(\href{https://github.com/numpy/numpydoc}{\texttt{numpydoc}}), embedded image
+handling
+(\href{https://github.com/sphinx-gallery/sphinx-gallery}{\texttt{sphinx-gallery}};
+\href{https://github.com/python-pillow}{\texttt{pillow}}), advanced
+documentation testing support
+(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
+and providing a custom documentation theme ideal for analysis packages
+(\href{https://github.com/astropy/sphinx-astropy}{\texttt{astropy-sphinx-theme}}).
+
+Community package testing infrastructure is supported through the
+\href{https://github.com/astropy/pytest-astropy}{\texttt{pytest-astropy}}
+meta-package, providing a unified testing framework with useful extensions
+compatible with both Astropy- and non-Astropy-affiliated community packages.
+This meta-package pulls in several
+\href{https://github.com/pytest-dev/pytest}{\texttt{pytest}} plugins to help
+with custom test headers
+(\href{https://github.com/astropy/pytest-astropy-header}{\texttt{pytest-astropy-header}}),
+accessing remotely-hosted data files in tests
+(\href{https://github.com/astropy/pytest-remotedata}{\texttt{pytest-remotedata}}),
+interoperability with documentation
+(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
+dangling file handle checking
+(\href{https://github.com/astropy/pytest-openfiles}{\texttt{pytest-openfiles}}),
+data array comparison support in tests
+(\href{https://github.com/astropy/pytest-arraydiff}{\texttt{pytest-arraydiff}}),
+sub-package command-line testing support
+(\href{https://github.com/astropy/pytest-filter-subpackage}{\texttt{pytest-filter-subpackage}}),
+improved mock object testing
+(\href{https://github.com/pytest-dev/pytest-mock}{\texttt{pytest-mock}}), test
+coverage reports and measurements
+(\href{https://github.com/pytest-dev/pytest-cov}{\texttt{pytest-cov}}), and
+configuring package for property-based testing
+(\href{https://github.com/HypothesisWorks/hypothesis}{\texttt{hypothesis}}).
+
+The Astropy Package Template helps facilitate the setup and creation of new
+Python packages leveraging the Astropy ecosystem. This tool utilizes the
+Cookiecutter project to walk users through the process of creating new
+packages complete with documentation and testing support. Additionally, the
+package template generation process includes the ability to setup
+interoperability with GitHub, allowing for easy repository access from
+documentation, as well as an example GitHub Actions workflow to demonstrate
+the use of GitHub's continuous integration tooling.
 
@@
 
-Summary of Astroquery and engagement with archives. Cite astroquery paper.
+% Summary of Astroquery and engagement with archives. Cite astroquery paper.
+Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
+Since \PaperII, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST; n{\'e}e Multimission Archive for at STSCI}, the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
+These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
+
+Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
+These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
+While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
+% note to coauthors: we need a section on pyvo
 
@@
 
-JWST
+\subsubsection{James Webb Space Telescope}
 \secauthor{Larry Bradley}
+
+The James Webb Space Telescope (JWST) is a 6.5-meter space-based
+infrared telescope that will provide unprecedented resolution and
+sensitivity from 0.6 -- 28 microns. JWST will enable a broad range
+of scientific investigations from exoplanets and their atmospheres
+to the formation of galaxies in the very early universe. Its four
+key scientific goals are to study the first light from stars and
+galaxies, the assembly and evolution of galaxies, the birth of stars and
+protoplanetary systems, and planetary systems and the origins of life.
+
+The telescope launched on an Ariane 5 rocket on 2021 December 25 from
+Kourou, French Guiana. After a series of successful deployments,
+including the sunshield and primary and secondary mirrors, JWST reached
+its orbit around the L2 Lagrange point on 2022 January 24. Commissioning
+of the telescope optics and science instruments will occur from January
+until the end of June 2022, when science operations are scheduled to
+begin.
+
+Software developers at the Space Telescope Science Institute (STScI),
+the operations center of JWST, have been developing Python-based tools
+for JWST since 2010 (starting with the JWST Calibration Reference Data
+System) and have provided major contributions to \astropy from its
+inception. The JWST instrument calibration pipelines, exposure-time
+calculators, and data analysis tools are all written in Python and
+depend on the \astropypkg core package and some coordinated and
+affiliated packages. For the \astropypkg core package, the JWST mission
+has provided extensive contributions to the \astropysubpkg{modeling},
+\astropysubpkg{units}, \astropysubpkg{coordinates}, \astropysubpkg{wcs},
+\astropysubpkg{io.fits}, \astropysubpkg{io.votable},
+\astropysubpkg{stats}, \astropysubpkg{visualization}, and
+\astropysubpkg{convolution} subpackages as well as to the general
+package infrastructure and maintenance.
+
+Likewise, JWST developers have provided significant contributions
+to the \package{photutils} \citep{photutils}, \package{specutils}
+\citep{specutils}, and \package{regions} \citep{regions} coordinated
+packages and the \package{gwcs} \citep{gwcs} and \package{synphot}
+\citep{synphot} affiliated packages. For example, development of the
+\package{photutils} coordinated package for source detection and
+photometry has largely been led by JWST contributions. JWST developers
+have also made significant contributions to the \package{specutils}
+coordinated package, which is used for analyzing spectroscopic data,
+and the \package{regions} coordinated package, which is used to
+handle geometric regions. The \package{gwcs} affiliated package for
+generalized world coordinate systems was created specifically to handle
+the complex world coordinate systems needed for JWST spectroscopic
+data. The \package{synphot} affiliated package for synthetic photometry
+was created at STScI and is a dependency of the JWST exposure-time
+calculators. Further, the \package{ASDF} (Advanced Scientific Data
+Format) package \citep{ASDF}, a next-generation interchange format for
+scientific data, was initially developed at STScI to serialize JWST WCS
+objects along with \astropypkg models, units, and coordinates. Over
+its mission lifetime, JWST will continue its support in developing and
+maintaining these critically dependent packages.
 
@@
 Cherenkov Telescope Array
-\secauthor{Axel Donath, Max Noethe}
+\secauthor{Axel Donath, Maximilian Nöthe}
+
+The Cherenkov Telescope (CTA) will be the next generation very-high-energy
+gamma-ray observatory.
+CTA will improve over the current generation of imaging atmospheric Cherenkov telescopes (IACTs)
+by a factor of five to ten in sensitivity and will be able to observe the whole sky from a combination of two sites:
+a northern site in La Palma, Spain, and a southern one in Paranal, Chile.
+CTA will be able to observe gamma rays in a broad energy range from around $20\,\mathrm{GeV}$ to over $300\,\mathrm{TeV}$
+using three different types of telescopes, in total over 100 telescopes are planned at the two sites.
+CTA will also be the first open gamma-ray observatory.
+
+The data analysis pipeline is developed as open source software and essentially split in two domains:
+\begin{enumerate}
+  \item In the low-level analysis, the properties of the recorded air-shower events
+    have to be estimated from the raw data.
+    The raw data consists of very short (\textasciitilde $40--100\,\mathrm{ns}$) videos recorded with the fast and
+    sensitive cameras of the telescopes.
+    This includes the energy, particle type and direction of origin of the particle that induced the air shower
+    and the time the shower was recorded.
+  \item In the higher-level analysis, these reconstructed event lists are used together with some
+    characterization of the instrument response to perform the actual scientific analysis.
+    This software will be delivered as CTA science tools to the future users of the Observatory.
+\end{enumerate}
+
+A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
+a python package developed to perform all the necessary tasks to from the raw data
+of Cherenkov telescopes to the reconstructed event lists.
+
+The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
+Gammapy~\citep{gammapy}.
+
+Both Gammapy and \texttt{ctapipe} are using astropy heavily, mainly for units, times, coordinate transformations,
+tables and FITS IO.
+
+As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
+it needs to perform a large number of coordinate transformations.
+To enable this, CTA member M.~Nöthe contributed a major performance improvement
+for large numbers of coordinates with different observation times,
+based on earlier work by B.~Winkel.
+
+Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
+a total of 107 merged pull requests were contributed to astropy.
 
@@
 
-\secauthor{Lía Corrales, David Shupe + Learn team}
+\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
+
+\subsubsection{Current status and scope}
+
+{\it Learn Astropy} is an umbrella term that acknowledges the broad educational
+efforts made by the Astropy Project, which are led by the Learn Team.
+The efforts focus on developing online content and workshops covering
+astro-specific coding tasks in Python.
+As described in \citet{astropy:2018}, There are four different types of
+Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
+lessons that are published in HTML format online; \textit{guides}, which are a
+series of lessons providing a foundational resource for performing certain
+type of astronomical analyses; \textit{examples}, which are snippets of code
+that showcase a short task that can be performed with Astropy or an affiliated
+package; and \textit{documentation}, which is contained within the code base.
+This categorization drives content development, infrastructure choices, and the
+appearance of the Learn Astropy website.
+The Learn Team meets weekly to work on creating, expanding, improving these
+educational resources.
 
 \begin{itemize}
-\item Current status and scope
-\item Vision for the future (review what we said in v2.0 paper)
-\item User forums and engagement with user base
-\item Workshops
+
+\item {\bf Learn Astropy website:} The Learn Team re-launched the website in
+2021 with a new infrastructure platform, built around full-text search and
+interactive filtering functionality, with the goal of making content
+discoverable as the Learn Astropy content catalog expands.
+This work has been supported in part by the Dunlap grant.
+We have adopted Algolia, a search-as-a-service cloud platform, to store the
+full-text and metadata records of Learn Astropy's content.
+The new Learn Astropy website is a JavaScript (Gatsby/React) application that
+uses the Algolia service to power its search and filtering user interface.
+Our Python-based application, Learn Astropy Librarian, populates data into the
+Algolia service. We tuned the Librarian around specific content formats (such as
+Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
+accurately index content and heuristically extract metadata.
+A consequence of the new platform is that we now maintain and compile content
+separately from the website application itself, enabling new content types.
+Tutorials, which as Jupyter Notebook-based, are now compiled into their own
+Learn Astropy sub-site using nbcollection.
+Guides, which are Jupyter Book-based, are also deployed as separate websites
+using GitHub pages. This architecture opens future possibilities of indexing
+third-party content, hosted elsewhere, such as on institutional websites.
+
+\item \textbf{Tutorials} Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
+
+\item \textbf{Guide} {\it ccdproc} Jupyter book has been completed.
+
+\item {\bf Workshops:} the Project has been conducting workshops at winter
+meetings of the American Astronomical Society since AAS 225 in January 2015.
+Up to the start of the coronavirus pandemic, these were full-day in-person
+workshops with as many as 90 participants and a dozen facilitators from the
+project.
+During the pandemic, these workshops were moved to an online format and split
+into basic and advanced sessions.
+Additionally, beginning with the AAS 238 online meeting, the workshops have been
+expanded to Summer AAS meetings.
+The Learn team finds that the workshop audience is best found as AAS meetings as
+opposed to more general Python meetings.
+
+The Astropy Project provided another mode of community engagement at AAS
+Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
+AAS Exhibit Hall.
+Funding for the exhibit hall was provided alternately by NumFOCUS and later by
+the Moore Foundation funding (\textbf{check this}).
+The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
+during the virtual AAS 237 meetings, to provide the general astronomy community
+information and access to experts on a variety of open source astronomical
+tools.
 \end{itemize}
+
+\subsubsection{Learn vision for the future}
+
+The Learn plan going forward is to continue to improve the tutorial website and
+facilate new content;
+to index third-party tutorials; and to look for supportable opportunities to
+expand the reach of Astropy workshops beyond the American Astronomical Society
+meetings.
 
@@
 
+\section{Community Engagement}
+
+\begin{itemize}
+\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+open-source projets.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+a Discourse site which is more open than the Facebook group and more user-oriented
+than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+developers are able to invoice as independent contractors the time they
+spend helping users on these forums.
+\end{itemize}
 \secunfilled
@@
           \package{matplotlib} \citep{matplotlib},
-          \package{Cython} \citep{cython}.
+          \package{Cython} \citep{cython},
+          \package{photutils} \citep{photutils},
+          \package{specutils} \citep{specutils},
+          \package{regions} \citep{regions},
+          \package{gwcs} \citep{gwcs},
+          \package{synphot} \citep{synphot},
+          \package{ASDF} \citep{ASDF}.
         %   \package{SOFA} \citep{sofa},

### 2022-01-21 src/ms.tex
Update ms.tex

--- 2022-01-19 src/ms.tex
@@
 
+All affiliated packages are listed on the astropy 
+website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
+\cite{astropy:2018}, there has been an explosion of gravitational-wave affiliated 
+packages including: \texttt{PyCBC} for exploring GW sources, lenstronomy for 
+modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing GW 
+probability maps, and \texttt{EinsteinPy} for general relativity and gravitational astronomy. 
+There have been several packages related to HEALPix: \texttt{astropy-healpix}
+for a BSD licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order 
+Coverage maps. \texttt{astroalign} as been introduced for astrometric registration, 
+and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
+For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
+VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
+We have \texttt{agnpy} for AGN astrophysics, and \texttt{statmorph} for fitting galactic 
+morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
+routines. For modeling interstellar dust extinction we have 
+\texttt{dust_extinction}, and for extracting features from time series we 
+have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
+pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
+inference, and \texttt{sbpy} does calculations for asteroid and cometary 
+astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
+
+Several of the affiliated packages described in [reference to paper 2] 
+have had substantial improvements. \texttt{Astroquery} [reference] has added access 
+to roughly a dozen new missions and data services, including JWST, and the 
+project has switched to a continuous release model. Every time a change 
+is committed to the main development branch it is published on the Python 
+Package Index (PyPI, ref) and available for installation. Formal releases 
+are still done a couple of times per year. \texttt{Photutils} released its first 
+stable version, indicating that the API will change less frequently, and 
+there have been several significant performance improvements.  \texttt{CCDproc} 
+also released a new major version, bringing better performance to some 
+image combination operations. The \texttt{regions} package, for manipulating 
+ds9-style [ref] region definitions, added new ways to manipulate regions 
+and introduced new region types. \texttt{Reproject}’s major new feature is a 
+function to align and co-add images to create a mosaic; better support 
+for parallelization was also added. \texttt{Specutils}, the package that defines 
+containers for 1D and 2D spectra, also had its first stable rele and 
+the addition of classes to read JWST data.
+
 \subsection{Connections with data archives}

### 2022-01-21 src/ms.tex
Add citations for coordinated packages

@@
 
-Several of the affiliated packages described in [reference to paper 2] 
-have had substantial improvements. \texttt{Astroquery} [reference] has added access 
-to roughly a dozen new missions and data services, including JWST, and the 
-project has switched to a continuous release model. Every time a change 
+Several of the affiliated packages described in [reference to paper 2] have
+had substantial improvements. \texttt{astroquery} \cite{astroquery} has added
+access to roughly a dozen new missions and data services, including JWST, and
+the project has switched to a continuous release model. Every time a change
 is committed to the main development branch it is published on the Python
-Package Index (PyPI, ref) and available for installation. Formal releases 
-are still done a couple of times per year. \texttt{Photutils} released its first 
-stable version, indicating that the API will change less frequently, and 
-there have been several significant performance improvements.  \texttt{CCDproc} 
-also released a new major version, bringing better performance to some 
-image combination operations. The \texttt{regions} package, for manipulating 
-ds9-style [ref] region definitions, added new ways to manipulate regions 
-and introduced new region types. \texttt{Reproject}’s major new feature is a 
-function to align and co-add images to create a mosaic; better support 
-for parallelization was also added. \texttt{Specutils}, the package that defines 
-containers for 1D and 2D spectra, also had its first stable rele and 
-the addition of classes to read JWST data.
+Package Index (PyPI, ref) and available for installation. Formal releases are
+still done a couple of times per year. \texttt{photutils} \cite
+{photutils} released its first stable version, indicating that the API will
+change less frequently, and there have been several significant performance
+improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
+version, bringing better performance to some image combination operations.
+The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
+region definitions \cite{ds9}, added new ways to manipulate regions and
+introduced new region types. \texttt{reproject}’s \cite{reproject} major new
+feature is a function to align and co-add images to create a mosaic; better
+support for parallelization was also added. \texttt{specutils}, the package
+that defines containers for 1D and 2D spectra \cite{specutils}, also had its
+first stable release and the addition of classes to read JWST data.
 

### 2022-01-22 src/ms.tex
Apply suggestions from code review
Co-authored-by: Lia Corrales <lia.astro@gmail.com>

--- 2022-01-21 src/ms.tex
@@
 
-All affiliated packages are listed on the astropy 
-website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
-\cite{astropy:2018}, there has been an explosion of gravitational-wave affiliated 
-packages including: \texttt{PyCBC} for exploring GW sources, lenstronomy for 
-modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing GW 
-probability maps, and \texttt{EinsteinPy} for general relativity and gravitational astronomy. 
-There have been several packages related to HEALPix: \texttt{astropy-healpix}
-for a BSD licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order 
-Coverage maps. \texttt{astroalign} as been introduced for astrometric registration, 
-and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
-For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
-We have \texttt{agnpy} for AGN astrophysics, and \texttt{statmorph} for fitting galactic 
-morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
-routines. For modeling interstellar dust extinction we have 
-\texttt{dust_extinction}, and for extracting features from time series we 
-have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
-pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
-inference, and \texttt{sbpy} does calculations for asteroid and cometary 
-astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
-
-Several of the affiliated packages described in [reference to paper 2] 
-have had substantial improvements. \texttt{Astroquery} [reference] has added access 
-to roughly a dozen new missions and data services, including JWST, and the 
-project has switched to a continuous release model. Every time a change 
-is committed to the main development branch it is published on the Python 
-Package Index (PyPI, ref) and available for installation. Formal releases 
-are still done a couple of times per year. \texttt{Photutils} released its first 
-stable version, indicating that the API will change less frequently, and 
-there have been several significant performance improvements.  \texttt{CCDproc} 
-also released a new major version, bringing better performance to some 
-image combination operations. The \texttt{regions} package, for manipulating 
-ds9-style [ref] region definitions, added new ways to manipulate regions 
-and introduced new region types. \texttt{Reproject}’s major new feature is a 
-function to align and co-add images to create a mosaic; better support 
-for parallelization was also added. \texttt{Specutils}, the package that defines 
-containers for 1D and 2D spectra, also had its first stable rele and 
-the addition of classes to read JWST data.
-
 \subsection{Connections with data archives}
@@
 
+\subsubsection{Current status and scope}
+
+{\it Learn Astropy} is an umbrella term that acknowledges the broad educational efforts made by the Astropy Project, which are led by the Astropy Learn Team. 
+As described in \textsl{[Astropy-ii paper]}, Learn Astropy content has been categorized along four axes: documentation, which is contained within the python code base; tutorials, consisting of Jupyter Notebook lessons that are published in HTML format online; guides, which are a series of lessons providing a foundational resource for performing a certain type of astronomical analyses; and examples, which are snippets of code that showcase a short task that can be performed with Astropy or its affiliated package.
+These categories continue to drive both the infrastructure choices and the appearance of the Learn Astropy website. Additionally, the Astropy Learn Team has expanded its efforts
+in user engagement and in conducting workshops. The Learn team meets
+weekly to drive these issues forward.
+
 \begin{itemize}
-\item Current status and scope
-\item Vision for the future (review what we said in v2.0 paper)
-\item User forums and engagement with user base
-\item Workshops
+
+\item {\bf Learn Astropy website:} The site has a new search interface powered by
+Algolia which is used by major shopping sites. This work has been
+supported in part by the Dunlap grant. Tutorials have been added and
+a {\it ccdproc} Jupyter book has been completed.
+
+\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+open-source projets.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+a Discourse site which is more open than the Facebook group and more user-oriented
+than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+developers are able to invoice as independent contractors the time they
+spend helping users on these forums.
+
+\item {\bf Workshops:} the Project has been conducting workshops at winter meetings
+of the American Astronomical Society since AAS 225 in January 2015. Up to the
+start of the coronavirus pandemic, these were full-day in-person workshops with
+as many as 90 participants and a dozen facilitators from the project. During the
+pandemic, these workshops were moved to an online format and split into
+basic and advanced sessions. Additionally, beginning with the AAS 238 online
+meeting, the workshops have been expanded to Summer AAS meetings. The Learn
+team finds that the workshop audience is best found as AAS meetings as opposed
+to more general Python meetings.
+
+The Astropy Project provided another mode of community engagement at AAS Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the AAS Exhibit Hall.
+Funding for the exhibit hall was provided alternately by NumFOCUS and later by the Moore Foundation funding (\textbf{check this}).
+The booth hosted a series of Q\&A special sessions during AAS 235 and webinars during the virtual AAS 237 meetings, to provide the general astronomy community information and access to experts on a variety of open source astronomical tools.
 \end{itemize}
+
+\subsubsection{Learn vision for the future}
+
+The Learn plan going forward is to continue to improve the tutorial website;
+to index third-party tutorials; to promote user engagement on the Discourse
+site; and to look for supportable opportunities to expand the reach of
+Astropy workshops beyond the American Astronomical Society meetings.
 

### 2022-01-22 src/ms.tex
Affiliated packages draft text
Committed in collaboration with @mwcraig.

--- 2022-01-21 src/ms.tex
@@
 
-Briefly summarize new procedures and governance structure, new CoCo, election
-process overview, etc.
+%Briefly summarize new procedures and governance structure, new CoCo, election
+%process overview, etc.
+
+As part of the process of developing Astropy into a long-term sustainable
+product, and to improve transparency and accountability, the Project agreed to
+write down and formalize our governance structure (partly supported by explicit
+funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
+Astropy Coordination Meeting, input was gathered from participants on what governance
+structures existed in the associated Open Source Software communities, and what
+would fit well with the needs of Astropy. This led into a ``retreat'' planned
+for March 2020, but due to the COVID-19 pandemic, this became a series of
+virtual meetings of the ``Astropy Governance Working Group''.  This group
+drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
+implemented by the ``Astropy Governance Implementation Working Group'' in Fall
+2021. While the process emphasized flexibility and the ability to adapt to
+changing circumstances, it is expected that this is the framework Astropy's
+governance will operate in for at least the medium-term future.
+
+The APE0 \citep{ape0} document lays out the principles of this governance
+structure, so we refer the reader to that document for a more thorough
+description.  However here we highlight some key elements. While many of these
+principles were already de facto true or have been discussed organically (and
+have been discussed in earlier papers in this series), the APE0-based governance
+aimed to provide a single place where the community can agree as a starting
+point. With this in mind, it highlights the developer and user community of the
+Astropy Project as the ultimate sources of authority, as well as the core
+principle of ``do-ocracy'' that those who do work for the Project (be it coding,
+training, or other less concrete contributions) gain more influence on the
+outputs of the Project by virtue of their effort.  APE0 adds, however, the
+concept of ``voting members'' - a self-governed part of that community who are
+entrusted to elect the Coordination Committee. While this committee has existed
+from the inception of the project, APE0 establishes a formal voting process for
+this committee, and explicitly outlines the rights and responsibilities of the
+Coordination Committee. This role is mainly to facilitate consensus and act as
+the decision maker when other mechanisms have failed.  However, it also includes
+powers that either require central authority or secrets (e.g., passwords), but
+APE0 also charges the Committee to devolve responsibilities and seek community
+input on these items as often as possible.
+
+The first Coordination Committee election under these rules took place in Fall
+2021, electing a mix of prior existing and new coordination committee members,
+and was contested in the sense of more candidates than available slots. This
+suggests the process is already working to serve the long-term interests of the
+committee to both spread the coordination effort, and to ensure it is not
+dominated by the same people for as long as the Project continues. While other,
+more fine-grained governance improvements are planned for the future, it is
+clear the foundation is now in place.
 

### 2022-01-25 src/ms.tex
wrap at 80 chars

--- 2022-01-21 src/ms.tex
@@
 \begin{abstract}
-The \astropy Project supports and fosters the development of open-source and openly-developed
-\python packages that provide commonly-needed functionality to the astronomical
-community.
-A key element of the \astropy Project is the core package \astropypkg, which serves as the
-foundation for more specialized projects and packages.
-In this article, we summarize key features in the core package as of the recent major
-release, version 5.0, and provide major updates for the project.
-We then discuss supporting a broader ecosystem of inter-operable packages,
-including connections with several astronomical observatories and missions.
-We also revisit the future outlook of the \astropy Project and the current
-status of Learn Astropy. 
-We conclude by raising and discussing the current and future challenges facing the project.
+
+    To be written...
+
 \end{abstract}
@@
 
-Briefly summarize new procedures and governance structure, new CoCo, election
-process overview, etc.
+%Briefly summarize new procedures and governance structure, new CoCo, election
+%process overview, etc.
+
+As part of the process of developing Astropy into a long-term sustainable
+product, and to improve transparency and accountability, the Project agreed to
+write down and formalize our governance structure (partly supported by explicit
+funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
+coordination meeting input was gathered from participants on what governance
+structures existed in the associated Open Source Software communities, and what
+would fit well with the needs of Astropy. This led into a ``retreat'' planned
+for March 2020, but due to the COVID-19 pandemic, this became a serious of
+virtual meetings of the ``Astropy Governance Working Group''.  This group
+drafted the APE0 document \citep{ape0}, which was then eventually ratified and
+implemented by the ``Astropy Governance Implementation Working Group'' in Fall
+2021. While the process emphasized flexibility and the ability to adapt to
+changing circumstances, it is expected that this is the framework Astropy's
+governance will operate in for at least the medium-term future.
+
+The APE0 \citep{ape0} document lays out the principles of this governance
+structure, so we refer the reader to that document for a more thorough
+description.  However here we highlight some key elements. While many of these
+principles were already de facto true or have been discussed organically (and
+have been discussed in earlier papers in this series), the APE0-based governance
+aimed to provide a single place where the community can agree as a starting
+point. With this in mind, it highlights the developer and user community of the
+Astropy Project as the ultimate sources of authority, as well as the core
+principle of ``do-ocracy'' that those who do work for the Project (be it coding,
+training, or other less concrete contributions) gain more influence on the
+outputs of the Project by virtue of their effort.  APE0 adds, however, the
+concept of ``voting members'' - a self-governed part of that community who are
+entrusted to elect the Coordination Committee. While this committee has existed
+from the inception of the project, APE0 establishes a formal voting process for
+this committee, and explicitly outlines the rights and responsibilities of the
+Coordination Committee. This role is mainly to facilitate consensus and act as
+the decision maker when other mechanisms have failed.  However, it also includes
+powers that either require central authority or secrets (e.g. passwords), but
+AOE0 also charges the Committee to devolve responsibilities and seek community
+input on these items as often as possible.
+
+The first coordinating committee election under these rules took place in fall
+2021, electing a mix of prior existing and new coordination committee members,
+and was contested in the sense of more candidates than available slots. This
+suggests the process is already working to serve the long-term interests of the
+committee to both spread the coordination effort, and to ensure it is not
+dominated by the same people for as long as the Project continues. While other,
+more fine-grained governance improvements are planned for the future, it is
+clear the foundation is now in place.
 
@@
 paper).
-
-All affiliated packages are listed on the astropy 
-website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
-\cite{astropy:2018}, there has been an explosion of gravitational-wave affiliated 
-packages including: \texttt{PyCBC} for exploring GW sources, lenstronomy for 
-modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing GW 
-probability maps, and \texttt{EinsteinPy} for general relativity and gravitational astronomy. 
-There have been several packages related to HEALPix: \texttt{astropy-healpix}
-for a BSD licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order 
-Coverage maps. \texttt{astroalign} as been introduced for astrometric registration, 
-and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
-For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
-We have \texttt{agnpy} for AGN astrophysics, and \texttt{statmorph} for fitting galactic 
-morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
-routines. For modeling interstellar dust extinction we have 
-\texttt{dust_extinction}, and for extracting features from time series we 
-have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
-pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
-inference, and \texttt{sbpy} does calculations for asteroid and cometary 
-astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
-
-Several of the affiliated packages described in [reference to paper 2] have
-had substantial improvements. \texttt{astroquery} \cite{astroquery} has added
-access to roughly a dozen new missions and data services, including JWST, and
-the project has switched to a continuous release model. Every time a change
-is committed to the main development branch it is published on the Python
-Package Index (PyPI, ref) and available for installation. Formal releases are
-still done a couple of times per year. \texttt{photutils} \cite
-{photutils} released its first stable version, indicating that the API will
-change less frequently, and there have been several significant performance
-improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
-version, bringing better performance to some image combination operations.
-The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
-region definitions \cite{ds9}, added new ways to manipulate regions and
-introduced new region types. \texttt{reproject}’s \cite{reproject} major new
-feature is a function to align and co-add images to create a mosaic; better
-support for parallelization was also added. \texttt{specutils}, the package
-that defines containers for 1D and 2D spectra \cite{specutils}, also had its
-first stable release and the addition of classes to read JWST data.
 

### 2022-01-25 src/ms.tex
Add initial JWST section

--- 2022-01-22 src/ms.tex
@@
 
-All affiliated packages are listed on the astropy 
-website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
-\cite{astropy:2018}, there has been an explosion of gravitational-wave affiliated 
-packages including: \texttt{PyCBC} for exploring GW sources, lenstronomy for 
-modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing GW 
-probability maps, and \texttt{EinsteinPy} for general relativity and gravitational astronomy. 
-There have been several packages related to HEALPix: \texttt{astropy-healpix}
-for a BSD licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order 
-Coverage maps. \texttt{astroalign} as been introduced for astrometric registration, 
-and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
-For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
-We have \texttt{agnpy} for AGN astrophysics, and \texttt{statmorph} for fitting galactic 
-morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
-routines. For modeling interstellar dust extinction we have 
-\texttt{dust_extinction}, and for extracting features from time series we 
-have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
-pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
-inference, and \texttt{sbpy} does calculations for asteroid and cometary 
-astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
-
-Several of the affiliated packages described in [reference to paper 2] have
-had substantial improvements. \texttt{astroquery} \cite{astroquery} has added
-access to roughly a dozen new missions and data services, including JWST, and
-the project has switched to a continuous release model. Every time a change
-is committed to the main development branch it is published on the Python
-Package Index (PyPI, ref) and available for installation. Formal releases are
-still done a couple of times per year. \texttt{photutils} \cite
-{photutils} released its first stable version, indicating that the API will
-change less frequently, and there have been several significant performance
-improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
-version, bringing better performance to some image combination operations.
-The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
-region definitions \cite{ds9}, added new ways to manipulate regions and
-introduced new region types. \texttt{reproject}’s \cite{reproject} major new
-feature is a function to align and co-add images to create a mosaic; better
-support for parallelization was also added. \texttt{specutils}, the package
-that defines containers for 1D and 2D spectra \cite{specutils}, also had its
-first stable release and the addition of classes to read JWST data.
-
 \subsection{Connections with data archives}
@@
 \astropysubpkg{units}, \astropysubpkg{coordinates}, \astropysubpkg{wcs},
-\astropysubpkg{io.fits}, \astropysubpkg{io.votable},
-\astropysubpkg{stats}, \astropysubpkg{visualization}, and
-\astropysubpkg{convolution} subpackages as well as to the general
-package infrastructure and maintenance.
+\astropysubpkg{fits}, \astropysubpkg{votable}, \astropysubpkg{stats},
+\astropysubpkg{visualization}, and \astropysubpkg{convolution}
+subpackages as well as to the general package infrastructure and
+maintenance.
 
-Likewise, JWST developers have provided significant contributions
-to the \package{photutils} \citep{photutils}, \package{specutils}
-\citep{specutils}, and \package{regions} \citep{regions} coordinated
-packages and the \package{gwcs} \citep{gwcs} and \package{synphot}
-\citep{synphot} affiliated packages. For example, development of the
-\package{photutils} coordinated package for source detection and
-photometry has largely been led by JWST contributions. JWST developers
-have also made significant contributions to the \package{specutils}
-coordinated package, which is used for analyzing spectroscopic data,
-and the \package{regions} coordinated package, which is used to
-handle geometric regions. The \package{gwcs} affiliated package for
-generalized world coordinate systems was created specifically to handle
-the complex world coordinate systems needed for JWST spectroscopic
-data. The \package{synphot} affiliated package for synthetic photometry
-was created at STScI and is a dependency of the JWST exposure-time
-calculators. Further, the \package{ASDF} (Advanced Scientific Data
-Format) package \citep{ASDF}, a next-generation interchange format for
-scientific data, was initially developed at STScI to serialize JWST WCS
-objects along with \astropypkg models, units, and coordinates. Over
+Likewise, JWST developers have provided significant contributions to
+the \package{photutils}, \package{specutils}, and \package{regions}
+coordinated packages and the \package{gwcs} affiliated package. For
+example, development of the \package{photutils} coordinated package
+for source detection and photometry has largely been led by JWST
+contributions. JWST developers have also made significant contributions
+to the \package{specutils} coordinated package, which is used for
+analyzing spectroscopic data, and the \package{regions} coordinated
+package, which is used to handle geometric regions. The \package{gwcs}
+affiliated package for generalized world coordinate systems was created
+specifically to handle the complex world coordinate systems needed
+for JWST spectroscopic data. Further, the \package{ASDF} (Advanced
+Scientific Data Format) package, a next-generation interchange format
+for scientific data, was initially developed at STScI to serialize JWST
+WCS objects along with \astropypkg models, units, and coordinates. Over
 its mission lifetime, JWST will continue its support in developing and
@@
           \package{matplotlib} \citep{matplotlib},
-          \package{Cython} \citep{cython},
-          \package{photutils} \citep{photutils},
-          \package{specutils} \citep{specutils},
-          \package{regions} \citep{regions},
-          \package{gwcs} \citep{gwcs},
-          \package{synphot} \citep{synphot},
-          \package{ASDF} \citep{ASDF}.
+          \package{Cython} \citep{cython}.
         %   \package{SOFA} \citep{sofa},

### 2022-01-25 src/ms.tex
Add citations

@@
 
-Likewise, JWST developers have provided significant contributions to
-the \package{photutils}, \package{specutils}, and \package{regions}
-coordinated packages and the \package{gwcs} affiliated package. For
+Likewise, JWST developers have provided significant contributions
+to the \package{photutils} \citep{photutils}, \package{specutils}
+\citep{specutils}, and \package{regions} \citep{regions} coordinated
+packages and the \package{gwcs} \citep{gwcs} affiliated package. For
 example, development of the \package{photutils} coordinated package
@@
 for JWST spectroscopic data. Further, the \package{ASDF} (Advanced
-Scientific Data Format) package, a next-generation interchange format
-for scientific data, was initially developed at STScI to serialize JWST
-WCS objects along with \astropypkg models, units, and coordinates. Over
-its mission lifetime, JWST will continue its support in developing and
-maintaining these critically dependent packages.
+Scientific Data Format) package \citep{ASDF}, a next-generation
+interchange format for scientific data, was initially developed at STScI
+to serialize JWST WCS objects along with \astropypkg models, units, and
+coordinates. Over its mission lifetime, JWST will continue its support
+in developing and maintaining these critically dependent packages.
 
@@
           \package{matplotlib} \citep{matplotlib},
-          \package{Cython} \citep{cython}.
+          \package{Cython} \citep{cython},
+          \package{photutils} \citep{photutils},
+          \package{specutils} \citep{specutils},
+          \package{regions} \citep{regions},
+          \package{gwcs} \citep{gwcs},
+          \package{ASDF} \citep{ASDF}.
         %   \package{SOFA} \citep{sofa},

### 2022-01-26 src/ms.tex
Fix subpackage names
astropy roadmap

--- 2022-01-25 src/ms.tex
@@
 \begin{abstract}
-
-    To be written...
-
+The \astropy Project supports and fosters the development of open-source and openly-developed
+\python packages that provide commonly-needed functionality to the astronomical
+community.
+A key element of the \astropy Project is the core package \astropypkg, which serves as the
+foundation for more specialized projects and packages.
+In this article, we summarize key features in the core package as of the recent major
+release, version 5.0, and provide major updates for the project.
+We then discuss supporting a broader ecosystem of inter-operable packages,
+including connections with several astronomical observatories and missions.
+We also revisit the future outlook of the \astropy Project and the current
+status of Learn Astropy. 
+We conclude by raising and discussing the current and future challenges facing the project.
 \end{abstract}
@@
 
-%Briefly summarize new procedures and governance structure, new CoCo, election
-%process overview, etc.
-
-As part of the process of developing Astropy into a long-term sustainable
-product, and to improve transparency and accountability, the Project agreed to
-write down and formalize our governance structure (partly supported by explicit
-funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
-coordination meeting input was gathered from participants on what governance
-structures existed in the associated Open Source Software communities, and what
-would fit well with the needs of Astropy. This led into a ``retreat'' planned
-for March 2020, but due to the COVID-19 pandemic, this became a serious of
-virtual meetings of the ``Astropy Governance Working Group''.  This group
-drafted the APE0 document \citep{ape0}, which was then eventually ratified and
-implemented by the ``Astropy Governance Implementation Working Group'' in Fall
-2021. While the process emphasized flexibility and the ability to adapt to
-changing circumstances, it is expected that this is the framework Astropy's
-governance will operate in for at least the medium-term future.
-
-The APE0 \citep{ape0} document lays out the principles of this governance
-structure, so we refer the reader to that document for a more thorough
-description.  However here we highlight some key elements. While many of these
-principles were already de facto true or have been discussed organically (and
-have been discussed in earlier papers in this series), the APE0-based governance
-aimed to provide a single place where the community can agree as a starting
-point. With this in mind, it highlights the developer and user community of the
-Astropy Project as the ultimate sources of authority, as well as the core
-principle of ``do-ocracy'' that those who do work for the Project (be it coding,
-training, or other less concrete contributions) gain more influence on the
-outputs of the Project by virtue of their effort.  APE0 adds, however, the
-concept of ``voting members'' - a self-governed part of that community who are
-entrusted to elect the Coordination Committee. While this committee has existed
-from the inception of the project, APE0 establishes a formal voting process for
-this committee, and explicitly outlines the rights and responsibilities of the
-Coordination Committee. This role is mainly to facilitate consensus and act as
-the decision maker when other mechanisms have failed.  However, it also includes
-powers that either require central authority or secrets (e.g. passwords), but
-AOE0 also charges the Committee to devolve responsibilities and seek community
-input on these items as often as possible.
-
-The first coordinating committee election under these rules took place in fall
-2021, electing a mix of prior existing and new coordination committee members,
-and was contested in the sense of more candidates than available slots. This
-suggests the process is already working to serve the long-term interests of the
-committee to both spread the coordination effort, and to ensure it is not
-dominated by the same people for as long as the Project continues. While other,
-more fine-grained governance improvements are planned for the future, it is
-clear the foundation is now in place.
+Briefly summarize new procedures and governance structure, new CoCo, election
+process overview, etc.
 
@@
 
-Overview of roadmap and some highlights.
+The Astropy Roadmap is a document in the astropy-project repository that captures high level actionable items that the Astropy project aims to undertake to improve the health and stability of the  project.
+It is a static document that is revisited regularly at the Astropy coordination meetings, to keep track of progress and write new versions as needed.
+There is a related project board linked to the Roadmap document, that holds specific issues and efforts related to Roadmap items.
+The project board is a living document that is continually updated as work is planned, assigned, and completed.
+
+All items in the Roadmap have been agreed to be priorities for the Astropy Project, and are color-coded based on resources (both time/effort and developers/expertise) needed to complete the item.
+Green items are well underway, have sufficient resources/support and a plan in place for completion.
+Orange items are well defined, and work for aquiring sufficient resources underway.
+Red items do not yet have a plan for implementation and need more resources.
+
+The Astropy Roadmap originates from the March 2021 Astropy Coordination meeting where it was first drafted before being handed off to a newly formed Astropy Roadmap working group for completion.
+The Roadmap in its currant form was adopted via pull request in December 2021.
 

### 2022-01-26 src/ms.tex
Include community-orientated infrastructure section

@@
 
-Infrastructure packages that exist to support and help others maintain their
-package infrastructure.
+The Astropy project supports the broader ecosystem by providing pre-configured 
+infrastructure packages that the community can use to support and maintain 
+their own software package infrastructure. These include tools to easily 
+generate documentation and setup automated testing, as well as provide 
+package scaffolding for new projects.
 
-This includes an overview of the sphinx-astropy tools, the pytest extensions
-maintained by astropy, and the package template.
+Sphinx is a common and useful tool for generating documentation for Python 
+packages. The Astropy project maintains a default Sphinx configuration along 
+with Astropy-specific extensions which can be easily added to community 
+projects via the \texttt{sphinx-astropy} meta package. This tool provides 
+a pre-configured Sphinx setup compatible with Astropy projects, which 
+includes several extensions useful for generating API documentation 
+(\texttt{sphinx-automodapi}), allowing for Numpy docstring parsing 
+(\texttt{numpydoc}), embedded image handling (\texttt{sphinx-gallery}; 
+\texttt{pillow}), advanced documentation testing support 
+(\texttt{pytest-doctestplus}), and providing a custom documentation theme 
+ideal for analysis packages (\texttt{astropy-sphinx-theme}).
+
+Community package testing infrastructure is supported through the 
+\texttt{pytest-astropy} meta-package, providing a unified testing framework 
+with useful extensions compatible with both Astropy- and 
+non-Astropy-affiliated community packages. This meta-package pulls in several 
+\texttt{pytest} plugins to help with custom test headers 
+(\texttt{pytest-astropy-header}), accessing remotely-hosted data files in 
+tests (\texttt{pytest-remotedata}), interoperability with documentation 
+(\texttt{pytest-doctestplus}), dangling file handle checking 
+(\texttt{pytest-openfiles}), data array comparison support in tests 
+(\texttt{pytest-arraydiff}), sub-package command-line testing support 
+(\texttt{pytest-filter-subpackage}), improved mock object testing 
+(\texttt{pytest-mock}), test coverage reports and measurements 
+(\texttt{pytest-cov}), and configuring package for property-based 
+testing (\texttt{hypothesis}).
+
+The Astropy Package Template helps facilitate the setup and creation of new 
+Python packages leveraging the Astropy ecosystem. This tool utilizes the 
+Cookiecutter project to walk users through the process of creating new 
+packages complete with documentation and testing support. Additionally, the 
+package template generation process includes the ability to setup 
+interoperability with GitHub, allowing for easy repository access from 
+documentation, as well as an example GitHub Actions workflow to demonstrate 
+the use of GitHub's continuous integration tooling.
 
@@
 
-The Astropy Roadmap is a document in the astropy-project repository that captures high level actionable items that the Astropy project aims to undertake to improve the health and stability of the  project.
-It is a static document that is revisited regularly at the Astropy coordination meetings, to keep track of progress and write new versions as needed.
-There is a related project board linked to the Roadmap document, that holds specific issues and efforts related to Roadmap items.
-The project board is a living document that is continually updated as work is planned, assigned, and completed.
-
-All items in the Roadmap have been agreed to be priorities for the Astropy Project, and are color-coded based on resources (both time/effort and developers/expertise) needed to complete the item.
-Green items are well underway, have sufficient resources/support and a plan in place for completion.
-Orange items are well defined, and work for aquiring sufficient resources underway.
-Red items do not yet have a plan for implementation and need more resources.
-
-The Astropy Roadmap originates from the March 2021 Astropy Coordination meeting where it was first drafted before being handed off to a newly formed Astropy Roadmap working group for completion.
-The Roadmap in its currant form was adopted via pull request in December 2021.
+Overview of roadmap and some highlights.
 

### 2022-01-27 src/ms.tex
Expanded a little the list of new features to highlight
Clarify WCS idea
lots of edits.

@@
 
-Major versions of the core package -- that is to say versions which add and/or
-modify functionality -- are released approximately every six months, and are
-then maintained with releases that fix issues, until the next major version
-is released. However, every two years a major release is designated as a
-long-term support (LTS) release which continues to be maintained for up to two
-years \citep{ape2}. The motivation for LTS releases is to provide longer-term stable
-versions of astropy that users requiring a high level of stability can make
-use of if they do not always need the latest features (this could include
-for example telescope pipelines and so on). The 5.0 release of the core package was designated as LTS,
-and since it was released at the end of 2021, it will be maintained until the
-end of 2023.
+Summarize a few notable things in v5.0 as compared to the last LTS release.
 
@@
 
-\secauthor{Lía Corrales, David Shupe + Learn team}
+\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
+
+\subsubsection{Current status and scope}
+
+{\it Learn Astropy} is an umbrella term that acknowledges the broad educational efforts made by the Astropy Project, which are led by the Learn Team.
+The efforts focus on developing online content and workshops covering astro-specific coding tasks in Python.
+As described in \textsl{[Astropy-ii paper]}, There are four different types of Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook lessons that are published in HTML format online; \textit{guides}, which are a series of lessons providing a foundational resource for performing a certain type of astronomical analyses; \textit{examples}, which are snippets of code that showcase a short task that can be performed with Astropy or an affiliated package; and \textit{documentation}, which is contained within the code base.
+This categorization drives content development, infrastructure choices, and the appearance of the Learn Astropy website.
+The Learn Team meets weekly to work on creating, expanding, improving these educational resources.
 
 \begin{itemize}
-\item Current status and scope
-\item Vision for the future (review what we said in v2.0 paper)
-\item User forums and engagement with user base
-\item Workshops
+
+\item {\bf Learn Astropy website:} The website has a new search interface powered by
+Algolia which is used by major e-commerce websites. This work has been
+supported in part by the Dunlap grant.
+
+\item \textbf{Tutorials} Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
+
+\item \textbf{Guide} {\it ccdproc} Jupyter book has been completed.
+
+\item {\bf Workshops:} the Project has been conducting workshops at winter meetings
+of the American Astronomical Society since AAS 225 in January 2015. Up to the
+start of the coronavirus pandemic, these were full-day in-person workshops with
+as many as 90 participants and a dozen facilitators from the project. During the
+pandemic, these workshops were moved to an online format and split into
+basic and advanced sessions. Additionally, beginning with the AAS 238 online
+meeting, the workshops have been expanded to Summer AAS meetings. The Learn
+team finds that the workshop audience is best found as AAS meetings as opposed
+to more general Python meetings.
+
+The Astropy Project provided another mode of community engagement at AAS Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the AAS Exhibit Hall.
+Funding for the exhibit hall was provided alternately by NumFOCUS and later by the Moore Foundation funding (\textbf{check this}).
+The booth hosted a series of Q\&A special sessions during AAS 235 and webinars during the virtual AAS 237 meetings, to provide the general astronomy community information and access to experts on a variety of open source astronomical tools.
 \end{itemize}
 
+\subsubsection{Learn vision for the future}
+
+The Learn plan going forward is to continue to improve the tutorial website and facilate new content;
+to index third-party tutorials; and to look for supportable opportunities to expand the reach of
+Astropy workshops beyond the American Astronomical Society meetings.
+
 \subsection{Current and Future Challenges}
+
+\section{Community Engagement}
+
+\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+open-source projets.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+a Discourse site which is more open than the Facebook group and more user-oriented
+than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+developers are able to invoice as independent contractors the time they
+spend helping users on these forums.
 

### 2022-01-27 src/ms.tex
first commit: IDE section outline

--- 2022-01-27 src/ms.tex
@@
 \secauthor{Lía Corrales}
+
+\begin{itemize}
+    \item Overview of Moore support for IDE programs, Astropy philosophy for selecting such works % Need help from CoCo or finance team to write
+    \item Astropy participation in NSBP and SACNAS % Lia can write with help of summaries in astropy-project repo
+    \item Funded IDE proposal: Outreachy % Some one from outreachy? Brigitta?
+    \item Funded IDE proposal: WoCCode % Lia write
+    \item NumFOCUS ``Contributor Diversification \& Retention'' (CDR) participation from Astropy % Lia and Adrian co-write?
+\end{itemize}
 

### 2022-01-27 src/ms.tex
Apply text about website and search from Jonathan
Co-authored-by: Jonathan Sick <jonathan@jsick.codes>

--- 2022-01-27 src/ms.tex
@@
 
-\item {\bf Learn Astropy website:} The website has a new search interface powered by
-Algolia which is used by major e-commerce websites. This work has been
-supported in part by the Dunlap grant.
+\item {\bf Learn Astropy website:} The Learn Team re-launched the website in 2021 with a new infrastructure platform, built around full-text search and interactive filtering functionality, with the goal of making content discoverable as the Learn Astropy content catalog expands.
+This work has been supported in part by the Dunlap grant.
+We have adopted Algolia, a search-as-a-service cloud platform, to store the full-text and metadata records of Learn Astropy's content.
+The new Learn Astropy website is a JavaScript (Gatsby/React) application that uses the Algolia service to power its search and filtering user interface.
+Our Python-based application, Learn Astropy Librarian, populates data into the Algolia service.
+We tuned the Librarian around specific content formats (such as Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more accurately index content and heuristically extract metadata.
+A consequence of the new platform is that we now maintain and compile content separately from the website application itself, enabling new content types.
+Tutorials, which as Jupyter Notebook-based, are now compiled into their own Learn Astropy sub-site using nbcollection.
+Guides, which are Jupyter Book-based, are also deployed as separate websites using GitHub pages.
+This architecture opens future possibilities of indexing third-party content, hosted elsewhere, such as on institutional websites.
 

### 2022-01-27 src/ms.tex
Add first version of CTA section

@@
 
-Infrastructure packages that exist to support and help others maintain their
-package infrastructure.
+The Astropy project supports the broader ecosystem by providing pre-configured 
+infrastructure packages that the community can use to support and maintain 
+their own software package infrastructure. These include tools to easily 
+generate documentation and setup automated testing, as well as provide 
+package scaffolding for new projects.
 
-This includes an overview of the sphinx-astropy tools, the pytest extensions
-maintained by astropy, and the package template.
+Sphinx is a common and useful tool for generating documentation for Python
+packages. The Astropy project maintains a default Sphinx configuration along
+with Astropy-specific extensions which can be easily added to community
+projects via the
+\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}} meta
+package. This tool provides a pre-configured Sphinx setup compatible with
+Astropy projects, which includes several extensions useful for generating API
+documentation
+(\href{https://github.com/astropy/sphinx-automodapi}{\texttt{sphinx-automodapi}}),
+allowing for Numpy docstring parsing
+(\href{https://github.com/numpy/numpydoc}{\texttt{numpydoc}}), embedded image
+handling
+(\href{https://github.com/sphinx-gallery/sphinx-gallery}{\texttt{sphinx-gallery}};
+\href{https://github.com/python-pillow}{\texttt{pillow}}), advanced
+documentation testing support
+(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
+and providing a custom documentation theme ideal for analysis packages
+(\href{https://github.com/astropy/sphinx-astropy}{\texttt{astropy-sphinx-theme}}).
+
+Community package testing infrastructure is supported through the
+\href{https://github.com/astropy/pytest-astropy}{\texttt{pytest-astropy}}
+meta-package, providing a unified testing framework with useful extensions
+compatible with both Astropy- and non-Astropy-affiliated community packages.
+This meta-package pulls in several
+\href{https://github.com/pytest-dev/pytest}{\texttt{pytest}} plugins to help
+with custom test headers
+(\href{https://github.com/astropy/pytest-astropy-header}{\texttt{pytest-astropy-header}}),
+accessing remotely-hosted data files in tests
+(\href{https://github.com/astropy/pytest-remotedata}{\texttt{pytest-remotedata}}),
+interoperability with documentation
+(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
+dangling file handle checking
+(\href{https://github.com/astropy/pytest-openfiles}{\texttt{pytest-openfiles}}),
+data array comparison support in tests
+(\href{https://github.com/astropy/pytest-arraydiff}{\texttt{pytest-arraydiff}}),
+sub-package command-line testing support
+(\href{https://github.com/astropy/pytest-filter-subpackage}{\texttt{pytest-filter-subpackage}}),
+improved mock object testing
+(\href{https://github.com/pytest-dev/pytest-mock}{\texttt{pytest-mock}}), test
+coverage reports and measurements
+(\href{https://github.com/pytest-dev/pytest-cov}{\texttt{pytest-cov}}), and
+configuring package for property-based testing
+(\href{https://github.com/HypothesisWorks/hypothesis}{\texttt{hypothesis}}).
+
+The Astropy Package Template helps facilitate the setup and creation of new 
+Python packages leveraging the Astropy ecosystem. This tool utilizes the 
+Cookiecutter project to walk users through the process of creating new 
+packages complete with documentation and testing support. Additionally, the 
+package template generation process includes the ability to setup 
+interoperability with GitHub, allowing for easy repository access from 
+documentation, as well as an example GitHub Actions workflow to demonstrate 
+the use of GitHub's continuous integration tooling.
 
@@
 Cherenkov Telescope Array
-\secauthor{Axel Donath, Max Noethe}
+\secauthor{Axel Donath, Maximilian Nöthe}
+
+The Cherenkov Telescope (CTA) will be the next generation very-high-energy
+gamma-ray observatory.
+It will improve over the current generation of imaging atmospheric Cherenkov telescopes (IACTs)
+by a factor of five to ten in sensitivity and it will be able to observe the whole sky from a combination of two sites:
+a northern site in La Palma, Spain, and a southern one in Paranal, Chile.
+CTA will be able to observe gamma rays in the energy range between $20\,\mathrm{GeV}$ and $300\,\mathrm{GeV}$
+using three different types of telescopes, in total over 100 telescopes are planned at the two sites.
+CTA will also be the first open gamma-ray observatory.
+
+The data analysis pipeline is developed as open source software and essentially split in two domains:
+\begin{enumerate}
+  \item In the low-level analysis, the properties of the recorded air shower events
+    have to be estimated from the raw data.
+    The raw data consists of very short (\textasciitilde $40\,\mathrm{ns}$) videos recorded with the fast and
+    sensitive cameras of the telescopes.
+    This includes the energy, particle type and direction of origin of the particle that induced the air shower
+    and the time the shower was recorded.
+  \item In the higher-level analysis, these reconstructed event lists are used together with the
+    characterization of the measurement process and the low-level analysis to perform the actual scientific analysis.
+    This includes energy spectra, light curves, sky maps and many more.
+\end{enumerate}
+
+A prototype for the low-level analysis is ctapipe~\cite{ctapipe},
+a python package developed to perform all the necessary tasks to from the raw data
+of Cherenkov telescopes to the reconstructed event lists.
+
+The high-level analysis will be based on the Gammapy~\cite{gammapy} library, which
+will be the core of the CTA science tools.
+
+Both Gammapy and \texttt{ctapipe} are using astropy heavily, mainly for units, coordinate transformations,
+tables and FITS IO.
+
+As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
+it needs to perform a large number of coordinate transformations.
+To enable this, CTA member M.~Nöthe contributed a major performance improvement
+for large numbers of coordinates with different observation times,
+based on earlier work by B.~Winkel.
+
+Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
+a total of 107 merged pull requests were contributed to ctapipe.
 
@@
 
-{\it Learn Astropy} is an umbrella term that acknowledges the broad educational efforts made by the Astropy Project, which are led by the Learn Team.
-The efforts focus on developing online content and workshops covering astro-specific coding tasks in Python.
-As described in \textsl{[Astropy-ii paper]}, There are four different types of Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook lessons that are published in HTML format online; \textit{guides}, which are a series of lessons providing a foundational resource for performing a certain type of astronomical analyses; \textit{examples}, which are snippets of code that showcase a short task that can be performed with Astropy or an affiliated package; and \textit{documentation}, which is contained within the code base.
-This categorization drives content development, infrastructure choices, and the appearance of the Learn Astropy website.
-The Learn Team meets weekly to work on creating, expanding, improving these educational resources.
+{\it Learn Astropy} is an umbrella term that acknowledges the broad educational
+efforts made by the Astropy Project, which are led by the Learn Team.
+The efforts focus on developing online content and workshops covering
+astro-specific coding tasks in Python.
+As described in \citet{astropy:2018}, There are four different types of
+Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
+lessons that are published in HTML format online; \textit{guides}, which are a
+series of lessons providing a foundational resource for performing certain
+type of astronomical analyses; \textit{examples}, which are snippets of code
+that showcase a short task that can be performed with Astropy or an affiliated
+package; and \textit{documentation}, which is contained within the code base.
+This categorization drives content development, infrastructure choices, and the
+appearance of the Learn Astropy website.
+The Learn Team meets weekly to work on creating, expanding, improving these
+educational resources.
 
@@
 
-\item {\bf Learn Astropy website:} The Learn Team re-launched the website in 2021 with a new infrastructure platform, built around full-text search and interactive filtering functionality, with the goal of making content discoverable as the Learn Astropy content catalog expands.
+\item {\bf Learn Astropy website:} The Learn Team re-launched the website in
+2021 with a new infrastructure platform, built around full-text search and
+interactive filtering functionality, with the goal of making content
+discoverable as the Learn Astropy content catalog expands.
 This work has been supported in part by the Dunlap grant.
-We have adopted Algolia, a search-as-a-service cloud platform, to store the full-text and metadata records of Learn Astropy's content.
-The new Learn Astropy website is a JavaScript (Gatsby/React) application that uses the Algolia service to power its search and filtering user interface.
-Our Python-based application, Learn Astropy Librarian, populates data into the Algolia service.
-We tuned the Librarian around specific content formats (such as Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more accurately index content and heuristically extract metadata.
-A consequence of the new platform is that we now maintain and compile content separately from the website application itself, enabling new content types.
-Tutorials, which as Jupyter Notebook-based, are now compiled into their own Learn Astropy sub-site using nbcollection.
-Guides, which are Jupyter Book-based, are also deployed as separate websites using GitHub pages.
-This architecture opens future possibilities of indexing third-party content, hosted elsewhere, such as on institutional websites.
+We have adopted Algolia, a search-as-a-service cloud platform, to store the
+full-text and metadata records of Learn Astropy's content.
+The new Learn Astropy website is a JavaScript (Gatsby/React) application that
+uses the Algolia service to power its search and filtering user interface.
+Our Python-based application, Learn Astropy Librarian, populates data into the
+Algolia service. We tuned the Librarian around specific content formats (such as
+Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
+accurately index content and heuristically extract metadata.
+A consequence of the new platform is that we now maintain and compile content
+separately from the website application itself, enabling new content types.
+Tutorials, which as Jupyter Notebook-based, are now compiled into their own
+Learn Astropy sub-site using nbcollection.
+Guides, which are Jupyter Book-based, are also deployed as separate websites
+using GitHub pages. This architecture opens future possibilities of indexing
+third-party content, hosted elsewhere, such as on institutional websites.
 
@@
 
-\item {\bf Workshops:} the Project has been conducting workshops at winter meetings
-of the American Astronomical Society since AAS 225 in January 2015. Up to the
-start of the coronavirus pandemic, these were full-day in-person workshops with
-as many as 90 participants and a dozen facilitators from the project. During the
-pandemic, these workshops were moved to an online format and split into
-basic and advanced sessions. Additionally, beginning with the AAS 238 online
-meeting, the workshops have been expanded to Summer AAS meetings. The Learn
-team finds that the workshop audience is best found as AAS meetings as opposed
-to more general Python meetings.
+\item {\bf Workshops:} the Project has been conducting workshops at winter
+meetings of the American Astronomical Society since AAS 225 in January 2015.
+Up to the start of the coronavirus pandemic, these were full-day in-person
+workshops with as many as 90 participants and a dozen facilitators from the
+project.
+During the pandemic, these workshops were moved to an online format and split
+into basic and advanced sessions.
+Additionally, beginning with the AAS 238 online meeting, the workshops have been
+expanded to Summer AAS meetings.
+The Learn team finds that the workshop audience is best found as AAS meetings as
+opposed to more general Python meetings.
 
-The Astropy Project provided another mode of community engagement at AAS Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the AAS Exhibit Hall.
-Funding for the exhibit hall was provided alternately by NumFOCUS and later by the Moore Foundation funding (\textbf{check this}).
-The booth hosted a series of Q\&A special sessions during AAS 235 and webinars during the virtual AAS 237 meetings, to provide the general astronomy community information and access to experts on a variety of open source astronomical tools.
+The Astropy Project provided another mode of community engagement at AAS
+Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
+AAS Exhibit Hall.
+Funding for the exhibit hall was provided alternately by NumFOCUS and later by
+the Moore Foundation funding (\textbf{check this}).
+The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
+during the virtual AAS 237 meetings, to provide the general astronomy community
+information and access to experts on a variety of open source astronomical
+tools.
 \end{itemize}
@@
 
-The Learn plan going forward is to continue to improve the tutorial website and facilate new content;
-to index third-party tutorials; and to look for supportable opportunities to expand the reach of
-Astropy workshops beyond the American Astronomical Society meetings.
+The Learn plan going forward is to continue to improve the tutorial website and
+facilate new content;
+to index third-party tutorials; and to look for supportable opportunities to
+expand the reach of Astropy workshops beyond the American Astronomical Society
+meetings.
 
@@
 
+\begin{itemize}
 \item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
@@
 spend helping users on these forums.
-
+\end{itemize}
 \secunfilled

### 2022-01-28 src/ms.tex
Implement suggestions by Axel
Co-authored-by: Axel Donath <axel.donath@mpi-hd.mpg.de>

@@
 a northern site in La Palma, Spain, and a southern one in Paranal, Chile.
-CTA will be able to observe gamma rays in the energy range between $20\,\mathrm{GeV}$ and $300\,\mathrm{GeV}$
+CTA will be able to observe gamma rays in the energy range between $20\,\mathrm{GeV}$ and $300\,\mathrm{TeV}$
 using three different types of telescopes, in total over 100 telescopes are planned at the two sites.
@@
     and the time the shower was recorded.
-  \item In the higher-level analysis, these reconstructed event lists are used together with the
-    characterization of the measurement process and the low-level analysis to perform the actual scientific analysis.
-    This includes energy spectra, light curves, sky maps and many more.
+  \item In the higher-level analysis, these reconstructed event lists are used together with some
+    characterization of the instrument response to perform the actual scientific analysis.
+    This software will be delivered as CTA science tools to the future users of the observatory.
 \end{enumerate}
@@
 
-The high-level analysis will be based on the Gammapy~\cite{gammapy} library, which
-will be the core of the CTA science tools.
+The high level analysis (or CTA science tools) will be based on the Astropy affiliated package
+Gammapy~\cite{gammapy}.
 
@@
 Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
-a total of 107 merged pull requests were contributed to ctapipe.
+a total of 107 merged pull requests were contributed to astropy.
 

### 2022-02-01 src/ms.tex
Mention times in CTA section
Apply suggestions from code review
Co-authored-by: P. L. Lim <2090236+pllim@users.noreply.github.com>

--- 2022-01-27 src/ms.tex
@@
 
-Briefly summarize new procedures and governance structure, new CoCo, election
-process overview, etc.
+%Briefly summarize new procedures and governance structure, new CoCo, election
+%process overview, etc.
+
+As part of the process of developing Astropy into a long-term sustainable
+product, and to improve transparency and accountability, the Project agreed to
+write down and formalize our governance structure (partly supported by explicit
+funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
+Astropy Coordination Meeting, input was gathered from participants on what governance
+structures existed in the associated Open Source Software communities, and what
+would fit well with the needs of Astropy. This led into a ``retreat'' planned
+for March 2020, but due to the COVID-19 pandemic, this became a series of
+virtual meetings of the ``Astropy Governance Working Group''.  This group
+drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
+implemented by the ``Astropy Governance Implementation Working Group'' in Fall
+2021. While the process emphasized flexibility and the ability to adapt to
+changing circumstances, it is expected that this is the framework Astropy's
+governance will operate in for at least the medium-term future.
+
+The APE0 \citep{ape0} document lays out the principles of this governance
+structure, so we refer the reader to that document for a more thorough
+description.  However here we highlight some key elements. While many of these
+principles were already de facto true or have been discussed organically (and
+have been discussed in earlier papers in this series), the APE0-based governance
+aimed to provide a single place where the community can agree as a starting
+point. With this in mind, it highlights the developer and user community of the
+Astropy Project as the ultimate sources of authority, as well as the core
+principle of ``do-ocracy'' that those who do work for the Project (be it coding,
+training, or other less concrete contributions) gain more influence on the
+outputs of the Project by virtue of their effort.  APE0 adds, however, the
+concept of ``voting members'' - a self-governed part of that community who are
+entrusted to elect the Coordination Committee. While this committee has existed
+from the inception of the project, APE0 establishes a formal voting process for
+this committee, and explicitly outlines the rights and responsibilities of the
+Coordination Committee. This role is mainly to facilitate consensus and act as
+the decision maker when other mechanisms have failed.  However, it also includes
+powers that either require central authority or secrets (e.g., passwords), but
+AOE0 also charges the Committee to devolve responsibilities and seek community
+input on these items as often as possible.
+
+The first Coordination Committee election under these rules took place in Fall
+2021, electing a mix of prior existing and new coordination committee members,
+and was contested in the sense of more candidates than available slots. This
+suggests the process is already working to serve the long-term interests of the
+committee to both spread the coordination effort, and to ensure it is not
+dominated by the same people for as long as the Project continues. While other,
+more fine-grained governance improvements are planned for the future, it is
+clear the foundation is now in place.
 
@@
 
-\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
-
-\subsubsection{Current status and scope}
-
-{\it Learn Astropy} is an umbrella term that acknowledges the broad educational efforts made by the Astropy Project, which are led by the Learn Team.
-The efforts focus on developing online content and workshops covering astro-specific coding tasks in Python.
-As described in \textsl{[Astropy-ii paper]}, There are four different types of Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook lessons that are published in HTML format online; \textit{guides}, which are a series of lessons providing a foundational resource for performing a certain type of astronomical analyses; \textit{examples}, which are snippets of code that showcase a short task that can be performed with Astropy or an affiliated package; and \textit{documentation}, which is contained within the code base.
-This categorization drives content development, infrastructure choices, and the appearance of the Learn Astropy website.
-The Learn Team meets weekly to work on creating, expanding, improving these educational resources.
+\secauthor{Lía Corrales, David Shupe + Learn team}
 
 \begin{itemize}
-
-\item {\bf Learn Astropy website:} The Learn Team re-launched the website in 2021 with a new infrastructure platform, built around full-text search and interactive filtering functionality, with the goal of making content discoverable as the Learn Astropy content catalog expands.
-This work has been supported in part by the Dunlap grant.
-We have adopted Algolia, a search-as-a-service cloud platform, to store the full-text and metadata records of Learn Astropy's content.
-The new Learn Astropy website is a JavaScript (Gatsby/React) application that uses the Algolia service to power its search and filtering user interface.
-Our Python-based application, Learn Astropy Librarian, populates data into the Algolia service.
-We tuned the Librarian around specific content formats (such as Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more accurately index content and heuristically extract metadata.
-A consequence of the new platform is that we now maintain and compile content separately from the website application itself, enabling new content types.
-Tutorials, which as Jupyter Notebook-based, are now compiled into their own Learn Astropy sub-site using nbcollection.
-Guides, which are Jupyter Book-based, are also deployed as separate websites using GitHub pages.
-This architecture opens future possibilities of indexing third-party content, hosted elsewhere, such as on institutional websites.
-
-\item \textbf{Tutorials} Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
-
-\item \textbf{Guide} {\it ccdproc} Jupyter book has been completed.
-
-\item {\bf Workshops:} the Project has been conducting workshops at winter meetings
-of the American Astronomical Society since AAS 225 in January 2015. Up to the
-start of the coronavirus pandemic, these were full-day in-person workshops with
-as many as 90 participants and a dozen facilitators from the project. During the
-pandemic, these workshops were moved to an online format and split into
-basic and advanced sessions. Additionally, beginning with the AAS 238 online
-meeting, the workshops have been expanded to Summer AAS meetings. The Learn
-team finds that the workshop audience is best found as AAS meetings as opposed
-to more general Python meetings.
-
-The Astropy Project provided another mode of community engagement at AAS Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the AAS Exhibit Hall.
-Funding for the exhibit hall was provided alternately by NumFOCUS and later by the Moore Foundation funding (\textbf{check this}).
-The booth hosted a series of Q\&A special sessions during AAS 235 and webinars during the virtual AAS 237 meetings, to provide the general astronomy community information and access to experts on a variety of open source astronomical tools.
+\item Current status and scope
+\item Vision for the future (review what we said in v2.0 paper)
+\item User forums and engagement with user base
+\item Workshops
 \end{itemize}
 
-\subsubsection{Learn vision for the future}
-
-The Learn plan going forward is to continue to improve the tutorial website and facilate new content;
-to index third-party tutorials; and to look for supportable opportunities to expand the reach of
-Astropy workshops beyond the American Astronomical Society meetings.
-
 \subsection{Current and Future Challenges}
-
-\section{Community Engagement}
-
-\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
-This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
-The Moore Foundation grant allows this Slack space to
-be on a paid plan; additionally NumFOCUS has negotiated a special rate for
-open-source projets.
-While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
-The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
-a Discourse site which is more open than the Facebook group and more user-oriented
-than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
-developers are able to invoice as independent contractors the time they
-spend helping users on these forums.
 

### 2022-02-01 src/ms.tex
Update src/ms.tex
Co-authored-by: Nicholas Earl <contact@nicholasearl.me>

@@
 powers that either require central authority or secrets (e.g., passwords), but
-AOE0 also charges the Committee to devolve responsibilities and seek community
+APE0 also charges the Committee to devolve responsibilities and seek community
 input on these items as often as possible.

### 2022-02-01 src/ms.tex
hard-wrap lines of text

@@
 
-%Briefly summarize new procedures and governance structure, new CoCo, election
-%process overview, etc.
-
-As part of the process of developing Astropy into a long-term sustainable
-product, and to improve transparency and accountability, the Project agreed to
-write down and formalize our governance structure (partly supported by explicit
-funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
-Astropy Coordination Meeting, input was gathered from participants on what governance
-structures existed in the associated Open Source Software communities, and what
-would fit well with the needs of Astropy. This led into a ``retreat'' planned
-for March 2020, but due to the COVID-19 pandemic, this became a series of
-virtual meetings of the ``Astropy Governance Working Group''.  This group
-drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
-implemented by the ``Astropy Governance Implementation Working Group'' in Fall
-2021. While the process emphasized flexibility and the ability to adapt to
-changing circumstances, it is expected that this is the framework Astropy's
-governance will operate in for at least the medium-term future.
-
-The APE0 \citep{ape0} document lays out the principles of this governance
-structure, so we refer the reader to that document for a more thorough
-description.  However here we highlight some key elements. While many of these
-principles were already de facto true or have been discussed organically (and
-have been discussed in earlier papers in this series), the APE0-based governance
-aimed to provide a single place where the community can agree as a starting
-point. With this in mind, it highlights the developer and user community of the
-Astropy Project as the ultimate sources of authority, as well as the core
-principle of ``do-ocracy'' that those who do work for the Project (be it coding,
-training, or other less concrete contributions) gain more influence on the
-outputs of the Project by virtue of their effort.  APE0 adds, however, the
-concept of ``voting members'' - a self-governed part of that community who are
-entrusted to elect the Coordination Committee. While this committee has existed
-from the inception of the project, APE0 establishes a formal voting process for
-this committee, and explicitly outlines the rights and responsibilities of the
-Coordination Committee. This role is mainly to facilitate consensus and act as
-the decision maker when other mechanisms have failed.  However, it also includes
-powers that either require central authority or secrets (e.g., passwords), but
-APE0 also charges the Committee to devolve responsibilities and seek community
-input on these items as often as possible.
-
-The first Coordination Committee election under these rules took place in Fall
-2021, electing a mix of prior existing and new coordination committee members,
-and was contested in the sense of more candidates than available slots. This
-suggests the process is already working to serve the long-term interests of the
-committee to both spread the coordination effort, and to ensure it is not
-dominated by the same people for as long as the Project continues. While other,
-more fine-grained governance improvements are planned for the future, it is
-clear the foundation is now in place.
+Briefly summarize new procedures and governance structure, new CoCo, election
+process overview, etc.
 
@@
 
-\secauthor{Lía Corrales, David Shupe + Learn team}
+\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
+
+\subsubsection{Current status and scope}
+
+{\it Learn Astropy} is an umbrella term that acknowledges the broad educational
+efforts made by the Astropy Project, which are led by the Learn Team.
+The efforts focus on developing online content and workshops covering
+astro-specific coding tasks in Python.
+As described in \textsl{[Astropy-ii paper]}, There are four different types of
+Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
+lessons that are published in HTML format online; \textit{guides}, which are a
+series of lessons providing a foundational resource for performing a certain
+type of astronomical analyses; \textit{examples}, which are snippets of code
+that showcase a short task that can be performed with Astropy or an affiliated
+package; and \textit{documentation}, which is contained within the code base.
+This categorization drives content development, infrastructure choices, and the
+appearance of the Learn Astropy website.
+The Learn Team meets weekly to work on creating, expanding, improving these
+educational resources.
 
 \begin{itemize}
-\item Current status and scope
-\item Vision for the future (review what we said in v2.0 paper)
-\item User forums and engagement with user base
-\item Workshops
+
+\item {\bf Learn Astropy website:} The Learn Team re-launched the website in
+2021 with a new infrastructure platform, built around full-text search and
+interactive filtering functionality, with the goal of making content
+discoverable as the Learn Astropy content catalog expands.
+This work has been supported in part by the Dunlap grant.
+We have adopted Algolia, a search-as-a-service cloud platform, to store the
+full-text and metadata records of Learn Astropy's content.
+The new Learn Astropy website is a JavaScript (Gatsby/React) application that
+uses the Algolia service to power its search and filtering user interface.
+Our Python-based application, Learn Astropy Librarian, populates data into the
+Algolia service. We tuned the Librarian around specific content formats (such as
+Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
+accurately index content and heuristically extract metadata.
+A consequence of the new platform is that we now maintain and compile content
+separately from the website application itself, enabling new content types.
+Tutorials, which as Jupyter Notebook-based, are now compiled into their own
+Learn Astropy sub-site using nbcollection.
+Guides, which are Jupyter Book-based, are also deployed as separate websites
+using GitHub pages. This architecture opens future possibilities of indexing
+third-party content, hosted elsewhere, such as on institutional websites.
+
+\item \textbf{Tutorials} Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
+
+\item \textbf{Guide} {\it ccdproc} Jupyter book has been completed.
+
+\item {\bf Workshops:} the Project has been conducting workshops at winter
+meetings of the American Astronomical Society since AAS 225 in January 2015.
+Up to the start of the coronavirus pandemic, these were full-day in-person
+workshops with as many as 90 participants and a dozen facilitators from the
+project.
+During the pandemic, these workshops were moved to an online format and split
+into basic and advanced sessions.
+Additionally, beginning with the AAS 238 online meeting, the workshops have been
+expanded to Summer AAS meetings.
+The Learn team finds that the workshop audience is best found as AAS meetings as
+opposed to more general Python meetings.
+
+The Astropy Project provided another mode of community engagement at AAS
+Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
+AAS Exhibit Hall.
+Funding for the exhibit hall was provided alternately by NumFOCUS and later by
+the Moore Foundation funding (\textbf{check this}).
+The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
+during the virtual AAS 237 meetings, to provide the general astronomy community
+information and access to experts on a variety of open source astronomical
+tools.
 \end{itemize}
 
+\subsubsection{Learn vision for the future}
+
+The Learn plan going forward is to continue to improve the tutorial website and
+facilate new content;
+to index third-party tutorials; and to look for supportable opportunities to
+expand the reach of Astropy workshops beyond the American Astronomical Society
+meetings.
+
 \subsection{Current and Future Challenges}
+
+\section{Community Engagement}
+
+\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+open-source projets.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+a Discourse site which is more open than the Facebook group and more user-oriented
+than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+developers are able to invoice as independent contractors the time they
+spend helping users on these forums.
 

### 2022-02-03 src/ms.tex
Apply suggestions from numeral
Co-authored-by: Nicholas Earl <contact@nicholasearl.me>

--- 2022-02-01 src/ms.tex
@@
 astro-specific coding tasks in Python.
-As described in \textsl{[Astropy-ii paper]}, There are four different types of
+As described in \citet{astropy:2018}, There are four different types of
 Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
 lessons that are published in HTML format online; \textit{guides}, which are a
-series of lessons providing a foundational resource for performing a certain
+series of lessons providing a foundational resource for performing certain
 type of astronomical analyses; \textit{examples}, which are snippets of code

### 2022-02-07 src/ms.tex
First draft of IDE section

--- 2022-02-01 src/ms.tex
@@
 
+With support from the Moore Foundation, the \astropy Project was
+allocated funding to support mentoring programs. In 2020 a call was
+made to submit proposals for IDE (Inclusion, Diversity, and
+Empowerment) initiatives for project-wide consideration on
+Github. This process was deemed the most ``open'' because it allowed
+for community wide feedback to focus and improve proposal
+initiatives. Two programs were selected via this process, described
+below.
+
+\textbf{Outreachy:} Describe the outreachy project
+
+\textbf{Women of Color Code (WoCCode)} is a peer-mentoring network for
+coders from traditionally marginalized groups, most notably women of
+color. Participants were invited to the WoCCode Slack space and
+encouraged to attend monthly webinars to share skills in the context
+of open source software libraries. Every other month, a guest speaker
+was invited to talk about their career path and share a skill. Program
+participants were solicited in the fall of 2020, yielding 73
+applications from 17 countries (48\% from the United States, 37\% from
+Africa, and 15\% from remaining continents). We selected participants
+who we identified as having a high potential for contributing to open
+source projects: intermediate to advanced programming skills with a
+vocal interest in contributing. Of the thirty applicants invited to
+the program, nineteen joined the community on Slack. Participants were
+organized into cohorts based on interest and each cohort was assigned
+one of three mentors that were also selected via open
+application. Mentors acted as a general resource to participants, gave
+one webinar, and organized a hack day. WoCCode also supported
+registration of two participants to attend the American Astronomical
+Society virtual summer 2021 meeting.
+
+Participants rated the impact of the program as very high. In the
+final webinar, participants reported a change in their perception of
+coding in general, for example, accepting coding as something they can
+do for fun. Participants reported generally feeling comfortable asking
+questions and interacting with a community where ``[e]veryone's
+thoughts are welcomed, no one is made to feel less important'' and
+where one can enjoy ``warm interactions with likeminded people.''
+WoCCode is continuing into 2022 by broadening participation in the
+Slack space and publicly advertising the guest webinar events.
+
+In addition to the supported mentoring initiatives, the \astropy
+project as a whole has taken steps to examine representation of
+marginalized groups within the project and search for avenues of
+improvement.
+
+% Much of this text was adapted from Robel Geda's report from NSBP and
+% SACNAS. Please include him on the author list.
+\textbf{\astropy representation at national diversity conferences:}
+Several members of the \astropy community attended virtual conferences
+of the National Society of Black Physicists (NSBP) and SACNAS
+(focusing on Hispanic and Native American scientists across all STEM
+fields). \astropy representatives noted that one underlying topic came
+up multiple times. A major barrier for persons who come from
+underrepresented communities is the lack of resources and expertise
+that are necessary to train students. Writing PEP 8 compliant software
+and understanding the Git and Github workflow is not part of the
+standard Physics and Astrophysics curriculum. Additionally, the larger
+Astronomnical community is still in the process of transitioning
+towards Python based tools, making it difficult for students not
+currently under advisement by some one with extensive Python expertise
+to get involved with the open source Python community. Projects such
+as Learn Astropy could have a profound impact by empowering
+underrepresented groups because it provides a free, searchable, and
+accessible introduction to Python tools for astrophysical
+research. Representatives also noted that offering training and
+teaching materials to PIs of Research Experience for Undergraduate
+(REU) programs throughout the United States could be helpful for
+encouraging advisors to teach students how to use Astropy, create
+their own libraries, and use version control on open source platforms
+like Github. Such materials could also be offered as a workshop at
+national diversity conferences themselves, as SACNAS solicits special
+session proposals each year.
+
+\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
+  initiative:} Describe more about this.
+
 \subsection{Funding} \label{sec:project-funding}
@@
 
-The Astropy project supports the broader ecosystem by providing pre-configured 
-infrastructure packages that the community can use to support and maintain 
-their own software package infrastructure. These include tools to easily 
-generate documentation and setup automated testing, as well as provide 
-package scaffolding for new projects.
+Infrastructure packages that exist to support and help others maintain their
+package infrastructure.
 
-Sphinx is a common and useful tool for generating documentation for Python
-packages. The Astropy project maintains a default Sphinx configuration along
-with Astropy-specific extensions which can be easily added to community
-projects via the
-\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}} meta
-package. This tool provides a pre-configured Sphinx setup compatible with
-Astropy projects, which includes several extensions useful for generating API
-documentation
-(\href{https://github.com/astropy/sphinx-automodapi}{\texttt{sphinx-automodapi}}),
-allowing for Numpy docstring parsing
-(\href{https://github.com/numpy/numpydoc}{\texttt{numpydoc}}), embedded image
-handling
-(\href{https://github.com/sphinx-gallery/sphinx-gallery}{\texttt{sphinx-gallery}};
-\href{https://github.com/python-pillow}{\texttt{pillow}}), advanced
-documentation testing support
-(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
-and providing a custom documentation theme ideal for analysis packages
-(\href{https://github.com/astropy/sphinx-astropy}{\texttt{astropy-sphinx-theme}}).
-
-Community package testing infrastructure is supported through the
-\href{https://github.com/astropy/pytest-astropy}{\texttt{pytest-astropy}}
-meta-package, providing a unified testing framework with useful extensions
-compatible with both Astropy- and non-Astropy-affiliated community packages.
-This meta-package pulls in several
-\href{https://github.com/pytest-dev/pytest}{\texttt{pytest}} plugins to help
-with custom test headers
-(\href{https://github.com/astropy/pytest-astropy-header}{\texttt{pytest-astropy-header}}),
-accessing remotely-hosted data files in tests
-(\href{https://github.com/astropy/pytest-remotedata}{\texttt{pytest-remotedata}}),
-interoperability with documentation
-(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
-dangling file handle checking
-(\href{https://github.com/astropy/pytest-openfiles}{\texttt{pytest-openfiles}}),
-data array comparison support in tests
-(\href{https://github.com/astropy/pytest-arraydiff}{\texttt{pytest-arraydiff}}),
-sub-package command-line testing support
-(\href{https://github.com/astropy/pytest-filter-subpackage}{\texttt{pytest-filter-subpackage}}),
-improved mock object testing
-(\href{https://github.com/pytest-dev/pytest-mock}{\texttt{pytest-mock}}), test
-coverage reports and measurements
-(\href{https://github.com/pytest-dev/pytest-cov}{\texttt{pytest-cov}}), and
-configuring package for property-based testing
-(\href{https://github.com/HypothesisWorks/hypothesis}{\texttt{hypothesis}}).
-
-The Astropy Package Template helps facilitate the setup and creation of new 
-Python packages leveraging the Astropy ecosystem. This tool utilizes the 
-Cookiecutter project to walk users through the process of creating new 
-packages complete with documentation and testing support. Additionally, the 
-package template generation process includes the ability to setup 
-interoperability with GitHub, allowing for easy repository access from 
-documentation, as well as an example GitHub Actions workflow to demonstrate 
-the use of GitHub's continuous integration tooling.
+This includes an overview of the sphinx-astropy tools, the pytest extensions
+maintained by astropy, and the package template.
 

### 2022-02-08 src/ms.tex

@@
 
-\secauthor{Lía Corrales, David Shupe + Learn team}
+\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
+
+\subsubsection{Current status and scope}
+
+{\it Learn Astropy} is an umbrella term that acknowledges the broad educational
+efforts made by the Astropy Project, which are led by the Learn Team.
+The efforts focus on developing online content and workshops covering
+astro-specific coding tasks in Python.
+As described in \citet{astropy:2018}, There are four different types of
+Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
+lessons that are published in HTML format online; \textit{guides}, which are a
+series of lessons providing a foundational resource for performing certain
+type of astronomical analyses; \textit{examples}, which are snippets of code
+that showcase a short task that can be performed with Astropy or an affiliated
+package; and \textit{documentation}, which is contained within the code base.
+This categorization drives content development, infrastructure choices, and the
+appearance of the Learn Astropy website.
+The Learn Team meets weekly to work on creating, expanding, improving these
+educational resources.
 
 \begin{itemize}
-\item Current status and scope
-\item Vision for the future (review what we said in v2.0 paper)
-\item User forums and engagement with user base
-\item Workshops
+
+\item {\bf Learn Astropy website:} The Learn Team re-launched the website in
+2021 with a new infrastructure platform, built around full-text search and
+interactive filtering functionality, with the goal of making content
+discoverable as the Learn Astropy content catalog expands.
+This work has been supported in part by the Dunlap grant.
+We have adopted Algolia, a search-as-a-service cloud platform, to store the
+full-text and metadata records of Learn Astropy's content.
+The new Learn Astropy website is a JavaScript (Gatsby/React) application that
+uses the Algolia service to power its search and filtering user interface.
+Our Python-based application, Learn Astropy Librarian, populates data into the
+Algolia service. We tuned the Librarian around specific content formats (such as
+Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
+accurately index content and heuristically extract metadata.
+A consequence of the new platform is that we now maintain and compile content
+separately from the website application itself, enabling new content types.
+Tutorials, which as Jupyter Notebook-based, are now compiled into their own
+Learn Astropy sub-site using nbcollection.
+Guides, which are Jupyter Book-based, are also deployed as separate websites
+using GitHub pages. This architecture opens future possibilities of indexing
+third-party content, hosted elsewhere, such as on institutional websites.
+
+\item \textbf{Tutorials} Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
+
+\item \textbf{Guide} {\it ccdproc} Jupyter book has been completed.
+
+\item {\bf Workshops:} the Project has been conducting workshops at winter
+meetings of the American Astronomical Society since AAS 225 in January 2015.
+Up to the start of the coronavirus pandemic, these were full-day in-person
+workshops with as many as 90 participants and a dozen facilitators from the
+project.
+During the pandemic, these workshops were moved to an online format and split
+into basic and advanced sessions.
+Additionally, beginning with the AAS 238 online meeting, the workshops have been
+expanded to Summer AAS meetings.
+The Learn team finds that the workshop audience is best found as AAS meetings as
+opposed to more general Python meetings.
+
+The Astropy Project provided another mode of community engagement at AAS
+Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
+AAS Exhibit Hall.
+Funding for the exhibit hall was provided alternately by NumFOCUS and later by
+the Moore Foundation funding (\textbf{check this}).
+The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
+during the virtual AAS 237 meetings, to provide the general astronomy community
+information and access to experts on a variety of open source astronomical
+tools.
 \end{itemize}
 
+\subsubsection{Learn vision for the future}
+
+The Learn plan going forward is to continue to improve the tutorial website and
+facilate new content;
+to index third-party tutorials; and to look for supportable opportunities to
+expand the reach of Astropy workshops beyond the American Astronomical Society
+meetings.
+
 \subsection{Current and Future Challenges}
+
+\section{Community Engagement}
+
+\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+open-source projets.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+a Discourse site which is more open than the Facebook group and more user-oriented
+than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+developers are able to invoice as independent contractors the time they
+spend helping users on these forums.
 

### 2022-02-08 src/ms.tex
First draft of funding section

--- 2022-02-08 src/ms.tex
@@
     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-    \item APE 14 WCS
-    \item A new SpectralCoord class for representing and transforming spectral quantities
-    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+    \item TODO: something about WCS, SpectralCoord?
 \end{itemize}
@@
 
-Summarize funding sources (Moore, NASA) and amounts and what this has been used
-for.
+% Summarize funding sources (Moore, NASA) and amounts and what this has been
+% used for.
+
+With the goal of establishing a sustainable financial model for the Astropy
+project, the Coordination Committee under the new governance charter (see \S
+\ref{sec:project-governance}) established a standing Finance committee.
+This committee overlooks the planning and disbursement of finances on behalf
+of the project. The Astropy project accepts funds from institutions as well as
+individuals through NUMFOCUS\footnote{NumFOCUS is a 501(c)(3) nonprofit that
+supports and promotes world-class, innovative, open source scientific computing.},
+the fiscal sponsor of the project. Previously, no direct financial support
+was available for the project development, and NUMFOCUS covered the incurred
+operational costs. However, transitioning to long-term financial stability meant
+having an influx of funds in the form of institutional support \citep{fundmodel},
+and hence the Astropy project applied for and successfully acquired several grants.
+The two major ones are briefly summarized below.
+
+\begin{itemize}
+    \item Moore: The Gordon and Betty Moore Foundation awarded the Astropy
+    project a \$900,000 grant in 2019 for the proposal ``Sustaining and Growing
+    the Astropy Project". Over a three-year term, this grant supported sustenance
+    of the project by providing funds for (1) the development and maintenance of
+    the \astropypkg and its infrastructure to existing project members and
+    targeted hires, (2) the transition of long-term users to contributors and
+    maintainers through mentorship efforts, (3) the formalization of a governance
+    structure (see \S \ref{sec:project-governance}), and (4) the improvement
+    of equity, diversity, and inclusion efforts in the project (see \S
+    \ref{sec:project-ide}). Some of the money supported travel for meetings,
+    conferences, and workshops. This generous grant helped reduce reliance on
+    volunteer-driven maintenance of the openly-developed package and paved
+    the way to acquire funding from federal agencies: a big milestone for the
+    success of the project.
+
+    \item NASA: A successful proposal to the NASA ROSES-2020 call, section E.7
+    (Support for Open Source Tools, Frameworks, and Libraries), provided the
+    Astropy project \$634,590 to support its work. Funded work for a term of
+    three years includes (1) project-wide infrastructure maintenance and
+    improvements, (2) targeted work on areas of the \astropy core and
+    coordinated packages that need support, (3) enhancements to the Learn
+    Astropy ecosystem and educational materials (see \S \ref{sec:learn}),
+    and (4) support for the \astropy affiliated packages. These goals
+    support the sustenance of the project by encouraging further
+    development of the package ecosystem and community engagement.
+\end{itemize}
+
+In addition to the two grants described above, the project benefited from
+a Gemini Observatory contract under its Science User Support Department,
+which funded development work to support both the Astropy and DRAGONS
+projects. The project also received a total of CAD 31,023 from the Dunlap
+Seed Funding program at the University of Toronto to develop educational
+resources for scientific software packaging within the Learn Astropy
+framework. All these successful proposals show that the Astropy project
+is vital for the astronomical infrastructure and community at large, and
+strengthen its promising and sustainable future.
 

### 2022-02-08 src/ms.tex
Fixed minor issue with a reference

@@
 operational costs. However, transitioning to long-term financial stability meant
-having an influx of funds in the form of institutional support \citep{fundmodel},
+having an influx of funds in the form of institutional support \citep{2016arXiv161003159M},
 and hence the Astropy project applied for and successfully acquired several grants.

### 2022-02-09 src/ms.tex
Implement SAPO changes

--- 2022-02-08 src/ms.tex
@@
 
-\subsubsection{James Webb Space Telescope}
+JWST
 \secauthor{Larry Bradley}
-
-The James Webb Space Telescope (JWST) is a 6.5-meter space-based
-infrared telescope that will provide unprecedented resolution and
-sensitivity from 0.6 -- 28 microns. JWST will enable a broad range
-of scientific investigations from exoplanets and their atmospheres
-to the formation of galaxies in the very early universe. Its four
-key scientific goals are to study the first light from stars and
-galaxies, the assembly and evolution of galaxies, the birth of stars and
-protoplanetary systems, and planetary systems and the origins of life.
-
-The telescope launched on an Ariane 5 rocket on 2021 December 25 from
-Kourou, French Guiana. After a series of successful deployments,
-including the sunshield and primary and secondary mirrors, JWST reached
-its orbit around the L2 Lagrange point on 2022 January 24. Commissioning
-of the telescope optics and science instruments will occur from January
-until the end of June 2022, when science operations are scheduled to
-begin.
-
-Software developers at the Space Telescope Science Institute (STScI),
-the operations center of JWST, have been developing Python-based tools
-for JWST since 2010 (starting with the JWST Calibration Reference Data
-System) and have provided major contributions to \astropy from its
-inception. The JWST instrument calibration pipelines, exposure-time
-calculators, and data analysis tools are all written in Python and
-depend on the \astropypkg core package and some coordinated and
-affiliated packages. For the \astropypkg core package, the JWST mission
-has provided extensive contributions to the \astropysubpkg{modeling},
-\astropysubpkg{units}, \astropysubpkg{coordinates}, \astropysubpkg{wcs},
-\astropysubpkg{io.fits}, \astropysubpkg{io.votable},
-\astropysubpkg{stats}, \astropysubpkg{visualization}, and
-\astropysubpkg{convolution} subpackages as well as to the general
-package infrastructure and maintenance.
-
-Likewise, JWST developers have provided significant contributions
-to the \package{photutils} \citep{photutils}, \package{specutils}
-\citep{specutils}, and \package{regions} \citep{regions} coordinated
-packages and the \package{gwcs} \citep{gwcs} and \package{synphot}
-\citep{synphot} affiliated packages. For example, development of the
-\package{photutils} coordinated package for source detection and
-photometry has largely been led by JWST contributions. JWST developers
-have also made significant contributions to the \package{specutils}
-coordinated package, which is used for analyzing spectroscopic data,
-and the \package{regions} coordinated package, which is used to
-handle geometric regions. The \package{gwcs} affiliated package for
-generalized world coordinate systems was created specifically to handle
-the complex world coordinate systems needed for JWST spectroscopic
-data. The \package{synphot} affiliated package for synthetic photometry
-was created at STScI and is a dependency of the JWST exposure-time
-calculators. Further, the \package{ASDF} (Advanced Scientific Data
-Format) package \citep{ASDF}, a next-generation interchange format for
-scientific data, was initially developed at STScI to serialize JWST WCS
-objects along with \astropypkg models, units, and coordinates. Over
-its mission lifetime, JWST will continue its support in developing and
-maintaining these critically dependent packages.
 
@@
 
-A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
+A prototype for the low-level analysis is ctapipe~\cite{ctapipe},
 a python package developed to perform all the necessary tasks to from the raw data
@@
 The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
-Gammapy~\citep{gammapy}.
+Gammapy~\cite{gammapy}.
 

### 2022-02-10 src/ms.tex
New features as itemized list

@@
 \newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
+\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
+\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
+\newcommand{\astropyunits}{\astropysubpkg{units}}
+\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
+\newcommand{\astropyconstants}{\astropysubpkg{constants}}
+\newcommand{\astropytime}{\astropysubpkg{time}}
+\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
 
 % Links to documentation: classes
-\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
-\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
-\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
-\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}
+\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
+\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}
+
+\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
+\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
+\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
+\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
+GalacticLSR}{GalacticLSR}}
+\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
+\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
+\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
+\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
+\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
+\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
+\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
+\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
+\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
+\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
+\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
+\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
+
+% suggestion: shortcut reference to Paper II
+\newcommand{\paperii}{\cite{astropy:2018}}
 
@@
 
-Summarize major pieces of development since the last paper (since v2.0). A few
-ideas (taken from What's New pages) below, but feel free to expand on or remove
-from this list:
+The \astropypkg core package is mature and stable, in that many features have
+been part of the core package for years with (purposefully) few changes to the
+software interface.
+This maturity and stability allows for the broader astronomy \python community
+to rely upon the \astropypkg core package and build specialized packages within
+the Astropy ecosystem.
+Even many non-astronomical \python libraries have come to rely on \astropypkg.
+
+The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
+% feels like ↑ could be expanded to 2 sentences
+
+% Maybe briefly outline the sectors of  
+% Using \astropypkg for
+% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
+% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
+% Astropy has grown to interoperate, e.g. NumPy
+
+Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
+
+% Summarize major pieces of development since the last paper (since v2.0). A few
+% ideas (taken from What's New pages) below, but feel free to expand on or remove
+% from this list:
+% \begin{itemize}
+%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+%     \item APE 14 WCS
+%     \item A new SpectralCoord class for representing and transforming spectral quantities
+%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+% \end{itemize}
+
+\subsubsection*{New Features} \label{sec:core-features-new}
+
 \begin{itemize}
-    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-    \item APE 14 WCS
+  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
+  \item Uncertainties and Distributions
+  \item ERFA module is now it’s own thing; Obliquely tie in that units
+    remains in astropy
+  \item ? Pytest plugin ecosystem
   \item A new SpectralCoord class for representing and transforming spectral quantities
-    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+    Velocity in Coordinates
+  \item WCS API : A shared Python interface for World Coordinate Systems
+        Include APE reference
 \end{itemize}
+
+\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
 
@@
 \secauthor{Lía Corrales}
+
+With support from the Moore Foundation, the \astropy Project was
+allocated funding to support mentoring programs. In 2020 a call was
+made to submit proposals for IDE (Inclusion, Diversity, and
+Empowerment) initiatives for project-wide consideration on
+Github. This process was deemed the most ``open'' because it allowed
+for community wide feedback to focus and improve proposal
+initiatives. Two programs were selected via this process, described
+below.
+
+\textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}
+
+\textbf{Women of Color Code (WoCCode)} is a peer-mentoring network for
+coders from traditionally marginalized groups, most notably women of
+color. Participants were invited to the WoCCode Slack space and
+encouraged to attend monthly webinars to share skills in the context
+of open source software libraries. Every other month, a guest speaker
+was invited to talk about their career path and share a skill. Program
+participants were solicited in the fall of 2020, yielding 73
+applications from 17 countries (48\% from the United States, 37\% from
+Africa, and 15\% from remaining continents). We selected participants
+who we identified as having a high potential for contributing to open
+source projects: intermediate to advanced programming skills with a
+vocal interest in contributing. Of the thirty applicants invited to
+the program, nineteen joined the community on Slack. Participants were
+organized into cohorts based on interest and each cohort was assigned
+one of three mentors that were also selected via open
+application. Mentors acted as a general resource to participants, gave
+one webinar, and organized a hack day. WoCCode also supported
+registration of two participants to attend the American Astronomical
+Society virtual summer 2021 meeting.
+
+Participants rated the impact of the program as very high. In the
+final webinar, participants reported a change in their perception of
+coding in general, for example, accepting coding as something they can
+do for fun. Participants reported generally feeling comfortable asking
+questions and interacting with a community where ``[e]veryone's
+thoughts are welcomed, no one is made to feel less important'' and
+where one can enjoy ``warm interactions with likeminded people.''
+WoCCode is continuing into 2022 by broadening participation in the
+Slack space and publicly advertising the guest webinar events.
+
+In addition to the supported mentoring initiatives, the \astropy
+project as a whole has taken steps to examine representation of
+marginalized groups within the project and search for avenues of
+improvement.
+
+% Much of this text was adapted from Robel Geda's report from NSBP and
+% SACNAS. Please include him on the author list.
+\textbf{\astropy representation at national diversity conferences:}
+Several members of the \astropy community attended virtual conferences
+of the National Society of Black Physicists (NSBP) and SACNAS
+(focusing on Hispanic and Native American scientists across all STEM
+fields). \astropy representatives noted that one underlying topic came
+up multiple times. A major barrier for persons who come from
+underrepresented communities is the lack of resources and expertise
+that are necessary to train students. Writing PEP 8 compliant software
+and understanding the Git and Github workflow is not part of the
+standard Physics and Astrophysics curriculum. Additionally, the larger
+Astronomnical community is still in the process of transitioning
+towards Python based tools, making it difficult for students not
+currently under advisement by some one with extensive Python expertise
+to get involved with the open source Python community. Projects such
+as Learn Astropy could have a profound impact by empowering
+underrepresented groups because it provides a free, searchable, and
+accessible introduction to Python tools for astrophysical
+research. Representatives also noted that offering training and
+teaching materials to PIs of Research Experience for Undergraduate
+(REU) programs throughout the United States could be helpful for
+encouraging advisors to teach students how to use Astropy, create
+their own libraries, and use version control on open source platforms
+like Github. Such materials could also be offered as a workshop at
+national diversity conferences themselves, as SACNAS solicits special
+session proposals each year.
+
+\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
+  initiative:} \todo{No author assigned}{Describe more about this.}
 
@@
 
-Summary of Astroquery and engagement with archives. Cite astroquery paper.
+Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
+It contains over 50 modules for querying astronomical databases, large and small.
+In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
+These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
+
+The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented.
+Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}].
+More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.
+
+Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
+These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
+While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
+% note to coauthors: we need a section on pyvo
 
@@
 
-A prototype for the low-level analysis is ctapipe~\cite{ctapipe},
+A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
 a python package developed to perform all the necessary tasks to from the raw data
@@
 The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
-Gammapy~\cite{gammapy}.
+Gammapy~\citep{gammapy}.
 
@@
 
-{\it Learn Astropy} is an umbrella term that acknowledges the broad educational
-efforts made by the Astropy Project, which are led by the Learn Team.
+\textit{Learn Astropy} is an umbrella term that acknowledges the broad
+educational efforts made by the Astropy Project, which are led by the Learn
+Team.
 The efforts focus on developing online content and workshops covering
-astro-specific coding tasks in Python.
-As described in \citet{astropy:2018}, There are four different types of
+astronomy-specific coding tasks in Python.
+As introduced in \citet{astropy:2018}, there are four different types of
 Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
@@
 
-\begin{itemize}
+\begin{description}
 
-\item {\bf Learn Astropy website:} The Learn Team re-launched the website in
-2021 with a new infrastructure platform, built around full-text search and
-interactive filtering functionality, with the goal of making content
-discoverable as the Learn Astropy content catalog expands.
-This work has been supported in part by the Dunlap grant.
+\item[Learn Astropy website] The Learn Team re-launched the main website and
+search interface for Learn Astropy in 2021 with a new infrastructure platform,
+built around full-text search and interactive filtering functionality, with the
+goal of making content discoverable as the Learn Astropy content catalog
+expands.
+This work has been supported in part by a grant from the Dunlap Institute.
 We have adopted Algolia, a search-as-a-service cloud platform, to store the
@@
 separately from the website application itself, enabling new content types.
-Tutorials, which as Jupyter Notebook-based, are now compiled into their own
-Learn Astropy sub-site using nbcollection.
-Guides, which are Jupyter Book-based, are also deployed as separate websites
-using GitHub pages. This architecture opens future possibilities of indexing
-third-party content, hosted elsewhere, such as on institutional websites.
+Tutorials, which are written as Jupyter Notebooks, are now compiled into their
+own Learn Astropy sub-site using
+\href{https://github.com/astropy/nbcollection}{\texttt{nbcollection}}.
+Guides, which utilize the \href{https://jupyterbook.org/}{Jupyter Book} build
+infrastructure, are also deployed as separate websites using GitHub pages.
+This architecture opens future possibilities of indexing third-party content,
+hosted elsewhere, such as on institutional websites.
 
-\item \textbf{Tutorials} Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
+\item[Tutorials] Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
 
-\item \textbf{Guide} {\it ccdproc} Jupyter book has been completed.
+\item[Guides] {\it ccdproc} Jupyter book has been completed.
 
-\item {\bf Workshops:} the Project has been conducting workshops at winter
+\item[Workshops] The Project has been conducting workshops at winter
 meetings of the American Astronomical Society since AAS 225 in January 2015.
@@
 tools.
-\end{itemize}
+\end{description}
 
@@
 be on a paid plan; additionally NumFOCUS has negotiated a special rate for
-open-source projets.
+open-source projects.
 While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
@@
 \end{itemize}
-\secunfilled
 

### 2022-02-10 src/ms.tex
Fix latex errors (#33)

--- 2022-02-09 src/ms.tex
@@
 Cherenkov Telescope Array
-\secauthor{Axel Donath, Maximilian Nöthe}
-
-The Cherenkov Telescope (CTA) will be the next generation very-high-energy
-gamma-ray observatory.
-CTA will improve over the current generation of imaging atmospheric Cherenkov telescopes (IACTs)
-by a factor of five to ten in sensitivity and will be able to observe the whole sky from a combination of two sites:
-a northern site in La Palma, Spain, and a southern one in Paranal, Chile.
-CTA will be able to observe gamma rays in a broad energy range from around $20\,\mathrm{GeV}$ to over $300\,\mathrm{TeV}$
-using three different types of telescopes, in total over 100 telescopes are planned at the two sites.
-CTA will also be the first open gamma-ray observatory.
-
-The data analysis pipeline is developed as open source software and essentially split in two domains:
-\begin{enumerate}
-  \item In the low-level analysis, the properties of the recorded air-shower events
-    have to be estimated from the raw data.
-    The raw data consists of very short (\textasciitilde $40--100\,\mathrm{ns}$) videos recorded with the fast and
-    sensitive cameras of the telescopes.
-    This includes the energy, particle type and direction of origin of the particle that induced the air shower
-    and the time the shower was recorded.
-  \item In the higher-level analysis, these reconstructed event lists are used together with some
-    characterization of the instrument response to perform the actual scientific analysis.
-    This software will be delivered as CTA science tools to the future users of the Observatory.
-\end{enumerate}
-
-A prototype for the low-level analysis is ctapipe~\cite{ctapipe},
-a python package developed to perform all the necessary tasks to from the raw data
-of Cherenkov telescopes to the reconstructed event lists.
-
-The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
-Gammapy~\cite{gammapy}.
-
-Both Gammapy and \texttt{ctapipe} are using astropy heavily, mainly for units, times, coordinate transformations,
-tables and FITS IO.
-
-As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
-it needs to perform a large number of coordinate transformations.
-To enable this, CTA member M.~Nöthe contributed a major performance improvement
-for large numbers of coordinates with different observation times,
-based on earlier work by B.~Winkel.
-
-Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
-a total of 107 merged pull requests were contributed to astropy.
+\secauthor{Axel Donath, Max Noethe}
 

### 2022-02-10 src/ms.tex
WCS high-level interface

--- 2022-02-10 src/ms.tex
@@
 \newcommand{\paperii}{\cite{astropy:2018}}
+
+% Maybe have command link to Zenodo reference instead, or include it?
+\newcommand{\astropyAPE}[1]{\href{https://github.com/astropy/astropy-APEs/blob/main/APE#1.rst}{APE #1}\xspace}
 
@@
 \begin{itemize}
-  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
-  \item Uncertainties and Distributions
+%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
+%  \item Uncertainties and Distributions
   \item ERFA module is now it’s own thing; Obliquely tie in that units
@@
   \item ? Pytest plugin ecosystem
-  \item A new SpectralCoord class for representing and transforming spectral quantities
-    Velocity in Coordinates
-  \item WCS API : A shared Python interface for World Coordinate Systems
-        Include APE reference
+%  \item A new SpectralCoord class for representing and transforming spectral quantities
+%    Velocity in Coordinates
+%  \item WCS API : A shared Python interface for World Coordinate Systems
+%        Include APE 14 reference
 \end{itemize}
+
+New features come in various forms, from new submodules providing
+wholly new capabilities, to larger additions to existing submodules,
+to large rewrites of the interface of modules.  This range is spanned
+by the four main features that entered the astropy core, which we
+describe below.
+
+\paragraph{Uncertainties and Distributions.} Early in our carreers, we
+were likely all taught that a measurement without a unit or an
+uncertainty is essentially meaningless.  \astropyunits already
+provides the ability to associate numbers with units, and propagate
+these correctly, with its \astropyQuantity\ class, but so far the
+ability to associate and propagate uncertainties is relatively
+limited: in \astropysubpkg{nddata}, there are options to associate
+errors with data arrays, but these are treated as independent, i.e.,
+covariances are not tracked.
+
+In general, proper error propagation is difficult, and often it is
+best to use Monte Carle methods.  With the new
+\astropysubpkg{uncertainties} sub-package, a first step is made to
+make this possible.  It allows one to generate, for each variable,
+randomly drawn samples in a \astropyDistribution, and then propagate
+these by passing them through the normal analysis, producing a
+\astropyDistribution\ of final results that can be inspected.
+
+This \astropyDistribution\ can be any type of array, including of
+course a \astropyQuantity, and the goal is to ensure it can be
+seamlessly used to instantiate also other astropy classes, such as
+\astropyskycoord\ and \astropyTime.
+
+\paragraph{Time series.}  From sampling a continuous variable at fixed
+times to counting events binned into time windows, many different
+areas of astrophysics require the manipulation of time series.  These
+are often logically tables, and the new \astropysubpkg{timeseries}
+sub-package extends the \astropyQTable\ class to tables of data as a
+function of time, where the data can either represent samples or
+averages over particular time bins.  The new classes offer a number of
+special methods to manipulate time series (folding, resampling, etc.)
+and to read different data formats (such as {\em Kepler} lightcurves,
+etc.).
+
+Also part of \astropysubpkg{timeseries} are common analysis routines,
+including Lomb-Scargle and box-least-squares periodogrammes.
+
+\paragraph{Spectral Coordinates.} Measurements are often taken at
+specific ``spectral coordinates'', be they frequencies, wavelengths,
+or photon energies.  The new \astropySpectralCoord\ class allows to
+transform between these as well as to equivalent Doppler velocities.
+It builds on the \astropyQuantity, which can also represent these
+kinds of physical quantities, and allows conversion via dedicated
+equivalencies, but aims to make this more straightforward.
+Furthermore, it can be made aware of the observer and target reference
+frames, allowing for instance transformation from telescope-centric (or
+topocentric) frames to Barycentric or Local Standard of Rest
+(LSRK and LSRD) velocity frames.
+
+\paragraph{A High-Level Interface to World Coordinate Systems.}
+Astronomical data are often provided with information about the
+``real-world'' coordinates that correspond to pixel coordinates. This
+mapping is the essence of the ``World Coordinate System (WCS)''
+concept.  From its inception, the \astropysubpkg{wcs} sub-package
+allowed access to WCS information provided in, e.g., FITS files, but
+it became clear that other WCS standards and representations had to be
+supported for new missions and observatories (e.g., the James Webb
+Space Telescope and the Rubin Observatory).  To harmonize these, a new
+high-level interface was created, based on a formal design first
+proposed via an \astropy\ Proposal for Enhancement (\astropyAPE{14}).
+The hope is that by having a formal design, it will be straightforward
+for other packages implementing WCS objects to either modify their
+classes to conform to the new interface or build thin wrappers that
+conform to it.  The implementation in \astropysubpkg{wcs} interacts
+well with other \astropypkg\ objects such as \astropyskycoord\ and
+\astropyTime.
 

### 2022-02-10 src/ms.tex
fix ampersands

--- 2022-02-10 src/ms.tex
@@
 \subsubsection*{New Features} \label{sec:core-features-new}
-
-\begin{itemize}
-  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
-  \item Uncertainties and Distributions
-  \item ERFA module is now it’s own thing; Obliquely tie in that units
-    remains in astropy
-  \item ? Pytest plugin ecosystem
-  \item A new SpectralCoord class for representing and transforming spectral quantities
-    Velocity in Coordinates
-  \item WCS API : A shared Python interface for World Coordinate Systems
-        Include APE reference
-\end{itemize}
 

### 2022-02-10 src/ms.tex
Remove pytest from whatsnew list
Expand on future plans of uncertainties

--- 2022-02-10 src/ms.tex
@@
     remains in astropy
-  \item ? Pytest plugin ecosystem
+%  \item ? Pytest plugin ecosystem -> done elsewhere
 %  \item A new SpectralCoord class for representing and transforming spectral quantities
@@
 \astropyskycoord\ and \astropyTime.
+
+Apart from Monte-Carlo error propagation, the plan for
+\astropysubpkg{uncertainties} is to also support the case where values
+have associated uncertainties, and these are propagated assuming that
+they are normally distributed, but tracking covariances.  One
+implementation problem to be solved is when to {\em stop} tracking
+covariances.  For instance, a simple operation such as subtracting the
+mean from $N$ data points implies that all data points are now
+covariant with each other, i.e., one has to carry $N\times N$
+covariances.  For a large image, that becomes not just pointless but
+also prohibitive in terms of memory and CPU usage.
 

### 2022-02-10 src/ms.tex
Expand intro to new features

@@
 \newcommand{\secunfilled}{{\color{red}Author:~\textit{Looking for volunteers!}}}
+
+\defcitealias{astropy:2013}{Paper I}
+\defcitealias{astropy:2018}{Paper II}
 
@@
 
-\subsubsection*{New Features} \label{sec:core-features-new}
+\subsubsection*{New and Planned Features} \label{sec:core-features-new}
 
@@
 to large rewrites of the interface of modules.  This range is spanned
-by the four main features that entered the astropy core, which we
-describe below.
+by the five main new features that entered the astropy core since the
+previous summary \citepalias{astropy:2018}.
+
+New features present a particular difficulty for a mature package such
+as \astropypkg, because by their nature they have not been tried out
+much and it is difficult to know in advance what the best interface
+will be, and hard to address all possible use cases in one go, even
+within \astropypkg\ itself.  To give a sense of this, we include in
+our descriptions how we hope the new features will evolve (realizing
+that, in the end, in a user-driven project like \astropy, what happens
+depends mostly on what is highest on the priority list of those that
+have time to contribute).
 

### 2022-02-10 src/ms.tex
Discuss removal of ERFA and .vo

@@
 
-The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
+The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citetalias{astropy:2018}.
 % feels like ↑ could be expanded to 2 sentences
@@
 Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
+
+Before discussing those, we should mention that sometimes one realizes
+features do not actually belong in the core package, either because
+they are of more general use outside of \astropypkg, or because they
+are too specialized.  An example of the former is the copy of the IAU
+Standards Of Fundamental Astronomy (SOFA) software
+Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
+astropy carried, as well as the python wrappers for it.  Since this is
+more basic infrastructure, with a release cycle set by SOFA, it made
+more sense to create separate packages that \astropypkg\ can depend
+on: \package{ERFA} \citep{erfa} and \package{PyERFA} \citep{pyerfa}.
+An example of something that was too specialized was a Virtual
+Observatory sub-package (now removed).  Of its content, the Simple
+Application Message Protocol made sense to keep
+(\astropysubpkg{samp}), but other parts belonged more properly in {\tt
+  Astroquery} (see Sect.~\ref{sec:astroquery}).
 
@@
 
-\begin{itemize}
+% \begin{itemize}
 %  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
 %  \item Uncertainties and Distributions
-  \item ERFA module is now it’s own thing; Obliquely tie in that units
-    remains in astropy
+%   \item ERFA module is now it’s own thing; Obliquely tie in that units
+%    remains in astropy -> above
 %  \item ? Pytest plugin ecosystem -> done elsewhere
@@
 %        Include APE 14 reference
-\end{itemize}
+% \end{itemize}
 
@@
 
+\label{sec:astroquery}
+% MHvK: feel free to move above label to where it makes most sense.
+
 \secauthor{Adam Ginsburg}

### 2022-02-11 src/ms.tex
cleanup section intro
Signed-off-by: nstarman <nstarkman@protonmail.com>

--- 2022-02-10 src/ms.tex
@@
   been part of the core package for years with (purposefully) few changes to the
-software interface.
-This maturity and stability allows for the broader astronomy \python community
-to rely upon the \astropypkg core package and build specialized packages within
-the Astropy ecosystem.
-Even many non-astronomical \python libraries have come to rely on \astropypkg.
+  software interface. This maturity and stability allows for the broader
+  astronomy \python community to rely upon the \astropypkg core package and
+  build specialized packages within the Astropy ecosystem. Even many
+  non-astronomical \python libraries have come to rely on \astropypkg.
 
-The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
-% feels like ↑ could be expanded to 2 sentences
+  The relationship between \astropy and community is reciprocal, with the
+  growing and evolving needs of the community and the maturation of the broader
+  scientific \python ecosystem now driving the development of the \astropypkg
+  core package.
 
-% Maybe briefly outline the sectors of  
-% Using \astropypkg for
-% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
-% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
-% Astropy has grown to interoperate, e.g. NumPy
+  One manifestation of this development realizing some features do not actually
+  belong in the core package, either because they are of more general use
+  outside of \astropypkg, or because they are too specialized. An example of the
+  former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
+  software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
+  \astropy carried, as well as the \python wrappers for it.  Since this is more
+  basic infrastructure, with a release cycle set by SOFA, it made more sense to
+  create separate packages that \astropypkg can depend on: \package{ERFA}
+  \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
+  was too specialized was a Virtual Observatory sub-package (now removed).  Of
+  its content, the Simple Application Message Protocol made sense to keep
+  (\astropysubpkg{samp}), but other parts belonged more properly in {\tt
+  Astroquery} (see Sect.~\ref{sec:astroquery}).
 
-Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
-
-% Summarize major pieces of development since the last paper (since v2.0). A few
-% ideas (taken from What's New pages) below, but feel free to expand on or remove
-% from this list:
-% \begin{itemize}
-%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-%     \item APE 14 WCS
-%     \item A new SpectralCoord class for representing and transforming spectral quantities
-%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
-% \end{itemize}
+  Within \astropypkg, development may be roughly split into a few categories:
+  new features added, performance enhancements, intra and inter-package
+  interoperability, and improvements to precision, accuracy, and
+  reproducibility. We discuss these in turn.
 
@@
 seamlessly used to instantiate also other astropy classes, such as
-\astropyskycoord\ and \astropyTime.
+\astropySkyCoord\ and \astropyTime.
 
@@
 conform to it.  The implementation in \astropysubpkg{wcs} interacts
-well with other \astropypkg\ objects such as \astropyskycoord\ and
+well with other \astropypkg\ objects such as \astropySkyCoord\ and
 \astropyTime.
@@
 \subsection{Connections with data archives}
+
+\label{sec:astroquery}
+% MHvK: feel free to move above label to where it makes most sense.
 

### 2022-02-12 src/ms.tex
Fix bib references

@@
 \newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
-\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
-\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
-\newcommand{\astropyunits}{\astropysubpkg{units}}
-\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
-\newcommand{\astropyconstants}{\astropysubpkg{constants}}
-\newcommand{\astropytime}{\astropysubpkg{time}}
-\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
 
 % Links to documentation: classes
-\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
-\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}
-
-\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
-\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
-\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
-\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
-GalacticLSR}{GalacticLSR}}
-\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
-\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
-\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
-\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
-\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
-\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
-\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
-\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
-\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
-\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
-\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
-\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
-
-% suggestion: shortcut reference to Paper II
-\newcommand{\paperii}{\cite{astropy:2018}}
-
-% Maybe have command link to Zenodo reference instead, or include it?
-\newcommand{\astropyAPE}[1]{\href{https://github.com/astropy/astropy-APEs/blob/main/APE#1.rst}{APE #1}\xspace}
+\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
+\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
+\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
+\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}
 
@@
 \newcommand{\secunfilled}{{\color{red}Author:~\textit{Looking for volunteers!}}}
-
-\defcitealias{astropy:2013}{Paper I}
-\defcitealias{astropy:2018}{Paper II}
 
@@
 
-  The \astropypkg core package is mature and stable, in that many features have
-  been part of the core package for years with (purposefully) few changes to the
-  software interface. This maturity and stability allows for the broader
-  astronomy \python community to rely upon the \astropypkg core package and
-  build specialized packages within the Astropy ecosystem. Even many
-  non-astronomical \python libraries have come to rely on \astropypkg.
-
-  The relationship between \astropy and community is reciprocal, with the
-  growing and evolving needs of the community and the maturation of the broader
-  scientific \python ecosystem now driving the development of the \astropypkg
-  core package.
-
-  One manifestation of this development realizing some features do not actually
-  belong in the core package, either because they are of more general use
-  outside of \astropypkg, or because they are too specialized. An example of the
-  former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
-  software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
-  \astropy carried, as well as the \python wrappers for it.  Since this is more
-  basic infrastructure, with a release cycle set by SOFA, it made more sense to
-  create separate packages that \astropypkg can depend on: \package{ERFA}
-  \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
-  was too specialized was a Virtual Observatory sub-package (now removed).  Of
-  its content, the Simple Application Message Protocol made sense to keep
-  (\astropysubpkg{samp}), but other parts belonged more properly in {\tt
-  Astroquery} (see Sect.~\ref{sec:astroquery}).
-
-  Within \astropypkg, development may be roughly split into a few categories:
-  new features added, performance enhancements, intra and inter-package
-  interoperability, and improvements to precision, accuracy, and
-  reproducibility. We discuss these in turn.
-
-\subsubsection*{New and Planned Features} \label{sec:core-features-new}
-
-% \begin{itemize}
-%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
-%  \item Uncertainties and Distributions
-%   \item ERFA module is now it’s own thing; Obliquely tie in that units
-%    remains in astropy -> above
-%  \item ? Pytest plugin ecosystem -> done elsewhere
-%  \item A new SpectralCoord class for representing and transforming spectral quantities
-%    Velocity in Coordinates
-%  \item WCS API : A shared Python interface for World Coordinate Systems
-%        Include APE 14 reference
-% \end{itemize}
-
-New features come in various forms, from new submodules providing
-wholly new capabilities, to larger additions to existing submodules,
-to large rewrites of the interface of modules.  This range is spanned
-by the five main new features that entered the astropy core since the
-previous summary \citepalias{astropy:2018}.
-
-New features present a particular difficulty for a mature package such
-as \astropypkg, because by their nature they have not been tried out
-much and it is difficult to know in advance what the best interface
-will be, and hard to address all possible use cases in one go, even
-within \astropypkg\ itself.  To give a sense of this, we include in
-our descriptions how we hope the new features will evolve (realizing
-that, in the end, in a user-driven project like \astropy, what happens
-depends mostly on what is highest on the priority list of those that
-have time to contribute).
-
-\paragraph{Uncertainties and Distributions.} Early in our carreers, we
-were likely all taught that a measurement without a unit or an
-uncertainty is essentially meaningless.  \astropyunits already
-provides the ability to associate numbers with units, and propagate
-these correctly, with its \astropyQuantity\ class, but so far the
-ability to associate and propagate uncertainties is relatively
-limited: in \astropysubpkg{nddata}, there are options to associate
-errors with data arrays, but these are treated as independent, i.e.,
-covariances are not tracked.
-
-In general, proper error propagation is difficult, and often it is
-best to use Monte Carle methods.  With the new
-\astropysubpkg{uncertainties} sub-package, a first step is made to
-make this possible.  It allows one to generate, for each variable,
-randomly drawn samples in a \astropyDistribution, and then propagate
-these by passing them through the normal analysis, producing a
-\astropyDistribution\ of final results that can be inspected.
-
-This \astropyDistribution\ can be any type of array, including of
-course a \astropyQuantity, and the goal is to ensure it can be
-seamlessly used to instantiate also other astropy classes, such as
-\astropySkyCoord\ and \astropyTime.
-
-Apart from Monte-Carlo error propagation, the plan for
-\astropysubpkg{uncertainties} is to also support the case where values
-have associated uncertainties, and these are propagated assuming that
-they are normally distributed, but tracking covariances.  One
-implementation problem to be solved is when to {\em stop} tracking
-covariances.  For instance, a simple operation such as subtracting the
-mean from $N$ data points implies that all data points are now
-covariant with each other, i.e., one has to carry $N\times N$
-covariances.  For a large image, that becomes not just pointless but
-also prohibitive in terms of memory and CPU usage.
-
-\paragraph{Time series.}  From sampling a continuous variable at fixed
-times to counting events binned into time windows, many different
-areas of astrophysics require the manipulation of time series.  These
-are often logically tables, and the new \astropysubpkg{timeseries}
-sub-package extends the \astropyQTable\ class to tables of data as a
-function of time, where the data can either represent samples or
-averages over particular time bins.  The new classes offer a number of
-special methods to manipulate time series (folding, resampling, etc.)
-and to read different data formats (such as {\em Kepler} lightcurves,
-etc.).
-
-Also part of \astropysubpkg{timeseries} are common analysis routines,
-including Lomb-Scargle and box-least-squares periodogrammes.
-
-\paragraph{Spectral Coordinates.} Measurements are often taken at
-specific ``spectral coordinates'', be they frequencies, wavelengths,
-or photon energies.  The new \astropySpectralCoord\ class allows to
-transform between these as well as to equivalent Doppler velocities.
-It builds on the \astropyQuantity, which can also represent these
-kinds of physical quantities, and allows conversion via dedicated
-equivalencies, but aims to make this more straightforward.
-Furthermore, it can be made aware of the observer and target reference
-frames, allowing for instance transformation from telescope-centric (or
-topocentric) frames to Barycentric or Local Standard of Rest
-(LSRK and LSRD) velocity frames.
-
-\paragraph{A High-Level Interface to World Coordinate Systems.}
-Astronomical data are often provided with information about the
-``real-world'' coordinates that correspond to pixel coordinates. This
-mapping is the essence of the ``World Coordinate System (WCS)''
-concept.  From its inception, the \astropysubpkg{wcs} sub-package
-allowed access to WCS information provided in, e.g., FITS files, but
-it became clear that other WCS standards and representations had to be
-supported for new missions and observatories (e.g., the James Webb
-Space Telescope and the Rubin Observatory).  To harmonize these, a new
-high-level interface was created, based on a formal design first
-proposed via an \astropy\ Proposal for Enhancement (\astropyAPE{14}).
-The hope is that by having a formal design, it will be straightforward
-for other packages implementing WCS objects to either modify their
-classes to conform to the new interface or build thin wrappers that
-conform to it.  The implementation in \astropysubpkg{wcs} interacts
-well with other \astropypkg\ objects such as \astropySkyCoord\ and
-\astropyTime.
-
-\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
+Summarize major pieces of development since the last paper (since v2.0). A few
+ideas (taken from What's New pages) below, but feel free to expand on or remove
+from this list:
+\begin{itemize}
+    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+    \item APE 14 WCS
+    \item A new SpectralCoord class for representing and transforming spectral quantities
+    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+\end{itemize}
 
@@
 \secauthor{Lía Corrales}
-
-With support from the Moore Foundation, the \astropy Project was
-allocated funding to support mentoring programs. In 2020 a call was
-made to submit proposals for IDE (Inclusion, Diversity, and
-Empowerment) initiatives for project-wide consideration on
-Github. This process was deemed the most ``open'' because it allowed
-for community wide feedback to focus and improve proposal
-initiatives. Two programs were selected via this process, described
-below.
-
-\textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}
-
-\textbf{Women of Color Code (WoCCode)} is a peer-mentoring network for
-coders from traditionally marginalized groups, most notably women of
-color. Participants were invited to the WoCCode Slack space and
-encouraged to attend monthly webinars to share skills in the context
-of open source software libraries. Every other month, a guest speaker
-was invited to talk about their career path and share a skill. Program
-participants were solicited in the fall of 2020, yielding 73
-applications from 17 countries (48\% from the United States, 37\% from
-Africa, and 15\% from remaining continents). We selected participants
-who we identified as having a high potential for contributing to open
-source projects: intermediate to advanced programming skills with a
-vocal interest in contributing. Of the thirty applicants invited to
-the program, nineteen joined the community on Slack. Participants were
-organized into cohorts based on interest and each cohort was assigned
-one of three mentors that were also selected via open
-application. Mentors acted as a general resource to participants, gave
-one webinar, and organized a hack day. WoCCode also supported
-registration of two participants to attend the American Astronomical
-Society virtual summer 2021 meeting.
-
-Participants rated the impact of the program as very high. In the
-final webinar, participants reported a change in their perception of
-coding in general, for example, accepting coding as something they can
-do for fun. Participants reported generally feeling comfortable asking
-questions and interacting with a community where ``[e]veryone's
-thoughts are welcomed, no one is made to feel less important'' and
-where one can enjoy ``warm interactions with likeminded people.''
-WoCCode is continuing into 2022 by broadening participation in the
-Slack space and publicly advertising the guest webinar events.
-
-In addition to the supported mentoring initiatives, the \astropy
-project as a whole has taken steps to examine representation of
-marginalized groups within the project and search for avenues of
-improvement.
-
-% Much of this text was adapted from Robel Geda's report from NSBP and
-% SACNAS. Please include him on the author list.
-\textbf{\astropy representation at national diversity conferences:}
-Several members of the \astropy community attended virtual conferences
-of the National Society of Black Physicists (NSBP) and SACNAS
-(focusing on Hispanic and Native American scientists across all STEM
-fields). \astropy representatives noted that one underlying topic came
-up multiple times. A major barrier for persons who come from
-underrepresented communities is the lack of resources and expertise
-that are necessary to train students. Writing PEP 8 compliant software
-and understanding the Git and Github workflow is not part of the
-standard Physics and Astrophysics curriculum. Additionally, the larger
-Astronomnical community is still in the process of transitioning
-towards Python based tools, making it difficult for students not
-currently under advisement by some one with extensive Python expertise
-to get involved with the open source Python community. Projects such
-as Learn Astropy could have a profound impact by empowering
-underrepresented groups because it provides a free, searchable, and
-accessible introduction to Python tools for astrophysical
-research. Representatives also noted that offering training and
-teaching materials to PIs of Research Experience for Undergraduate
-(REU) programs throughout the United States could be helpful for
-encouraging advisors to teach students how to use Astropy, create
-their own libraries, and use version control on open source platforms
-like Github. Such materials could also be offered as a workshop at
-national diversity conferences themselves, as SACNAS solicits special
-session proposals each year.
-
-\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
-  initiative:} \todo{No author assigned}{Describe more about this.}
 
@@
 
-\label{sec:astroquery}
-% MHvK: feel free to move above label to where it makes most sense.
-
 \secauthor{Adam Ginsburg}
 
-Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
-It contains over 50 modules for querying astronomical databases, large and small.
-In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
-These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
-
-The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented.
-Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}].
-More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.
-
-Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
-These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
-While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
-% note to coauthors: we need a section on pyvo
+Summary of Astroquery and engagement with archives. Cite astroquery paper.
 
@@
 
-\textit{Learn Astropy} is an umbrella term that acknowledges the broad
-educational efforts made by the Astropy Project, which are led by the Learn
-Team.
+{\it Learn Astropy} is an umbrella term that acknowledges the broad educational
+efforts made by the Astropy Project, which are led by the Learn Team.
 The efforts focus on developing online content and workshops covering
-astronomy-specific coding tasks in Python.
-As introduced in \citet{astropy:2018}, there are four different types of
+astro-specific coding tasks in Python.
+As described in \citet{astropy:2018}, There are four different types of
 Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
@@
 
-\begin{description}
+\begin{itemize}
 
-\item[Learn Astropy website] The Learn Team re-launched the main website and
-search interface for Learn Astropy in 2021 with a new infrastructure platform,
-built around full-text search and interactive filtering functionality, with the
-goal of making content discoverable as the Learn Astropy content catalog
-expands.
-This work has been supported in part by a grant from the Dunlap Institute.
+\item {\bf Learn Astropy website:} The Learn Team re-launched the website in
+2021 with a new infrastructure platform, built around full-text search and
+interactive filtering functionality, with the goal of making content
+discoverable as the Learn Astropy content catalog expands.
+This work has been supported in part by the Dunlap grant.
 We have adopted Algolia, a search-as-a-service cloud platform, to store the
@@
 separately from the website application itself, enabling new content types.
-Tutorials, which are written as Jupyter Notebooks, are now compiled into their
-own Learn Astropy sub-site using
-\href{https://github.com/astropy/nbcollection}{\texttt{nbcollection}}.
-Guides, which utilize the \href{https://jupyterbook.org/}{Jupyter Book} build
-infrastructure, are also deployed as separate websites using GitHub pages.
-This architecture opens future possibilities of indexing third-party content,
-hosted elsewhere, such as on institutional websites.
+Tutorials, which as Jupyter Notebook-based, are now compiled into their own
+Learn Astropy sub-site using nbcollection.
+Guides, which are Jupyter Book-based, are also deployed as separate websites
+using GitHub pages. This architecture opens future possibilities of indexing
+third-party content, hosted elsewhere, such as on institutional websites.
 
-\item[Tutorials] Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
+\item \textbf{Tutorials} Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
 
-\item[Guides] {\it ccdproc} Jupyter book has been completed.
+\item \textbf{Guide} {\it ccdproc} Jupyter book has been completed.
 
-\item[Workshops] The Project has been conducting workshops at winter
+\item {\bf Workshops:} the Project has been conducting workshops at winter
 meetings of the American Astronomical Society since AAS 225 in January 2015.
@@
 tools.
-\end{description}
+\end{itemize}
 
@@
 be on a paid plan; additionally NumFOCUS has negotiated a special rate for
-open-source projects.
+open-source projets.
 While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
@@
 \end{itemize}
+\secunfilled
 

### 2022-02-15 src/ms.tex
more text about astroquery

@@
 
-Summary of Astroquery and engagement with archives. Cite astroquery paper.
+Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
+It contains over 50 modules for querying astronomical databases, large and small.
+In particular, since \PaperII, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
+These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
+
+The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented.
+Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}].
+More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.
+
+Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
+These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
+While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
+% note to coauthors: we need a section on pyvo
 

### 2022-02-17 src/ms.tex
more documentation links
And a clean consistent means to define more.

Signed-off-by: nstarman <nstarkman@protonmail.com>

--- 2022-02-12 src/ms.tex
@@
 \newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
+\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
+\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
+\newcommand{\astropyunits}{\astropysubpkg{units}}
+\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
+\newcommand{\astropyconstants}{\astropysubpkg{constants}}
+\newcommand{\astropytime}{\astropysubpkg{time}}
+\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
 
 % Links to documentation: classes
-\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
-\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
-\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
-\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}
+\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
+\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}
+
+\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
+\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
+\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
+\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
+GalacticLSR}{GalacticLSR}}
+\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
+\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
+\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
+\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
+\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
+\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
+\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
+\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
+\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
+\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
+\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
+\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
 

### 2022-02-17 src/ms.tex
Text for uncertainties and distributions

--- 2022-02-15 src/ms.tex
@@
 
-Summarize major pieces of development since the last paper (since v2.0). A few
-ideas (taken from What's New pages) below, but feel free to expand on or remove
-from this list:
+The \astropypkg core package is mature and stable, in that many features have
+been part of the core package for years with (purposefully) few changes to the
+software interface.
+This maturity and stability allows for the broader astronomy \python community
+to rely upon the \astropypkg core package and build specialized packages within
+the Astropy ecosystem.
+Even many non-astronomical \python libraries have come to rely on \astropypkg.
+
+The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
+% feels like ↑ could be expanded to 2 sentences
+
+% Maybe briefly outline the sectors of  
+% Using \astropypkg for
+% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
+% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
+% Astropy has grown to interoperate, e.g. NumPy
+
+Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
+
+% Summarize major pieces of development since the last paper (since v2.0). A few
+% ideas (taken from What's New pages) below, but feel free to expand on or remove
+% from this list:
+% \begin{itemize}
+%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+%     \item APE 14 WCS
+%     \item A new SpectralCoord class for representing and transforming spectral quantities
+%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+% \end{itemize}
+
+\subsubsection*{New Features} \label{sec:core-features-new}
+
 \begin{itemize}
-    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-    \item APE 14 WCS
+  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
+%  \item Uncertainties and Distributions
+  \item ERFA module is now it’s own thing; Obliquely tie in that units
+    remains in astropy
+  \item ? Pytest plugin ecosystem
   \item A new SpectralCoord class for representing and transforming spectral quantities
-    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+    Velocity in Coordinates
+  \item WCS API : A shared Python interface for World Coordinate Systems
+        Include APE reference
 \end{itemize}
+
+New features come in various forms, from new submodules providing
+wholly new capabilities, to larger additions to existing submodules,
+to large rewrites of the interface of modules.  This range is spanned
+by the four main features that entered the astropy core, which we
+describe below.
+
+\paragraph{Uncertainties and Distributions.} Early in our carreers, we
+were likely all taught that a measurement without a unit or an
+uncertainty is essentially meaningless.  \astropyunits already
+provides the ability to associate numbers with units, and propagate
+these correctly, with its \astropyQuantity\ class, but so far the
+ability to associate and propagate uncertainties is relatively
+limited: in \astropysubpkg{nddata}, there are options to associate
+errors with data arrays, but these are treated as independent, i.e.,
+covariances are not tracked.
+
+In general, proper error propagation is difficult, and often it is
+best to use Monte Carle methods.  With the new
+\astropysubpkg{uncertainties} sub-package, a first step is made to
+make this possible.  It allows one to generate, for each variable,
+randomly drawn samples in a \astropyDistribution, and then propagate
+these by passing them through the normal analysis, producing a
+\astropyDistribution\ of final results that can be inspected.
+
+This \astropyDistribution\ can be any type of array, including of
+course a \astropyQuantity, and the goal is to ensure it can be
+seamlessly used to instantiate also other astropy classes, such as
+\astropyskycoord\ and \astropyTime.
+
+\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
 
@@
 
-\textbf{Outreachy:} Describe the outreachy project
+\textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}
 
@@
 \textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
-  initiative:} Describe more about this.
+  initiative:} \todo{No author assigned}{Describe more about this.}
 
@@
 It contains over 50 modules for querying astronomical databases, large and small.
-In particular, since \PaperII, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
+In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
 These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.

### 2022-02-17 src/ms.tex
Time series and box-least-squares

@@
 \begin{itemize}
-  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
+%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
 %  \item Uncertainties and Distributions
@@
 \astropyskycoord\ and \astropyTime.
+
+\paragraph{Time series.}  From sampling a continuous variable at fixed
+times to counting events binned into time windows, many different
+areas of astrophysics require the manipulation of time series.  These
+are often logically tables, and the new \astropysubpkg{timeseries}
+sub-package extends the \astropyQTable\ class to tables of data as a
+function of time, where the data can either represent samples or
+averages over particular time bins.  The new classes offer a number of
+special methods to manipulate time series (folding, resampling, etc.)
+and to read different data formats (such as {\em Kepler} lightcurves,
+etc.).
+
+Also part of \astropysubpkg{timeseries} are common analysis routines,
+including Lomb-Scargle and box-least-squares periodogrammes.
 

### 2022-02-17 src/ms.tex
capitalize SkyCoord shortcut
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
 
+% Maybe have command link to Zenodo reference instead, or include it?
+\newcommand{\astropyAPE}[1]{\href{https://github.com/astropy/astropy-APEs/blob/main/APE#1.rst}{APE #1}\xspace}
+
 \newcommand{\secauthor}[1]{{\color{blue}Author:~\textit{#1}}}
 \newcommand{\secunfilled}{{\color{red}Author:~\textit{Looking for volunteers!}}}
+
+\defcitealias{astropy:2013}{Paper I}
+\defcitealias{astropy:2018}{Paper II}
 
@@
 
-\subsubsection*{New Features} \label{sec:core-features-new}
+\subsubsection*{New and Planned Features} \label{sec:core-features-new}
 
-\begin{itemize}
+% \begin{itemize}
 %  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
 %  \item Uncertainties and Distributions
-  \item ERFA module is now it’s own thing; Obliquely tie in that units
-    remains in astropy
-  \item ? Pytest plugin ecosystem
-  \item A new SpectralCoord class for representing and transforming spectral quantities
-    Velocity in Coordinates
-  \item WCS API : A shared Python interface for World Coordinate Systems
-        Include APE reference
-\end{itemize}
+%   \item ERFA module is now it’s own thing; Obliquely tie in that units
+%    remains in astropy -> above
+%  \item ? Pytest plugin ecosystem -> done elsewhere
+%  \item A new SpectralCoord class for representing and transforming spectral quantities
+%    Velocity in Coordinates
+%  \item WCS API : A shared Python interface for World Coordinate Systems
+%        Include APE 14 reference
+% \end{itemize}
 
@@
 seamlessly used to instantiate also other astropy classes, such as
-\astropyskycoord\ and \astropyTime.
+\astropySkyCoord\ and \astropyTime.
+
+Apart from Monte-Carlo error propagation, the plan for
+\astropysubpkg{uncertainties} is to also support the case where values
+have associated uncertainties, and these are propagated assuming that
+they are normally distributed, but tracking covariances.  One
+implementation problem to be solved is when to {\em stop} tracking
+covariances.  For instance, a simple operation such as subtracting the
+mean from $N$ data points implies that all data points are now
+covariant with each other, i.e., one has to carry $N\times N$
+covariances.  For a large image, that becomes not just pointless but
+also prohibitive in terms of memory and CPU usage.
 
@@
 including Lomb-Scargle and box-least-squares periodogrammes.
+
+\paragraph{Spectral Coordinates.} Measurements are often taken at
+specific ``spectral coordinates'', be they frequencies, wavelengths,
+or photon energies.  The new \astropySpectralCoord\ class allows to
+transform between these as well as to equivalent Doppler velocities.
+It builds on the \astropyQuantity, which can also represent these
+kinds of physical quantities, and allows conversion via dedicated
+equivalencies, but aims to make this more straightforward.
+Furthermore, it can be made aware of the observer and target reference
+frames, allowing for instance transformation from telescope-centric (or
+topocentric) frames to Barycentric or Local Standard of Rest
+(LSRK and LSRD) velocity frames.
+
+\paragraph{A High-Level Interface to World Coordinate Systems.}
+Astronomical data are often provided with information about the
+``real-world'' coordinates that correspond to pixel coordinates. This
+mapping is the essence of the ``World Coordinate System (WCS)''
+concept.  From its inception, the \astropysubpkg{wcs} sub-package
+allowed access to WCS information provided in, e.g., FITS files, but
+it became clear that other WCS standards and representations had to be
+supported for new missions and observatories (e.g., the James Webb
+Space Telescope and the Rubin Observatory).  To harmonize these, a new
+high-level interface was created, based on a formal design first
+proposed via an \astropy\ Proposal for Enhancement (\astropyAPE{14}).
+The hope is that by having a formal design, it will be straightforward
+for other packages implementing WCS objects to either modify their
+classes to conform to the new interface or build thin wrappers that
+conform to it.  The implementation in \astropysubpkg{wcs} interacts
+well with other \astropypkg\ objects such as \astropySkyCoord\ and
+\astropyTime.
 

### 2022-02-17 src/ms.tex
draft draft section on interoperability
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
   been part of the core package for years with (purposefully) few changes to the
-software interface.
-This maturity and stability allows for the broader astronomy \python community
-to rely upon the \astropypkg core package and build specialized packages within
-the Astropy ecosystem.
-Even many non-astronomical \python libraries have come to rely on \astropypkg.
-
-The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citetalias{astropy:2018}.
-% feels like ↑ could be expanded to 2 sentences
-
-% Maybe briefly outline the sectors of  
-% Using \astropypkg for
-% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
-% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
-% Astropy has grown to interoperate, e.g. NumPy
-
-Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
-
-Before discussing those, we should mention that sometimes one realizes
-features do not actually belong in the core package, either because
-they are of more general use outside of \astropypkg, or because they
-are too specialized.  An example of the former is the copy of the IAU
-Standards Of Fundamental Astronomy (SOFA) software
-Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
-astropy carried, as well as the python wrappers for it.  Since this is
-more basic infrastructure, with a release cycle set by SOFA, it made
-more sense to create separate packages that \astropypkg\ can depend
-on: \package{ERFA} \citep{erfa} and \package{PyERFA} \citep{pyerfa}.
-An example of something that was too specialized was a Virtual
-Observatory sub-package (now removed).  Of its content, the Simple
-Application Message Protocol made sense to keep
+  software interface. This maturity and stability allows for the broader
+  astronomy \python community to rely upon the \astropypkg core package and
+  build specialized packages within the Astropy ecosystem. Even many
+  non-astronomical \python libraries have come to rely on \astropypkg.
+
+  The relationship between \astropy and community is reciprocal, with the
+  growing and evolving needs of the community and the maturation of the broader
+  scientific \python ecosystem now driving the development of the \astropypkg
+  core package.
+
+  One manifestation of this development realizing some features do not actually
+  belong in the core package, either because they are of more general use
+  outside of \astropypkg, or because they are too specialized. An example of the
+  former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
+  software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
+  \astropy carried, as well as the \python wrappers for it.  Since this is more
+  basic infrastructure, with a release cycle set by SOFA, it made more sense to
+  create separate packages that \astropypkg can depend on: \package{ERFA}
+  \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
+  was too specialized was a Virtual Observatory sub-package (now removed).  Of
+  its content, the Simple Application Message Protocol made sense to keep
   (\astropysubpkg{samp}), but other parts belonged more properly in {\tt
@@
 
-% Summarize major pieces of development since the last paper (since v2.0). A few
-% ideas (taken from What's New pages) below, but feel free to expand on or remove
-% from this list:
-% \begin{itemize}
-%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-%     \item APE 14 WCS
-%     \item A new SpectralCoord class for representing and transforming spectral quantities
-%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
-% \end{itemize}
+  Within \astropypkg, development may be roughly split into a few categories:
+  new features added, performance enhancements, intra and inter-package
+  interoperability, and improvements to precision, accuracy, and
+  reproducibility. We discuss these in turn.
 
@@
 
+  % New features capture a lot of attention. While exciting, the arguably more
+  % significant improvements are those of interoperability -- ensuring existing
+  % features connect smoothly together.
+
+  The modular nature of \astropypkg means that not all sub-packages deeply
+  integrate with one another -- and similarly between separate \python
+  libraries. Because \astropy is situated at the nexus between scientific
+  computing into astronomy, interoperability has been and continues to be an
+  important focus for \astropy, both within the various \astropypkg sub-packages
+  and with other important \python libraries. While increasing inter and
+  intra-package integration will be a perpetual goal -- with feature additions
+  in one place requiring updates in another -- \astropy has made impactful
+  progress with units, tables, and I/O.
+  % Better list? Also better transition
+  If done right, all a user notices is that code using many different python
+  features and libraries ``just works''. Here we will draw notice to these
+  efforts to make your code ``just work''.
+
+  \paragraph{\package{numpy} on Units}
+
+    \astropyQuantity\ is a backbone of \astropy, leveraging the power of
+    \package{numpy} and adding units. Previously, many \package{numpy} functions
+    would strip a \astropyQuantity\ of its units (or fail outright), limiting
+    \astropyQuantity's usefulness. Now, advances in \package{numpy} function
+    overloading (see for example
+    \href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
+    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.7+)
+    functions. These advances have similarly benefitted many
+    \package{numpy}-dependent packages, like \package{scipy} and their sparse
+    matrices. And again \astropypkg benefits since more \package{scipy}
+    functions now work with \astropyQuantity.
+
+    In the remaining gaps, the \package{numpy} and \astropyQuantity\
+    interoperability efforts are ongoing. For some functions, such as in
+    \href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
+    module} for manipulating structured arrays, compatibility only requires
+    exending the existing \package{numpy}-\astropypkg bridge framework.
+    Community interest, in the form of a Feature Request (or better yet Pull
+    Request) would be sufficient to see this realized. For other functions,
+    discussed further in the
+    \href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
+    documentation}, the \package{numpy} framework does not yet allow for full
+    interoperability with \astropy. A goal of \astropy, and the scientific
+    \python community is to enhance and implement the frameworks, allowing
+    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem of
+    \package{numpy}-like projects.
+
+  \paragraph{Astropy in Time and on Units}
+
+    % Bad sentence ↓. Rewrite.
+    Not only did \astropyunits gain powerful integrations with \package{numpy},
+    there were many intra-package improvements.
+
+    \astropymodeling can add units to otherwise unit-less models.
+
+    A new module –
+    \astropycosmologyunits
+    – has been added to the cosmology sub-package for defining and collecting
+    cosmological units and equivalencies. The unit \texttt{littleh} and
+    equivalency
+    \astropyapidoc{cosmology.units.with_H0}{with\_H0}
+    are deprecated from the main \astropyunits sub-package and moved to
+    \astropycosmologyunits. A new unit, \texttt{redshift}, is added for tracking
+    factors of cosmological redshift.
+    % As this is a pseudo-unit an equivalency
+    % \astropyapidoc{cosmology.units.dimensionless_redshift}{dimensionless\_redshift}
+    % is added (and enabled by default) to allow for redshift - dimensionless
+    % conversions.
+    To convert between redshift and other cosmological distance
+    measures, e.g. CMB temperature or comoving distance, the equivalency
+    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
+    is also added. This equivalency is actually a composite of other
+    equivalencies,
+    % : \astropyapidoc{cosmology.units.redshift_distance}{redshift\_distance},
+    % \astropyapidoc{cosmology.units.redshift_hubble}{redshift\_hubble}, and
+    % \astropyapidoc{cosmology.units.redshift_temperature}{redshift\_temperature},
+    which may be used separately.
+
+    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
+    The FITS Time standard paper defines the formats and keywords used to
+    represent timing information in FITS files. The Astropy FITS package now
+    provides support for reading and writing native Time table columns using
+    this format. This is described within the FITS unified I/O interface and
+    examples of usage can be found in the Astropy native objects (mixin columns)
+    section.
+
+    Implementation of this feature was made possible through the support of the
+    Google Summer of Code 2017 program. To learn more about this project please
+    see the final report, A mixin protocol for seamless interoperability.
+
+  \paragraph{Table Mixin Columns}
+
+    Numerous \python libraries have custom table implementations
+    \begin{itemize}
+      \item (Support use of SkyCoord as Column happened before) and in table vstack, dstack, and insert\_row
+      \item Converting SkyCoord to QTable (1 sentence addition to above)
+      \item Added support for dask arrays in tables
+      \item Added support for registering array-like objects as mixin columns
+    \end{itemize}
+
+  \paragraph{Unified I/O architecture}
+
+    \astropyiosubpkg{registry} is a powerful way to define input and output
+    (I/O) functions for \astropypkg objects, such as reading or writing to a
+    file. \astropy uses this internally for the I/O methods in \astropyTable and
+    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
+    extend the options on these classes.
+
+    As of Astropy 5.0 the I/O registry submodule has switched to a class-based
+    architecture and been refactored to allow read-only, write-only, and
+    read-write registries. Additionally, the registry submodule no longer
+    assumes an object has \astropy-style \texttt{read} and \texttt{write}
+    methods when trying to read or write that object. These two improvements
+    enable a number of new use cases for \astropyiosubpkg{registry}, some
+    highlighted here, but others we hope to see have community adoption.
+
+    First, the class-based architecture allows for the creation of custom
+    registries. An application of this is in \astropycosmology, which
+    has two different \textit{kinds} of registries for \astropyCosmology: one
+    for reading and writing files, and another for converting between python
+    objects.
+
+    Second, \textit{any} kind of object is now compatible with
+    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
+    classes means a user does not have to create a whole new I/O system and can
+    offer a unified and familiar API.
+
+  \paragraph{Persistent Storage and I/O}
+
+    % TODO Better transition sentence
+    The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.
+
+    \begin{itemize}
+      \item ASDF : JSWST  and models can be stored with ASDF.
+      \item ECSV : Support for multidimensional and object columns in ECSV
+      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
+      \item Support for reading, writing, and converting Cosmology
+      \item MRT Support for reading and writing tables to MRT format : GSOC
+      \item Support for reading and writing tables to Parquet format
+      \item Support for reading and writing tables to QDP format
+      \item ASDF read write
+    \end{itemize}
+
 \subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
 
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
+  An important focus of \astropy has been to increase the precision and accuracy
+  of its functionality, while, where appropriate, making sure results obtained
+  using older versions of \astropypkg are reproducible.
+  % Something about ScienceState
+  Notable improvements have been made to \astropytime, \astropyconstants,
+  \astropycoordinates, and \astropycosmology.
+
+  \paragraph{Time}
+
+    In astronomy, time accuracy down to the second is frequently important.
+    Whether for planning observations or crunching data, missing seconds can
+    meaningfully impact results. Before, when using \astropyTime in \astropytime
+    leap seconds had to be manually applied; now leap seconds are applied
+    automatically. Moreover, \astropypkg updates internal time data-files to
+    ensure that the correct leap second adjustment is always used. For
+    reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
+    manually applied.
+
+  \paragraph{Constants}
+
+    Measurements of physical constants (and the units defined from them) improve
+    over time. Periodically the standardized systems of units and constants are
+    updated to reflect these improvements, for instance in 2019 the SI system was
+    redefined \citep{NIST2019}, with an accompanying update to the physical
+    constants in
+    \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
+    2018}}. \astropypkg contains two modules -- \astropyconstants and
+    \astropyunits -- for working with constants and units, respectively.
+    \astropyconstants now defaults to use the \texttt{SI/CODATA 2018} values, with
+    the relation between units in \astropyunits based upon these constants. Most
+    of \astropypkg and affiliate packages build upon \astropyunits, so this update
+    has far-reaching effects across the \astropy ecosystem and astronomy
+    community.
+
+    For reproducibility, \astropypkg allows the constants (and therefore units)
+    definitions to be rolled back to prior values, for instance the
+    \texttt{SI/CODATA 2014} definitions.
+    % CITATION
+    The constants and units definitions are set by configuration file as well as
+    \astropyScienceState (an runtime configuration controller). For work where
+    updates to the values of the fundamental constants significantly alters
+    results we recommend including with the work an astropy configuration file,
+    which specifies the set of constants used.
+
+  \paragraph{Cosmology}
+    % TODO! At least one good sentence about "what is the benefit of astropy.cosmology"
+
+    \astropycosmology contains classes for representing cosmological models.
+    % currently all homogeneous and isotropic variations of the
+    % Friedmann-Lemaitre-Robertson-Walker metric.
+    Bundled with the classes are commonly used \astropyCosmology realizations, for
+    example best-fit measurements from
+    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
+    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
+    \citep{PlanckMission:2006}.
+
+    When these important missions publish new measurements, \astropycosmology is
+    updated to include these results as realizations of the appropriate class
+    (generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
+    always exactly correspond to \astropycosmology classes, for instance the
+    Planck 2018 results \citep{Planck2018VI} include massive neutrinos in
+    $\Omega_{matter,0}$, while in \astropyFlatLambdaCDM this mass contribution is
+    stored in a separate parameter. Consequently, while some parameter values
+    appear different from the source paper, the \astropyCosmology realizations
+    correctly reproduce the cosmological models.
+
+    Since \citealt{astropy:2018} the following cosmology realizations have been
+    added:
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
+    from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
+    Lya]{WMAP1Year:2003};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
+    from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
+    from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
+    ext]{Planck2015XIII:2016};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
+    from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE,
+    EE + lowE + lensing + BAO]{Planck2018VI:2020}.
+
+    Astropy provides a
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
+    default cosmology}. All calculations done in a cosmological context will use
+    this default cosmology. The default \astropyCosmology has been updated to the
+    Planck 2018 parameters. For reproducibility, the default cosmology may be set
+    to old defaults, such as the WMAP Three Year values. The default cosmology is
+    dynamically configurable using \astropyScienceState, meaning a set of
+    calculations may be run using different assumed cosmologies, and results
+    compared between the two.
+    % Repeat the word default less.
+
+  \paragraph{Coordinates}
+
+    Being able to describe the position of objects, i.e. their coordinates, is
+    fundamental across many knowledge domains. \astropypkg provides
+    \astropycoordinates for working with low-level positional and velocity data
+    all the way to high-level coordinate objects with reference frames,
+    atmospheric information, and more. A central feature of \astropycoordinates is
+    the ability to transform data between reference frames, e.g. \astropyICRS
+    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
+    there have been some historical limitations for spatially proximate
+    transformations, impacting the usefulness of \astropycoordinates for
+    ground-based telescopes and intra-solar astronomy. Transformations in the
+    \astropyAltAz frame were reasonably precise for very distant objects, but
+    wrong by up around several arcseconds for e.g. the location of the moon. Now
+    these transformations are much more precise, down to the milli-arcsecond
+    level. Similar precision improvements were made to the
+    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
+    transformations. Additionally,
+    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
+    and associated transformations have been updated to correctly reflect the
+    “true” (includes precession but not nutation) and “mean” (precession and
+    nutation) terminology.
+
+    For Galactic astronomers, the \astropyGalactocentric
+    \astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults} have
+    been updated to include more recent measurements (tabulated in the
+    \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
+    attribute). For reproducibility, old definitions are still available, and will
+    continue to be so as \astropypkg is updated with forthcoming measurements of
+    the Galactic center, solar motion, etc.
+    % may be used by setting the default value to ``pre-v4.0'' in the
+    % \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html}{\texttt{galactocentric\_frame\_defaults}}
+    % \astropyScienceState.
+
+\subsection*{Performance} \label{sec:core-features-performance}
+
+\subsection*{Future} \label{sec:core-features-future}
 

### 2022-02-20 src/ms.tex
Small grammar correction

@@
 
-  One manifestation of this development realizing some features do not actually
+One manifestation of this was the realization that some features do not actually
 belong in the core package, either because they are of more general use

### 2022-02-20 src/ms.tex
Describe new masked arrays.

@@
 \newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
+\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
 \newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
@@
 also prohibitive in terms of memory and CPU usage.
+
+\paragraph{Masked quantities.}  Data can be known to be bad, in which
+case it is useful to mask them.  Support for masked data involves a
+number of choices, such as whether one indicates data values are bad
+by a separate flag or by replacing them by a special value, such as
+``Not a Number'' (NaN).  In \astropypkg, both approaches have been
+used: a flag for masked columns in \astropyTable, mask flags and
+bitmaps for N-dimensional data in \astropysubpkg{nddata}, and
+replacing elements with NaN in \astropyTime.  A particular problem was
+that no masked quantities were available, making it more difficult to
+migrate to \astropyQTable, in which \astropyQuantity\ is used for all
+columns with units, thus propagating those correctly (instead of
+treating the unit merely as metadata).
+
+Unfortunately, it turned out that the \texttt{MaskedArray} class from
+\package{numpy} could not be used, as its design made it a wrapper
+around any subclass, rather than allow another class to use it as an
+alternative to a regular data array.  For quantities, this does not
+work, as one can see from the fact that a masked value does not mean
+the unit is not known.  Given this, a new \astropyMasked\ class was
+designed, based on a framework very similar to that of
+\astropyDistribution, making it easy to create masked instances of
+other classes.  In particular, this can be used to create masked
+quantities for \astropyQTable, including full support for reading and
+writing to the different file types.
+
+The masked quantities work mostly without any changes in higher-level
+objects such as \astropySkyCoord, but work is still underway to expose
+the mask in those classes, as well as to use masked arrays in
+\astropyTime\ instead of replacing data with NaN.
 

### 2022-02-20 src/ms.tex
Correct numpy version that introduces __array_function__

@@
     \href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
-    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.7+)
+    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.17+)
     functions. These advances have similarly benefitted many

### 2022-02-20 src/ms.tex
Performance section

@@
 
-\subsection*{Future} \label{sec:core-features-future}
+As features are added and corner cases dealt with, performance often
+suffers.  Hence, pull requests that improve performance are welcomed
+generally, and were even the main goal for one release cycle.  For
+most packages, this meant improvements in the \python\ code, but for
+some the most time-critical pieces were rewritten in~C (e.g., for
+convolution of images, sigma-clipping, and converting time strings to
+binary).  Furthermore, a particular effort was made to ensure
+\astropypkg\ is thread-safe, so that it can be used on supercomputer
+clusters.
+
+% \subsection*{Future} \label{sec:core-features-future}
+% MHvK: Merged with the new feature section
 

### 2022-02-20 src/ms.tex
Put performance last also in intro.

@@
 
-Within \astropypkg, development may be roughly split into a few categories:
-new features added, performance enhancements, intra and inter-package
-interoperability, and improvements to precision, accuracy, and
-reproducibility. We discuss these in turn.
+Within \astropypkg, development may be roughly split into a few
+categories: new features added, intra and inter-package
+interoperability, improvements to precision, accuracy, and
+reproducibility, and performance enhancements. We discuss these in turn.
 

### 2022-02-20 src/ms.tex
Some comments on numpy and units

@@
 
+  % MHvK: this repeats the same thing 3 times; try to use just one sentence!
   The modular nature of \astropypkg means that not all sub-packages deeply
@@
   If done right, all a user notices is that code using many different python
-  features and libraries ``just works''. Here we will draw notice to these
-  efforts to make your code ``just work''.
+  features and libraries ``just works''.  Here, we describe a few
+  particular efforts towards this goal.
 
@@
     mean \astropyQuantity\ works with almost all \package{numpy} (version 1.17+)
-    functions. These advances have similarly benefitted many
-    \package{numpy}-dependent packages, like \package{scipy} and their sparse
-    matrices. And again \astropypkg benefits since more \package{scipy}
-    functions now work with \astropyQuantity.
+    functions.
+    % MHvK: I don't think the following is true, so commented it out.
+    % These advances have similarly benefitted many
+    % \package{numpy}-dependent packages, like \package{scipy} and their sparse
+    % matrices. And again \astropypkg benefits since more \package{scipy}
+    % functions now work with \astropyQuantity.
 
@@
     Community interest, in the form of a Feature Request (or better yet Pull
-    Request) would be sufficient to see this realized. For other functions,
+    Request) would be sufficient to see this realized. For a few remaining functions,
     discussed further in the

### 2022-02-23 src/ms.tex
Few more comments
Addressing comments, thanks @cosimoNigro, @matteobachetti, @sibirrer

--- 2022-02-20 src/ms.tex
@@
 \newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
-\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
-\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
-\newcommand{\astropyunits}{\astropysubpkg{units}}
-\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
-\newcommand{\astropyconstants}{\astropysubpkg{constants}}
-\newcommand{\astropytime}{\astropysubpkg{time}}
-\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
 
 % Links to documentation: classes
-\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
-\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}
-
-\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
-\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
-\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
-\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
-GalacticLSR}{GalacticLSR}}
-\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
-\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
-\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
-\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
-\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
-\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
-\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
-\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
-\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
-\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
-\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
-\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
-\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
-
-% suggestion: shortcut reference to Paper II
-\newcommand{\paperii}{\cite{astropy:2018}}
-
-% Maybe have command link to Zenodo reference instead, or include it?
-\newcommand{\astropyAPE}[1]{\href{https://github.com/astropy/astropy-APEs/blob/main/APE#1.rst}{APE #1}\xspace}
+\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
+\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
+\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
+\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}
 
@@
 
-The \astropypkg core package is mature and stable, in that many features have
-been part of the core package for years with (purposefully) few changes to the
-software interface. This maturity and stability allows for the broader
-astronomy \python community to rely upon the \astropypkg core package and
-build specialized packages within the Astropy ecosystem. Even many
-non-astronomical \python libraries have come to rely on \astropypkg.
-
-The relationship between \astropy and community is reciprocal, with the
-growing and evolving needs of the community and the maturation of the broader
-scientific \python ecosystem now driving the development of the \astropypkg
-core package.
-
-One manifestation of this was the realization that some features do not actually
-belong in the core package, either because they are of more general use
-outside of \astropypkg, or because they are too specialized. An example of the
-former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
-software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
-\astropy carried, as well as the \python wrappers for it.  Since this is more
-basic infrastructure, with a release cycle set by SOFA, it made more sense to
-create separate packages that \astropypkg can depend on: \package{ERFA}
-\citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
-was too specialized was a Virtual Observatory sub-package (now removed).  Of
-its content, the Simple Application Message Protocol made sense to keep
-(\astropysubpkg{samp}), but other parts belonged more properly in {\tt
-Astroquery} (see Sect.~\ref{sec:astroquery}).
-
-Within \astropypkg, development may be roughly split into a few
-categories: new features added, intra and inter-package
-interoperability, improvements to precision, accuracy, and
-reproducibility, and performance enhancements. We discuss these in turn.
-
-\subsubsection*{New and Planned Features} \label{sec:core-features-new}
-
-% \begin{itemize}
-%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
-%  \item Uncertainties and Distributions
-%   \item ERFA module is now it’s own thing; Obliquely tie in that units
-%    remains in astropy -> above
-%  \item ? Pytest plugin ecosystem -> done elsewhere
-%  \item A new SpectralCoord class for representing and transforming spectral quantities
-%    Velocity in Coordinates
-%  \item WCS API : A shared Python interface for World Coordinate Systems
-%        Include APE 14 reference
-% \end{itemize}
-
-New features come in various forms, from new submodules providing
-wholly new capabilities, to larger additions to existing submodules,
-to large rewrites of the interface of modules.  This range is spanned
-by the five main new features that entered the astropy core since the
-previous summary \citepalias{astropy:2018}.
-
-New features present a particular difficulty for a mature package such
-as \astropypkg, because by their nature they have not been tried out
-much and it is difficult to know in advance what the best interface
-will be, and hard to address all possible use cases in one go, even
-within \astropypkg\ itself.  To give a sense of this, we include in
-our descriptions how we hope the new features will evolve (realizing
-that, in the end, in a user-driven project like \astropy, what happens
-depends mostly on what is highest on the priority list of those that
-have time to contribute).
-
-\paragraph{Uncertainties and Distributions.} Early in our carreers, we
-were likely all taught that a measurement without a unit or an
-uncertainty is essentially meaningless.  \astropyunits already
-provides the ability to associate numbers with units, and propagate
-these correctly, with its \astropyQuantity\ class, but so far the
-ability to associate and propagate uncertainties is relatively
-limited: in \astropysubpkg{nddata}, there are options to associate
-errors with data arrays, but these are treated as independent, i.e.,
-covariances are not tracked.
-
-In general, proper error propagation is difficult, and often it is
-best to use Monte Carle methods.  With the new
-\astropysubpkg{uncertainties} sub-package, a first step is made to
-make this possible.  It allows one to generate, for each variable,
-randomly drawn samples in a \astropyDistribution, and then propagate
-these by passing them through the normal analysis, producing a
-\astropyDistribution\ of final results that can be inspected.
-
-This \astropyDistribution\ can be any type of array, including of
-course a \astropyQuantity, and the goal is to ensure it can be
-seamlessly used to instantiate also other astropy classes, such as
-\astropySkyCoord\ and \astropyTime.
-
-Apart from Monte-Carlo error propagation, the plan for
-\astropysubpkg{uncertainties} is to also support the case where values
-have associated uncertainties, and these are propagated assuming that
-they are normally distributed, but tracking covariances.  One
-implementation problem to be solved is when to {\em stop} tracking
-covariances.  For instance, a simple operation such as subtracting the
-mean from $N$ data points implies that all data points are now
-covariant with each other, i.e., one has to carry $N\times N$
-covariances.  For a large image, that becomes not just pointless but
-also prohibitive in terms of memory and CPU usage.
-
-\paragraph{Masked quantities.}  Data can be known to be bad, in which
-case it is useful to mask them.  Support for masked data involves a
-number of choices, such as whether one indicates data values are bad
-by a separate flag or by replacing them by a special value, such as
-``Not a Number'' (NaN).  In \astropypkg, both approaches have been
-used: a flag for masked columns in \astropyTable, mask flags and
-bitmaps for N-dimensional data in \astropysubpkg{nddata}, and
-replacing elements with NaN in \astropyTime.  A particular problem was
-that no masked quantities were available, making it more difficult to
-migrate to \astropyQTable, in which \astropyQuantity\ is used for all
-columns with units, thus propagating those correctly (instead of
-treating the unit merely as metadata).
-
-Unfortunately, it turned out that the \texttt{MaskedArray} class from
-\package{numpy} could not be used, as its design made it a wrapper
-around any subclass, rather than allow another class to use it as an
-alternative to a regular data array.  For quantities, this does not
-work, as one can see from the fact that a masked value does not mean
-the unit is not known.  Given this, a new \astropyMasked\ class was
-designed, based on a framework very similar to that of
-\astropyDistribution, making it easy to create masked instances of
-other classes.  In particular, this can be used to create masked
-quantities for \astropyQTable, including full support for reading and
-writing to the different file types.
-
-The masked quantities work mostly without any changes in higher-level
-objects such as \astropySkyCoord, but work is still underway to expose
-the mask in those classes, as well as to use masked arrays in
-\astropyTime\ instead of replacing data with NaN.
-
-\paragraph{Time series.}  From sampling a continuous variable at fixed
-times to counting events binned into time windows, many different
-areas of astrophysics require the manipulation of time series.  These
-are often logically tables, and the new \astropysubpkg{timeseries}
-sub-package extends the \astropyQTable\ class to tables of data as a
-function of time, where the data can either represent samples or
-averages over particular time bins.  The new classes offer a number of
-special methods to manipulate time series (folding, resampling, etc.)
-and to read different data formats (such as {\em Kepler} lightcurves,
-etc.).
-
-Also part of \astropysubpkg{timeseries} are common analysis routines,
-including Lomb-Scargle and box-least-squares periodogrammes.
-
-\paragraph{Spectral Coordinates.} Measurements are often taken at
-specific ``spectral coordinates'', be they frequencies, wavelengths,
-or photon energies.  The new \astropySpectralCoord\ class allows to
-transform between these as well as to equivalent Doppler velocities.
-It builds on the \astropyQuantity, which can also represent these
-kinds of physical quantities, and allows conversion via dedicated
-equivalencies, but aims to make this more straightforward.
-Furthermore, it can be made aware of the observer and target reference
-frames, allowing for instance transformation from telescope-centric (or
-topocentric) frames to Barycentric or Local Standard of Rest
-(LSRK and LSRD) velocity frames.
-
-\paragraph{A High-Level Interface to World Coordinate Systems.}
-Astronomical data are often provided with information about the
-``real-world'' coordinates that correspond to pixel coordinates. This
-mapping is the essence of the ``World Coordinate System (WCS)''
-concept.  From its inception, the \astropysubpkg{wcs} sub-package
-allowed access to WCS information provided in, e.g., FITS files, but
-it became clear that other WCS standards and representations had to be
-supported for new missions and observatories (e.g., the James Webb
-Space Telescope and the Rubin Observatory).  To harmonize these, a new
-high-level interface was created, based on a formal design first
-proposed via an \astropy\ Proposal for Enhancement (\astropyAPE{14}).
-The hope is that by having a formal design, it will be straightforward
-for other packages implementing WCS objects to either modify their
-classes to conform to the new interface or build thin wrappers that
-conform to it.  The implementation in \astropysubpkg{wcs} interacts
-well with other \astropypkg\ objects such as \astropySkyCoord\ and
-\astropyTime.
-
-\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
-
-  % New features capture a lot of attention. While exciting, the arguably more
-  % significant improvements are those of interoperability -- ensuring existing
-  % features connect smoothly together.
-
-  The modular nature of \astropypkg means that not all sub-packages deeply
-  integrate with one another -- and similarly between separate \python
-  libraries. Because \astropy is situated at the nexus between scientific
-  computing into astronomy, interoperability has been and continues to be an
-  important focus for \astropy, both within the various \astropypkg sub-packages
-  and with other important \python libraries. While increasing inter and
-  intra-package integration will be a perpetual goal -- with feature additions
-  in one place requiring updates in another -- \astropy has made impactful
-  progress with units, tables, and I/O.
-  % Better list? Also better transition
-  If done right, all a user notices is that code using many different python
-  features and libraries ``just works''. Here we will draw notice to these
-  efforts to make your code ``just work''.
-
-  \paragraph{\package{numpy} on Units}
-
-    \astropyQuantity\ is a backbone of \astropy, leveraging the power of
-    \package{numpy} and adding units. Previously, many \package{numpy} functions
-    would strip a \astropyQuantity\ of its units (or fail outright), limiting
-    \astropyQuantity's usefulness. Now, advances in \package{numpy} function
-    overloading (see for example
-    \href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
-    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.17+)
-    functions. These advances have similarly benefitted many
-    \package{numpy}-dependent packages, like \package{scipy} and their sparse
-    matrices. And again \astropypkg benefits since more \package{scipy}
-    functions now work with \astropyQuantity.
-
-    In the remaining gaps, the \package{numpy} and \astropyQuantity\
-    interoperability efforts are ongoing. For some functions, such as in
-    \href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
-    module} for manipulating structured arrays, compatibility only requires
-    exending the existing \package{numpy}-\astropypkg bridge framework.
-    Community interest, in the form of a Feature Request (or better yet Pull
-    Request) would be sufficient to see this realized. For other functions,
-    discussed further in the
-    \href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
-    documentation}, the \package{numpy} framework does not yet allow for full
-    interoperability with \astropy. A goal of \astropy, and the scientific
-    \python community is to enhance and implement the frameworks, allowing
-    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem of
-    \package{numpy}-like projects.
-
-  \paragraph{Astropy in Time and on Units}
-
-    % Bad sentence ↓. Rewrite.
-    Not only did \astropyunits gain powerful integrations with \package{numpy},
-    there were many intra-package improvements.
-
-    \astropymodeling can add units to otherwise unit-less models.
-
-    A new module –
-    \astropycosmologyunits
-    – has been added to the cosmology sub-package for defining and collecting
-    cosmological units and equivalencies. The unit \texttt{littleh} and
-    equivalency
-    \astropyapidoc{cosmology.units.with_H0}{with\_H0}
-    are deprecated from the main \astropyunits sub-package and moved to
-    \astropycosmologyunits. A new unit, \texttt{redshift}, is added for tracking
-    factors of cosmological redshift.
-    % As this is a pseudo-unit an equivalency
-    % \astropyapidoc{cosmology.units.dimensionless_redshift}{dimensionless\_redshift}
-    % is added (and enabled by default) to allow for redshift - dimensionless
-    % conversions.
-    To convert between redshift and other cosmological distance
-    measures, e.g. CMB temperature or comoving distance, the equivalency
-    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
-    is also added. This equivalency is actually a composite of other
-    equivalencies,
-    % : \astropyapidoc{cosmology.units.redshift_distance}{redshift\_distance},
-    % \astropyapidoc{cosmology.units.redshift_hubble}{redshift\_hubble}, and
-    % \astropyapidoc{cosmology.units.redshift_temperature}{redshift\_temperature},
-    which may be used separately.
-
-    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
-    The FITS Time standard paper defines the formats and keywords used to
-    represent timing information in FITS files. The Astropy FITS package now
-    provides support for reading and writing native Time table columns using
-    this format. This is described within the FITS unified I/O interface and
-    examples of usage can be found in the Astropy native objects (mixin columns)
-    section.
-
-    Implementation of this feature was made possible through the support of the
-    Google Summer of Code 2017 program. To learn more about this project please
-    see the final report, A mixin protocol for seamless interoperability.
-
-  \paragraph{Table Mixin Columns}
-
-    Numerous \python libraries have custom table implementations
+Summarize major pieces of development since the last paper (since v2.0). A few
+ideas (taken from What's New pages) below, but feel free to expand on or remove
+from this list:
 \begin{itemize}
-      \item (Support use of SkyCoord as Column happened before) and in table vstack, dstack, and insert\_row
-      \item Converting SkyCoord to QTable (1 sentence addition to above)
-      \item Added support for dask arrays in tables
-      \item Added support for registering array-like objects as mixin columns
+    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+    \item APE 14 WCS
+    \item A new SpectralCoord class for representing and transforming spectral quantities
+    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
 \end{itemize}
-
-  \paragraph{Unified I/O architecture}
-
-    \astropyiosubpkg{registry} is a powerful way to define input and output
-    (I/O) functions for \astropypkg objects, such as reading or writing to a
-    file. \astropy uses this internally for the I/O methods in \astropyTable and
-    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
-    extend the options on these classes.
-
-    As of Astropy 5.0 the I/O registry submodule has switched to a class-based
-    architecture and been refactored to allow read-only, write-only, and
-    read-write registries. Additionally, the registry submodule no longer
-    assumes an object has \astropy-style \texttt{read} and \texttt{write}
-    methods when trying to read or write that object. These two improvements
-    enable a number of new use cases for \astropyiosubpkg{registry}, some
-    highlighted here, but others we hope to see have community adoption.
-
-    First, the class-based architecture allows for the creation of custom
-    registries. An application of this is in \astropycosmology, which
-    has two different \textit{kinds} of registries for \astropyCosmology: one
-    for reading and writing files, and another for converting between python
-    objects.
-
-    Second, \textit{any} kind of object is now compatible with
-    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
-    classes means a user does not have to create a whole new I/O system and can
-    offer a unified and familiar API.
-
-  \paragraph{Persistent Storage and I/O}
-
-    % TODO Better transition sentence
-    The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.
-
-    \begin{itemize}
-      \item ASDF : JSWST  and models can be stored with ASDF.
-      \item ECSV : Support for multidimensional and object columns in ECSV
-      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
-      \item Support for reading, writing, and converting Cosmology
-      \item MRT Support for reading and writing tables to MRT format : GSOC
-      \item Support for reading and writing tables to Parquet format
-      \item Support for reading and writing tables to QDP format
-      \item ASDF read write
-    \end{itemize}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
-
-  An important focus of \astropy has been to increase the precision and accuracy
-  of its functionality, while, where appropriate, making sure results obtained
-  using older versions of \astropypkg are reproducible.
-  % Something about ScienceState
-  Notable improvements have been made to \astropytime, \astropyconstants,
-  \astropycoordinates, and \astropycosmology.
-
-  \paragraph{Time}
-
-    In astronomy, time accuracy down to the second is frequently important.
-    Whether for planning observations or crunching data, missing seconds can
-    meaningfully impact results. Before, when using \astropyTime in \astropytime
-    leap seconds had to be manually applied; now leap seconds are applied
-    automatically. Moreover, \astropypkg updates internal time data-files to
-    ensure that the correct leap second adjustment is always used. For
-    reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
-    manually applied.
-
-  \paragraph{Constants}
-
-    Measurements of physical constants (and the units defined from them) improve
-    over time. Periodically the standardized systems of units and constants are
-    updated to reflect these improvements, for instance in 2019 the SI system was
-    redefined \citep{NIST2019}, with an accompanying update to the physical
-    constants in
-    \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
-    2018}}. \astropypkg contains two modules -- \astropyconstants and
-    \astropyunits -- for working with constants and units, respectively.
-    \astropyconstants now defaults to use the \texttt{SI/CODATA 2018} values, with
-    the relation between units in \astropyunits based upon these constants. Most
-    of \astropypkg and affiliate packages build upon \astropyunits, so this update
-    has far-reaching effects across the \astropy ecosystem and astronomy
-    community.
-
-    For reproducibility, \astropypkg allows the constants (and therefore units)
-    definitions to be rolled back to prior values, for instance the
-    \texttt{SI/CODATA 2014} definitions.
-    % CITATION
-    The constants and units definitions are set by configuration file as well as
-    \astropyScienceState (an runtime configuration controller). For work where
-    updates to the values of the fundamental constants significantly alters
-    results we recommend including with the work an astropy configuration file,
-    which specifies the set of constants used.
-
-  \paragraph{Cosmology}
-    % TODO! At least one good sentence about "what is the benefit of astropy.cosmology"
-
-    \astropycosmology contains classes for representing cosmological models.
-    % currently all homogeneous and isotropic variations of the
-    % Friedmann-Lemaitre-Robertson-Walker metric.
-    Bundled with the classes are commonly used \astropyCosmology realizations, for
-    example best-fit measurements from
-    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
-    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
-    \citep{PlanckMission:2006}.
-
-    When these important missions publish new measurements, \astropycosmology is
-    updated to include these results as realizations of the appropriate class
-    (generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
-    always exactly correspond to \astropycosmology classes, for instance the
-    Planck 2018 results \citep{Planck2018VI} include massive neutrinos in
-    $\Omega_{matter,0}$, while in \astropyFlatLambdaCDM this mass contribution is
-    stored in a separate parameter. Consequently, while some parameter values
-    appear different from the source paper, the \astropyCosmology realizations
-    correctly reproduce the cosmological models.
-
-    Since \citealt{astropy:2018} the following cosmology realizations have been
-    added:
-    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
-    from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
-    Lya]{WMAP1Year:2003};
-    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
-    from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
-    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
-    from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
-    ext]{Planck2015XIII:2016};
-    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
-    from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE,
-    EE + lowE + lensing + BAO]{Planck2018VI:2020}.
-
-    Astropy provides a
-    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
-    default cosmology}. All calculations done in a cosmological context will use
-    this default cosmology. The default \astropyCosmology has been updated to the
-    Planck 2018 parameters. For reproducibility, the default cosmology may be set
-    to old defaults, such as the WMAP Three Year values. The default cosmology is
-    dynamically configurable using \astropyScienceState, meaning a set of
-    calculations may be run using different assumed cosmologies, and results
-    compared between the two.
-    % Repeat the word default less.
-
-  \paragraph{Coordinates}
-
-    Being able to describe the position of objects, i.e. their coordinates, is
-    fundamental across many knowledge domains. \astropypkg provides
-    \astropycoordinates for working with low-level positional and velocity data
-    all the way to high-level coordinate objects with reference frames,
-    atmospheric information, and more. A central feature of \astropycoordinates is
-    the ability to transform data between reference frames, e.g. \astropyICRS
-    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
-    there have been some historical limitations for spatially proximate
-    transformations, impacting the usefulness of \astropycoordinates for
-    ground-based telescopes and intra-solar astronomy. Transformations in the
-    \astropyAltAz frame were reasonably precise for very distant objects, but
-    wrong by up around several arcseconds for e.g. the location of the moon. Now
-    these transformations are much more precise, down to the milli-arcsecond
-    level. Similar precision improvements were made to the
-    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
-    transformations. Additionally,
-    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
-    and associated transformations have been updated to correctly reflect the
-    “true” (includes precession but not nutation) and “mean” (precession and
-    nutation) terminology.
-
-    For Galactic astronomers, the \astropyGalactocentric
-    \astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults} have
-    been updated to include more recent measurements (tabulated in the
-    \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
-    attribute). For reproducibility, old definitions are still available, and will
-    continue to be so as \astropypkg is updated with forthcoming measurements of
-    the Galactic center, solar motion, etc.
-    % may be used by setting the default value to ``pre-v4.0'' in the
-    % \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html}{\texttt{galactocentric\_frame\_defaults}}
-    % \astropyScienceState.
-
-\subsection*{Performance} \label{sec:core-features-performance}
-
-As features are added and corner cases dealt with, performance often
-suffers.  Hence, pull requests that improve performance are welcomed
-generally, and were even the main goal for one release cycle.  For
-most packages, this meant improvements in the \python\ code, but for
-some the most time-critical pieces were rewritten in~C (e.g., for
-convolution of images, sigma-clipping, and converting time strings to
-binary).  Furthermore, a particular effort was made to ensure
-\astropypkg\ is thread-safe, so that it can be used on supercomputer
-clusters.
-
-% \subsection*{Future} \label{sec:core-features-future}
-% MHvK: Merged with the new feature section
 
@@
 
+All coordinated and affiliated packages are listed on the astropy
+website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
+\cite{astropy:2018}, there has been an expansion of affiliated
+packages for gravitational astrophysics including:
+\texttt{PyCBC} for exploring gravitational wave signals, lenstronomy for
+modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
+gravitational wave probability maps, and \texttt{EinsteinPy} for general
+relativity and gravitational astronomy. There have been several packages
+added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
+for a BSD licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order 
+Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
+and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
+For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
+VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
+We have \texttt{agnpy} for AGN astrophysics, and \texttt{statmorph} for fitting galactic 
+morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
+routines. For modeling interstellar dust extinction we have 
+\texttt{dust_extinction}, and for extracting features from time series we 
+have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
+pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
+inference, and \texttt{sbpy} does calculations for asteroid and cometary 
+astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
+
+Several of the coordinated and affiliated packages described in [reference to
+paper 2] have had substantial improvements. \texttt{astroquery}
+\cite{astroquery} has added
+access to roughly a dozen new missions and data services, including JWST, and
+the project has switched to a continuous release model. Every time a change
+is committed to the main development branch it is published on the Python
+Package Index (PyPI, ref) and available for installation. Formal releases are
+still done a couple of times per year. \texttt{photutils} \cite
+{photutils} released its first stable version, indicating that the API will
+change less frequently, and there have been several significant performance
+improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
+version, bringing better performance to some image combination operations.
+The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
+region definitions \cite{ds9}, added new ways to manipulate regions and
+introduced new region types. \texttt{reproject}’s \cite{reproject} major new
+feature is a function to align and co-add images to create a mosaic; better
+support for parallelization was also added. \texttt{specutils}, the package
+that defines containers for 1D and 2D spectra \cite{specutils}, also had its
+first stable release and the addition of classes to read JWST data.
+\texttt{stingray} has also had a major performance overhaul.
+
 \subsection{Connections with data archives}
-
-\label{sec:astroquery}
-% MHvK: feel free to move above label to where it makes most sense.
 
@@
 
-Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
-It contains over 50 modules for querying astronomical databases, large and small.
-In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
-These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
-
-The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented.
-Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}].
-More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.
-
-Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
-These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
-While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
-% note to coauthors: we need a section on pyvo
+Summary of Astroquery and engagement with archives. Cite astroquery paper.
 

### 2022-02-23 src/ms.tex
Add summary of gammapy changes
Address review comments

@@
 packages for gravitational astrophysics including:
-\texttt{PyCBC} for exploring gravitational wave signals, lenstronomy for
+\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
 modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
@@
 added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
-for a BSD licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order 
+for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
 Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
@@
 VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
-We have \texttt{agnpy} for AGN astrophysics, and \texttt{statmorph} for fitting galactic 
+We have \texttt{agnpy} for AGN jets, and \texttt{statmorph} for fitting galactic
 morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
@@
 pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
-inference, and \texttt{sbpy} does calculations for asteroid and cometary 
+inference, and \texttt{sbpy} \cite{Mommert2019} does calculations for asteroid and cometary
 astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
 
-Several of the coordinated and affiliated packages described in [reference to
-paper 2] have had substantial improvements. \texttt{astroquery}
+Several of the coordinated and affiliated packages described in \cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
 \cite{astroquery} has added
@@
 first stable release and the addition of classes to read JWST data.
-\texttt{stingray} has also had a major performance overhaul.
+\texttt{stingray} has also had a major performance overhaul. The package
+\texttt{gammapy} \cite{gammapy} also has major performance improvements,
+has unified its API in preparation for its first stable release.
 

### 2022-02-24 src/ms.tex
Fix latex error
continue draft section on interoperability
Signed-off-by: nstarman <nstarkman@protonmail.com>

--- 2022-02-20 src/ms.tex
@@
 \newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
+\newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
 \newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
@@
 
-  % New features capture a lot of attention. While exciting, the arguably more
-  % significant improvements are those of interoperability -- ensuring existing
-  % features connect smoothly together.
-
-  % MHvK: this repeats the same thing 3 times; try to use just one sentence!
-  The modular nature of \astropypkg means that not all sub-packages deeply
-  integrate with one another -- and similarly between separate \python
-  libraries. Because \astropy is situated at the nexus between scientific
-  computing into astronomy, interoperability has been and continues to be an
-  important focus for \astropy, both within the various \astropypkg sub-packages
-  and with other important \python libraries. While increasing inter and
-  intra-package integration will be a perpetual goal -- with feature additions
-  in one place requiring updates in another -- \astropy has made impactful
-  progress with units, tables, and I/O.
-  % Better list? Also better transition
-  If done right, all a user notices is that code using many different python
-  features and libraries ``just works''.  Here, we describe a few
-  particular efforts towards this goal.
+  Because \astropy is modular and situated at the nexus between scientific
+  computing and astronomy, interoperability is an important focus for \astropy,
+  both within the various \astropypkg sub-packages and with other important
+  \python libraries. If done right, all a user notices is that code using many
+  different python features and libraries ``just works''. Here, we describe a
+  few particular efforts towards this goal.
 
@@
     functions.
-    % MHvK: I don't think the following is true, so commented it out.
-    % These advances have similarly benefitted many
-    % \package{numpy}-dependent packages, like \package{scipy} and their sparse
-    % matrices. And again \astropypkg benefits since more \package{scipy}
-    % functions now work with \astropyQuantity.
 
@@
     module} for manipulating structured arrays, compatibility only requires
-    exending the existing \package{numpy}-\astropypkg bridge framework.
+    extending the existing \package{numpy}-\astropypkg bridge framework.
     Community interest, in the form of a Feature Request (or better yet Pull
-    Request) would be sufficient to see this realized. For a few remaining functions,
-    discussed further in the
+    Request) would be sufficient to see this realized. For a few remaining
+    functions, discussed further in the
     \href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
@@
     \python community is to enhance and implement the frameworks, allowing
-    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem of
-    \package{numpy}-like projects.
-
-  \paragraph{Astropy in Time and on Units}
-
-    % Bad sentence ↓. Rewrite.
-    Not only did \astropyunits gain powerful integrations with \package{numpy},
-    there were many intra-package improvements.
-
-    \astropymodeling can add units to otherwise unit-less models.
-
+    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem
+    of \package{numpy}-like projects.
+
+  \paragraph{\astropypkg on Units}
+
+    Units were also integrated further within \astropypkg. For already-defined
+    \astropyModel, for instance imported from another library, \astropymodeling
+    can now add units to otherwise unit-less models.
+    % TODO! transition words
     A new module –
-    \astropycosmologyunits
-    – has been added to the cosmology sub-package for defining and collecting
-    cosmological units and equivalencies. The unit \texttt{littleh} and
-    equivalency
-    \astropyapidoc{cosmology.units.with_H0}{with\_H0}
-    are deprecated from the main \astropyunits sub-package and moved to
-    \astropycosmologyunits. A new unit, \texttt{redshift}, is added for tracking
-    factors of cosmological redshift.
-    % As this is a pseudo-unit an equivalency
-    % \astropyapidoc{cosmology.units.dimensionless_redshift}{dimensionless\_redshift}
-    % is added (and enabled by default) to allow for redshift - dimensionless
-    % conversions.
-    To convert between redshift and other cosmological distance
-    measures, e.g. CMB temperature or comoving distance, the equivalency
+    \astropycosmologyunits – has been added to the cosmology sub-package for
+    defining and collecting cosmological units and equivalencies. The unit
+    \texttt{littleh} and equivalency
+    \astropyapidoc{cosmology.units.with_H0}{with\_H0} are deprecated from the
+    main \astropyunits sub-package and moved to \astropycosmologyunits. A new
+    unit, \texttt{redshift}, is added for tracking factors of cosmological
+    redshift. To correctly use redshift units in dimensional analysis,
+    \texttt{redshift} is
+    \astropyapidoc{cosmology.units.dimensionless_redshift}{treated as
+    dimensionless} by default. To convert between redshift and other
+    cosmological distance measures, e.g. CMB temperature or comoving distance,
+    the equivalency
     \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
     is also added. This equivalency is actually a composite of other
-    equivalencies,
-    % : \astropyapidoc{cosmology.units.redshift_distance}{redshift\_distance},
-    % \astropyapidoc{cosmology.units.redshift_hubble}{redshift\_hubble}, and
-    % \astropyapidoc{cosmology.units.redshift_temperature}{redshift\_temperature},
-    which may be used separately.
-
+    equivalencies, which may be used separately.
+
+  \paragraph{Astropy Fits in Time}
     % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
+
     The FITS Time standard paper defines the formats and keywords used to
@@
 
-  \paragraph{Unified I/O architecture}
-
-    \astropyiosubpkg{registry} is a powerful way to define input and output
-    (I/O) functions for \astropypkg objects, such as reading or writing to a
-    file. \astropy uses this internally for the I/O methods in \astropyTable and
-    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
-    extend the options on these classes.
-
-    As of Astropy 5.0 the I/O registry submodule has switched to a class-based
-    architecture and been refactored to allow read-only, write-only, and
-    read-write registries. Additionally, the registry submodule no longer
-    assumes an object has \astropy-style \texttt{read} and \texttt{write}
-    methods when trying to read or write that object. These two improvements
-    enable a number of new use cases for \astropyiosubpkg{registry}, some
-    highlighted here, but others we hope to see have community adoption.
-
-    First, the class-based architecture allows for the creation of custom
-    registries. An application of this is in \astropycosmology, which
-    has two different \textit{kinds} of registries for \astropyCosmology: one
-    for reading and writing files, and another for converting between python
-    objects.
-
-    Second, \textit{any} kind of object is now compatible with
-    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
-    classes means a user does not have to create a whole new I/O system and can
-    offer a unified and familiar API.
-
   \paragraph{Persistent Storage and I/O}
 
-    % TODO Better transition sentence
+  % MHvK: maybe do this before the structure? Change in structure
+  % partially driven by this?
     The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.
@@
 
+  \paragraph{Unified I/O architecture}
+
+    \astropyiosubpkg{registry} is a powerful way to define input and output
+    (I/O) functions for \astropypkg objects, such as reading or writing to a
+    file. \astropy uses this internally for the I/O methods in \astropyTable and
+    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
+    extend the options on these classes.
+
+    As of Astropy 5.0 the I/O registry submodule has switched to a class-based
+    architecture and been refactored to allow read-only, write-only, and
+    read-write registries. Additionally, the registry submodule no longer
+    assumes an object has \astropy-style \texttt{read} and \texttt{write}
+    methods when trying to read or write that object. These two improvements
+    enable a number of new use cases for \astropyiosubpkg{registry}, some
+    highlighted here, but others we hope to see have community adoption.
+
+    First, the class-based architecture allows for the creation of custom
+    registries. An application of this is in \astropycosmology, which has two
+    different \textit{kinds} of registries for \astropyCosmology: one for
+    reading and writing files, and another for converting between python
+    objects. Second, \textit{any} kind of object is now compatible with
+    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
+    classes means a user does not have to create a whole new I/O system and can
+    offer a unified and familiar API.
+
 \subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
@@
   of its functionality, while, where appropriate, making sure results obtained
-  using older versions of \astropypkg are reproducible.
-  % Something about ScienceState
-  Notable improvements have been made to \astropytime, \astropyconstants,
-  \astropycoordinates, and \astropycosmology.
+  using older versions of \astropypkg are reproducible. Within each section we
+  include in our descriptions how old results may be reproduced. The primary
+  means are with configuration file options, and settings on runtime
+  configuration objects called \astropyScienceState, described further in the
+  documentation. Notable improvements to \astropypkg have been made to
+  \astropytime, \astropyconstants, \astropycoordinates, and \astropycosmology,
+  and we describe each in turn.
 
   \paragraph{Time}
-
-    In astronomy, time accuracy down to the second is frequently important.
-    Whether for planning observations or crunching data, missing seconds can
-    meaningfully impact results. Before, when using \astropyTime in \astropytime
-    leap seconds had to be manually applied; now leap seconds are applied
-    automatically. Moreover, \astropypkg updates internal time data-files to
-    ensure that the correct leap second adjustment is always used. For
-    reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
+    In astronomy, time accuracy down to the (nano)second is frequently
+    important. Whether for planning observations, crunching pulsar or VLBI data,
+    missing seconds meaningfully impact results. Before, when using \astropyTime
+    in \astropytime leap seconds had to be manually applied; now leap seconds
+    are applied automatically. Moreover, \astropypkg updates internal time
+    data-files to ensure that the correct leap second adjustment is always used.
+    For reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
     manually applied.

### 2022-02-24 src/ms.tex
spelling and fixes
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
   growing and evolving needs of the community and the maturation of the broader
-scientific \python ecosystem now driving the development of the \astropypkg
-core package.
-
-One manifestation of this was the realization that some features do not actually
-belong in the core package, either because they are of more general use
-outside of \astropypkg, or because they are too specialized. An example of the
-former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
+  scientific \python ecosystem driving much of the development of the
+  \astropypkg core package.
+
+  One manifestation of this was the realization that some features do not
+  actually belong in the core package, either because they are of more general
+  use outside of \astropypkg, or because they are too specialized. An example of
+  the former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
   software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
@@
   its content, the Simple Application Message Protocol made sense to keep
-(\astropysubpkg{samp}), but other parts belonged more properly in {\tt
-Astroquery} (see Sect.~\ref{sec:astroquery}).
-
-Within \astropypkg, development may be roughly split into a few
-categories: new features added, intra and inter-package
-interoperability, improvements to precision, accuracy, and
-reproducibility, and performance enhancements. We discuss these in turn.
+  (\astropysubpkg{samp}), but other parts belonged more properly in \package{
+  astroquery} (see Sect.~\ref{sec:astroquery}).
+
+  Within \astropypkg, development may be roughly split into a few categories:
+  new features added, intra and inter-package interoperability, improvements to
+  precision, accuracy, and reproducibility, and performance enhancements. We
+  discuss these in turn.
 
 \subsubsection*{New and Planned Features} \label{sec:core-features-new}
-
-% \begin{itemize}
-%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
-%  \item Uncertainties and Distributions
-%   \item ERFA module is now it’s own thing; Obliquely tie in that units
-%    remains in astropy -> above
-%  \item ? Pytest plugin ecosystem -> done elsewhere
-%  \item A new SpectralCoord class for representing and transforming spectral quantities
-%    Velocity in Coordinates
-%  \item WCS API : A shared Python interface for World Coordinate Systems
-%        Include APE 14 reference
-% \end{itemize}
 
@@
 to large rewrites of the interface of modules.  This range is spanned
-by the five main new features that entered the astropy core since the
+by the five main new features that entered the \astropypkg core since the
 previous summary \citepalias{astropy:2018}.
@@
 
-\paragraph{Uncertainties and Distributions.} Early in our carreers, we
+\paragraph{Uncertainties and Distributions.} Early in our careers, we
 were likely all taught that a measurement without a unit or an
@@
 In general, proper error propagation is difficult, and often it is
-best to use Monte Carle methods.  With the new
+best to use Monte Carlo methods.  With the new
 \astropysubpkg{uncertainties} sub-package, a first step is made to
@@
 course a \astropyQuantity, and the goal is to ensure it can be
-seamlessly used to instantiate also other astropy classes, such as
+seamlessly used to instantiate also other \astropypkg classes, such as
 \astropySkyCoord\ and \astropyTime.
@@
       \item Converting SkyCoord to QTable (1 sentence addition to above)
-      \item Added support for dask arrays in tables
+      \item Added support for \package{dask} arrays in tables
       \item Added support for registering array-like objects as mixin columns
@@
   % partially driven by this?
-    The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.
+    The improvements to \astropy's I/O architecture are mirrored in the increased support for I/O options.
 
@@
     updates to the values of the fundamental constants significantly alters
-    results we recommend including with the work an astropy configuration file,
+    results we recommend including with the work an \astropy configuration file,
     which specifies the set of constants used.
@@
     \astropycosmology contains classes for representing cosmological models.
-    % currently all homogeneous and isotropic variations of the
-    % Friedmann-Lemaitre-Robertson-Walker metric.
     Bundled with the classes are commonly used \astropyCosmology realizations, for
@@
     wrong by up around several arcseconds for e.g. the location of the moon. Now
-    these transformations are much more precise, down to the milli-arcsecond
+    these transformations are much more precise, down to the milliarcsecond
     level. Similar precision improvements were made to the
@@
 clusters.
-
-% \subsection*{Future} \label{sec:core-features-future}
-% MHvK: Merged with the new feature section
 

### 2022-02-24 src/ms.tex
Start section introduction
Signed-off-by: nstarman <nstarkman@protonmail.com>

--- 2022-02-23 src/ms.tex
@@
 
-Summarize major pieces of development since the last paper (since v2.0). A few
-ideas (taken from What's New pages) below, but feel free to expand on or remove
-from this list:
-\begin{itemize}
-    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-    \item APE 14 WCS
-    \item A new SpectralCoord class for representing and transforming spectral quantities
-    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
-\end{itemize}
+The \astropypkg core package is mature and stable, in that many features have
+been part of the core package for years with (purposefully) few changes to the
+software interface.
+This maturity and stability allows for the broader astronomy \python community
+to rely upon the \astropypkg core package and build specialized packages within
+the Astropy ecosystem.
+Even many non-astronomical \python libraries have come to rely on \astropypkg.
+
+The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
+% feels like ↑ could be expanded to 2 sentences
+
+% Maybe briefly outline the sectors of  
+% Using \astropypkg for
+% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
+% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
+% Astropy has grown to interoperate, e.g. NumPy
+
+Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
+
+% Summarize major pieces of development since the last paper (since v2.0). A few
+% ideas (taken from What's New pages) below, but feel free to expand on or remove
+% from this list:
+% \begin{itemize}
+%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+%     \item APE 14 WCS
+%     \item A new SpectralCoord class for representing and transforming spectral quantities
+%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+% \end{itemize}
+
+\subsubsection*{New Features} \label{sec:core-features-new}
+
+\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
+
+\subsection*{Precision, Accuracy, & Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
+
+\subsection*{Precision, Accuracy, & Reproducibility} \label{sec:core-features-performance}
+
+\subsection*{Precision, Accuracy, & Reproducibility} \label{sec:core-features-future}
 
@@
 
-All coordinated and affiliated packages are listed on the astropy
-website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
-\cite{astropy:2018}, there has been an expansion of affiliated
-packages for gravitational astrophysics including:
-\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
-modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
-gravitational wave probability maps, and \texttt{EinsteinPy} for general
-relativity and gravitational astronomy. There have been several packages
-added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
-for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
-Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
-and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
-For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
-We have \texttt{agnpy} for AGN jets, and \texttt{statmorph} for fitting galactic
-morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
-routines. For modeling interstellar dust extinction we have 
-\texttt{dust_extinction}, and for extracting features from time series we 
-have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
-pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
-inference, and \texttt{sbpy} \cite{Mommert2019} does calculations for asteroid and cometary
-astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
-
-Several of the coordinated and affiliated packages described in \cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
-\cite{astroquery} has added
-access to roughly a dozen new missions and data services, including JWST, and
-the project has switched to a continuous release model. Every time a change
-is committed to the main development branch it is published on the Python
-Package Index (PyPI, ref) and available for installation. Formal releases are
-still done a couple of times per year. \texttt{photutils} \cite
-{photutils} released its first stable version, indicating that the API will
-change less frequently, and there have been several significant performance
-improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
-version, bringing better performance to some image combination operations.
-The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
-region definitions \cite{ds9}, added new ways to manipulate regions and
-introduced new region types. \texttt{reproject}’s \cite{reproject} major new
-feature is a function to align and co-add images to create a mosaic; better
-support for parallelization was also added. \texttt{specutils}, the package
-that defines containers for 1D and 2D spectra \cite{specutils}, also had its
-first stable release and the addition of classes to read JWST data.
-\texttt{stingray} has also had a major performance overhaul. The package
-\texttt{gammapy} \cite{gammapy} also has major performance improvements,
-has unified its API in preparation for its first stable release.
-
 \subsection{Connections with data archives}

### 2022-02-24 src/ms.tex
start FITS <-> Time
Signed-off-by: nstarman <nstarkman@protonmail.com>

--- 2022-02-24 src/ms.tex
@@
 
-  \paragraph{Astropy Fits in Time}
-    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
-
-    The FITS Time standard paper defines the formats and keywords used to
-    represent timing information in FITS files. The Astropy FITS package now
-    provides support for reading and writing native Time table columns using
-    this format. This is described within the FITS unified I/O interface and
-    examples of usage can be found in the Astropy native objects (mixin columns)
-    section.
-
-    Implementation of this feature was made possible through the support of the
-    Google Summer of Code 2017 program. To learn more about this project please
-    see the final report, A mixin protocol for seamless interoperability.
+  \paragraph{Astropy Fits in Time}    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
+
+    As anyone who has worked with FITS files knows, the FITS standard is both
+    detailed and very flexible. While this means FITS is useful across many
+    domains, working with the files can be involved. Expanding on the WCS and
+    related work, the \astropyfits package now provides support for reading and
+    writing native \astropyTime table columns using the FITS Time standard
+    \citep{FITS-Time:2015}. For backward compatibility, this feature may be
+    turned off. Implementation of this feature was made possible through the
+    support of the Google Summer of Code 2017 program.
+    % TODO: is there a way to cite this
+    To learn more about this project please see the final report,
+    \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
+    mixin protocol for seamless interoperability}.
 

### 2022-02-27 src/ms.tex
Start Table Mixin Columns
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
 \newcommand{\astropyconstants}{\astropysubpkg{constants}}
+\newcommand{\astropytable}{\astropysubpkg{table}}
 \newcommand{\astropytime}{\astropysubpkg{time}}
 \newcommand{\astropymodeling}{\astropysubpkg{modeling}}
+\newcommand{\astropywcs}{\astropysubpkg{wcs}}
+\newcommand{\astropyfits}{\astropyiosubpkg{fits}}
 
@@
 \newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
+\newcommand{\astropyFitsColumn}{\astropyapidoc{io.fits.Column}{Column}}
 \newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
+\newcommand{\astropyTableColumn}{\astropyapidoc{table.Column}{Column}}
 \newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
@@
 
-  \paragraph{Astropy Fits in Time}    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
-
-    As anyone who has worked with FITS files knows, the FITS standard is both
-    detailed and very flexible. While this means FITS is useful across many
-    domains, working with the files can be involved. Expanding on the WCS and
-    related work, the \astropyfits package now provides support for reading and
-    writing native \astropyTime table columns using the FITS Time standard
-    \citep{FITS-Time:2015}. For backward compatibility, this feature may be
-    turned off. Implementation of this feature was made possible through the
-    support of the Google Summer of Code 2017 program.
-    % TODO: is there a way to cite this
-    To learn more about this project please see the final report,
+  \paragraph{Table Mixin Columns}
+
+    % TODO: this is very rough. Probably needs a bit of a re-write.
+
+    Within \astropypkg and \python there are numerous ways to represent and
+    store array-valued data. Some of the differences are historical:
+    \astropyapidoc{table.Column}{table.Column} and
+    \astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
+    differences are computational: \package{numpy}, \package{cupy}, and
+    \package{dask} arrays are almost API identical, but are optimized for
+    different compute needs. Lastly, some differences are inherent:
+    \astropyQuantity, \astropyTime and \astropySkyCoord represent fundamentally
+    different types of objects.
+
+    \astropy believes however array-valued data is stored, it should be usable
+    as a column in a table: historical differences are baked into stable API,
+    computational needs can differ, and some objects have extra information. The
+    \astropytable and \astropyfits table implementations allow for mixin-columns
+    (in general a mixin column may contain multiple data components as well as
+    object attributes beyond the standard column). A well defined protocol has
+    been developed to allow the usage of these mixin columns while allowing the
+    object to ``round-trip'' through tables with no loss of data or attributes.
+    With this protocol, it is now possible to store not only standard
+    \astropyTableColumn objects to a FITS table
+    \href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}, but
+    also any \astropy native objects within a \astropyTable. This includes
+    \astropyTime, \astropyQuantity, \astropySkyCoord, and many others. For
+    objects not already covered by the mixin protocol, functions can be
+    registered with \astropyTable to convert, on-the-fly, any array-like object
+    into a `mixin' column that can be used in a table. As an example, this is
+    used to provide the seamless integration of \package{dask} arrays in
+    \astropyTable, allowing computer cluster-scale data to be stored in a
+    \astropyTable.
+
+  \paragraph{Astropy Fits in Time}
+    % TODO! maybe merge with previous section
+
+    Time as a dimension in astronomical data presents challenges in its
+    representation in FITS files. The standard has therefore been extended to
+    describe rigorously the time coordinate in the World Coordinate System (WCS)
+    framework \citep{FITS-Time:2015}.
+
+    \astropyfits can store this data in a standard \astropyFitsColumn (for
+    details, see the
+    \href{https://docs.astropy.org/en/stable/io/fits/index.html}{FITS
+    documentation}), but will not ``understand'' the metadata and attributes.
+    Parsing the column to a useful time representation would be done manually.
+    Thankfully, through the support of the Google Summer of Code 2017 program,
+    % TODO: is there a way to cite this?
+    the \astropyfits package now allows \astropyTime as a `mixin' column.
+    \astropyTime columns in FITS tables ``understand'' the metadata -- scale,
+    reference, position, offset, etc. -- while also providing a similar
+    interface as the regular \astropyFitsColumn objects. For backward
+    compatibility, this feature may be turned off. To learn more about this
+    project please see the final report,
     \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
@@
 
-  \paragraph{Table Mixin Columns}
-
-    Numerous \python libraries have custom table implementations
+  \paragraph{Persistent Storage and I/O}
+
+    % TODO! need good transition sentence. This one is bad.
+    The column mixin options in tables are mirrored in the increased support for
+    I/O options.
+
     \begin{itemize}
-      \item (Support use of SkyCoord as Column happened before) and in table vstack, dstack, and insert\_row
-      \item Converting SkyCoord to QTable (1 sentence addition to above)
-      \item Added support for \package{dask} arrays in tables
-      \item Added support for registering array-like objects as mixin columns
-    \end{itemize}
-
-  \paragraph{Persistent Storage and I/O}
-
-  % MHvK: maybe do this before the structure? Change in structure
-  % partially driven by this?
-    The improvements to \astropy's I/O architecture are mirrored in the increased support for I/O options.
-
-    \begin{itemize}
-      \item ASDF : JSWST  and models can be stored with ASDF.
       \item ECSV : Support for multidimensional and object columns in ECSV
       \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
-      \item Support for reading, writing, and converting Cosmology
       \item MRT Support for reading and writing tables to MRT format : GSOC
@@
 
+    Also stuff that isn't a table
+
+    \begin{itemize}
+      \item Support for reading, writing, and converting Cosmology
+    \end{itemize}
+
   \paragraph{Unified I/O architecture}
 
+    % TODO! better transition
     \astropyiosubpkg{registry} is a powerful way to define input and output

### 2022-03-01 src/ms.tex
more documentation links (#35)
And a clean consistent means to define more.

Signed-off-by: nstarman <nstarkman@protonmail.com>

--- 2022-02-24 src/ms.tex
@@
 \newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
-
-% suggestion: shortcut reference to Paper II
-\newcommand{\paperii}{\cite{astropy:2018}}
 
@@
 
-The \astropypkg core package is mature and stable, in that many features have
-been part of the core package for years with (purposefully) few changes to the
-software interface.
-This maturity and stability allows for the broader astronomy \python community
-to rely upon the \astropypkg core package and build specialized packages within
-the Astropy ecosystem.
-Even many non-astronomical \python libraries have come to rely on \astropypkg.
-
-The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
-% feels like ↑ could be expanded to 2 sentences
-
-% Maybe briefly outline the sectors of  
-% Using \astropypkg for
-% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
-% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
-% Astropy has grown to interoperate, e.g. NumPy
-
-Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
-
-% Summarize major pieces of development since the last paper (since v2.0). A few
-% ideas (taken from What's New pages) below, but feel free to expand on or remove
-% from this list:
-% \begin{itemize}
-%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-%     \item APE 14 WCS
-%     \item A new SpectralCoord class for representing and transforming spectral quantities
-%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
-% \end{itemize}
-
-\subsubsection*{New Features} \label{sec:core-features-new}
-
-\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
-
-\subsection*{Precision, Accuracy, & Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
-
-\subsection*{Precision, Accuracy, & Reproducibility} \label{sec:core-features-performance}
-
-\subsection*{Precision, Accuracy, & Reproducibility} \label{sec:core-features-future}
+Summarize major pieces of development since the last paper (since v2.0). A few
+ideas (taken from What's New pages) below, but feel free to expand on or remove
+from this list:
+\begin{itemize}
+    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+    \item APE 14 WCS
+    \item A new SpectralCoord class for representing and transforming spectral quantities
+    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+\end{itemize}
 

### 2022-03-01 src/ms.tex
minor formatting edits to Learn section

--- 2022-03-01 src/ms.tex
@@
 \newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
-\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
-\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
-\newcommand{\astropyunits}{\astropysubpkg{units}}
-\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
-\newcommand{\astropyconstants}{\astropysubpkg{constants}}
-\newcommand{\astropytime}{\astropysubpkg{time}}
-\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
 
 % Links to documentation: classes
-\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
-\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}
-
-\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
-\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
-\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
-\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
-GalacticLSR}{GalacticLSR}}
-\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
-\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
-\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
-\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
-\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
-\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
-\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
-\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
-\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
-\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
-\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
-\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
+\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
+\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
+\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
+\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}
 
@@
     \begin{centering}
-      % \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
+        \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
         \caption{
@@
     \begin{centering}
-      % \includegraphics{figures/contributor-summary.pdf}
+        \includegraphics{figures/contributor-summary.pdf}
         \caption{Placeholder figure!}
@@
 Cherenkov Telescope Array
-\secauthor{Axel Donath, Maximilian Nöthe}
-
-The Cherenkov Telescope (CTA) will be the next generation very-high-energy
-gamma-ray observatory.
-CTA will improve over the current generation of imaging atmospheric Cherenkov telescopes (IACTs)
-by a factor of five to ten in sensitivity and will be able to observe the whole sky from a combination of two sites:
-a northern site in La Palma, Spain, and a southern one in Paranal, Chile.
-CTA will be able to observe gamma rays in a broad energy range from around $20\,\mathrm{GeV}$ to over $300\,\mathrm{TeV}$
-using three different types of telescopes, in total over 100 telescopes are planned at the two sites.
-CTA will also be the first open gamma-ray observatory.
-
-The data analysis pipeline is developed as open source software and essentially split in two domains:
-\begin{enumerate}
-  \item In the low-level analysis, the properties of the recorded air-shower events
-    have to be estimated from the raw data.
-    The raw data consists of very short (\textasciitilde $40--100\,\mathrm{ns}$) videos recorded with the fast and
-    sensitive cameras of the telescopes.
-    This includes the energy, particle type and direction of origin of the particle that induced the air shower
-    and the time the shower was recorded.
-  \item In the higher-level analysis, these reconstructed event lists are used together with some
-    characterization of the instrument response to perform the actual scientific analysis.
-    This software will be delivered as CTA science tools to the future users of the Observatory.
-\end{enumerate}
-
-A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
-a python package developed to perform all the necessary tasks to from the raw data
-of Cherenkov telescopes to the reconstructed event lists.
-
-The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
-Gammapy~\citep{gammapy}.
-
-Both Gammapy and \texttt{ctapipe} are using astropy heavily, mainly for units, times, coordinate transformations,
-tables and FITS IO.
-
-As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
-it needs to perform a large number of coordinate transformations.
-To enable this, CTA member M.~Nöthe contributed a major performance improvement
-for large numbers of coordinates with different observation times,
-based on earlier work by B.~Winkel.
-
-Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
-a total of 107 merged pull requests were contributed to astropy.
+\secauthor{Axel Donath, Max Noethe}
 
@@
 
-\begin{itemize}
-\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+{\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
 This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
@@
 spend helping users on these forums.
-\end{itemize}
+
 \secunfilled

### 2022-03-01 src/ms.tex

--- 2022-03-01 src/ms.tex
@@
 \newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
+
+% suggestion: shortcut reference to Paper II
+\newcommand{\paperii}{\cite{astropy:2018}}
 

### 2022-03-01 src/ms.tex
improvements to Astropy Fit + Time
Co-authored-by: Aarya Patil <aaryapatil1996@gmail.com>
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
 \newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
+\newcommand{\astropyFitsColumn}{\astropyapidoc{io.fits.Column}{Column}}
+\newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
+\newcommand{\astropyTableColumn}{\astropyapidoc{table.Column}{Column}}
 \newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
@@
 
-Summarize major pieces of development since the last paper (since v2.0). A few
-ideas (taken from What's New pages) below, but feel free to expand on or remove
-from this list:
+  The \astropypkg core package is mature and stable, in that many features have
+  been part of the core package for years with (purposefully) few changes to the
+  software interface. This maturity and stability allows for the broader
+  astronomy \python community to rely upon the \astropypkg core package and
+  build specialized packages within the Astropy ecosystem. Even many
+  non-astronomical \python libraries have come to rely on \astropypkg.
+
+  The relationship between \astropy and community is reciprocal, with the
+  growing and evolving needs of the community and the maturation of the broader
+  scientific \python ecosystem driving much of the development of the
+  \astropypkg core package.
+
+  One manifestation of this was the realization that some features do not
+  actually belong in the core package, either because they are of more general
+  use outside of \astropypkg, or because they are too specialized. An example of
+  the former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
+  software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
+  \astropy carried, as well as the \python wrappers for it.  Since this is more
+  basic infrastructure, with a release cycle set by SOFA, it made more sense to
+  create separate packages that \astropypkg can depend on: \package{ERFA}
+  \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
+  was too specialized was a Virtual Observatory sub-package (now removed).  Of
+  its content, the Simple Application Message Protocol made sense to keep
+  (\astropysubpkg{samp}), but other parts belonged more properly in \package{
+  astroquery} (see Sect.~\ref{sec:astroquery}).
+
+  Within \astropypkg, development may be roughly split into a few categories:
+  new features added, intra and inter-package interoperability, improvements to
+  precision, accuracy, and reproducibility, and performance enhancements. We
+  discuss these in turn.
+
+\subsubsection*{New and Planned Features} \label{sec:core-features-new}
+
+New features come in various forms, from new submodules providing
+wholly new capabilities, to larger additions to existing submodules,
+to large rewrites of the interface of modules.  This range is spanned
+by the five main new features that entered the \astropypkg core since the
+previous summary \citepalias{astropy:2018}.
+
+New features present a particular difficulty for a mature package such
+as \astropypkg, because by their nature they have not been tried out
+much and it is difficult to know in advance what the best interface
+will be, and hard to address all possible use cases in one go, even
+within \astropypkg\ itself.  To give a sense of this, we include in
+our descriptions how we hope the new features will evolve (realizing
+that, in the end, in a user-driven project like \astropy, what happens
+depends mostly on what is highest on the priority list of those that
+have time to contribute).
+
+\paragraph{Uncertainties and Distributions.} Early in our careers, we
+were likely all taught that a measurement without a unit or an
+uncertainty is essentially meaningless.  \astropyunits already
+provides the ability to associate numbers with units, and propagate
+these correctly, with its \astropyQuantity\ class, but so far the
+ability to associate and propagate uncertainties is relatively
+limited: in \astropysubpkg{nddata}, there are options to associate
+errors with data arrays, but these are treated as independent, i.e.,
+covariances are not tracked.
+
+In general, proper error propagation is difficult, and often it is
+best to use Monte Carlo methods.  With the new
+\astropysubpkg{uncertainties} sub-package, a first step is made to
+make this possible.  It allows one to generate, for each variable,
+randomly drawn samples in a \astropyDistribution, and then propagate
+these by passing them through the normal analysis, producing a
+\astropyDistribution\ of final results that can be inspected.
+
+This \astropyDistribution\ can be any type of array, including of
+course a \astropyQuantity, and the goal is to ensure it can be
+seamlessly used to instantiate also other \astropypkg classes, such as
+\astropySkyCoord\ and \astropyTime.
+
+Apart from Monte-Carlo error propagation, the plan for
+\astropysubpkg{uncertainties} is to also support the case where values
+have associated uncertainties, and these are propagated assuming that
+they are normally distributed, but tracking covariances.  One
+implementation problem to be solved is when to {\em stop} tracking
+covariances.  For instance, a simple operation such as subtracting the
+mean from $N$ data points implies that all data points are now
+covariant with each other, i.e., one has to carry $N\times N$
+covariances.  For a large image, that becomes not just pointless but
+also prohibitive in terms of memory and CPU usage.
+
+\paragraph{Masked quantities.}  Data can be known to be bad, in which
+case it is useful to mask them.  Support for masked data involves a
+number of choices, such as whether one indicates data values are bad
+by a separate flag or by replacing them by a special value, such as
+``Not a Number'' (NaN).  In \astropypkg, both approaches have been
+used: a flag for masked columns in \astropyTable, mask flags and
+bitmaps for N-dimensional data in \astropysubpkg{nddata}, and
+replacing elements with NaN in \astropyTime.  A particular problem was
+that no masked quantities were available, making it more difficult to
+migrate to \astropyQTable, in which \astropyQuantity\ is used for all
+columns with units, thus propagating those correctly (instead of
+treating the unit merely as metadata).
+
+Unfortunately, it turned out that the \texttt{MaskedArray} class from
+\package{numpy} could not be used, as its design made it a wrapper
+around any subclass, rather than allow another class to use it as an
+alternative to a regular data array.  For quantities, this does not
+work, as one can see from the fact that a masked value does not mean
+the unit is not known.  Given this, a new \astropyMasked\ class was
+designed, based on a framework very similar to that of
+\astropyDistribution, making it easy to create masked instances of
+other classes.  In particular, this can be used to create masked
+quantities for \astropyQTable, including full support for reading and
+writing to the different file types.
+
+The masked quantities work mostly without any changes in higher-level
+objects such as \astropySkyCoord, but work is still underway to expose
+the mask in those classes, as well as to use masked arrays in
+\astropyTime\ instead of replacing data with NaN.
+
+\paragraph{Time series.}  From sampling a continuous variable at fixed
+times to counting events binned into time windows, many different
+areas of astrophysics require the manipulation of time series.  These
+are often logically tables, and the new \astropysubpkg{timeseries}
+sub-package extends the \astropyQTable\ class to tables of data as a
+function of time, where the data can either represent samples or
+averages over particular time bins.  The new classes offer a number of
+special methods to manipulate time series (folding, resampling, etc.)
+and to read different data formats (such as {\em Kepler} lightcurves,
+etc.).
+
+Also part of \astropysubpkg{timeseries} are common analysis routines,
+including Lomb-Scargle and box-least-squares periodogrammes.
+
+\paragraph{Spectral Coordinates.} Measurements are often taken at
+specific ``spectral coordinates'', be they frequencies, wavelengths,
+or photon energies.  The new \astropySpectralCoord\ class allows to
+transform between these as well as to equivalent Doppler velocities.
+It builds on the \astropyQuantity, which can also represent these
+kinds of physical quantities, and allows conversion via dedicated
+equivalencies, but aims to make this more straightforward.
+Furthermore, it can be made aware of the observer and target reference
+frames, allowing for instance transformation from telescope-centric (or
+topocentric) frames to Barycentric or Local Standard of Rest
+(LSRK and LSRD) velocity frames.
+
+\paragraph{A High-Level Interface to World Coordinate Systems.}
+Astronomical data are often provided with information about the
+``real-world'' coordinates that correspond to pixel coordinates. This
+mapping is the essence of the ``World Coordinate System (WCS)''
+concept.  From its inception, the \astropysubpkg{wcs} sub-package
+allowed access to WCS information provided in, e.g., FITS files, but
+it became clear that other WCS standards and representations had to be
+supported for new missions and observatories (e.g., the James Webb
+Space Telescope and the Rubin Observatory).  To harmonize these, a new
+high-level interface was created, based on a formal design first
+proposed via an \astropy\ Proposal for Enhancement (\astropyAPE{14}).
+The hope is that by having a formal design, it will be straightforward
+for other packages implementing WCS objects to either modify their
+classes to conform to the new interface or build thin wrappers that
+conform to it.  The implementation in \astropysubpkg{wcs} interacts
+well with other \astropypkg\ objects such as \astropySkyCoord\ and
+\astropyTime.
+
+\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
+
+  Because \astropy is modular and situated at the nexus between scientific
+  computing and astronomy, interoperability is an important focus for \astropy,
+  both within the various \astropypkg sub-packages and with other important
+  \python libraries. If done right, all a user notices is that code using many
+  different python features and libraries ``just works''. Here, we describe a
+  few particular efforts towards this goal.
+
+  \paragraph{\package{numpy} on Units}
+
+    \astropyQuantity\ is a backbone of \astropy, leveraging the power of
+    \package{numpy} and adding units. Previously, many \package{numpy} functions
+    would strip a \astropyQuantity\ of its units (or fail outright), limiting
+    \astropyQuantity's usefulness. Now, advances in \package{numpy} function
+    overloading (see for example
+    \href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
+    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.17+)
+    functions.
+
+    In the remaining gaps, the \package{numpy} and \astropyQuantity\
+    interoperability efforts are ongoing. For some functions, such as in
+    \href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
+    module} for manipulating structured arrays, compatibility only requires
+    extending the existing \package{numpy}-\astropypkg bridge framework.
+    Community interest, in the form of a Feature Request (or better yet Pull
+    Request) would be sufficient to see this realized. For a few remaining
+    functions, discussed further in the
+    \href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
+    documentation}, the \package{numpy} framework does not yet allow for full
+    interoperability with \astropy. A goal of \astropy, and the scientific
+    \python community is to enhance and implement the frameworks, allowing
+    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem
+    of \package{numpy}-like projects.
+
+  \paragraph{\astropypkg on Units}
+
+    Units were also integrated further within \astropypkg. For already-defined
+    \astropyModel, for instance imported from another library, \astropymodeling
+    can now add units to otherwise unit-less models.
+    % TODO! transition words
+    A new module – \astropycosmologyunits – has been added to the cosmology
+    sub-package for defining and collecting cosmological units and
+    equivalencies. The unit \texttt{littleh} and equivalency
+    \astropyapidoc{cosmology.units.with_H0}{with\_H0} are deprecated from the
+    main \astropyunits sub-package and moved to \astropycosmologyunits. A new
+    unit, \texttt{redshift}, is added for tracking factors of cosmological
+    redshift. To correctly use redshift units in dimensional analysis,
+    \texttt{redshift} is
+    \astropyapidoc{cosmology.units.dimensionless_redshift}{treated as
+    dimensionless} by default. To convert between redshift and other
+    cosmological distance measures, e.g. CMB temperature or comoving distance,
+    the equivalency
+    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
+    is also added. This equivalency is actually a composite of other
+    equivalencies, which may be used separately.
+
+  \paragraph{Table Mixin Columns}
+
+    % TODO: this is very rough. Probably needs a bit of a re-write.
+
+    Within \astropypkg and \python there are numerous ways to represent and
+    store array-valued data. Some of the differences are historical:
+    \astropyapidoc{table.Column}{table.Column} and
+    \astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
+    differences are computational: \package{numpy}, \package{cupy}, and
+    \package{dask} arrays are almost API identical, but are optimized for
+    different compute needs. Lastly, some differences are inherent:
+    \astropyQuantity, \astropyTime and \astropySkyCoord represent fundamentally
+    different types of objects.
+
+    \astropy believes however array-valued data is stored, it should be usable
+    as a column in a table: historical differences are baked into stable API,
+    computational needs can differ, and some objects have extra information. The
+    \astropytable and \astropyfits table implementations allow for mixin-columns
+    (in general a mixin column may contain multiple data components as well as
+    object attributes beyond the standard column). A well defined protocol has
+    been developed to allow the usage of these mixin columns while allowing the
+    object to ``round-trip'' through tables with no loss of data or attributes.
+    With this protocol, it is now possible to store not only standard
+    \astropyTableColumn objects to a FITS table
+    \href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}, but
+    also any \astropy native objects within a \astropyTable. This includes
+    \astropyTime, \astropyQuantity, \astropySkyCoord, and many others. For
+    objects not already covered by the mixin protocol, functions can be
+    registered with \astropyTable to convert, on-the-fly, any array-like object
+    into a `mixin' column that can be used in a table. As an example, this is
+    used to provide the seamless integration of \package{dask} arrays in
+    \astropyTable, allowing computer cluster-scale data to be stored in a
+    \astropyTable.
+
+  \paragraph{Astropy Fits in Time}
+
+    The FITS standard was extended to rigorously describe time coordinates in
+    the World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared
+    to other coordinates in the WCS framework, time requires more metadata --
+    format, scale, position, reference etc. -- which had to be manually
+    specified and understood. \astropyfits can store this data in a standard
+    \astropyFitsColumn (for details, see the
+    \href{https://docs.astropy.org/en/stable/io/fits/index.html}{FITS
+    documentation}), but will not ``understand'' the metadata and attributes.
+    Thankfully, through the support of the
+    \href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
+    Summer of Code} 2017 program, the \astropyfits package now allows
+    \astropyTime as a `mixin' column. \astropyTime columns in FITS tables
+    ``understand'' the metadata while also providing a similar interface as the
+    regular \astropyFitsColumn objects. With \astropyTime columns, \astropy
+    ensures that time coordinates are stored according to the FITS standard in a
+    consistent manner and can be read/written without loss of information. For
+    backward compatibility with manual systems, this feature may be turned off,
+    but it is recommended to update. To learn more about this project please see
+    the final report,
+    \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
+    mixin protocol for seamless interoperability}.
+
+  \paragraph{Persistent Storage and I/O}
+
+    % TODO! need good transition sentence. This one is bad.
+    The column mixin options in tables are mirrored in the increased support for
+    I/O options.
+
     \begin{itemize}
-    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-    \item APE 14 WCS
-    \item A new SpectralCoord class for representing and transforming spectral quantities
-    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+      \item ECSV : Support for multidimensional and object columns in ECSV
+      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
+      \item MRT Support for reading and writing tables to MRT format : GSOC
+      \item Support for reading and writing tables to Parquet format
+      \item Support for reading and writing tables to QDP format
+      \item ASDF read write
     \end{itemize}
+
+    Also stuff that isn't a table
+
+    \begin{itemize}
+      \item Support for reading, writing, and converting Cosmology
+    \end{itemize}
+
+  \paragraph{Unified I/O architecture}
+
+    % TODO! better transition
+    \astropyiosubpkg{registry} is a powerful way to define input and output
+    (I/O) functions for \astropypkg objects, such as reading or writing to a
+    file. \astropy uses this internally for the I/O methods in \astropyTable and
+    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
+    extend the options on these classes.
+
+    As of Astropy 5.0 the I/O registry submodule has switched to a class-based
+    architecture and been refactored to allow read-only, write-only, and
+    read-write registries. Additionally, the registry submodule no longer
+    assumes an object has \astropy-style \texttt{read} and \texttt{write}
+    methods when trying to read or write that object. These two improvements
+    enable a number of new use cases for \astropyiosubpkg{registry}, some
+    highlighted here, but others we hope to see have community adoption.
+
+    First, the class-based architecture allows for the creation of custom
+    registries. An application of this is in \astropycosmology, which has two
+    different \textit{kinds} of registries for \astropyCosmology: one for
+    reading and writing files, and another for converting between python
+    objects. Second, \textit{any} kind of object is now compatible with
+    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
+    classes means a user does not have to create a whole new I/O system and can
+    offer a unified and familiar API.
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
+
+  An important focus of \astropy has been to increase the precision and accuracy
+  of its functionality, while, where appropriate, making sure results obtained
+  using older versions of \astropypkg are reproducible. Within each section we
+  include in our descriptions how old results may be reproduced. The primary
+  means are with configuration file options, and settings on runtime
+  configuration objects called \astropyScienceState, described further in the
+  documentation. Notable improvements to \astropypkg have been made to
+  \astropytime, \astropyconstants, \astropycoordinates, and \astropycosmology,
+  and we describe each in turn.
+
+  \paragraph{Time}
+    In astronomy, time accuracy down to the (nano)second is frequently
+    important. Whether for planning observations, crunching pulsar or VLBI data,
+    missing seconds meaningfully impact results. Before, when using \astropyTime
+    in \astropytime leap seconds had to be manually applied; now leap seconds
+    are applied automatically. Moreover, \astropypkg updates internal time
+    data-files to ensure that the correct leap second adjustment is always used.
+    For reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
+    manually applied.
+
+  \paragraph{Constants}
+
+    Measurements of physical constants (and the units defined from them) improve
+    over time. Periodically the standardized systems of units and constants are
+    updated to reflect these improvements, for instance in 2019 the SI system was
+    redefined \citep{NIST2019}, with an accompanying update to the physical
+    constants in
+    \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
+    2018}}. \astropypkg contains two modules -- \astropyconstants and
+    \astropyunits -- for working with constants and units, respectively.
+    \astropyconstants now defaults to use the \texttt{SI/CODATA 2018} values, with
+    the relation between units in \astropyunits based upon these constants. Most
+    of \astropypkg and affiliate packages build upon \astropyunits, so this update
+    has far-reaching effects across the \astropy ecosystem and astronomy
+    community.
+
+    For reproducibility, \astropypkg allows the constants (and therefore units)
+    definitions to be rolled back to prior values, for instance the
+    \texttt{SI/CODATA 2014} definitions.
+    % CITATION
+    The constants and units definitions are set by configuration file as well as
+    \astropyScienceState (an runtime configuration controller). For work where
+    updates to the values of the fundamental constants significantly alters
+    results we recommend including with the work an \astropy configuration file,
+    which specifies the set of constants used.
+
+  \paragraph{Cosmology}
+    % TODO! At least one good sentence about "what is the benefit of astropy.cosmology"
+
+    \astropycosmology contains classes for representing cosmological models.
+    Bundled with the classes are commonly used \astropyCosmology realizations, for
+    example best-fit measurements from
+    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
+    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
+    \citep{PlanckMission:2006}.
+
+    When these important missions publish new measurements, \astropycosmology is
+    updated to include these results as realizations of the appropriate class
+    (generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
+    always exactly correspond to \astropycosmology classes, for instance the
+    Planck 2018 results \citep{Planck2018VI} include massive neutrinos in
+    $\Omega_{matter,0}$, while in \astropyFlatLambdaCDM this mass contribution is
+    stored in a separate parameter. Consequently, while some parameter values
+    appear different from the source paper, the \astropyCosmology realizations
+    correctly reproduce the cosmological models.
+
+    Since \citealt{astropy:2018} the following cosmology realizations have been
+    added:
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
+    from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
+    Lya]{WMAP1Year:2003};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
+    from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
+    from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
+    ext]{Planck2015XIII:2016};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
+    from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE,
+    EE + lowE + lensing + BAO]{Planck2018VI:2020}.
+
+    Astropy provides a
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
+    default cosmology}. All calculations done in a cosmological context will use
+    this default cosmology. The default \astropyCosmology has been updated to the
+    Planck 2018 parameters. For reproducibility, the default cosmology may be set
+    to old defaults, such as the WMAP Three Year values. The default cosmology is
+    dynamically configurable using \astropyScienceState, meaning a set of
+    calculations may be run using different assumed cosmologies, and results
+    compared between the two.
+    % Repeat the word default less.
+
+  \paragraph{Coordinates}
+
+    Being able to describe the position of objects, i.e. their coordinates, is
+    fundamental across many knowledge domains. \astropypkg provides
+    \astropycoordinates for working with low-level positional and velocity data
+    all the way to high-level coordinate objects with reference frames,
+    atmospheric information, and more. A central feature of \astropycoordinates is
+    the ability to transform data between reference frames, e.g. \astropyICRS
+    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
+    there have been some historical limitations for spatially proximate
+    transformations, impacting the usefulness of \astropycoordinates for
+    ground-based telescopes and intra-solar astronomy. Transformations in the
+    \astropyAltAz frame were reasonably precise for very distant objects, but
+    wrong by up around several arcseconds for e.g. the location of the moon. Now
+    these transformations are much more precise, down to the milliarcsecond
+    level. Similar precision improvements were made to the
+    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
+    transformations. Additionally,
+    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
+    and associated transformations have been updated to correctly reflect the
+    “true” (includes precession but not nutation) and “mean” (precession and
+    nutation) terminology.
+
+    For Galactic astronomers, the \astropyGalactocentric
+    \astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults} have
+    been updated to include more recent measurements (tabulated in the
+    \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
+    attribute). For reproducibility, old definitions are still available, and will
+    continue to be so as \astropypkg is updated with forthcoming measurements of
+    the Galactic center, solar motion, etc.
+    % may be used by setting the default value to ``pre-v4.0'' in the
+    % \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html}{\texttt{galactocentric\_frame\_defaults}}
+    % \astropyScienceState.
+
+\subsection*{Performance} \label{sec:core-features-performance}
+
+As features are added and corner cases dealt with, performance often
+suffers.  Hence, pull requests that improve performance are welcomed
+generally, and were even the main goal for one release cycle.  For
+most packages, this meant improvements in the \python\ code, but for
+some the most time-critical pieces were rewritten in~C (e.g., for
+convolution of images, sigma-clipping, and converting time strings to
+binary).  Furthermore, a particular effort was made to ensure
+\astropypkg\ is thread-safe, so that it can be used on supercomputer
+clusters.
 

### 2022-03-01 src/ms.tex
Persistent storage section
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
 
-    Units were also integrated further within \astropypkg. For already-defined
+    Units were also integrated further within \astropypkg. For an already-defined
     \astropyModel, for instance imported from another library, \astropymodeling
@@
 
-    % TODO: this is very rough. Probably needs a bit of a re-write.
-
     Within \astropypkg and \python there are numerous ways to represent and
@@
 
-    \astropy believes however array-valued data is stored, it should be usable
-    as a column in a table: historical differences are baked into stable API,
-    computational needs can differ, and some objects have extra information. The
-    \astropytable and \astropyfits table implementations allow for mixin-columns
-    (in general a mixin column may contain multiple data components as well as
-    object attributes beyond the standard column). A well defined protocol has
-    been developed to allow the usage of these mixin columns while allowing the
-    object to ``round-trip'' through tables with no loss of data or attributes.
-    With this protocol, it is now possible to store not only standard
-    \astropyTableColumn objects to a FITS table
-    \href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}, but
+    \astropy believes that however array-valued data is stored, it should still
+    be usable as a column in a table. A well defined protocol for mixin-columns
+    has been developed for \astropytable and \astropyfits table, allowing the
+    original object to be used as a column with a familiar API, and to
+    ``round-trip'' through tables with no loss of data or attributes. With this
+    protocol, it is now possible to store standard \astropyTableColumn objects
+    to a FITS table
+    \href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}, and
     also any \astropy native objects within a \astropyTable. This includes
@@
     objects not already covered by the mixin protocol, functions can be
-    registered with \astropyTable to convert, on-the-fly, any array-like object
-    into a `mixin' column that can be used in a table. As an example, this is
-    used to provide the seamless integration of \package{dask} arrays in
-    \astropyTable, allowing computer cluster-scale data to be stored in a
-    \astropyTable.
+    registered with \astropyTable to convert on-the-fly any array-like object
+    into a mixin column. As an example, this is used to provide the seamless
+    integration of \package{dask} arrays with \astropyTable, allowing cloud-stored
+    or cluster-scale data to be used as a column in a \astropyTable.
 
@@
 
-  \paragraph{Persistent Storage and I/O}
-
-    % TODO! need good transition sentence. This one is bad.
-    The column mixin options in tables are mirrored in the increased support for
-    I/O options.
-
-    \begin{itemize}
-      \item ECSV : Support for multidimensional and object columns in ECSV
-      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
-      \item MRT Support for reading and writing tables to MRT format : GSOC
-      \item Support for reading and writing tables to Parquet format
-      \item Support for reading and writing tables to QDP format
-      \item ASDF read write
-    \end{itemize}
-
-    Also stuff that isn't a table
-
-    \begin{itemize}
-      \item Support for reading, writing, and converting Cosmology
-    \end{itemize}
+  \paragraph{Persistent Storage}
+
+    \astropy's efforts to increase support for column types in a table is
+    mirrored by efforts to expand storage format options. New formats have been
+    added, and existing formats updated to support more column types. We detail
+    the changes, starting with the new formats.
+
+    \astropy now supports reading and writing tables in the American
+    Astronomical Society Journals' Machine-Readable Table (MRT) format. This has
+    been a long-missing ASCII format, and we are pleased that this year's Google
+    Summer of Code project adds MRT to \astropyTable I/O.  
+    For details see the
+    \href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{Machine-Readable
+    Table Format} section. In addition, \astropyTable may also read and write to
+    \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
+    \href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
+    and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.
+
+    The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
+    standard} has been updated to version 1.0, adding support for reading and
+    writing \astropyTable with three additional data subtypes. First,
+    multidimensional column data (both masked and unmasked) with fixed
+    dimensions in all table cells. Second, multidimensional column data with
+    variable-dimension arrays similar to FITS variable-length arrays. Third,
+    object-type columns with simple \python objects.
+
+    As mentioned previously, \package{dask} arrays may be used as a mixin column
+    in \astropyTable. \package{dask} may also be the data array
+    \astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the
+    array will be computed while written, avoiding excessive memory use.
+    \astropyTable can now be appended to an existing FITS file, and
+    \package{dask} mixin columns interoperate seamlessly between the two table
+    types.
 

### 2022-03-01 src/ms.tex
Spectral coordinates

--- 2022-03-01 src/ms.tex
@@
 
-Summarize major pieces of development since the last paper (since v2.0). A few
-ideas (taken from What's New pages) below, but feel free to expand on or remove
-from this list:
+The \astropypkg core package is mature and stable, in that many features have
+been part of the core package for years with (purposefully) few changes to the
+software interface.
+This maturity and stability allows for the broader astronomy \python community
+to rely upon the \astropypkg core package and build specialized packages within
+the Astropy ecosystem.
+Even many non-astronomical \python libraries have come to rely on \astropypkg.
+
+The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
+% feels like ↑ could be expanded to 2 sentences
+
+% Maybe briefly outline the sectors of  
+% Using \astropypkg for
+% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
+% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
+% Astropy has grown to interoperate, e.g. NumPy
+
+Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
+
+% Summarize major pieces of development since the last paper (since v2.0). A few
+% ideas (taken from What's New pages) below, but feel free to expand on or remove
+% from this list:
+% \begin{itemize}
+%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+%     \item APE 14 WCS
+%     \item A new SpectralCoord class for representing and transforming spectral quantities
+%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+% \end{itemize}
+
+\subsubsection*{New Features} \label{sec:core-features-new}
+
 \begin{itemize}
-    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-    \item APE 14 WCS
-    \item A new SpectralCoord class for representing and transforming spectral quantities
-    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
+%  \item Uncertainties and Distributions
+  \item ERFA module is now it’s own thing; Obliquely tie in that units
+    remains in astropy
+  \item ? Pytest plugin ecosystem
+%  \item A new SpectralCoord class for representing and transforming spectral quantities
+%    Velocity in Coordinates
+  \item WCS API : A shared Python interface for World Coordinate Systems
+        Include APE reference
 \end{itemize}
+
+New features come in various forms, from new submodules providing
+wholly new capabilities, to larger additions to existing submodules,
+to large rewrites of the interface of modules.  This range is spanned
+by the four main features that entered the astropy core, which we
+describe below.
+
+\paragraph{Uncertainties and Distributions.} Early in our carreers, we
+were likely all taught that a measurement without a unit or an
+uncertainty is essentially meaningless.  \astropyunits already
+provides the ability to associate numbers with units, and propagate
+these correctly, with its \astropyQuantity\ class, but so far the
+ability to associate and propagate uncertainties is relatively
+limited: in \astropysubpkg{nddata}, there are options to associate
+errors with data arrays, but these are treated as independent, i.e.,
+covariances are not tracked.
+
+In general, proper error propagation is difficult, and often it is
+best to use Monte Carle methods.  With the new
+\astropysubpkg{uncertainties} sub-package, a first step is made to
+make this possible.  It allows one to generate, for each variable,
+randomly drawn samples in a \astropyDistribution, and then propagate
+these by passing them through the normal analysis, producing a
+\astropyDistribution\ of final results that can be inspected.
+
+This \astropyDistribution\ can be any type of array, including of
+course a \astropyQuantity, and the goal is to ensure it can be
+seamlessly used to instantiate also other astropy classes, such as
+\astropyskycoord\ and \astropyTime.
+
+\paragraph{Time series.}  From sampling a continuous variable at fixed
+times to counting events binned into time windows, many different
+areas of astrophysics require the manipulation of time series.  These
+are often logically tables, and the new \astropysubpkg{timeseries}
+sub-package extends the \astropyQTable\ class to tables of data as a
+function of time, where the data can either represent samples or
+averages over particular time bins.  The new classes offer a number of
+special methods to manipulate time series (folding, resampling, etc.)
+and to read different data formats (such as {\em Kepler} lightcurves,
+etc.).
+
+Also part of \astropysubpkg{timeseries} are common analysis routines,
+including Lomb-Scargle and box-least-squares periodogrammes.
+
+\paragraph{Spectral Coordinates.} Measurements are often taken at
+specific ``spectral coordinates'', be they frequencies, wavelengths,
+or photon energies.  The new \astropySpectralCoord\ class allows to
+transform between these as well as to equivalent Doppler velocities.
+It builds on the \astropyQuantity, which can also represent these
+kinds of physical quantities, and allows conversion via dedicated
+equivalencies, but aims to make this more straightforward.
+Furthermore, it can be made aware of the observer and target reference
+frames, allowing for instance transformation from telescope-centric (or
+topocentric) frames to Barycentric or Local Standard of Rest
+(LSRK and LSRD) velocity frames.
+
+\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}
+
+\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
 

### 2022-03-01 src/ms.tex
draft subsection on precision
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
   been part of the core package for years with (purposefully) few changes to the
-software interface.
-This maturity and stability allows for the broader astronomy \python community
-to rely upon the \astropypkg core package and build specialized packages within
-the Astropy ecosystem.
-Even many non-astronomical \python libraries have come to rely on \astropypkg.
+  software interface. This maturity and stability allows for the broader
+  astronomy \python community to rely upon the \astropypkg core package and
+  build specialized packages within the Astropy ecosystem. Even many
+  non-astronomical \python libraries have come to rely on \astropypkg.
 
-The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
-% feels like ↑ could be expanded to 2 sentences
+  The relationship between \astropy and community is reciprocal, with the
+  growing and evolving needs of the community and the maturation of the broader
+  scientific \python ecosystem now driving the development of the \astropypkg
+  core package.
 
-% Maybe briefly outline the sectors of  
-% Using \astropypkg for
-% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
-% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
-% Astropy has grown to interoperate, e.g. NumPy
+  One manifestation of this development realizing some features do not actually
+  belong in the core package, either because they are of more general use
+  outside of \astropypkg, or because they are too specialized. An example of the
+  former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
+  software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
+  \astropy carried, as well as the \python wrappers for it.  Since this is more
+  basic infrastructure, with a release cycle set by SOFA, it made more sense to
+  create separate packages that \astropypkg can depend on: \package{ERFA}
+  \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
+  was too specialized was a Virtual Observatory sub-package (now removed).  Of
+  its content, the Simple Application Message Protocol made sense to keep
+  (\astropysubpkg{samp}), but other parts belonged more properly in {\tt
+  Astroquery} (see Sect.~\ref{sec:astroquery}).
 
-Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
+  Within \astropypkg, development may be roughly split into a few categories:
+  new features added, performance enhancements, intra and inter-package
+  interoperability, and improvements to precision, accuracy, and
+  reproducibility. We discuss these in turn.
 
-% Summarize major pieces of development since the last paper (since v2.0). A few
-% ideas (taken from What's New pages) below, but feel free to expand on or remove
-% from this list:
+\subsubsection*{New and Planned Features} \label{sec:core-features-new}
+
 % \begin{itemize}
-%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-%     \item APE 14 WCS
-%     \item A new SpectralCoord class for representing and transforming spectral quantities
-%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
-% \end{itemize}
-
-\subsubsection*{New Features} \label{sec:core-features-new}
-
-\begin{itemize}
 %  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
 %  \item Uncertainties and Distributions
-  \item ERFA module is now it’s own thing; Obliquely tie in that units
-    remains in astropy
-  \item ? Pytest plugin ecosystem
+%   \item ERFA module is now it’s own thing; Obliquely tie in that units
+%    remains in astropy -> above
+%  \item ? Pytest plugin ecosystem -> done elsewhere
 %  \item A new SpectralCoord class for representing and transforming spectral quantities
 %    Velocity in Coordinates
-  \item WCS API : A shared Python interface for World Coordinate Systems
-        Include APE reference
-\end{itemize}
+%  \item WCS API : A shared Python interface for World Coordinate Systems
+%        Include APE 14 reference
+% \end{itemize}
 
@@
 
+\paragraph{A High-Level Interface to World Coordinate Systems.}
+Astronomical data are often provided with information about the
+``real-world'' coordinates that correspond to pixel coordinates. This
+mapping is the essence of the ``World Coordinate System (WCS)''
+concept.  From its inception, the \astropysubpkg{wcs} sub-package
+allowed access to WCS information provided in, e.g., FITS files, but
+it became clear that other WCS standards and representations had to be
+supported for new missions and observatories (e.g., the James Webb
+Space Telescope and the Rubin Observatory).  To harmonize these, a new
+high-level interface was created, based on a formal design first
+proposed via an \astropy\ Proposal for Enhancement (\astropyAPE{14}).
+The hope is that by having a formal design, it will be straightforward
+for other packages implementing WCS objects to either modify their
+classes to conform to the new interface or build thin wrappers that
+conform to it.  The implementation in \astropysubpkg{wcs} interacts
+well with other \astropypkg\ objects such as \astropySkyCoord\ and
+\astropyTime.
+
 \subsubsection*{Interoperability} \label{sec:core-features-interoperability}
@@
 \subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
+
+\subsection*{Precision, Accuracy, and Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
+
+  An important focus of \astropy has been to increase the precision and accuracy
+  of its functionality, while, where appropriate, making sure results obtained
+  using older versions of \astropypkg are reproducible.
+  % Something about ScienceState
+  Notable improvements have been made to \astropytime, \astropyconstants,
+  \astropycoordinates, and \astropycosmology.
+
+  \paragraph{Time}
+
+    In astronomy, time accuracy down to the second is frequently important.
+    Whether for planning observations or crunching data, missing seconds can
+    meaningfully impact results. Before, when using \astropyTime in \astropytime
+    leap seconds had to be manually applied; now leap seconds are applied
+    automatically. Moreover, \astropypkg updates internal time data-files to
+    ensure that the correct leap second adjustment is always used. For
+    reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
+    manually applied.
+
+  \paragraph{Constants}
+
+    Measurements of physical constants (and the units defined from them) improve
+    over time. Periodically the standardized systems of units and constants are
+    updated to reflect these improvements, for instance in 2019 the SI system was
+    redefined \citep{NIST2019}, with an accompanying update to the physical
+    constants in
+    \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
+    2018}}. \astropypkg contains two modules -- \astropyconstants and
+    \astropyunits -- for working with constants and units, respectively.
+    \astropyconstants now defaults to use the \texttt{SI/CODATA 2018} values, with
+    the relation between units in \astropyunits based upon these constants. Most
+    of \astropypkg and affiliate packages build upon \astropyunits, so this update
+    has far-reaching effects across the \astropy ecosystem and astronomy
+    community.
+
+    For reproducibility, \astropypkg allows the constants (and therefore units)
+    definitions to be rolled back to prior values, for instance the
+    \texttt{SI/CODATA 2014} definitions.
+    % CITATION
+    The constants and units definitions are set by configuration file as well as
+    \astropyScienceState (an runtime configuration controller). For work where
+    updates to the values of the fundamental constants significantly alters
+    results we recommend including with the work an astropy configuration file,
+    which specifies the set of constants used.
+
+  \paragraph{Cosmology}
+    % TODO! At least one good sentence about "what is the benefit of astropy.cosmology"
+
+    \astropycosmology contains classes for representing cosmological models.
+    % currently all homogeneous and isotropic variations of the
+    % Friedmann-Lemaitre-Robertson-Walker metric.
+    Bundled with the classes are commonly used \astropyCosmology realizations, for
+    example best-fit measurements from
+    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
+    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
+    \citep{PlanckMission:2006}.
+
+    When these important missions publish new measurements, \astropycosmology is
+    updated to include these results as realizations of the appropriate class
+    (generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
+    always exactly correspond to \astropycosmology classes, for instance the
+    Planck 2018 results \citep{Planck2018VI} include massive neutrinos in
+    $\Omega_{matter,0}$, while in \astropyFlatLambdaCDM this mass contribution is
+    stored in a separate parameter. Consequently, while some parameter values
+    appear different from the source paper, the \astropyCosmology realizations
+    correctly reproduce the cosmological models.
+
+    Since \citealt{astropy:2018} the following cosmology realizations have been
+    added:
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
+    from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
+    Lya]{WMAP1Year:2003};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
+    from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
+    from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
+    ext]{Planck2015XIII:2016};
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
+    from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE,
+    EE + lowE + lensing + BAO]{Planck2018VI:2020}.
+
+    Astropy provides a
+    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
+    default cosmology}. All calculations done in a cosmological context will use
+    this default cosmology. The default \astropyCosmology has been updated to the
+    Planck 2018 parameters. For reproducibility, the default cosmology may be set
+    to old defaults, such as the WMAP Three Year values. The default cosmology is
+    dynamically configurable using \astropyScienceState, meaning a set of
+    calculations may be run using different assumed cosmologies, and results
+    compared between the two.
+    % Repeat the word default less.
+
+  \paragraph{Coordinates}
+
+    Being able to describe the position of objects, i.e. their coordinates, is
+    fundamental across many knowledge domains. \astropypkg provides
+    \astropycoordinates for working with low-level positional and velocity data
+    all the way to high-level coordinate objects with reference frames,
+    atmospheric information, and more. A central feature of \astropycoordinates is
+    the ability to transform data between reference frames, e.g. \astropyICRS
+    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
+    there have been some historical limitations for spatially proximate
+    transformations, impacting the usefulness of \astropycoordinates for
+    ground-based telescopes and intra-solar astronomy. Transformations in the
+    \astropyAltAz frame were reasonably precise for very distant objects, but
+    wrong by up around several arcseconds for e.g. the location of the moon. Now
+    these transformations are much more precise, down to the milli-arcsecond
+    level. Similar precision improvements were made to the
+    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
+    transformations. Additionally,
+    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
+    and associated transformations have been updated to correctly reflect the
+    “true” (includes precession but not nutation) and “mean” (precession and
+    nutation) terminology.
+
+    For Galactic astronomers, the \astropyGalactocentric
+    \astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults} have
+    been updated to include more recent measurements (tabulated in the
+    \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
+    attribute). For reproducibility, old definitions are still available, and will
+    continue to be so as \astropypkg is updated with forthcoming measurements of
+    the Galactic center, solar motion, etc.
+    % may be used by setting the default value to ``pre-v4.0'' in the
+    % \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html}{\texttt{galactocentric\_frame\_defaults}}
+    % \astropyScienceState.
 

### 2022-03-01 src/ms.tex
Rview comments addresssed

--- 2022-03-01 src/ms.tex
@@
 
-The \astropypkg core package is mature and stable, in that many features have
-been part of the core package for years with (purposefully) few changes to the
-software interface.
-This maturity and stability allows for the broader astronomy \python community
-to rely upon the \astropypkg core package and build specialized packages within
-the Astropy ecosystem.
-Even many non-astronomical \python libraries have come to rely on \astropypkg.
-
-The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
-% feels like ↑ could be expanded to 2 sentences
-
-% Maybe briefly outline the sectors of  
-% Using \astropypkg for
-% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
-% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
-% Astropy has grown to interoperate, e.g. NumPy
-
-Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
-
-% Summarize major pieces of development since the last paper (since v2.0). A few
-% ideas (taken from What's New pages) below, but feel free to expand on or remove
-% from this list:
-% \begin{itemize}
-%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-%     \item APE 14 WCS
-%     \item A new SpectralCoord class for representing and transforming spectral quantities
-%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
-% \end{itemize}
-
-\subsubsection*{New Features} \label{sec:core-features-new}
-
+Summarize major pieces of development since the last paper (since v2.0). A few
+ideas (taken from What's New pages) below, but feel free to expand on or remove
+from this list:
 \begin{itemize}
-%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
-%  \item Uncertainties and Distributions
-  \item ERFA module is now it’s own thing; Obliquely tie in that units
-    remains in astropy
-  \item ? Pytest plugin ecosystem
-%  \item A new SpectralCoord class for representing and transforming spectral quantities
-%    Velocity in Coordinates
-  \item WCS API : A shared Python interface for World Coordinate Systems
-        Include APE reference
+    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+    \item TODO: something about WCS, SpectralCoord?
 \end{itemize}
-
-New features come in various forms, from new submodules providing
-wholly new capabilities, to larger additions to existing submodules,
-to large rewrites of the interface of modules.  This range is spanned
-by the four main features that entered the astropy core, which we
-describe below.
-
-\paragraph{Uncertainties and Distributions.} Early in our carreers, we
-were likely all taught that a measurement without a unit or an
-uncertainty is essentially meaningless.  \astropyunits already
-provides the ability to associate numbers with units, and propagate
-these correctly, with its \astropyQuantity\ class, but so far the
-ability to associate and propagate uncertainties is relatively
-limited: in \astropysubpkg{nddata}, there are options to associate
-errors with data arrays, but these are treated as independent, i.e.,
-covariances are not tracked.
-
-In general, proper error propagation is difficult, and often it is
-best to use Monte Carle methods.  With the new
-\astropysubpkg{uncertainties} sub-package, a first step is made to
-make this possible.  It allows one to generate, for each variable,
-randomly drawn samples in a \astropyDistribution, and then propagate
-these by passing them through the normal analysis, producing a
-\astropyDistribution\ of final results that can be inspected.
-
-This \astropyDistribution\ can be any type of array, including of
-course a \astropyQuantity, and the goal is to ensure it can be
-seamlessly used to instantiate also other astropy classes, such as
-\astropyskycoord\ and \astropyTime.
-
-\paragraph{Time series.}  From sampling a continuous variable at fixed
-times to counting events binned into time windows, many different
-areas of astrophysics require the manipulation of time series.  These
-are often logically tables, and the new \astropysubpkg{timeseries}
-sub-package extends the \astropyQTable\ class to tables of data as a
-function of time, where the data can either represent samples or
-averages over particular time bins.  The new classes offer a number of
-special methods to manipulate time series (folding, resampling, etc.)
-and to read different data formats (such as {\em Kepler} lightcurves,
-etc.).
-
-Also part of \astropysubpkg{timeseries} are common analysis routines,
-including Lomb-Scargle and box-least-squares periodogrammes.
-
-\paragraph{Spectral Coordinates.} Measurements are often taken at
-specific ``spectral coordinates'', be they frequencies, wavelengths,
-or photon energies.  The new \astropySpectralCoord\ class allows to
-transform between these as well as to equivalent Doppler velocities.
-It builds on the \astropyQuantity, which can also represent these
-kinds of physical quantities, and allows conversion via dedicated
-equivalencies, but aims to make this more straightforward.
-Furthermore, it can be made aware of the observer and target reference
-frames, allowing for instance transformation from telescope-centric (or
-topocentric) frames to Barycentric or Local Standard of Rest
-(LSRK and LSRD) velocity frames.
-
-\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
 
@@
 
-With support from the Moore Foundation, the \astropy Project was
-allocated funding to support mentoring programs. In 2020 a call was
-made to submit proposals for IDE (Inclusion, Diversity, and
-Empowerment) initiatives for project-wide consideration on
-Github. This process was deemed the most ``open'' because it allowed
-for community wide feedback to focus and improve proposal
-initiatives. Two programs were selected via this process, described
-below.
-
-\textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}
-
-\textbf{Women of Color Code (WoCCode)} is a peer-mentoring network for
-coders from traditionally marginalized groups, most notably women of
-color. Participants were invited to the WoCCode Slack space and
-encouraged to attend monthly webinars to share skills in the context
-of open source software libraries. Every other month, a guest speaker
-was invited to talk about their career path and share a skill. Program
-participants were solicited in the fall of 2020, yielding 73
-applications from 17 countries (48\% from the United States, 37\% from
-Africa, and 15\% from remaining continents). We selected participants
-who we identified as having a high potential for contributing to open
-source projects: intermediate to advanced programming skills with a
-vocal interest in contributing. Of the thirty applicants invited to
-the program, nineteen joined the community on Slack. Participants were
-organized into cohorts based on interest and each cohort was assigned
-one of three mentors that were also selected via open
-application. Mentors acted as a general resource to participants, gave
-one webinar, and organized a hack day. WoCCode also supported
-registration of two participants to attend the American Astronomical
-Society virtual summer 2021 meeting.
-
-Participants rated the impact of the program as very high. In the
-final webinar, participants reported a change in their perception of
-coding in general, for example, accepting coding as something they can
-do for fun. Participants reported generally feeling comfortable asking
-questions and interacting with a community where ``[e]veryone's
-thoughts are welcomed, no one is made to feel less important'' and
-where one can enjoy ``warm interactions with likeminded people.''
-WoCCode is continuing into 2022 by broadening participation in the
-Slack space and publicly advertising the guest webinar events.
-
-In addition to the supported mentoring initiatives, the \astropy
-project as a whole has taken steps to examine representation of
-marginalized groups within the project and search for avenues of
-improvement.
-
-% Much of this text was adapted from Robel Geda's report from NSBP and
-% SACNAS. Please include him on the author list.
-\textbf{\astropy representation at national diversity conferences:}
-Several members of the \astropy community attended virtual conferences
-of the National Society of Black Physicists (NSBP) and SACNAS
-(focusing on Hispanic and Native American scientists across all STEM
-fields). \astropy representatives noted that one underlying topic came
-up multiple times. A major barrier for persons who come from
-underrepresented communities is the lack of resources and expertise
-that are necessary to train students. Writing PEP 8 compliant software
-and understanding the Git and Github workflow is not part of the
-standard Physics and Astrophysics curriculum. Additionally, the larger
-Astronomnical community is still in the process of transitioning
-towards Python based tools, making it difficult for students not
-currently under advisement by some one with extensive Python expertise
-to get involved with the open source Python community. Projects such
-as Learn Astropy could have a profound impact by empowering
-underrepresented groups because it provides a free, searchable, and
-accessible introduction to Python tools for astrophysical
-research. Representatives also noted that offering training and
-teaching materials to PIs of Research Experience for Undergraduate
-(REU) programs throughout the United States could be helpful for
-encouraging advisors to teach students how to use Astropy, create
-their own libraries, and use version control on open source platforms
-like Github. Such materials could also be offered as a workshop at
-national diversity conferences themselves, as SACNAS solicits special
-session proposals each year.
-
-\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
-  initiative:} \todo{No author assigned}{Describe more about this.}
-
 \subsection{Funding} \label{sec:project-funding}
@@
 
-Summarize funding sources (Moore, NASA) and amounts and what this has been used
-for.
+% Summarize funding sources (Moore, NASA) and amounts and what this has been
+% used for.
+
+With the goal of establishing a sustainable financial model for the Astropy
+Project, the Coordination Committee under the new governance charter (see \S
+\ref{sec:project-governance}) established a standing Finance committee.
+This committee oversee the planning and allocation of finances on behalf
+of the project. The Astropy Project accepts funds from institutions as well as
+individuals through NUMFOCUS\footnote{NUMFOCUS is a 501(c)(3) nonprofit that
+supports and promotes world-class, innovative, open source scientific computing.}.
+Previously, no direct financial support was available for the project development,
+and NUMFOCUS covered most of the incurred operational costs. However, transitioning
+to long-term financial stability meant having an influx of funds in the form of
+institutional support, and hence the Astropy Project applied for and successfully
+acquired four grants. These grants are administered by NumFOCUS on behalf of the
+project, providing a ``neutral'' space not tied to any particular astronomy
+institution. The two major ones are briefly summarized below.
+
+\begin{itemize}
+    \item Moore: The Gordon and Betty Moore Foundation awarded the Astropy
+    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
+    Growing the Astropy Project". Over a three-year term, this grant supported
+    sustenance of the project by providing funds for (1) the development and
+    maintenance of the \astropypkg and its infrastructure to existing project
+    members and targeted hires, (2) the transition of long-term users to
+    contributors and maintainers through mentorship efforts, (3) the formalization
+    of a governance structure (see \S \ref{sec:project-governance}), and (4) the
+    improvement of equity, diversity, and inclusion efforts in the project (see \S
+    \ref{sec:project-ide}). Some of the money supported travel for meetings,
+    conferences, and workshops. This generous grant helped reduce reliance on
+    volunteer-driven maintenance of the openly-developed package and paved
+    the way to acquire funding from federal agencies: a big milestone for the
+    success of the project.
+
+    \item NASA: A successful proposal to the NASA ROSES-2020
+    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
+    (Support for Open Source Tools, Frameworks, and Libraries), granted the
+    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
+    term of three years includes (1) project-wide infrastructure maintenance and
+    improvements, (2) targeted work on areas of the \astropy core and
+    coordinated packages that need support, (3) enhancements to the Learn
+    Astropy ecosystem and educational materials (see \S \ref{sec:learn}),
+    and (4) support for the \astropy affiliated packages. These goals
+    support the sustenance of the project by encouraging further
+    development of the package ecosystem and community engagement.
+\end{itemize}
+
+In addition to the two grants described above, the project benefited from
+a Gemini Observatory contract under its Science User Support Department,
+which awarded funds for development work that supports both the Astropy and
+DRAGONS projects. The project also received financial support from the Dunlap
+Seed Funding program at the University of Toronto to develop educational
+resources for scientific software packaging within the Learn Astropy
+framework. All these successful proposals show that the Astropy project
+is vital for the astronomical infrastructure and community at large, and
+strengthen its promising and sustainable future.
+
+% ADD text for the new finance APE once it has a doi
 
@@
 
-\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
+\secauthor{Lía Corrales, David Shupe + Learn team}
 
-\subsubsection{Current status and scope}
-
-\textit{Learn Astropy} is an umbrella term that acknowledges the broad
-educational efforts made by the Astropy Project, which are led by the Learn
-Team.
-The efforts focus on developing online content and workshops covering
-astronomy-specific coding tasks in Python.
-As introduced in \citet{astropy:2018}, there are four different types of
-Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
-lessons that are published in HTML format online; \textit{guides}, which are a
-series of lessons providing a foundational resource for performing certain
-type of astronomical analyses; \textit{examples}, which are snippets of code
-that showcase a short task that can be performed with Astropy or an affiliated
-package; and \textit{documentation}, which is contained within the code base.
-This categorization drives content development, infrastructure choices, and the
-appearance of the Learn Astropy website.
-The Learn Team meets weekly to work on creating, expanding, improving these
-educational resources.
-
-\begin{description}
-
-\item[Learn Astropy website] The Learn Team re-launched the main website and
-search interface for Learn Astropy in 2021 with a new infrastructure platform,
-built around full-text search and interactive filtering functionality, with the
-goal of making content discoverable as the Learn Astropy content catalog
-expands.
-This work has been supported in part by a grant from the Dunlap Institute.
-We have adopted Algolia, a search-as-a-service cloud platform, to store the
-full-text and metadata records of Learn Astropy's content.
-The new Learn Astropy website is a JavaScript (Gatsby/React) application that
-uses the Algolia service to power its search and filtering user interface.
-Our Python-based application, Learn Astropy Librarian, populates data into the
-Algolia service. We tuned the Librarian around specific content formats (such as
-Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
-accurately index content and heuristically extract metadata.
-A consequence of the new platform is that we now maintain and compile content
-separately from the website application itself, enabling new content types.
-Tutorials, which are written as Jupyter Notebooks, are now compiled into their
-own Learn Astropy sub-site using
-\href{https://github.com/astropy/nbcollection}{\texttt{nbcollection}}.
-Guides, which utilize the \href{https://jupyterbook.org/}{Jupyter Book} build
-infrastructure, are also deployed as separate websites using GitHub pages.
-This architecture opens future possibilities of indexing third-party content,
-hosted elsewhere, such as on institutional websites.
-
-\item[Tutorials] Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
-
-\item[Guides] {\it ccdproc} Jupyter book has been completed.
-
-\item[Workshops] The Project has been conducting workshops at winter
-meetings of the American Astronomical Society since AAS 225 in January 2015.
-Up to the start of the coronavirus pandemic, these were full-day in-person
-workshops with as many as 90 participants and a dozen facilitators from the
-project.
-During the pandemic, these workshops were moved to an online format and split
-into basic and advanced sessions.
-Additionally, beginning with the AAS 238 online meeting, the workshops have been
-expanded to Summer AAS meetings.
-The Learn team finds that the workshop audience is best found as AAS meetings as
-opposed to more general Python meetings.
-
-The Astropy Project provided another mode of community engagement at AAS
-Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
-AAS Exhibit Hall.
-Funding for the exhibit hall was provided alternately by NumFOCUS and later by
-the Moore Foundation funding (\textbf{check this}).
-The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
-during the virtual AAS 237 meetings, to provide the general astronomy community
-information and access to experts on a variety of open source astronomical
-tools.
-\end{description}
-
-\subsubsection{Learn vision for the future}
-
-The Learn plan going forward is to continue to improve the tutorial website and
-facilate new content;
-to index third-party tutorials; and to look for supportable opportunities to
-expand the reach of Astropy workshops beyond the American Astronomical Society
-meetings.
+\begin{itemize}
+\item Current status and scope
+\item Vision for the future (review what we said in v2.0 paper)
+\item User forums and engagement with user base
+\item Workshops
+\end{itemize}
 
@@
 
-\section{Community Engagement}
-
-\begin{itemize}
-\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
-This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
-The Moore Foundation grant allows this Slack space to
-be on a paid plan; additionally NumFOCUS has negotiated a special rate for
-open-source projects.
-While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
-The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
-a Discourse site which is more open than the Facebook group and more user-oriented
-than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
-developers are able to invoice as independent contractors the time they
-spend helping users on these forums.
-\end{itemize}
+\secunfilled
 

### 2022-03-03 src/ms.tex
formatting, grammar, and clarity
Signed-off-by: nstarman <nstarkman@protonmail.com>

--- 2022-03-01 src/ms.tex
@@
 been part of the core package for years with (purposefully) few changes to the
-software interface.
-This maturity and stability allows for the broader astronomy \python community
-to rely upon the \astropypkg core package and build specialized packages within
-the Astropy ecosystem.
-Even many non-astronomical \python libraries have come to rely on \astropypkg.
-
-The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
-% feels like ↑ could be expanded to 2 sentences
-
-% Maybe briefly outline the sectors of  
-% Using \astropypkg for
-% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
-% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
-% Astropy has grown to interoperate, e.g. NumPy
-
-Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.
-
-% Summarize major pieces of development since the last paper (since v2.0). A few
-% ideas (taken from What's New pages) below, but feel free to expand on or remove
-% from this list:
-% \begin{itemize}
-%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
-%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
-%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
-%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
-%     \item APE 14 WCS
-%     \item A new SpectralCoord class for representing and transforming spectral quantities
-%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
-% \end{itemize}
-
-\subsubsection*{New Features} \label{sec:core-features-new}
-
-\begin{itemize}
-%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
-%  \item Uncertainties and Distributions
-  \item ERFA module is now it’s own thing; Obliquely tie in that units
-    remains in astropy
-  \item ? Pytest plugin ecosystem
-%  \item A new SpectralCoord class for representing and transforming spectral quantities
-%    Velocity in Coordinates
-  \item WCS API : A shared Python interface for World Coordinate Systems
-        Include APE reference
-\end{itemize}
-
-New features come in various forms, from new submodules providing
-wholly new capabilities, to larger additions to existing submodules,
-to large rewrites of the interface of modules.  This range is spanned
-by the four main features that entered the astropy core, which we
-describe below.
-
-\paragraph{Uncertainties and Distributions.} Early in our carreers, we
-were likely all taught that a measurement without a unit or an
-uncertainty is essentially meaningless.  \astropyunits already
-provides the ability to associate numbers with units, and propagate
-these correctly, with its \astropyQuantity\ class, but so far the
-ability to associate and propagate uncertainties is relatively
-limited: in \astropysubpkg{nddata}, there are options to associate
-errors with data arrays, but these are treated as independent, i.e.,
-covariances are not tracked.
-
-In general, proper error propagation is difficult, and often it is
-best to use Monte Carle methods.  With the new
-\astropysubpkg{uncertainties} sub-package, a first step is made to
-make this possible.  It allows one to generate, for each variable,
-randomly drawn samples in a \astropyDistribution, and then propagate
-these by passing them through the normal analysis, producing a
-\astropyDistribution\ of final results that can be inspected.
-
-This \astropyDistribution\ can be any type of array, including of
-course a \astropyQuantity, and the goal is to ensure it can be
-seamlessly used to instantiate also other astropy classes, such as
-\astropyskycoord\ and \astropyTime.
-
-\paragraph{Time series.}  From sampling a continuous variable at fixed
-times to counting events binned into time windows, many different
-areas of astrophysics require the manipulation of time series.  These
-are often logically tables, and the new \astropysubpkg{timeseries}
-sub-package extends the \astropyQTable\ class to tables of data as a
-function of time, where the data can either represent samples or
-averages over particular time bins.  The new classes offer a number of
-special methods to manipulate time series (folding, resampling, etc.)
-and to read different data formats (such as {\em Kepler} lightcurves,
-etc.).
-
-Also part of \astropysubpkg{timeseries} are common analysis routines,
-including Lomb-Scargle and box-least-squares periodogrammes.
-
-\paragraph{Spectral Coordinates.} Measurements are often taken at
-specific ``spectral coordinates'', be they frequencies, wavelengths,
-or photon energies.  The new \astropySpectralCoord\ class allows to
-transform between these as well as to equivalent Doppler velocities.
-It builds on the \astropyQuantity, which can also represent these
-kinds of physical quantities, and allows conversion via dedicated
-equivalencies, but aims to make this more straightforward.
-Furthermore, it can be made aware of the observer and target reference
-frames, allowing for instance transformation from telescope-centric (or
-topocentric) frames to Barycentric or Local Standard of Rest
-(LSRK and LSRD) velocity frames.
+software interface.  This maturity and stability allows for the broader
+astronomy \python community to rely upon the \astropypkg core package and build
+specialized packages within the Astropy ecosystem.  Even many non-astronomical
+\python libraries have come to rely on \astropypkg. The relationship between
+\astropy and the \python-using community is reciprocal, with the evolving needs
+of the community and the maturation of the broader scientific \python ecosystem
+driving much of the development of the \astropypkg core package.
+
+Within \astropypkg, development may be roughly split into a few
+categories: new features added, intra and inter-package interoperability,
+improvements to precision, accuracy, and reproducibility, and performance
+enhancements. We discuss these in turn.
+
+Additionally, some features have been moved from the core package, either
+because they are of more general use outside of \astropypkg, or because they are
+too specialized and belong in an Astropy-affiliated package. An example of the
+former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA) software
+Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
+carried, as well as the \python wrappers for it. Since this is basic
+infrastructure, with a release cycle set by SOFA, it made more sense to create
+separate packages that serve as \astropypkg dependencies: \package{ERFA}
+\citep{erfa} and \package{PyERFA} \citep{pyerfa}.  An example of something that
+was too specialized was a Virtual Observatory sub-package; the Simple
+Application Message Protocol made sense to keep (\astropysubpkg{samp}), but
+other parts more properly belonged in \package{astroquery} (see
+Sect.~\ref{sec:astroquery}).
+
+\subsubsection*{New and Planned Features} \label{sec:core-features-new}
+
+New features present a particular difficulty for a mature package such as
+\astropypkg, because it is difficult to know {\em a priori} what the best
+interface will be, and hard to address all possible use cases in one round of
+development, even within \astropypkg\ itself.
+
+New features come in various forms, from new submodules providing wholly new
+capabilities, to larger additions to existing submodules, to large rewrites of
+the interface of modules. This range is spanned by five main new features that
+entered the \astropypkg core since the previous summary
+\citepalias{astropy:2018}. In our descriptions of each we include how we hope
+the new feature will evolve. Ultimately, in a user-driven project like \astropy,
+development depends primarily on the priorities of contributors.
+
+\paragraph{Uncertainties and Distributions} Early in our careers we were likely
+all taught that measurements without units and uncertainties are meaningless.
+\astropyunits provides the ability to associate numbers with units, and
+propagate these correctly, using the \astropyQuantity\ class.  However, the
+ability to associate and propagate uncertainties is relatively limited: in
+\astropysubpkg{nddata}, there are options to associate errors with data arrays,
+but these are treated as independent, i.e., covariances are not tracked.
+
+In general, proper error propagation is difficult, and often it is best to use
+Monte Carlo methods.  The new \astropysubpkg{uncertainties} sub-package is our
+first step towards enabling seamless error propagation.  The package allows one
+to generate, for each variable, randomly drawn samples in a
+\astropyDistribution, and then propagate these by passing them through the
+normal analysis, producing a \astropyDistribution\ of final results that can be
+inspected.
+
+This \astropyDistribution\ can be any type of array, including a
+\astropyQuantity. The goal is to ensure \astropyDistribution\ can be seamlessly
+used to instantiate other \astropypkg classes as well, such as \astropySkyCoord,
+\astropyTime, and \astropyCosmology.
+
+A plan for future development of \astropysubpkg{uncertainties} is to support
+error propagation, tracking covariances, where the uncertainties are assumed to
+be normally distributed rather than approximated by Monte-Carlo means. One
+implementation problem to be solved is when to {\em stop} tracking covariances.
+For instance, a simple operation such as subtracting the mean from $N$ data
+points implies that all data points are now covariant with each other, i.e., one
+has to carry $N\times N$ covariances.  For a large image, that becomes not just
+pointless but also prohibitive in terms of memory and CPU usage.
+
+\paragraph{Masked quantities}  It can be useful to mask bad data.  How one
+supports masked data involves a number of choices, such as whether one indicates
+data values are bad by a separate flag or by replacing them with a special
+value, such as ``Not a Number'' (NaN).  In \astropypkg, both approaches have
+been used: a flag for masked columns in \astropyTable, mask flags and bitmaps
+for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
+NaN in \astropyTime.  However masked quantities were poorly supported, making it
+difficult to use masks in a \astropyQTable (in which \astropyQuantity\ is used
+for all columns with units).
+
+Unfortunately, the \texttt{MaskedArray} class from \package{numpy} could not be
+used because its design made it a wrapper around any subclass. Therefore
+another class could not use it as an alternative to a regular data array.
+% Nathaniel: I find this topic sentence a bit confusing.
+For quantities, \texttt{MaskedArray} does not work because a masked value does
+not mean the unit is unknown. Consequently, a new \astropyMasked\ class was
+designed, based on a framework very similar to that of \astropyDistribution,
+making it easy to create masked instances of other classes.  In particular,
+\astropyMasked\ can be used to create masked quantities for \astropyQTable,
+enabling full support for reading and writing different file types.
+
+The masked quantities work largely without any changes in higher-level objects
+such as \astropySkyCoord, but work is still underway to expose the mask in those
+classes, as well as to use masked arrays in \astropyTime\ instead of replacing
+data with NaN.
+
+\paragraph{Time series}  From sampling a continuous variable at fixed times to
+counting events binned into time windows, many different areas of astrophysics
+require the manipulation of time series.  The new \astropysubpkg{timeseries}
+sub-package extends the \astropyQTable\ class to tables of data as a function of
+time, where the data can either represent samples or averages over particular
+time bins. The new classes offer a number of special methods to manipulate time
+series (folding, resampling, etc.) and to read different data formats (such as
+{\em Kepler} lightcurves, etc.).
+
+Also part of \astropysubpkg{timeseries} are common analysis routines, including
+Lomb-Scargle and box-least-squares periodogrammes.
+
+\paragraph{Spectral Coordinates} Measurements are often taken at specific
+``spectral coordinates,'' be they frequencies, wavelengths, or photon energies.
+\astropyQuantity\ can represent these physical quantities and convert between
+them via dedicated equivalencies. The new \astropySpectralCoord\ builds on
+\astropyQuantity\ by providing a more straightforward interface: baking in these
+equivalencies and those for Doppler velocities. Furthermore,
+\astropySpectralCoord\ can be made aware of the observer and target reference
+frames, allowing transformation from telescope-centric (or topocentric) frames
+to Barycentric or Local Standard of Rest (LSRK and LSRD) velocity frames.
+
+\paragraph{A High-Level Interface to World Coordinate Systems}
+Astronomical data are often provided alongside information about the
+correspondence between ``real-world'' and pixel coordinates. This mapping is the
+essence of the ``World Coordinate System (WCS)'' concept. From its inception,
+the \astropysubpkg{wcs} sub-package allowed access to WCS information provided
+in, e.g., FITS files, but it became clear that other WCS standards and
+representations had to be supported for new missions and observatories (e.g.,
+the James Webb Space Telescope and the Rubin Observatory).  To harmonize these
+needs, a new high-level interface was created based on a formal design first
+proposed in an \astropy\ Proposal for Enhancement (\astropyAPE{14}). The hope is
+that by having a formal design, other packages implementing WCS objects can
+straightforwardly modify their classes to conform to the new interface or build
+thin wrappers that conform.  The implementation in \astropysubpkg{wcs} interacts
+well with other \astropypkg\ objects such as \astropySkyCoord\ and \astropyTime.
 
@@
 
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}
-
-\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
+Because \astropy is modular and situated at the nexus between scientific
+computing and astronomy, interoperability is an important focus for \astropy,
+both within the various \astropypkg sub-packages and with other \python
+libraries. If done right, all a person notices is that code using many different
+python features and libraries ``just works.'' Here, we describe a few particular
+efforts towards this goal.
+
+\paragraph{\package{numpy} on Units}
+
+\astropyQuantity\ is a backbone of \astropy, leveraging the power of
+\package{numpy} and adding units. Previously, many \package{numpy} functions
+would strip a \astropyQuantity\ of its units (or fail outright), limiting
+\astropyQuantity's usefulness. Now, advances in \package{numpy} function
+overloading (e.g.,
+\href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
+mean \astropyQuantity\ works with almost all \package{numpy} (v1.17+)
+functions.
+
+In the remaining gaps, the \package{numpy} and \astropyQuantity\
+interoperability efforts are ongoing. For some functions, such as in
+\href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
+module} for manipulating structured arrays, compatibility only requires
+extending the existing \package{numpy}-\astropypkg bridge frameworks. Community
+interest, in the form of a Feature Request (or better yet Pull Request) would be
+sufficient to see this compatibility completed. For a few remaining functions,
+discussed further in the
+\href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
+documentation}, the \package{numpy} framework does not yet allow for full
+interoperability with \astropy. A goal of \astropy and the scientific \python
+community is to enhance and implement the frameworks, allowing \astropyQuantity\
+to propagate units seamlessly across the whole ecosystem of \package{numpy}-like
+projects.
+
+\paragraph{\astropypkg on Units}
+
+Units were also integrated further within \astropypkg. For an already-defined unit-less
+\astropyModel, e.g. those imported from another library, \astropymodeling can
+now coerce units.
+% TODO! transition words
+A new module – \astropycosmologyunits – has been added to the cosmology
+sub-package for defining and collecting cosmological units and equivalencies.
+The unit \texttt{littleh} and equivalency
+\astropyapidoc{cosmology.units.with_H0}{with\_H0} are deprecated from the main
+\astropyunits sub-package and moved to \astropycosmologyunits. A new unit,
+\texttt{redshift}, has been added for tracking factors of cosmological redshift. To
+correctly use redshift units in dimensional analysis, \texttt{redshift} is
+\astropyapidoc{cosmology.units.dimensionless_redshift}{treated as dimensionless}
+by default. To convert between redshift and other cosmological distance
+measures, e.g. CMB temperature or comoving distance, the equivalency
+\astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
+has also been added. This equivalency is actually a composite of other equivalencies,
+which may be used separately.
+
+\paragraph{Table Mixin Columns}
+
+Within \astropypkg and \python there are numerous ways to represent and store
+array-valued data. Some of the differences are historical:
+\astropyapidoc{table.Column}{table.Column} and
+\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
+differences are computational: \package{numpy}, \package{cupy}, and
+\package{dask} arrays are almost API identical, but are optimized for different
+compute needs. Lastly, some differences are inherent: \astropyQuantity,
+\astropyTime\, and \astropySkyCoord\ represent fundamentally different types of
+objects.
+
+\astropy believes that however array-valued data is stored, it should still be
+usable as a column in a table. A well defined protocol for mixin-columns has
+been developed for \astropytable and \astropyfits tables, allowing the original
+object to be used as a column with a familiar API, and to ``round-trip'' through
+tables with no loss of data or attributes. With this protocol, it is now
+possible to store standard \astropyTableColumn\ objects to a FITS table
+\href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}. It is also
+possible to store \astropy native objects -- including \astropyTime,
+\astropyQuantity, and \astropySkyCoord --  within a \astropyTable. For objects
+not already covered by the mixin protocol, functions can be registered with
+\astropyTable\ to convert on-the-fly any array-like object into a mixin column.
+As an example, the mixin functions are used to integrate \package{dask} arrays
+with \astropyTable, allowing cloud-stored or cluster-scale data to be used as a
+column in a \astropyTable.
+
+\paragraph{Astropy FITS in Time}
+
+The FITS standard was extended to rigorously describe time coordinates in the
+World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared to
+other coordinates in the WCS framework, time requires more metadata: format,
+scale, position, reference, etc. This metadata had to be manually specified and
+the nuances understood by the user. \astropyfits could store this data in a
+standard \astropyFitsColumn\ (for details, see the
+\href{https://docs.astropy.org/en/stable/io/fits/index.html}{FITS
+documentation}), but would not ``understand'' the metadata and attributes.
+Thankfully, through the support of the
+\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
+Summer of Code} 2017 program, the \astropyfits package now allows \astropyTime\
+as a `mixin' column. \astropyTime\ columns in FITS tables ``understand'' the
+metadata while also providing a similar interface as the regular
+\astropyFitsColumn\ objects. With \astropyTime\ columns, \astropy ensures that
+time coordinates are stored according to the FITS standard in a consistent
+manner and can be read/written without loss of information. For backward
+compatibility with manual systems, this feature may be turned off, but it is
+recommended to update. To learn more about this project, please see the final
+report,
+\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
+mixin protocol for seamless interoperability}.
+
+\paragraph{Persistent Storage}
+
+\astropy's efforts to increase support for column types in a table is mirrored
+by efforts to expand storage format options. New formats have been added, and
+existing formats updated to support more column types.
+
+\astropy now supports reading and writing tables in the American Astronomical
+Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
+long been missing, and we are pleased that this year's Google Summer of Code
+project adds MRT to \astropyTable\ I/O.  
+For details, see the
+\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{Machine-Readable
+Table Format} section of \astropy's documentation. In addition, \astropyTable\ may also read from and write to
+\href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
+\href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
+and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.
+
+The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
+standard} has been updated to version 1.0, adding support for reading and
+writing \astropyTable\ with three additional data subtypes. First, it adds
+multidimensional column data (both masked and unmasked) with fixed dimensions in
+all table cells. Second, it adds multidimensional column data with
+variable-dimension arrays similar to FITS variable-length arrays. Third, it adds
+object-type columns with simple \python objects.
+
+As mentioned previously, \package{dask} arrays may be used as a mixin column in
+\astropyTable. Now, \package{dask} may also be the data array in
+\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the array
+will be computed while written, avoiding excessive memory use. \astropyTable\ can
+now be appended to an existing FITS file, and \package{dask} mixin columns
+interoperate seamlessly between the two table types.
+
+\paragraph{Unified I/O architecture}
+
+% TODO! better transition
+\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
+functions for \astropypkg objects, such as reading from or writing to a file.
+\astropy uses this internally for the I/O methods in \astropyTable\ and
+\astropyCosmology\ (I/O added v5.0+), and users can register custom I/O to extend
+the options on these classes.
+
+As of Astropy 5.0, the I/O registry submodule has switched to a class-based
+architecture and has been refactored to allow read-only, write-only, and
+read-write registries. Additionally, the registry submodule no longer assumes an
+object has \astropy-style \texttt{read} and \texttt{write} methods when trying
+to read or write that object. These two improvements enable a number of new use
+cases for \astropyiosubpkg{registry}, which we hope have community adoption.
+First, the class-based architecture allows for the creation of custom
+registries. One application is in \astropycosmology, which has two different
+\textit{kinds} of registries for \astropyCosmology: one for reading and writing
+files, and another for converting between python objects. Second, \textit{any}
+kind of object is now compatible with \astropyiosubpkg{registry}. Using
+\astropyiosubpkg{registry} for custom classes means a developer does not have to
+create a whole new I/O system and can offer to the user a unified and familiar
+API.
+
+\subsection*{Precision, Accuracy, \& Reproducibility}
+\label{sec:core-features-precision_accuracy_repreducibility}
+
+An important focus of \astropy has been to increase the precision and accuracy
+of its functionality, while, where appropriate, making sure results obtained
+using older versions of \astropypkg are reproducible. Within each section below
+we include in our descriptions how old results may be reproduced. The primary
+means are with configuration file options, and settings on runtime configuration
+objects called \astropyScienceState. Notable improvements to \astropypkg have
+been made to \astropytime, \astropyconstants, \astropycoordinates, and
+\astropycosmology, and we describe each in turn.
+
+\paragraph{Time}
+In astronomy, time accuracy down to the (nano)second is frequently important.
+Whether for planning observations, crunching pulsar or VLBI data, missing
+seconds meaningfully impact results. Before, when using \astropyTime\ in
+\astropytime leap seconds had to be manually applied; now, leap seconds are
+applied automatically. Moreover, \astropypkg updates internal time data-files to
+ensure that the correct leap second adjustment is always used. For
+reproducibility, e.g. with an old code, \astropyLeapSeconds\ may also be
+manually applied.
+
+\paragraph{Constants}
+
+Measurements of physical constants (and the units defined from them) improve
+over time. Periodically the standardized systems of units and constants are
+updated to reflect these improvements. For instance, in 2019 the SI system was
+redefined \citep{NIST2019}, with an accompanying update to the physical
+constants in
+\href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
+2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
+values, with the relation between units in \astropyunits based upon these
+constants. Most of \astropypkg and affiliate packages build upon \astropyunits,
+so this update has far-reaching effects across the \astropy ecosystem and
+astronomy community.
+
+For reproducibility, \astropypkg allows the constants' (and therefore units')
+definitions to be rolled back to prior values, e.g. to the
+\texttt{SI/CODATA 2014} values.
+% CITATION
+The constants' and units' definitions are set by configuration file as well as a
+\astropyScienceState\. For work sensitive to the values of the fundamental
+constants we recommend including with the work an \astropy configuration file,
+which specifies the set of constants used.
+
+\paragraph{Cosmology}
+% TODO! At least one good sentence about "what is the benefit of
+% astropy.cosmology"
+
+\astropycosmology contains classes for representing cosmological models. Bundled
+with the classes are commonly used \astropyCosmology\ realizations, e.g.,
+best-fit measurements from
+\href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
+\citep{WMAP2003} and
+\href{https://www.nasa.gov/mission_pages/planck}{Planck}
+\citep{PlanckMission:2006}.
+
+When these important missions publish new measurements, \astropycosmology is
+updated to include these results as realizations of the appropriate class
+(generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
+always exactly correspond to \astropycosmology classes. For instance, the Planck
+2018 results \citep{Planck2018VI:2020} include massive neutrinos in
+$\Omega_{matter,0}$, while in \astropyFlatLambdaCDM\ this mass contribution is
+stored in a separate parameter. Consequently, while some parameter values appear
+different from the source paper, the \astropyCosmology\ realizations correctly
+reproduce the cosmological models.
+
+Since \citealt{astropy:2018} the following cosmology realizations have been
+added:
+\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
+from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
+Lya]{WMAP1Year:2003};
+\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
+from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
+\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
+from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
+ext]{Planck2015XIII:2016};
+and \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
+from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE, EE
++ lowE + lensing + BAO]{Planck2018VI:2020}.
+
+Astropy provides a
+\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
+default cosmology}, which is used in calculations done in a cosmological
+context. The default \astropyCosmology\ has been updated to the Planck 2018
+parameters. For reproducibility, this cosmology may be set to old defaults, such
+as the WMAP Three Year values. The default cosmology is dynamically configurable
+using \astropyScienceState, meaning a set of calculations may be run using
+different assumed cosmologies, and results compared between the two.
+
+\paragraph{Coordinates}
+
+Being able to describe the position of objects, i.e. their coordinates, is
+fundamental across many knowledge domains. \astropypkg provides
+\astropycoordinates for working with low-level positional and velocity data all
+the way to high-level coordinate objects with reference frames, atmospheric
+information, and more. A central feature of \astropycoordinates is the ability
+to transform data between reference frames, e.g. \astropyICRS\ \citep{ICRS:1997}
+to \astropyGalacticLSR\ \citep{GalacticLSR:2010}. However, there have been some
+historical limitations for spatially proximate transformations, impacting the
+usefulness of \astropycoordinates for ground-based telescopes and intra-solar
+astronomy. Transformations in the \astropyAltAz\ frame were reasonably precise
+for very distant objects, but wrong by up to several arcseconds for e.g., the
+location of the moon. Now these transformations are much more precise, down to
+the milliarcsecond level. Similar precision improvements were made to the
+\astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination}
+frame transformations. Additionally,
+\astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
+and associated transformations have been updated to correctly reflect the “true”
+(precession but not nutation) and “mean” (both precession and nutation)
+terminology.
+
+For Galactic astronomers, the \astropyGalactocentric\
+\astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults}
+have been updated to include more recent measurements (tabulated in the
+\href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
+attribute). For reproducibility, old definitions are still available, and will
+continue to be so as \astropypkg is updated with forthcoming measurements of the
+Galactic center, solar motion, etc.
+
+\subsection*{Performance} \label{sec:core-features-performance}
+
+As features are added and corner cases dealt with, performance often suffers.
+Hence, pull requests that improve performance are welcomed, and were even the
+main goal for one release cycle.  For most sub-packages, performance was
+improved in the \python\ code, but for some, the most time-critical pieces were
+rewritten in~C (e.g., for convolution of images, sigma-clipping, and converting
+time strings to binary).  Furthermore, a particular effort was made to ensure
+\astropypkg\ is thread-safe, so that it can be used on supercomputer clusters.
 

### 2022-03-03 src/ms.tex
editing session
Co-authored-by: Marten van Kerkwijk <mhvk@astro.utoronto.ca>
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
 The \astropypkg core package is mature and stable, in that many features have
-been part of the core package for years with (purposefully) few changes to the
+been part of the core package for years with purposefully few changes to the
 software interface.  This maturity and stability allows for the broader
@@
 interface will be, and hard to address all possible use cases in one round of
-development, even within \astropypkg\ itself.
-
-New features come in various forms, from new submodules providing wholly new
-capabilities, to larger additions to existing submodules, to large rewrites of
-the interface of modules. This range is spanned by five main new features that
-entered the \astropypkg core since the previous summary
-\citepalias{astropy:2018}. In our descriptions of each we include how we hope
-the new feature will evolve. Ultimately, in a user-driven project like \astropy,
-development depends primarily on the priorities of contributors.
+development, even within \astropypkg\ itself. New features come in various forms,
+from new submodules providing wholly new capabilities, to larger additions to
+existing submodules, to large rewrites of the interface of modules. This range
+is spanned by five main new features that entered the \astropypkg core since the
+previous summary \citepalias{astropy:2018}. In our descriptions of each we
+include how we hope the new feature will evolve. Ultimately, in a user-driven
+project like \astropy, development depends primarily on the priorities of
+contributors.
 
@@
 
-In general, proper error propagation is difficult, and often it is best to use
-Monte Carlo methods.  The new \astropysubpkg{uncertainties} sub-package is our
-first step towards enabling seamless error propagation.  The package allows one
-to generate, for each variable, randomly drawn samples in a
-\astropyDistribution, and then propagate these by passing them through the
-normal analysis, producing a \astropyDistribution\ of final results that can be
-inspected.
-
-This \astropyDistribution\ can be any type of array, including a
-\astropyQuantity. The goal is to ensure \astropyDistribution\ can be seamlessly
-used to instantiate other \astropypkg classes as well, such as \astropySkyCoord,
+Error propagation is difficult, and often is best approximated using Monte Carlo
+methods.  The new \astropysubpkg{uncertainties} sub-package is our first step
+towards enabling seamless error propagation.  The package allows one to
+generate, for each variable, randomly drawn samples in a \astropyDistribution,
+and then propagate these by passing them through the normal analysis, producing
+a \astropyDistribution\ of final results that can be inspected. This
+\astropyDistribution\ can be any type of array, including a \astropyQuantity.
+The goal is to ensure \astropyDistribution\ can be seamlessly used to
+instantiate other \astropypkg classes as well, such as \astropySkyCoord,
 \astropyTime, and \astropyCosmology.
 
-A plan for future development of \astropysubpkg{uncertainties} is to support
-error propagation, tracking covariances, where the uncertainties are assumed to
-be normally distributed rather than approximated by Monte-Carlo means. One
+A plan for future development of \astropysubpkg{uncertainties} is to also
+support error propagation, tracking covariances, for the case that uncertainties
+are normally distributed rather than approximated by Monte-Carlo means. One
 implementation problem to be solved is when to {\em stop} tracking covariances.
@@
 supports masked data involves a number of choices, such as whether one indicates
-data values are bad by a separate flag or by replacing them with a special
+data values are bad by a separate flag or by replacing bad data with a special
 value, such as ``Not a Number'' (NaN).  In \astropypkg, both approaches have
@@
 NaN in \astropyTime.  However masked quantities were poorly supported, making it
-difficult to use masks in a \astropyQTable (in which \astropyQuantity\ is used
+difficult to use masks in a \astropyQTable\ (in which \astropyQuantity\ is used
 for all columns with units).
 
-Unfortunately, the \texttt{MaskedArray} class from \package{numpy} could not be
-used because its design made it a wrapper around any subclass. Therefore
-another class could not use it as an alternative to a regular data array.
-% Nathaniel: I find this topic sentence a bit confusing.
-For quantities, \texttt{MaskedArray} does not work because a masked value does
-not mean the unit is unknown. Consequently, a new \astropyMasked\ class was
-designed, based on a framework very similar to that of \astropyDistribution,
-making it easy to create masked instances of other classes.  In particular,
-\astropyMasked\ can be used to create masked quantities for \astropyQTable,
-enabling full support for reading and writing different file types.
+Unfortunately, the \texttt{MaskedArray} class from \package{numpy} only works
+well with plain data arrays, hiding other metadata and attributes, for example
+the \texttt{unit} of a \astropyQuantity. Consequently, a new \astropyMasked\
+class was designed, based on a framework very similar to that of
+\astropyDistribution, making it easy to create masked instances of other
+classes.  In particular, \astropyMasked\ can be used to create masked quantities
+for \astropyQTable, enabling I/O of masked data from different file types.
 
@@
 Also part of \astropysubpkg{timeseries} are common analysis routines, including
-Lomb-Scargle and box-least-squares periodogrammes.
+\href{https://docs.astropy.org/en/stable/timeseries/lombscargle.html}{Lomb-Scargle}
+and
+\href{https://docs.astropy.org/en/stable/timeseries/bls.html}{box-least-squares}
+periodogrammes.
 
@@
 ``spectral coordinates,'' be they frequencies, wavelengths, or photon energies.
-\astropyQuantity\ can represent these physical quantities and convert between
-them via dedicated equivalencies. The new \astropySpectralCoord\ builds on
-\astropyQuantity\ by providing a more straightforward interface: baking in these
-equivalencies and those for Doppler velocities. Furthermore,
-\astropySpectralCoord\ can be made aware of the observer and target reference
-frames, allowing transformation from telescope-centric (or topocentric) frames
-to Barycentric or Local Standard of Rest (LSRK and LSRD) velocity frames.
+\astropyQuantity\ can represent these and convert between them via dedicated
+equivalencies. The new \astropySpectralCoord\ builds on \astropyQuantity\ by
+providing a more straightforward interface: baking in these equivalencies and
+those for Doppler velocities. Furthermore, \astropySpectralCoord\ can be made
+aware of the observer and target reference frames, allowing transformation from
+telescope-centric (or topocentric) frames to Barycentric or Local Standard of
+Rest (LSRK and LSRD) velocity frames.
 
@@
 The unit \texttt{littleh} and equivalency
-\astropyapidoc{cosmology.units.with_H0}{with\_H0} are deprecated from the main
-\astropyunits sub-package and moved to \astropycosmologyunits. A new unit,
-\texttt{redshift}, has been added for tracking factors of cosmological redshift. To
-correctly use redshift units in dimensional analysis, \texttt{redshift} is
-\astropyapidoc{cosmology.units.dimensionless_redshift}{treated as dimensionless}
+\astropyapidoc{cosmology.units.with_H0}{with\_H0} was moved from the main
+\astropyunits sub-package to \astropycosmologyunits. A new unit,
+\texttt{redshift}, has been added for tracking factors of cosmological redshift.
+To correctly use redshift units in dimensional analysis, \texttt{redshift} is
+\astropyapi{cosmology.units.dimensionless_redshift}{treated as dimensionless}
 by default. To convert between redshift and other cosmological distance
@@
 differences are computational: \package{numpy}, \package{cupy}, and
-\package{dask} arrays are almost API identical, but are optimized for different
-compute needs. Lastly, some differences are inherent: \astropyQuantity,
+\package{dask} arrays have almost identical APIs, but are optimized for
+different use cases. Lastly, some differences are inherent: \astropyQuantity,
 \astropyTime\, and \astropySkyCoord\ represent fundamentally different types of
@@
 
-\astropy believes that however array-valued data is stored, it should still be
-usable as a column in a table. A well defined protocol for mixin-columns has
-been developed for \astropytable and \astropyfits tables, allowing the original
-object to be used as a column with a familiar API, and to ``round-trip'' through
-tables with no loss of data or attributes. With this protocol, it is now
-possible to store standard \astropyTableColumn\ objects to a FITS table
-\href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}. It is also
-possible to store \astropy native objects -- including \astropyTime,
-\astropyQuantity, and \astropySkyCoord --  within a \astropyTable. For objects
-not already covered by the mixin protocol, functions can be registered with
-\astropyTable\ to convert on-the-fly any array-like object into a mixin column.
-As an example, the mixin functions are used to integrate \package{dask} arrays
-with \astropyTable, allowing cloud-stored or cluster-scale data to be used as a
-column in a \astropyTable.
+We aim to make it possible for any array-valued data to be used as a column in a
+table. A well defined protocol for mixin-columns has been developed for
+\astropytable, allowing the original object to be used as a column with a
+familiar API, and to ``round-trip'' through tables with no loss of data or
+attributes. With this protocol, it is now possible to store \astropy native
+objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord --
+within a \astropyTable and write these to various file formats, such as FITs.
+For objects not already covered by the mixin protocol, functions can be
+registered with \astropyTable\ to convert any array-like object into a mixin
+column. As an example, the mixin functions are used to integrate \package{dask}
+arrays with \astropyTable, allowing cloud-stored or cluster-scale data to be
+used as a column in a \astropyTable.
 
@@
 World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared to
-other coordinates in the WCS framework, time requires more metadata: format,
-scale, position, reference, etc. This metadata had to be manually specified and
-the nuances understood by the user. \astropyfits could store this data in a
-standard \astropyFitsColumn\ (for details, see the
-\href{https://docs.astropy.org/en/stable/io/fits/index.html}{FITS
-documentation}), but would not ``understand'' the metadata and attributes.
-Thankfully, through the support of the
+other types of coordinates in WCS, time requires more metadata: format, scale,
+position, reference, etc. This metadata had to be manually specified and the
+nuances understood by the user. \astropyfits could read this data as a standard
+\astropyFitsColumn\, but would not interpret the time-related metadata and
+attributes. Through the support of the
 \href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
-Summer of Code} 2017 program, the \astropyfits package now allows \astropyTime\
-as a `mixin' column. \astropyTime\ columns in FITS tables ``understand'' the
-metadata while also providing a similar interface as the regular
-\astropyFitsColumn\ objects. With \astropyTime\ columns, \astropy ensures that
-time coordinates are stored according to the FITS standard in a consistent
-manner and can be read/written without loss of information. For backward
-compatibility with manual systems, this feature may be turned off, but it is
-recommended to update. To learn more about this project, please see the final
-report,
+Summer of Code} 2017 program\footnote{ To learn more about this project, please
+see the final report,
 \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
-mixin protocol for seamless interoperability}.
+mixin protocol for seamless interoperability}}, the \astropyfits package now
+interprets time data correctly, using \astropyTime\ as a `mixin' column.  For
+backward compatibility with manual systems, this feature may be turned off.
 
@@
 Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
-long been missing, and we are pleased that this year's Google Summer of Code
-project adds MRT to \astropyTable\ I/O.  
-For details, see the
-\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{Machine-Readable
-Table Format} section of \astropy's documentation. In addition, \astropyTable\ may also read from and write to
-\href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
+long been missing and as added in this year's Google Summer of Code project adds
+\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}
+to \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
+to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
 \href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
@@
 The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
-standard} has been updated to version 1.0, adding support for reading and
-writing \astropyTable\ with three additional data subtypes. First, it adds
-multidimensional column data (both masked and unmasked) with fixed dimensions in
-all table cells. Second, it adds multidimensional column data with
-variable-dimension arrays similar to FITS variable-length arrays. Third, it adds
-object-type columns with simple \python objects.
+standard} has been updated to version 1.0, adding support three additional data
+subtypes. First, it adds multidimensional column data (both masked and unmasked)
+with fixed dimensions in all table cells. Second, it adds multidimensional
+column data with variable-dimension arrays similar to FITS variable-length
+arrays. Third, it adds object-type columns with simple \python objects.
 
@@
 \astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
-functions for \astropypkg objects, such as reading from or writing to a file.
-\astropy uses this internally for the I/O methods in \astropyTable\ and
-\astropyCosmology\ (I/O added v5.0+), and users can register custom I/O to extend
-the options on these classes.
-
-As of Astropy 5.0, the I/O registry submodule has switched to a class-based
-architecture and has been refactored to allow read-only, write-only, and
-read-write registries. Additionally, the registry submodule no longer assumes an
-object has \astropy-style \texttt{read} and \texttt{write} methods when trying
-to read or write that object. These two improvements enable a number of new use
-cases for \astropyiosubpkg{registry}, which we hope have community adoption.
-First, the class-based architecture allows for the creation of custom
-registries. One application is in \astropycosmology, which has two different
-\textit{kinds} of registries for \astropyCosmology: one for reading and writing
-files, and another for converting between python objects. Second, \textit{any}
-kind of object is now compatible with \astropyiosubpkg{registry}. Using
-\astropyiosubpkg{registry} for custom classes means a developer does not have to
-create a whole new I/O system and can offer to the user a unified and familiar
-API.
+functions for \astropypkg objects, such as reading from or writing to a file,
+following a given standard. \astropy uses this internally for the I/O methods in
+\astropyTable\ and \astropyCosmology, and users can register custom I/O to
+extend the options on these classes.
+
+As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
+number of new use cases. for instance, the class-based architecture allows for
+the creation of custom registries. One application is in \astropycosmology,
+which has two different \textit{kinds} of registries for \astropyCosmology: one
+for reading and writing files, and another for converting between python
+objects. Using \astropyiosubpkg{registry} for custom classes means a developer
+does not have to create a whole new I/O system and can offer to the user a
+unified and familiar API.
 
@@
 we include in our descriptions how old results may be reproduced. The primary
-means are with configuration file options, and settings on runtime configuration
+means are with options in configuration files and settings on runtime configuration
 objects called \astropyScienceState. Notable improvements to \astropypkg have
@@
 2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
-values, with the relation between units in \astropyunits based upon these
-constants. Most of \astropypkg and affiliate packages build upon \astropyunits,
-so this update has far-reaching effects across the \astropy ecosystem and
-astronomy community.
+values, with the units in \astropyunits based on these constants. Most of
+\astropypkg and affiliate packages build upon \astropyunits, so this update
+effects the entire \astropy ecosystem.
 
 For reproducibility, \astropypkg allows the constants' (and therefore units')
-definitions to be rolled back to prior values, e.g. to the
-\texttt{SI/CODATA 2014} values.
-% CITATION
-The constants' and units' definitions are set by configuration file as well as a
-\astropyScienceState\. For work sensitive to the values of the fundamental
-constants we recommend including with the work an \astropy configuration file,
-which specifies the set of constants used.
+definitions to be rolled back to prior values, e.g. to the \texttt{SI/CODATA
+2014} values. For work sensitive to the values of the fundamental constants we
+recommend including with the work an \astropy configuration file, which
+specifies the set of constants used.
 
@@
 added:
-\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
-from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
-Lya]{WMAP1Year:2003};
-\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
-from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
-\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
-from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
-ext]{Planck2015XIII:2016};
-and \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
-from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE, EE
-+ lowE + lensing + BAO]{Planck2018VI:2020}.
+\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{WMAP First Year}
+\citep{WMAP1Year:2003};
+\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{WMAP
+Three Year} \citep{WMAP3Year:2007};
+\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{Planck 2015}
+\citep{Planck2015XIII:2016}; and the
+\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{Planck
+2018} best-fit cosmological parameters \citep{Planck2018VI:2020}.
 
@@
 
-Being able to describe the position of objects, i.e. their coordinates, is
-fundamental across many knowledge domains. \astropypkg provides
-\astropycoordinates for working with low-level positional and velocity data all
-the way to high-level coordinate objects with reference frames, atmospheric
-information, and more. A central feature of \astropycoordinates is the ability
-to transform data between reference frames, e.g. \astropyICRS\ \citep{ICRS:1997}
-to \astropyGalacticLSR\ \citep{GalacticLSR:2010}. However, there have been some
-historical limitations for spatially proximate transformations, impacting the
-usefulness of \astropycoordinates for ground-based telescopes and intra-solar
-astronomy. Transformations in the \astropyAltAz\ frame were reasonably precise
-for very distant objects, but wrong by up to several arcseconds for e.g., the
-location of the moon. Now these transformations are much more precise, down to
-the milliarcsecond level. Similar precision improvements were made to the
+Being able to describe the position of objects, i.e., their coordinates, is
+fundamental across many knowledge domains. \astropycoordinates allows one to
+work with low-level positional and velocity data all the way to high-level
+coordinate objects with reference frames, atmospheric information, and more. A
+central feature of \astropycoordinates is the ability to transform data between
+reference frames, e.g. \astropyICRS\ \citep{ICRS:1997} to \astropyGalacticLSR\
+\citep{GalacticLSR:2010}. However, there have been some historical limitations
+for spatially proximate transformations, impacting the usefulness of
+\astropycoordinates for ground-based telescopes and astronometry within the
+solar system. Transformations in the \astropyAltAz\ frame were reasonably
+precise for very distant objects, but wrong by up to several arcseconds for
+e.g., the location of the moon. Now these transformations are much more precise,
+down to the milliarcsecond level. Similar precision improvements were made to
+the
 \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination}
@@
 and associated transformations have been updated to correctly reflect the “true”
-(precession but not nutation) and “mean” (both precession and nutation)
-terminology.
+and “mean” terminology.
 
@@
 \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
-attribute). For reproducibility, old definitions are still available, and will
-continue to be so as \astropypkg is updated with forthcoming measurements of the
-Galactic center, solar motion, etc.
+attribute). For reproducibility, old definitions are still available.
 

### 2022-03-08 src/ms.tex
Apply suggestions from code review

--- 2022-03-01 src/ms.tex
@@
 
-% Summarize funding sources (Moore, NASA) and amounts and what this has been
-% used for.
-
-With the goal of establishing a sustainable financial model for the Astropy
-Project, the Coordination Committee under the new governance charter (see \S
-\ref{sec:project-governance}) established a standing Finance committee.
-This committee oversee the planning and allocation of finances on behalf
-of the project. The Astropy Project accepts funds from institutions as well as
-individuals through NUMFOCUS\footnote{NUMFOCUS is a 501(c)(3) nonprofit that
-supports and promotes world-class, innovative, open source scientific computing.}.
-Previously, no direct financial support was available for the project development,
-and NUMFOCUS covered most of the incurred operational costs. However, transitioning
-to long-term financial stability meant having an influx of funds in the form of
-institutional support, and hence the Astropy Project applied for and successfully
-acquired four grants. These grants are administered by NumFOCUS on behalf of the
-project, providing a ``neutral'' space not tied to any particular astronomy
-institution. The two major ones are briefly summarized below.
-
-\begin{itemize}
-    \item Moore: The Gordon and Betty Moore Foundation awarded the Astropy
-    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
-    Growing the Astropy Project". Over a three-year term, this grant supported
-    sustenance of the project by providing funds for (1) the development and
-    maintenance of the \astropypkg and its infrastructure to existing project
-    members and targeted hires, (2) the transition of long-term users to
-    contributors and maintainers through mentorship efforts, (3) the formalization
-    of a governance structure (see \S \ref{sec:project-governance}), and (4) the
-    improvement of equity, diversity, and inclusion efforts in the project (see \S
-    \ref{sec:project-ide}). Some of the money supported travel for meetings,
-    conferences, and workshops. This generous grant helped reduce reliance on
-    volunteer-driven maintenance of the openly-developed package and paved
-    the way to acquire funding from federal agencies: a big milestone for the
-    success of the project.
-
-    \item NASA: A successful proposal to the NASA ROSES-2020
-    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
-    (Support for Open Source Tools, Frameworks, and Libraries), granted the
-    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
-    term of three years includes (1) project-wide infrastructure maintenance and
-    improvements, (2) targeted work on areas of the \astropy core and
-    coordinated packages that need support, (3) enhancements to the Learn
-    Astropy ecosystem and educational materials (see \S \ref{sec:learn}),
-    and (4) support for the \astropy affiliated packages. These goals
-    support the sustenance of the project by encouraging further
-    development of the package ecosystem and community engagement.
-\end{itemize}
-
-In addition to the two grants described above, the project benefited from
-a Gemini Observatory contract under its Science User Support Department,
-which awarded funds for development work that supports both the Astropy and
-DRAGONS projects. The project also received financial support from the Dunlap
-Seed Funding program at the University of Toronto to develop educational
-resources for scientific software packaging within the Learn Astropy
-framework. All these successful proposals show that the Astropy project
-is vital for the astronomical infrastructure and community at large, and
-strengthen its promising and sustainable future.
-
-% ADD text for the new finance APE once it has a doi
+Summarize funding sources (Moore, NASA) and amounts and what this has been used
+for.
 
@@
 
-Overview of roadmap and some highlights.
+The Astropy Roadmap\footnote{\url{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}} is a document in the astropy-project repository that captures high level actionable items that the Astropy project aims to undertake to improve the health and stability of the  project.
+It is a static document that is revisited regularly at the Astropy coordination meetings, to keep track of progress and write new versions as needed.
+There is a related project board linked to the Roadmap document, that holds specific issues and efforts related to Roadmap items.
+The project board is a living document that is continually updated as work is planned, assigned, and completed.
+
+All items in the Roadmap have been agreed to be priorities for the Astropy Project, and are color-coded based on resources (both time/effort and developers/expertise) needed to complete the item.
+Green items are well underway, have sufficient resources/support and a plan in place for completion.
+Orange items are well defined, and work for aquiring sufficient resources underway.
+Red items do not yet have a plan for implementation and need more resources.
+\todo{addition}{Add some examples of high priority items from the roadmap in a new paragraph here}
+The Astropy Roadmap originates from the March 2021 Astropy Coordination meeting where it was first drafted before being handed off to a newly formed Astropy Roadmap working group for completion.
+The Roadmap in its currant form was adopted via pull request in December 2021.
 

### 2022-03-08 src/ms.tex
Add text for Astropy Roadmap section (#22)
* astropy roadmap

* Apply suggestions from code review

Co-authored-by: Adrian Price-Whelan <adrianmpw@gmail.com>

--- 2022-03-03 src/ms.tex
@@
 \newcommand{\astropyconstants}{\astropysubpkg{constants}}
-\newcommand{\astropytable}{\astropysubpkg{table}}
 \newcommand{\astropytime}{\astropysubpkg{time}}
 \newcommand{\astropymodeling}{\astropysubpkg{modeling}}
-\newcommand{\astropywcs}{\astropysubpkg{wcs}}
-\newcommand{\astropyfits}{\astropyiosubpkg{fits}}
 
@@
 \newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
-\newcommand{\astropyFitsColumn}{\astropyapidoc{io.fits.Column}{Column}}
-\newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
-\newcommand{\astropyTableColumn}{\astropyapidoc{table.Column}{Column}}
 \newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
@@
 \newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
-\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
 \newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
@@
 
-% Maybe have command link to Zenodo reference instead, or include it?
-\newcommand{\astropyAPE}[1]{\href{https://github.com/astropy/astropy-APEs/blob/main/APE#1.rst}{APE #1}\xspace}
-
 \newcommand{\secauthor}[1]{{\color{blue}Author:~\textit{#1}}}
 \newcommand{\secunfilled}{{\color{red}Author:~\textit{Looking for volunteers!}}}
-
-\defcitealias{astropy:2013}{Paper I}
-\defcitealias{astropy:2018}{Paper II}
 
@@
 
-The \astropypkg core package is mature and stable, in that many features have
-been part of the core package for years with purposefully few changes to the
-software interface.  This maturity and stability allows for the broader
-astronomy \python community to rely upon the \astropypkg core package and build
-specialized packages within the Astropy ecosystem.  Even many non-astronomical
-\python libraries have come to rely on \astropypkg. The relationship between
-\astropy and the \python-using community is reciprocal, with the evolving needs
-of the community and the maturation of the broader scientific \python ecosystem
-driving much of the development of the \astropypkg core package.
-
-Within \astropypkg, development may be roughly split into a few
-categories: new features added, intra and inter-package interoperability,
-improvements to precision, accuracy, and reproducibility, and performance
-enhancements. We discuss these in turn.
-
-Additionally, some features have been moved from the core package, either
-because they are of more general use outside of \astropypkg, or because they are
-too specialized and belong in an Astropy-affiliated package. An example of the
-former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA) software
-Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
-carried, as well as the \python wrappers for it. Since this is basic
-infrastructure, with a release cycle set by SOFA, it made more sense to create
-separate packages that serve as \astropypkg dependencies: \package{ERFA}
-\citep{erfa} and \package{PyERFA} \citep{pyerfa}.  An example of something that
-was too specialized was a Virtual Observatory sub-package; the Simple
-Application Message Protocol made sense to keep (\astropysubpkg{samp}), but
-other parts more properly belonged in \package{astroquery} (see
-Sect.~\ref{sec:astroquery}).
-
-\subsubsection*{New and Planned Features} \label{sec:core-features-new}
-
-New features present a particular difficulty for a mature package such as
-\astropypkg, because it is difficult to know {\em a priori} what the best
-interface will be, and hard to address all possible use cases in one round of
-development, even within \astropypkg\ itself. New features come in various forms,
-from new submodules providing wholly new capabilities, to larger additions to
-existing submodules, to large rewrites of the interface of modules. This range
-is spanned by five main new features that entered the \astropypkg core since the
-previous summary \citepalias{astropy:2018}. In our descriptions of each we
-include how we hope the new feature will evolve. Ultimately, in a user-driven
-project like \astropy, development depends primarily on the priorities of
-contributors.
-
-\paragraph{Uncertainties and Distributions} Early in our careers we were likely
-all taught that measurements without units and uncertainties are meaningless.
-\astropyunits provides the ability to associate numbers with units, and
-propagate these correctly, using the \astropyQuantity\ class.  However, the
-ability to associate and propagate uncertainties is relatively limited: in
-\astropysubpkg{nddata}, there are options to associate errors with data arrays,
-but these are treated as independent, i.e., covariances are not tracked.
-
-Error propagation is difficult, and often is best approximated using Monte Carlo
-methods.  The new \astropysubpkg{uncertainties} sub-package is our first step
-towards enabling seamless error propagation.  The package allows one to
-generate, for each variable, randomly drawn samples in a \astropyDistribution,
-and then propagate these by passing them through the normal analysis, producing
-a \astropyDistribution\ of final results that can be inspected. This
-\astropyDistribution\ can be any type of array, including a \astropyQuantity.
-The goal is to ensure \astropyDistribution\ can be seamlessly used to
-instantiate other \astropypkg classes as well, such as \astropySkyCoord,
-\astropyTime, and \astropyCosmology.
-
-A plan for future development of \astropysubpkg{uncertainties} is to also
-support error propagation, tracking covariances, for the case that uncertainties
-are normally distributed rather than approximated by Monte-Carlo means. One
-implementation problem to be solved is when to {\em stop} tracking covariances.
-For instance, a simple operation such as subtracting the mean from $N$ data
-points implies that all data points are now covariant with each other, i.e., one
-has to carry $N\times N$ covariances.  For a large image, that becomes not just
-pointless but also prohibitive in terms of memory and CPU usage.
-
-\paragraph{Masked quantities}  It can be useful to mask bad data.  How one
-supports masked data involves a number of choices, such as whether one indicates
-data values are bad by a separate flag or by replacing bad data with a special
-value, such as ``Not a Number'' (NaN).  In \astropypkg, both approaches have
-been used: a flag for masked columns in \astropyTable, mask flags and bitmaps
-for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
-NaN in \astropyTime.  However masked quantities were poorly supported, making it
-difficult to use masks in a \astropyQTable\ (in which \astropyQuantity\ is used
-for all columns with units).
-
-Unfortunately, the \texttt{MaskedArray} class from \package{numpy} only works
-well with plain data arrays, hiding other metadata and attributes, for example
-the \texttt{unit} of a \astropyQuantity. Consequently, a new \astropyMasked\
-class was designed, based on a framework very similar to that of
-\astropyDistribution, making it easy to create masked instances of other
-classes.  In particular, \astropyMasked\ can be used to create masked quantities
-for \astropyQTable, enabling I/O of masked data from different file types.
-
-The masked quantities work largely without any changes in higher-level objects
-such as \astropySkyCoord, but work is still underway to expose the mask in those
-classes, as well as to use masked arrays in \astropyTime\ instead of replacing
-data with NaN.
-
-\paragraph{Time series}  From sampling a continuous variable at fixed times to
-counting events binned into time windows, many different areas of astrophysics
-require the manipulation of time series.  The new \astropysubpkg{timeseries}
-sub-package extends the \astropyQTable\ class to tables of data as a function of
-time, where the data can either represent samples or averages over particular
-time bins. The new classes offer a number of special methods to manipulate time
-series (folding, resampling, etc.) and to read different data formats (such as
-{\em Kepler} lightcurves, etc.).
-
-Also part of \astropysubpkg{timeseries} are common analysis routines, including
-\href{https://docs.astropy.org/en/stable/timeseries/lombscargle.html}{Lomb-Scargle}
-and
-\href{https://docs.astropy.org/en/stable/timeseries/bls.html}{box-least-squares}
-periodogrammes.
-
-\paragraph{Spectral Coordinates} Measurements are often taken at specific
-``spectral coordinates,'' be they frequencies, wavelengths, or photon energies.
-\astropyQuantity\ can represent these and convert between them via dedicated
-equivalencies. The new \astropySpectralCoord\ builds on \astropyQuantity\ by
-providing a more straightforward interface: baking in these equivalencies and
-those for Doppler velocities. Furthermore, \astropySpectralCoord\ can be made
-aware of the observer and target reference frames, allowing transformation from
-telescope-centric (or topocentric) frames to Barycentric or Local Standard of
-Rest (LSRK and LSRD) velocity frames.
-
-\paragraph{A High-Level Interface to World Coordinate Systems}
-Astronomical data are often provided alongside information about the
-correspondence between ``real-world'' and pixel coordinates. This mapping is the
-essence of the ``World Coordinate System (WCS)'' concept. From its inception,
-the \astropysubpkg{wcs} sub-package allowed access to WCS information provided
-in, e.g., FITS files, but it became clear that other WCS standards and
-representations had to be supported for new missions and observatories (e.g.,
-the James Webb Space Telescope and the Rubin Observatory).  To harmonize these
-needs, a new high-level interface was created based on a formal design first
-proposed in an \astropy\ Proposal for Enhancement (\astropyAPE{14}). The hope is
-that by having a formal design, other packages implementing WCS objects can
-straightforwardly modify their classes to conform to the new interface or build
-thin wrappers that conform.  The implementation in \astropysubpkg{wcs} interacts
-well with other \astropypkg\ objects such as \astropySkyCoord\ and \astropyTime.
-
-\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
-
-Because \astropy is modular and situated at the nexus between scientific
-computing and astronomy, interoperability is an important focus for \astropy,
-both within the various \astropypkg sub-packages and with other \python
-libraries. If done right, all a person notices is that code using many different
-python features and libraries ``just works.'' Here, we describe a few particular
-efforts towards this goal.
-
-\paragraph{\package{numpy} on Units}
-
-\astropyQuantity\ is a backbone of \astropy, leveraging the power of
-\package{numpy} and adding units. Previously, many \package{numpy} functions
-would strip a \astropyQuantity\ of its units (or fail outright), limiting
-\astropyQuantity's usefulness. Now, advances in \package{numpy} function
-overloading (e.g.,
-\href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
-mean \astropyQuantity\ works with almost all \package{numpy} (v1.17+)
-functions.
-
-In the remaining gaps, the \package{numpy} and \astropyQuantity\
-interoperability efforts are ongoing. For some functions, such as in
-\href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
-module} for manipulating structured arrays, compatibility only requires
-extending the existing \package{numpy}-\astropypkg bridge frameworks. Community
-interest, in the form of a Feature Request (or better yet Pull Request) would be
-sufficient to see this compatibility completed. For a few remaining functions,
-discussed further in the
-\href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
-documentation}, the \package{numpy} framework does not yet allow for full
-interoperability with \astropy. A goal of \astropy and the scientific \python
-community is to enhance and implement the frameworks, allowing \astropyQuantity\
-to propagate units seamlessly across the whole ecosystem of \package{numpy}-like
-projects.
-
-\paragraph{\astropypkg on Units}
-
-Units were also integrated further within \astropypkg. For an already-defined unit-less
-\astropyModel, e.g. those imported from another library, \astropymodeling can
-now coerce units.
-% TODO! transition words
-A new module – \astropycosmologyunits – has been added to the cosmology
-sub-package for defining and collecting cosmological units and equivalencies.
-The unit \texttt{littleh} and equivalency
-\astropyapidoc{cosmology.units.with_H0}{with\_H0} was moved from the main
-\astropyunits sub-package to \astropycosmologyunits. A new unit,
-\texttt{redshift}, has been added for tracking factors of cosmological redshift.
-To correctly use redshift units in dimensional analysis, \texttt{redshift} is
-\astropyapi{cosmology.units.dimensionless_redshift}{treated as dimensionless}
-by default. To convert between redshift and other cosmological distance
-measures, e.g. CMB temperature or comoving distance, the equivalency
-\astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
-has also been added. This equivalency is actually a composite of other equivalencies,
-which may be used separately.
-
-\paragraph{Table Mixin Columns}
-
-Within \astropypkg and \python there are numerous ways to represent and store
-array-valued data. Some of the differences are historical:
-\astropyapidoc{table.Column}{table.Column} and
-\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
-differences are computational: \package{numpy}, \package{cupy}, and
-\package{dask} arrays have almost identical APIs, but are optimized for
-different use cases. Lastly, some differences are inherent: \astropyQuantity,
-\astropyTime\, and \astropySkyCoord\ represent fundamentally different types of
-objects.
-
-We aim to make it possible for any array-valued data to be used as a column in a
-table. A well defined protocol for mixin-columns has been developed for
-\astropytable, allowing the original object to be used as a column with a
-familiar API, and to ``round-trip'' through tables with no loss of data or
-attributes. With this protocol, it is now possible to store \astropy native
-objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord --
-within a \astropyTable and write these to various file formats, such as FITs.
-For objects not already covered by the mixin protocol, functions can be
-registered with \astropyTable\ to convert any array-like object into a mixin
-column. As an example, the mixin functions are used to integrate \package{dask}
-arrays with \astropyTable, allowing cloud-stored or cluster-scale data to be
-used as a column in a \astropyTable.
-
-\paragraph{Astropy FITS in Time}
-
-The FITS standard was extended to rigorously describe time coordinates in the
-World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared to
-other types of coordinates in WCS, time requires more metadata: format, scale,
-position, reference, etc. This metadata had to be manually specified and the
-nuances understood by the user. \astropyfits could read this data as a standard
-\astropyFitsColumn\, but would not interpret the time-related metadata and
-attributes. Through the support of the
-\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
-Summer of Code} 2017 program\footnote{ To learn more about this project, please
-see the final report,
-\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
-mixin protocol for seamless interoperability}}, the \astropyfits package now
-interprets time data correctly, using \astropyTime\ as a `mixin' column.  For
-backward compatibility with manual systems, this feature may be turned off.
-
-\paragraph{Persistent Storage}
-
-\astropy's efforts to increase support for column types in a table is mirrored
-by efforts to expand storage format options. New formats have been added, and
-existing formats updated to support more column types.
-
-\astropy now supports reading and writing tables in the American Astronomical
-Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
-long been missing and as added in this year's Google Summer of Code project adds
-\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}
-to \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
-to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
-\href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
-and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.
-
-The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
-standard} has been updated to version 1.0, adding support three additional data
-subtypes. First, it adds multidimensional column data (both masked and unmasked)
-with fixed dimensions in all table cells. Second, it adds multidimensional
-column data with variable-dimension arrays similar to FITS variable-length
-arrays. Third, it adds object-type columns with simple \python objects.
-
-As mentioned previously, \package{dask} arrays may be used as a mixin column in
-\astropyTable. Now, \package{dask} may also be the data array in
-\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the array
-will be computed while written, avoiding excessive memory use. \astropyTable\ can
-now be appended to an existing FITS file, and \package{dask} mixin columns
-interoperate seamlessly between the two table types.
-
-\paragraph{Unified I/O architecture}
-
-% TODO! better transition
-\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
-functions for \astropypkg objects, such as reading from or writing to a file,
-following a given standard. \astropy uses this internally for the I/O methods in
-\astropyTable\ and \astropyCosmology, and users can register custom I/O to
-extend the options on these classes.
-
-As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
-number of new use cases. for instance, the class-based architecture allows for
-the creation of custom registries. One application is in \astropycosmology,
-which has two different \textit{kinds} of registries for \astropyCosmology: one
-for reading and writing files, and another for converting between python
-objects. Using \astropyiosubpkg{registry} for custom classes means a developer
-does not have to create a whole new I/O system and can offer to the user a
-unified and familiar API.
-
-\subsection*{Precision, Accuracy, \& Reproducibility}
-\label{sec:core-features-precision_accuracy_repreducibility}
-
-An important focus of \astropy has been to increase the precision and accuracy
-of its functionality, while, where appropriate, making sure results obtained
-using older versions of \astropypkg are reproducible. Within each section below
-we include in our descriptions how old results may be reproduced. The primary
-means are with options in configuration files and settings on runtime configuration
-objects called \astropyScienceState. Notable improvements to \astropypkg have
-been made to \astropytime, \astropyconstants, \astropycoordinates, and
-\astropycosmology, and we describe each in turn.
-
-\paragraph{Time}
-In astronomy, time accuracy down to the (nano)second is frequently important.
-Whether for planning observations, crunching pulsar or VLBI data, missing
-seconds meaningfully impact results. Before, when using \astropyTime\ in
-\astropytime leap seconds had to be manually applied; now, leap seconds are
-applied automatically. Moreover, \astropypkg updates internal time data-files to
-ensure that the correct leap second adjustment is always used. For
-reproducibility, e.g. with an old code, \astropyLeapSeconds\ may also be
-manually applied.
-
-\paragraph{Constants}
-
-Measurements of physical constants (and the units defined from them) improve
-over time. Periodically the standardized systems of units and constants are
-updated to reflect these improvements. For instance, in 2019 the SI system was
-redefined \citep{NIST2019}, with an accompanying update to the physical
-constants in
-\href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
-2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
-values, with the units in \astropyunits based on these constants. Most of
-\astropypkg and affiliate packages build upon \astropyunits, so this update
-effects the entire \astropy ecosystem.
-
-For reproducibility, \astropypkg allows the constants' (and therefore units')
-definitions to be rolled back to prior values, e.g. to the \texttt{SI/CODATA
-2014} values. For work sensitive to the values of the fundamental constants we
-recommend including with the work an \astropy configuration file, which
-specifies the set of constants used.
-
-\paragraph{Cosmology}
-% TODO! At least one good sentence about "what is the benefit of
-% astropy.cosmology"
-
-\astropycosmology contains classes for representing cosmological models. Bundled
-with the classes are commonly used \astropyCosmology\ realizations, e.g.,
-best-fit measurements from
-\href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
-\citep{WMAP2003} and
-\href{https://www.nasa.gov/mission_pages/planck}{Planck}
-\citep{PlanckMission:2006}.
-
-When these important missions publish new measurements, \astropycosmology is
-updated to include these results as realizations of the appropriate class
-(generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
-always exactly correspond to \astropycosmology classes. For instance, the Planck
-2018 results \citep{Planck2018VI:2020} include massive neutrinos in
-$\Omega_{matter,0}$, while in \astropyFlatLambdaCDM\ this mass contribution is
-stored in a separate parameter. Consequently, while some parameter values appear
-different from the source paper, the \astropyCosmology\ realizations correctly
-reproduce the cosmological models.
-
-Since \citealt{astropy:2018} the following cosmology realizations have been
-added:
-\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{WMAP First Year}
-\citep{WMAP1Year:2003};
-\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{WMAP
-Three Year} \citep{WMAP3Year:2007};
-\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{Planck 2015}
-\citep{Planck2015XIII:2016}; and the
-\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{Planck
-2018} best-fit cosmological parameters \citep{Planck2018VI:2020}.
-
-Astropy provides a
-\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
-default cosmology}, which is used in calculations done in a cosmological
-context. The default \astropyCosmology\ has been updated to the Planck 2018
-parameters. For reproducibility, this cosmology may be set to old defaults, such
-as the WMAP Three Year values. The default cosmology is dynamically configurable
-using \astropyScienceState, meaning a set of calculations may be run using
-different assumed cosmologies, and results compared between the two.
-
-\paragraph{Coordinates}
-
-Being able to describe the position of objects, i.e., their coordinates, is
-fundamental across many knowledge domains. \astropycoordinates allows one to
-work with low-level positional and velocity data all the way to high-level
-coordinate objects with reference frames, atmospheric information, and more. A
-central feature of \astropycoordinates is the ability to transform data between
-reference frames, e.g. \astropyICRS\ \citep{ICRS:1997} to \astropyGalacticLSR\
-\citep{GalacticLSR:2010}. However, there have been some historical limitations
-for spatially proximate transformations, impacting the usefulness of
-\astropycoordinates for ground-based telescopes and astronometry within the
-solar system. Transformations in the \astropyAltAz\ frame were reasonably
-precise for very distant objects, but wrong by up to several arcseconds for
-e.g., the location of the moon. Now these transformations are much more precise,
-down to the milliarcsecond level. Similar precision improvements were made to
-the
-\astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination}
-frame transformations. Additionally,
-\astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
-and associated transformations have been updated to correctly reflect the “true”
-and “mean” terminology.
-
-For Galactic astronomers, the \astropyGalactocentric\
-\astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults}
-have been updated to include more recent measurements (tabulated in the
-\href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
-attribute). For reproducibility, old definitions are still available.
-
-\subsection*{Performance} \label{sec:core-features-performance}
-
-As features are added and corner cases dealt with, performance often suffers.
-Hence, pull requests that improve performance are welcomed, and were even the
-main goal for one release cycle.  For most sub-packages, performance was
-improved in the \python\ code, but for some, the most time-critical pieces were
-rewritten in~C (e.g., for convolution of images, sigma-clipping, and converting
-time strings to binary).  Furthermore, a particular effort was made to ensure
-\astropypkg\ is thread-safe, so that it can be used on supercomputer clusters.
+Summarize major pieces of development since the last paper (since v2.0). A few
+ideas (taken from What's New pages) below, but feel free to expand on or remove
+from this list:
+\begin{itemize}
+    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
+    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
+    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
+    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
+    \item APE 14 WCS
+    \item A new SpectralCoord class for representing and transforming spectral quantities
+    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
+\end{itemize}
 
@@
 
-\label{sec:astroquery}
-% MHvK: feel free to move above label to where it makes most sense.
-
 \secauthor{Adam Ginsburg}

### 2022-03-09 src/figures/contributor-summary.py
add contributor section figure script

@@ -1,13 +1,322 @@
-"""
-TODO: This currently just produces a placeholder figure!
-"""
-import numpy as np
+# -*- coding: utf-8 -*-
+# Ideas for plots:
+# - Commits per month
+# - Commiters per month
+# - Lines of code added/removed per month (harder)
+# - Contributors x commits in bins of astropy eras
+# - Breakdown of career stage of top N committers in different eras (early, mid, present)
+#     - Eras: 2012–2015, 2015–2018, 2018–present
+#     - Or maybe only include committers that contribute a certain number of line changes?
+#
+#
+# Stretch goals:
+# - For committers, did they start by opening an issue or go right to PR?
+
+# ---
+
+# +
+from collections import defaultdict
+import datetime
+import pathlib
+
+from astropy.time import Time
+import astropy.units as u
+from git import Repo
 import matplotlib.pyplot as plt
+import numpy as np
+# -
 
-rng = np.random.default_rng(42)
+astropy_repo_path = pathlib.Path(
+    '/Users/apricewhelan/projects/astropy-all/astropy')
 
-fig, ax = plt.subplots(1, 1, figsize=(6, 6))
-xy = rng.uniform(size=(2, 128))
-ax.plot(*xy, linestyle='none')
+repo = Repo(astropy_repo_path)
 
-fig.savefig("contributor-summary.pdf", bbox_inches="tight")
+first_commit_hash = repo.git.log('--pretty=format:%h', '--reverse').split('\n')[0]
+first_commit = list(repo.iter_commits(max_count=1, rev=first_commit_hash))[0]
+first_date = Time(first_commit.committed_date, format='unix')
+first_date.datetime
+
+# Date of last commit in v5.0
+v50_date = Time(datetime.datetime(2021, 11, 19))
+total_days = int((v50_date - first_date).jd)
+total_days
+
+# ### Time ranges:
+
+month_bins = (
+    first_date + np.arange(0, total_days, 30).astype(int) * u.day
+)
+
+# +
+# 4 bin edges for 3 eras
+era_edges = (
+    first_date + np.linspace(0, total_days, 4).astype(int) * u.day
+)
+
+eras = {
+    'early': (era_edges[0], era_edges[1]),
[…]

### 2022-03-09 src/ms.tex
add figures to tex file

--- 2022-03-03 src/ms.tex
@@
 \input{preamble.tex}
-
-% Links to documentation: subpackages
-\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
-\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
-\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
-\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
-\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
-\newcommand{\astropyunits}{\astropysubpkg{units}}
-\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
-\newcommand{\astropyconstants}{\astropysubpkg{constants}}
-\newcommand{\astropytable}{\astropysubpkg{table}}
-\newcommand{\astropytime}{\astropysubpkg{time}}
-\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
-\newcommand{\astropywcs}{\astropysubpkg{wcs}}
-\newcommand{\astropyfits}{\astropyiosubpkg{fits}}
-
-% Links to documentation: classes
-\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
-\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}
-
-\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
-\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
-\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
-\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
-GalacticLSR}{GalacticLSR}}
-\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
-\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
-\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
-\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
-\newcommand{\astropyFitsColumn}{\astropyapidoc{io.fits.Column}{Column}}
-\newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
-\newcommand{\astropyTableColumn}{\astropyapidoc{table.Column}{Column}}
-\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
-\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
-\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
-\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
-\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
-\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
-\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
-\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
-\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
-
-% suggestion: shortcut reference to Paper II
-\newcommand{\paperii}{\cite{astropy:2018}}
-
-% Maybe have command link to Zenodo reference instead, or include it?
-\newcommand{\astropyAPE}[1]{\href{https://github.com/astropy/astropy-APEs/blob/main/APE#1.rst}{APE #1}\xspace}
 
@@
 
-\secauthor{Adrian Price-Whelan}
+% \secauthor{Adrian Price-Whelan}
 
@@
 computational domains, technological sectors, and fields of research.
-Despite claims to the contrary \citep{Portegies-Zwart:2020}, \python enables
+This recent and rapid adoption of \python stems from the fact that it enables
 scalable, time- and energy-efficient code execution \citep[e.g.,][]{Augier:2021}
@@
 Over the last decade, \python has grown enormously in popularity to become a
-dominant programming language in the astronomical and broader scientific
+dominant programming language also in the astronomical and broader scientific
 communities.
@@
 exponential growth in popularity.
-The rapid adoption of \python by astronomy researchers, students, observatories,
-and technical staff combined with an associated increase in awareness and
-interest about open-source software tools is contributing to a paradigm shift in
-the way research is done, data is analyzed, and results are shared in astronomy
-and beyond.
+The adoption of \python by astronomy researchers, students, observatories, and
+technical staff combined with an associated increase in awareness and interest
+about open-source software tools is contributing to a paradigm shift in the way
+research is done, data is analyzed, and results are shared in astronomy and
+beyond.
 
@@
 An early description of the core functionality in \astropypkg can be found in
-the first Astropy paper \citep{astropy:2013} or in detail in the core package
+the first \astropy paper \citep{astropy:2013} or in detail in the core package
 documentation.\footnote{\url{https://docs.astropy.org/}}
-The \astropypkg core package is now largely stable in that the software
-interface does not change without sufficient and significant motivation, and the
-addition of new features into the core package has slowed significantly as
-compared to the first years of its development.
+The codebase in the \astropypkg core package is now largely stable, in that the
+software interface does not change without sufficient and significant
+motivation, and the addition of new features into the core package has slowed
+as compared to the first years of its development.
 This is largely driven by the fact that the core package now represents just one
@@
 is now happening in more specialized packages that are expanding the
-capabilities of the Astropy ecosystem by building on top of the foundations laid
-by the \astropypkg core package.
-Because of this natural expansion, the name Astropy has grown in scope beyond a
+capabilities of the \astropy ecosystem by building on top of the foundations
+laid by the \astropypkg core package.
+Because of this natural expansion, the name \astropy has grown in scope beyond a
 single \python library to become ``the Astropy Project.''
 
-The Astropy Project is a community effort that represents the union of the
+\begin{figure}[th]
+    \begin{centering}
+      \includegraphics[width=\textwidth]{python-mentions.pdf}
+        \caption{
+            Yearly full-text mentions of programming languages (indicated in the
+            figure legend) in refereed publications in the astronomical
+            literature database in the Astrophysics Data System (ADS).
+            \python has rapidly become the dominant programming language
+            mentioned in refereed articles over the last 10 years.
+        }
+        \label{fig:python-mentions}
+    \end{centering}
+\end{figure}
+
+The \astropy Project is a community effort that represents the union of the
 \astropypkg core package, the ecosystem of astronomy-specific software tools
@@
 
-\begin{figure}
-    \begin{centering}
-      % \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
-        \caption{
-            Yearly full-text mentions of programming languages (indicated in the
-            figure legend) in refereed publications in the astronomical
-            literature database in the Astrophysics Data System (ADS;
-            \citealt{ads}).
-            \python has rapidly become the dominant programming language
-            mentioned in refereed articles over the last 10 years.
-        }
-        \label{fig:python-mentions}
-    \end{centering}
-\end{figure}
-
 \section{Major Updates to the Astropy Core Package} \label{sec:core-updates}
@@
 
-\secauthor{Tom Robitaille}
-
-Major versions of the core package -- that is to say versions which add and/or
-modify functionality -- are released approximately every six months, and are
-then maintained with releases that fix issues, until the next major version
-is released. However, every two years a major release is designated as a
-long-term support (LTS) release which continues to be maintained for up to two
-years \citep{ape2}. The motivation for LTS releases is to provide longer-term stable
-versions of astropy that users requiring a high level of stability can make
-use of if they do not always need the latest features (this could include
-for example telescope pipelines and so on). The 5.0 release of the core package was designated as LTS,
-and since it was released at the end of 2021, it will be maintained until the
-end of 2023.
+% \secauthor{Tom Robitaille}
+
+Major versions of the core package --- that is, versions that add and/or modify
+functionality --- are released approximately every six months, and are then
+maintained with releases that fix issues until the next major version is
+released. However, every two years a major release is designated as a long-term
+support (LTS) release, which continues to be maintained for up to two years
+\citep{ape2}. The motivation for LTS releases is to provide longer-term stable
+versions of \astropypkg that users that require a high level of stability can
+make use of if they do not always need the latest features (this could include,
+for example, telescope operation or data reduction pipelines, etc.). The v5.0
+release of the core package was designated as LTS, and since it was released at
+the end of 2021, it will be maintained until the end of 2023.
 
@@
 
-\secauthor{Nathaniel Starkman, Marten van Kerkwijk}
+% \secauthor{Nathaniel Starkman, Marten van Kerkwijk}
 
@@
 Within \astropypkg, development may be roughly split into a few
-categories: new features added, intra and inter-package interoperability,
+categories: new features added, intra- and inter-package interoperability,
 improvements to precision, accuracy, and reproducibility, and performance
-enhancements. We discuss these in turn.
+enhancements. We discuss these in the subsections below.
 
@@
 was too specialized was a Virtual Observatory sub-package; the Simple
-Application Message Protocol made sense to keep (\astropysubpkg{samp}), but
-other parts more properly belonged in \package{astroquery} (see
-Sect.~\ref{sec:astroquery}).
-
-\subsubsection*{New and Planned Features} \label{sec:core-features-new}
+Application Message Protocol was kept in the core package
+(\astropysubpkg{samp}), but other parts more properly belonged in
+\package{astroquery} (see Section~\ref{sec:astroquery}).
+
+\subsubsection{New and Planned Features} \label{sec:core-features-new}
 
@@
 require the manipulation of time series. The new \astropysubpkg{timeseries}
-sub-package extends the \astropyQTable\ class to tables of data as a function of
-time, where the data can either represent samples or averages over particular
-time bins. The new classes offer a number of special methods to manipulate time
-series (folding, resampling, etc.) and to read different data formats (such as
-{\em Kepler} lightcurves, etc.).
+sub-package extends the \astropyQTable\ class to support tables of data as a
+function of time, where the data can either represent samples or averages over
+particular time bins. The new classes offer a number of special methods to
+manipulate time series (folding, resampling, etc.) and to read different data
+formats (such as \mission{Kepler} light curves, etc.).
 
@@
 \href{https://docs.astropy.org/en/stable/timeseries/bls.html}{box-least-squares}
-periodogrammes.
+periodograms.
 
@@
 
-\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
+\subsubsection{Interoperability} \label{sec:core-features-interoperability}
 
@@
 objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord --
-within a \astropyTable and write these to various file formats, such as FITs.
+within a \astropyTable\ and write these to various file formats, such as FITS.
 For objects not already covered by the mixin protocol, functions can be
@@
 \href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
-Summer of Code} 2017 program\footnote{ To learn more about this project, please
-see the final report,
+Summer of Code} 2017 program,\footnote{ To learn more about this project, please
+see the final report
 \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
-mixin protocol for seamless interoperability}}, the \astropyfits package now
+mixin protocol for seamless interoperability}.}, the \astropyfits package now
 interprets time data correctly, using \astropyTime\ as a `mixin' column. For
@@
 
-% TODO! better transition
 \astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
@@
 
-\subsection*{Precision, Accuracy, \& Reproducibility}
+\subsection{Precision, Accuracy, \& Reproducibility}
 \label{sec:core-features-precision_accuracy_repreducibility}
@@
 Whether for planning observations, crunching pulsar or VLBI data, missing
-seconds meaningfully impact results. Before, when using \astropyTime\ in
-\astropytime leap seconds had to be manually applied; now, leap seconds are
-applied automatically. Moreover, \astropypkg updates internal time data-files to
-ensure that the correct leap second adjustment is always used. For
-reproducibility, e.g. with an old code, \astropyLeapSeconds\ may also be
+seconds meaningfully impact results. In past versions of \astropypkg, when using
+\astropyTime\ in \astropytime leap seconds had to be manually applied; now, leap
+seconds are applied automatically. Moreover, \astropypkg updates internal time
+data-files to ensure that the correct leap second adjustment is always used. For
+reproducibility, e.g., with an old code, \astropyLeapSeconds\ may also be
 manually applied.
@@
 For reproducibility, \astropypkg allows the constants' (and therefore units')
-definitions to be rolled back to prior values, e.g. to the \texttt{SI/CODATA
+definitions to be rolled back to prior values, e.g., to the \texttt{SI/CODATA
 2014} values. For work sensitive to the values of the fundamental constants we
-recommend including with the work an \astropy configuration file, which
+recommend including an \astropy configuration file with the work, which
 specifies the set of constants used.
@@
 \paragraph{Cosmology}
-% TODO! At least one good sentence about "what is the benefit of
-% astropy.cosmology"
 
@@
 
-Since \citealt{astropy:2018} the following cosmology realizations have been
+Since \citet{astropy:2018}, the following cosmology realizations have been
 added:
@@
 central feature of \astropycoordinates is the ability to transform data between
-reference frames, e.g. \astropyICRS\ \citep{ICRS:1997} to \astropyGalacticLSR\
+reference frames, e.g., \astropyICRS\ \citep{ICRS:1997} to \astropyGalacticLSR\
 \citep{GalacticLSR:2010}. However, there have been some historical limitations
 for spatially proximate transformations, impacting the usefulness of
-\astropycoordinates for ground-based telescopes and astronometry within the
+\astropycoordinates for ground-based telescopes and astrometry within the
 solar system. Transformations in the \astropyAltAz\ frame were reasonably
@@
 
-\subsection*{Performance} \label{sec:core-features-performance}
+\subsection{Performance} \label{sec:core-features-performance}
 
@@
 
+Performance in \astropypkg is addressed based on feedback and requests, and
+through contributions from community members: If you find that parts of
+\astropypkg are not performant, we encourage you to open a \github issue to
+start a discussion with the \astropypkg core maintainers.
+
 \section{Major Updates in the Astropy Project} \label{sec:project-updates}
@@
 
-\secauthor{Erik Tollerud}
+% \secauthor{Erik Tollerud}
 
@@
 
-As part of the process of developing Astropy into a long-term sustainable
+As part of the process of developing \astropypkg into a long-term sustainable
 product, and to improve transparency and accountability, the Project agreed to
 write down and formalize our governance structure (partly supported by explicit
-funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
+funding for this purpose; see Section~\ref{sec:project-funding}). At the 2019
 Astropy Coordination Meeting, input was gathered from participants on what governance
 structures existed in the associated Open Source Software communities, and what
-would fit well with the needs of Astropy. This led into a ``retreat'' planned
+would fit well with the needs of \astropy. This led into a ``retreat'' planned
 for March 2020, but due to the COVID-19 pandemic, this became a series of
-virtual meetings of the ``Astropy Governance Working Group''.  This group
+virtual meetings of the ``Astropy Governance Working Group.'' This group
 drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
@@
 structure, so we refer the reader to that document for a more thorough
-description.  However here we highlight some key elements. While many of these
+description, and here we only highlight some key elements. While many of these
 principles were already de facto true or have been discussed organically (and
 have been discussed in earlier papers in this series), the APE0-based governance
-aimed to provide a single place where the community can agree as a starting
+aims to provide a single place where the community can agree as a starting
 point. With this in mind, it highlights the developer and user community of the
@@
 principle of ``do-ocracy'' that those who do work for the Project (be it coding,
-training, or other less concrete contributions) gain more influence on the
-outputs of the Project by virtue of their effort.  APE0 adds, however, the
-concept of ``voting members'' - a self-governed part of that community who are
-entrusted to elect the Coordination Committee. While this committee has existed
+training, or other harder-to-quantify contributions) gain more influence on the
+outputs of the Project by virtue of their effort. However, APE0 adds the
+concept of ``voting members'': a self-governed part of that community who are
+entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
 from the inception of the project, APE0 establishes a formal voting process for
-this committee, and explicitly outlines the rights and responsibilities of the
-Coordination Committee. This role is mainly to facilitate consensus and act as
-the decision maker when other mechanisms have failed.  However, it also includes
-powers that either require central authority or secrets (e.g., passwords), but
-APE0 also charges the Committee to devolve responsibilities and seek community
-input on these items as often as possible.
+CoCo members, and explicitly outlines the rights and responsibilities of the
+CoCo. This role is mainly to facilitate consensus and act as the decision maker
+when other mechanisms have failed, and includes powers that either
+require central authority or secrets (e.g., passwords) However, APE0 also
+charges the Committee to devolve responsibilities and seek community input on
+these items as often as possible.
 
 The first Coordination Committee election under these rules took place in Fall
-2021, electing a mix of prior existing and new coordination committee members,
+2021, electing a mix of prior and new coordination committee members,
 and was contested in the sense of more candidates than available slots. This
@@
     \begin{centering}
-      % \includegraphics{figures/contributor-summary.pdf}
-        \caption{Placeholder figure!}
-        \label{fig:contributor-summary}
+      \includegraphics{figures/commits_committers_per_month.pdf}
+        \caption{
+            \todo{Adrian}{Write caption@}
+        }
+        \label{fig:contributor-summary:commits_committers_per_month}
     \end{centering}
@@
 
+\begin{figure*}
+    \begin{centering}
+      \includegraphics{figures/Ncommitters_vs_frac_commits.pdf}
+        \caption{
+            \todo{Adrian}{Write caption@}
+        }
+        \label{fig:contributor-summary:Ncommitters_vs_frac_commits}
+    \end{centering}
+\end{figure*}
+
+\begin{figure*}
+    \begin{centering}
+      \includegraphics{figures/top_contributors_career.pdf}
+        \caption{
+            \todo{Adrian}{Write caption@}
+        }
+        \label{fig:contributor-summary:top_contributors_career}
+    \end{centering}
+\end{figure*}
+
 % Transition of needs: From new features, to community development, sustainability, ...
@@
 
-\secauthor{Lía Corrales}
+% \secauthor{Lía Corrales}
 
@@
 Github. This process was deemed the most ``open'' because it allowed
-for community wide feedback to focus and improve proposal
-initiatives. Two programs were selected via this process, described
-below.
-
-\textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}
+for community-wide feedback to focus and improve proposal
+initiatives. Two highlighted programs that were selected and implemented via
+this process are described below.
+
+% \textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}
 
@@
 \textbf{\astropy representation at national diversity conferences:}
-Several members of the \astropy community attended virtual conferences
-of the National Society of Black Physicists (NSBP) and SACNAS
-(focusing on Hispanic and Native American scientists across all STEM
-fields). \astropy representatives noted that one underlying topic came
-up multiple times. A major barrier for persons who come from
-underrepresented communities is the lack of resources and expertise
-that are necessary to train students. Writing PEP 8 compliant software
-and understanding the Git and Github workflow is not part of the
-standard Physics and Astrophysics curriculum. Additionally, the larger
-Astronomnical community is still in the process of transitioning
-towards Python based tools, making it difficult for students not
-currently under advisement by some one with extensive Python expertise
-to get involved with the open source Python community. Projects such
-as Learn Astropy could have a profound impact by empowering
-underrepresented groups because it provides a free, searchable, and
-accessible introduction to Python tools for astrophysical
-research. Representatives also noted that offering training and
-teaching materials to PIs of Research Experience for Undergraduate
-(REU) programs throughout the United States could be helpful for
-encouraging advisors to teach students how to use Astropy, create
-their own libraries, and use version control on open source platforms
-like Github. Such materials could also be offered as a workshop at
-national diversity conferences themselves, as SACNAS solicits special
-session proposals each year.
-
-\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
-  initiative:} \todo{No author assigned}{Describe more about this.}
-
-\subsection{Funding} \label{sec:project-funding}
-
-\secauthor{Aarya Patil}
-
-Summarize funding sources (Moore, NASA) and amounts and what this has been used
-for.
+Several members of the \astropy community attended virtual conferences of the
+National Society of Black Physicists (NSBP) and SACNAS (focusing on Hispanic and
+Native American scientists across all STEM fields). \astropy representatives
+noted that one underlying topic came up multiple times. A major barrier for
+persons who come from underrepresented communities is the lack of resources and
+expertise that are necessary to train students. Writing PEP 8 compliant software
+and understanding the Git and \github workflow is not part of the standard
+Physics and Astrophysics curriculum. Additionally, the larger astronomical
+community is still in the process of transitioning towards \python-based tools,
+making it difficult for students not currently under advisement by some one with
+extensive \python expertise to get involved with the open source \python
+community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
+have a profound impact by empowering underrepresented groups because it provides
+a free, searchable, and accessible introduction to \python tools for
+astrophysical research. Representatives also noted that offering training and
+teaching materials to PIs of Research Experience for Undergraduate (REU)
+programs throughout the United States could be helpful for encouraging advisors
+to teach students how to use Astropy, create their own libraries, and use
+version control on open source platforms like \github. Such materials could also
+be offered as a workshop at national diversity conferences themselves, as SACNAS
+solicits special session proposals each year.
+
+% \textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
+%   initiative:} \todo{No author assigned}{Describe more about this.}
+
+\subsection{Project funding} \label{sec:project-funding}
+
+% \secauthor{Aarya Patil}
+
+% Summarize funding sources (Moore, NASA) and amounts and what this has been
+% used for.
+
+With the goal of establishing a sustainable financial model for the Astropy
+Project, the Coordination Committee, under the new governance charter (see
+Section~\ref{sec:project-governance}), established a standing Finance Committee.
+This committee oversees the planning and allocation of finances on behalf
+of the Project. The Astropy Project accepts funds from institutions as well as
+individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
+supports and promotes world-class, innovative, open source scientific computing.}
+Previously, no direct financial support was available for the project development,
+and NumFOCUS covered most of the incurred operational costs. However, transitioning
+to long-term financial stability meant having an influx of funds in the form of
+institutional support, and hence the Astropy Project applied for and successfully
+acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
+Project, providing a ``neutral'' space not tied to any particular astronomy
+institution. The two major grants are briefly summarized below.
+
+\begin{description}
+    \item[Moore] The Gordon and Betty Moore Foundation awarded the Astropy
+    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
+    Growing the Astropy Project.'' Over a three-year term, this grant supported
+    sustenance of the Project by providing funds for (1) the development and
+    maintenance of the \astropypkg and its infrastructure, provided to existing
+    project members and targeted hires, (2) the transition of long-term users to
+    contributors and maintainers through mentorship efforts, (3) the
+    formalization of a governance structure (see
+    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
+    diversity, and inclusion efforts in the project (see
+    Section~\ref{sec:project-ide}). Some of this money supported travel for
+    meetings, conferences, and workshops. This generous grant helped reduce
+    reliance on volunteer-driven maintenance of the openly-developed package and
+    paved the way to acquire funding from federal agencies, representing a major
+    milestone for the success of the project.
+
+    \item[NASA] A successful proposal to the NASA ROSES-2020
+    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
+    (Support for Open Source Tools, Frameworks, and Libraries), granted the
+    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
+    term of three years includes (1) project-wide infrastructure maintenance and
+    improvements, (2) targeted work on areas of the \astropy core and
+    coordinated packages, (3) enhancements to the Learn
+    Astropy ecosystem and educational materials (see Section~\ref{sec:learn}),
+    and (4) support for the \astropy affiliated packages. These goals
+    support the sustenance of the project by encouraging further
+    development of the package ecosystem and community engagement.
+\end{description}
+
+In addition to the two grants described above, the project benefited from
+a Gemini Observatory contract under its Science User Support Department,
+which awarded funds for development work that supports both the Astropy and
+DRAGONS projects. The project also received financial support from the Dunlap
+Seed Funding program at the University of Toronto to develop educational
+resources for scientific software packaging within the Learn Astropy
+framework. All of these successful proposals show that the Astropy Project
+is vital for the astronomical infrastructure and community at large, and
+strengthen its promising and sustainable future.
+
+% ADD text for the new finance APE once it has a doi
 
@@
 
-\secauthor{Nicholas Earl}
+% \secauthor{Nicholas Earl}
 
 The Astropy project supports the broader ecosystem by providing pre-configured
-infrastructure packages that the community can use to support and maintain
-their own software package infrastructure. These include tools to easily
-generate documentation and setup automated testing, as well as provide
-package scaffolding for new projects.
+infrastructure packages that the community can use to support and maintain their
+own software package infrastructure. These include tools and extensions that
+enable easily generating documentation and setting up automated testing, as well
+as providing package scaffolding for new projects.
 
@@
 packages. The Astropy project maintains a default Sphinx configuration along
-with Astropy-specific extensions which can be easily added to community
+with Astropy-specific extensions that can be easily added to community
 projects via the
@@
 (\href{https://github.com/pytest-dev/pytest-cov}{\texttt{pytest-cov}}), and
-configuring package for property-based testing
+configuring packages for property-based testing
 (\href{https://github.com/HypothesisWorks/hypothesis}{\texttt{hypothesis}}).
@@
 package template generation process includes the ability to setup
-interoperability with GitHub, allowing for easy repository access from
-documentation, as well as an example GitHub Actions workflow to demonstrate
-the use of GitHub's continuous integration tooling.
-
-\subsection{Affiliated packages}
-
-\secauthor{Matt Craig, Brett Morris}
-
-Highlight a few new affiliated packages and major updates to existing ones.
-Include a big table of all affiliated packages and references (as in v2.0
-paper).
+interoperability with \github, allowing for easy repository access from
+documentation, as well as an example \github Actions workflow to demonstrate
+the use of \github's continuous integration tooling.
+
+\subsection{Astropy affiliated packages}
+
+% \secauthor{Matt Craig, Brett Morris}
+
+% Highlight a few new affiliated packages and major updates to existing ones.
+% Include a big table of all affiliated packages and references (as in v2.0
+% paper).
+
+As defined in \paperii, Astropy affiliated packages are astronomy-related
+\python packages that provide functionality that builds upon and extends the
+\astropypkg core package that have requested to be included as part of the
+Astropy Project community. These packages support the goals and vision of
+Astropy of improving code re-use, interoperability, and embracing good coding
+practices such as testing and thorough documentation.
+All Astropy-coordinated and other affiliated packages are listed with detailed
+information on the \href{{https://www.astropy.org/affiliated}}{\astropy
+website}; A table summarizing key aspects of the affiliated packages is included
+in the Appendix below (Table~\ref{tab:affiliated-registry}).
+
+Since \paperii, there has been an expansion of affiliated
+packages for gravitational astrophysics, including:
+\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
+modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
+gravitational wave probability maps, and \texttt{EinsteinPy} for general
+relativity and gravitational astronomy. There have also been several packages
+added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
+for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
+Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
+and \texttt{python-cpl} has been added for ESO pipelines and VLT data products.
+For ground-based astronomy, \texttt{baseband} has added IO capabilities for
+VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
+Other new affiliated packages include \texttt{agnpy} for AGN jets,
+\texttt{statmorph} for fitting galactic morphological diagnostics,
+\texttt{dust\_extinction} for modeling interstellar dust extinction,
+\texttt{feets} for extracting features from time series data, and
+\texttt{corral} for managing data intensive parallel pipelines.
+\texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
+\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
+and \texttt{sbpy} enables calculations for asteroid and
+cometary astrophysics. Finally, \texttt{synphot} provides an interface for
+synthetic photometry.
+% \citep{Mommert2019}
+
+Several of the coordinated and affiliated packages described in
+\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
+\citep{astroquery} has added access to roughly a dozen new missions and data
+services, including the James Webb Space Telescope (JWST) archive, and the
+project has switched to a continuous release model: Every time a change is
+committed to the main development branch it is published on the Python Package
+Index (PyPI) and available for installation. (However, formal releases are still
+done a few times per year.) \texttt{photutils} \cite {photutils} released its
+first stable version, indicating that the API will change less frequently, and
+there have been several significant performance improvements. \texttt{ccdproc}
+\cite{ccdproc} also released a new major version, bringing better performance to
+some image combination operations. The \texttt{regions} package
+\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
+added new ways to manipulate regions and introduced new region types.
+\texttt{reproject}'s \cite{reproject} major new feature is a function to align
+and co-add images to create a mosaic; better support for parallelization was
+also added. \texttt{specutils}, the package that defines containers for 1D and
+2D spectra \cite{specutils}, also had its first stable release and the addition
+of classes to read JWST data. \texttt{stingray} has also had a major performance
+overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
+improvements and has unified its API in preparation for its first stable
+release.
 
 \subsection{Connections with data archives}
-
 \label{sec:astroquery}
-% MHvK: feel free to move above label to where it makes most sense.
-
-\secauthor{Adam Ginsburg}
-
-Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
-It contains over 50 modules for querying astronomical databases, large and small.
-In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
-These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
-
-The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented.
-Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}].
-More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.
-
-Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
-These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
-While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
-% note to coauthors: we need a section on pyvo
+
+% \secauthor{Adam Ginsburg}
+
+\texttt{astroquery} \citep{Ginsburg2019} is the Astropy-coordinated package for
+interacting with online archives of astronomical and related data. It contains
+over 50 modules for querying astronomical databases, large and small. In
+particular, since \paperii, significant contributions to \texttt{astroquery}
+have come from several of the major archives, including the European Space
+Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space
+Telescope Science Institute (STScI), the Infrared Science Archive at the NASA
+Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian
+Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter
+Array archive (ALMA). These contributions represent a formal acknowledgement of
+the utility of a centralized tool suite for archive interaction from \python.
+The widespread usage of \texttt{astroquery} is apparent from the range of
+keywords represented in the journal articles that cite the \texttt{astroquery}
+paper \citep{Ginsburg2019}; everything from asteroids to galaxies is
+represented. Additionally \texttt{astroquery} functionality has been built into
+several special-purpose \python packages, such as \texttt{LightKurve}
+\citep{LightKurve} and SORA \citep{SORA}]. More than 4,000 repositories on
+\github make use of \texttt{astroquery} in some way, and \texttt{astroquery}
+features in a large number of tutorials on various facets of astronomical
+analysis.
+
+Many of the existing and newly-contributed tools rely on Virtual Observatory
+(VO) tools. These use the underlying package \texttt{pyvo}, which has also
+recently become an Astropy-coordinated package. While many or all of the
+functions provided in \texttt{astroquery} can be achieved through direct use of
+VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more
+closely resemble the web interfaces that are more familiar to most users.
 
@@
 
-Brief summary of efforts in observatory- or mission-driven development that have
-contributed to Astropy, and vice versa.
+In this section, we highlight a few efforts in observatory- or mission-driven
+development that have contributed to Astropy, and vice versa.
 
 \subsubsection{James Webb Space Telescope}
-\secauthor{Larry Bradley}
+% \secauthor{Larry Bradley}
 
@@
 infrared telescope that will provide unprecedented resolution and
-sensitivity from 0.6 -- 28 microns. JWST will enable a broad range
+sensitivity from 0.6--28 microns. JWST will enable a broad range
 of scientific investigations from exoplanets and their atmospheres
@@
 Software developers at the Space Telescope Science Institute (STScI),
-the operations center of JWST, have been developing Python-based tools
+the operations center of JWST, have been developing \python-based tools
 for JWST since 2010 (starting with the JWST Calibration Reference Data
@@
 
-Gemini
-\secunfilled
-
-Cherenkov Telescope Array
-\secauthor{Axel Donath, Maximilian Nöthe}
+% Gemini
+% \secunfilled
+
+\subsubsection{Cherenkov Telescope Array}
+% \secauthor{Axel Donath, Maximilian Nöthe}
 
@@
 of Cherenkov telescopes to the reconstructed event lists.
-
-The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
+The high-level analysis (or CTA science tools) will be based on the Astropy affiliated package
 Gammapy~\citep{gammapy}.
-
-Both Gammapy and \texttt{ctapipe} are using astropy heavily, mainly for units, times, coordinate transformations,
-tables and FITS IO.
-
+Both Gammapy and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
+tables and FITS I/O.
 As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
@@
 based on earlier work by B.~Winkel.
-
 Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
@@
 
-Rubin Observatory
-\secunfilled
-
-LIGO/Virgo/KAGRA
-\secauthor{Leo Singer}
-\section{Future Plans for the Astropy Project} \label{sec:future}
-
-\subsection{Roadmap}
-
-\secauthor{Clara Brasseur}
-
-Overview of roadmap and some highlights.
-
-\section{Learn Astropy} \label{sec:learn}
+% Rubin Observatory
+% \secunfilled
+
+% LIGO/Virgo/KAGRA
+% \secauthor{Leo Singer}
+
+\subsection{Learn Astropy} \label{sec:learn}
 
@@
 
-\subsubsection{Current status and scope}
+% \subsubsection{Current status and scope}
 
@@
 
-\subsubsection{Learn vision for the future}
+% \subsubsection{Learn vision for the future}
 
@@
 
+\section{Future Plans for the Astropy Project} \label{sec:future}
+
+\subsection{A roadmap for future priorities}
+% \secauthor{Clara Brasseur}
+
+The
+\href{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}{Astropy Roadmap} is a document in the \texttt{astropy-project} \github
+repository within the \astropy organization that captures high level actionable
+items that the Astropy project aims to undertake to improve the health and
+stability of the project. It is a static document that is revisited regularly at
+the Astropy coordination meetings, intended to keep track of progress and write
+new versions as needed. There is a related project board linked to the Roadmap
+document that holds specific issues and efforts related to Roadmap items. The
+project board is a living document that is continually updated as work is
+planned, assigned, and completed.
+
+All items in the Roadmap have been agreed to be priorities for the Astropy
+Project, and are color-coded based on resources (both time/effort and
+developers/expertise) needed to complete the item. Green items are well
+underway, have sufficient resources/support and a plan in place for completion;
+For example, efforts to improve the discoverability of documentation and
+educational materials by overhauling the Learn website, which is underway (see
+Section~\ref{sec:learn}).
+Orange items are well defined, and work for acquiring sufficient resources
+underway; For example, the goal of providing next-generation spectroscopic
+reduction, analysis, and visualization tools usable by individual researchers
+and larger surveys.
+Red items do not yet have a plan for implementation and need more resources; For
+example, the goal of improving interoperability with performant I/O file formats
+and libraries such as HDF5 and Dask.
+The Astropy Roadmap originates from the March 2021 Astropy Coordination meeting
+where it was first drafted before being handed off to a newly formed Astropy
+Roadmap working group for completion. The Roadmap in its current form was
+adopted via pull request in December 2021.
+
 \subsection{Current and Future Challenges}
 
-\section{Community Engagement}
+Community Engagement
 
@@
 
-We would like to thank the members of the community that have contributed to
-\astropy, that have opened issues and provided feedback, and have supported the
+We would like to thank the members of the community who have contributed to
+\astropy, who have opened issues and provided feedback, and have supported the
 project in a number of different ways.
 
-The \astropy community is supported by and makes use
-of a number of organizations and services outside the traditional
-academic community. We thank Google for financing and organizing the
+The \astropy community is supported by and makes use of a number of
+organizations and services outside the traditional academic community.
+We thank Google for financing and organizing the
 Google Summer of Code (GSoC) program, that has funded severals
 students per year to work on \astropy related projects over the
-summer. These students often turn into long-term contributors. We also
-thank NumFOCUS and the Python Software Foundation for financial
-support. Within the academic community, we thank
-institutions that make it possible that astronomers and other developers on
-their staff can contribute their time to the development of
-\astropy projects.  We would like acknowledge the support of the
-Space Telescope Science Institute, Harvard--Smithsonian Center for Astrophysics,
-and the South African Astronomical Observatory.
+summer. These students often turn into long-term contributors.
+We also thank NumFOCUS and the Python Software Foundation for financial
+support.
+Within the academic community, we thank institutions that make it possible that
+astronomers and other developers on their staff can contribute their time to the
+development of \astropy projects.
+We would like acknowledge the support of the Space Telescope Science Institute,
+Harvard--Smithsonian Center for Astrophysics, and the South African Astronomical
+Observatory.
+
+Supported by the international Gemini Observatory, a program of NSF's NOIRLab,
+which is managed by the Association of Universities for Research in Astronomy
+(AURA) under a cooperative agreement with the National Science Foundation, on
+behalf of the Gemini partnership of Argentina, Brazil, Canada, Chile, the
+Republic of Korea, and the United States of America.
 
@@
 a number of web services for code hosting, continuous integration, and
-documentation; in particular, \astropy heavily relies on GitHub, Travis CI,
-Appveyor, CircleCI, and Read the Docs.
+documentation; in particular, \astropy heavily relies on GitHub, Azure
+Pipelines, CircleCI, and Read the Docs.
+
+This research has made use of NASA's Astrophysics Data System.
 
@@
 \bibliographystyle{aasjournal}
-\bibliography{refs}
-
-% \appendix
-
-% \section{List of Affiliated Packages}
-
-% \begin{longrotatetable}
-%     \begin{deluxetable*}{cccp{3in}c}
-%     \tablecaption{Registry of affiliated packages.}
-%     \label{tab:registry}
-%     \tablehead{
-%         \colhead{Package Name} &
+\bibliography{refs, static/affiliated-refs}
+
+\appendix
+
+\section{List of Affiliated Packages}
+
+\begin{longrotatetable}
+    \begin{deluxetable*}{ccp{2in}p{2.5in}}
+    \tablecaption{Registry of affiliated packages.}
+    \label{tab:affiliated-registry}
+    \tablehead{
+        \colhead{Package Name} &
         % \colhead{Stable} &
-%         \colhead{PyPI Name} &
-%         \colhead{Maintainer} &
-%         \colhead{Citation}
-%       }
-%       \startdata
-%         \input{registry.tex}
-%       \enddata
-%   \end{deluxetable*}
-% \end{longrotatetable}
+        \colhead{PyPI Name} &
+        \colhead{Maintainer(s)} &
+        \colhead{Citation(s)}
+      }
+      \startdata
+        \input{generated/affiliated-table.tex}
+      \enddata
+  \end{deluxetable*}
+\end{longrotatetable}
 

### 2022-03-09 src/ms.tex
fix figures build

@@
 % To allow putting figures in a subdir
-\graphicspath{{figures/}}
+% \graphicspath{{figures/}}
 
@@
     \begin{centering}
-      \includegraphics{figures/commits_committers_per_month.pdf}
+      \includegraphics[width=0.5\textwidth]{commits_committers_per_month.pdf}
         \caption{
-            \todo{Adrian}{Write caption@}
+            todo
         }
-        \label{fig:contributor-summary:commits_committers_per_month}
+        \label{fig:contributor-summary:per-month}
     \end{centering}
@@
 
-\begin{figure*}
+\begin{figure}
     \begin{centering}
-      \includegraphics{figures/Ncommitters_vs_frac_commits.pdf}
+      \includegraphics[width=\textwidth]{Ncommitters_vs_frac_commits.pdf}
         \caption{
-            \todo{Adrian}{Write caption@}
+            todo
         }
-        \label{fig:contributor-summary:Ncommitters_vs_frac_commits}
+        \label{fig:contributor-summary:Ncommitters}
     \end{centering}
-\end{figure*}
-
-\begin{figure*}
+\end{figure}
+
+\begin{figure}
     \begin{centering}
-      \includegraphics{figures/top_contributors_career.pdf}
+      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
         \caption{
-            \todo{Adrian}{Write caption@}
+            todo
         }
-        \label{fig:contributor-summary:top_contributors_career}
+        \label{fig:contributor-summary:top}
     \end{centering}
-\end{figure*}
+\end{figure}
 

### 2022-03-11 src/ms.tex
new figure and words

@@
 
-\subsection{Contributor base} \label{sec:project-contributors}
+\subsection{Core package contributor base} \label{sec:project-contributors}
 
@@
 
-Overview and statistics of contributors. Highlight changes since v2.0.
-
-\begin{figure}
+% TODO: move this subsection to "updates to core package" section above?
+
+\begin{figure}[ht]
+    \begin{centering}
+      \includegraphics[width=0.8\textwidth]{Ncommitters_vs_frac_commits.pdf}
+        \caption{
+            todo
+        }
+        \label{fig:contributor-summary:Ncommitters}
+    \end{centering}
+\end{figure}
+
+As described in the introduction above, the \astropy Project is broader than the
+core \astropypkg package as it also represents a number of core infrastructure
+packages, affiliated packages, educational initiatives, and user communities.
+In total, well over 1500 people have contributed in some way to the Project,
+whether, for example, through contributing code or documentation, opening issues
+on \github, providing infrastructure support, writing tutorials, or helping to
+coordinate workshops that relate to Astropy.
+The vast majority of these contributors and participants have interacted as
+volunteers choosing to engage with the open-source and open-development software
+community.
+However, while the total number of contributors is large, the number of
+contributions and bulk of the maintenance of the \astropy Project components has
+primarily been sustained by a small number of heavily-engaged or
+explicitly-employed contributors \citep[see, e.g., ][]{astropy:2018}, as is the
+case in many large, open-source software communities \citep[e.g.,][]{numpy}.
+
+As an example of this effect within the \astropy Project, here we focus on
+contributions made to the \astropypkg core package.
+Figure~\ref{fig:contributor-summary:Ncommitters} shows the number of
+contributors who have contributed a given fraction of the total number of Git
+commits, separated into three separate ``eras'' in the history of the
+Project.\footnote{Note, however, that measuring these statistics in terms of Git
+commits is not a perfect way of encapsulating ``effort,'' but we use this metric
+for its simplicity and clear definition.}
+The annotations show that, within the early (July 2011--January 2015), mid
+(January 2015--June 2018), and recent (June 2018--November 2021) eras, 90\% of
+the commits to the \astropypkg core package have come from 18, 34, and 44
+contributors.
+While the trend is encouraging --- a larger pool of contributors are
+contributing more, fractionally --- this emphasizes that, still, the vast
+majority of the work in \astropypkg is done my a small subset of the
+contributors.
+
+\begin{figure}[ht]
     \begin{centering}
@@
 
-\begin{figure}
-    \begin{centering}
-      \includegraphics[width=\textwidth]{Ncommitters_vs_frac_commits.pdf}
-        \caption{
-            todo
-        }
-        \label{fig:contributor-summary:Ncommitters}
-    \end{centering}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[ht]
     \begin{centering}

### 2022-03-11 src/figures/contributor-summary.py

@@ -70,6 +70,7 @@ for name, era in eras.items():
 
 figsize = (8.5, 6)
 style = dict(color='k')
+era_colors = ['#1b9e77', '#d95f02', '#7570b3']
 
 # ## Commits per month
 
@@ -272,7 +273,49 @@ ax.set_yticklabels([str(tick) for tick in yticks])
 
 ax.set_ylim(0.5, 200)
 
-fig.savefig('Ncommitters_vs_frac_commits.pdf')
+# +
+fig, ax = plt.subplots(figsize=figsize)
+for i, (name, _n_commits) in enumerate(era_n_commits.items()):
+    y = np.cumsum(_n_commits) / np.sum(_n_commits)
+    x = np.arange(1, len(y)+1)
+    ax.semilogx(x, y, 
+                marker='', color=era_colors[i], lw=2,
+                label=name)
+    
+    _idx = np.abs(y - 0.9).argmin()
+    
+    btm = 0.9 - 0.4 - i * 0.15
+    ax.plot(
+        [x[_idx]]*2, [btm, 0.9], 
+        marker='', ls='--', 
+        alpha=0.6, color=era_colors[i]
+    )
+    
+    _val = np.where(y > 0.9)[0][0]
+    ax.text(1.15 * x[_idx], btm, 
+            f'{_val} committers',
+            fontsize=16,
+            ha='right', va='top',
+            color=era_colors[i]
+           )
+    
+    print(name, np.where(y > 0.5)[0][0], np.where(y > 0.9)[0][0])
+    
+ax.set_xlim(1, 300)
+ax.set_ylim(0, 1.04)
+ax.legend(
+    loc='lower right', fontsize=18,
+    title='Era:', title_fontsize=19
+)
+
+ax.text(1.1, 0.9, '90% of commits', ha='left', va='bottom', 
+        fontsize=16, color='#666666')
+ax.axhline(0.9, zorder=-10, color='#aaaaaa', ls='--')
+
+ax.set_xlabel('Number of committers')
+ax.set_ylabel('Cumulative fraction of commits\nin each time period')
+
+fig.savefig('Ncommitters_vs_frac_commits.pdf', bbox_inches='tight')
 # -
 
 # ## Career stages
@@ -305,10 +348,8 @@ fig, ax = plt.subplots(figsize=figsize, constrained_layout=True)
 ind = np.arange(len(labels))
 width = 0.25
 
-colors = ['C1', 'C2', 'C3']
-
[…]

### 2022-03-11 src/ms.tex
numbers

@@
 
-As an example of this effect within the \astropy Project, here we focus on
+As an example of this within the \astropy Project, here we focus on
 contributions made to the \astropypkg core package.
@@
 contributors who have contributed a given fraction of the total number of Git
-commits, separated into three separate ``eras'' in the history of the
-Project.\footnote{Note, however, that measuring these statistics in terms of Git
-commits is not a perfect way of encapsulating ``effort,'' but we use this metric
-for its simplicity and clear definition.}
+commits, separated into three separate time spans (``eras'') in the history of
+the Project.\footnote{Note, however, that measuring these statistics in terms of
+Git commits is not a perfect way of encapsulating ``effort,'' but we use this
+metric for its simplicity and clear definition.}
 The annotations show that, within the early (July 2011--January 2015), mid
@@
 contributing more, fractionally --- this emphasizes that, still, the vast
-majority of the work in \astropypkg is done my a small subset of the
-contributors.
+majority of the work in \astropypkg is done my a small subset ($<20\%$) of the
+115, 216, and 234 contributors, respectively, within each era.
+
+Since its inception, the most active contributors to the \astropypkg package
+have spanned a range of career stages and career paths.
+
+\begin{figure}[ht]
+    \begin{centering}
+      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
+        \caption{
+            todo
+        }
+        \label{fig:contributor-summary:top}
+    \end{centering}
+\end{figure}
 
@@
         \label{fig:contributor-summary:per-month}
-    \end{centering}
-\end{figure}
-
-\begin{figure}[ht]
-    \begin{centering}
-      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
-        \caption{
-            todo
-        }
-        \label{fig:contributor-summary:top}
     \end{centering}

### 2022-03-11 src/figures/contributor-summary.py

@@ -233,6 +233,8 @@ for name, era_lim in eras.items():
     ).split('\n')
     era_shortlogs[name] = _shortlog
     era_n_commits[name] = np.array([int(x.split('\t')[0]) for x in _shortlog])
+    
+    print(f"era={name} had {len(era_n_commits[name])} committers")
 
 # +
 _bins = np.logspace(-5, 0, 25)

### 2022-03-11 src/ms.tex
Whatsnew section (#32)
* Start section introduction

Signed-off-by: nstarman <nstarkman@protonmail.com>

* fix ampersands

* New features as itemized list

* Text for uncertainties and distributions

* Time series and box-least-squares

* Spectral coordinates

* WCS high-level interface

* Remove pytest from whatsnew list

* Expand on future plans of uncertainties

* Expand intro to new features

* Discuss removal of ERFA and .vo

* capitalize SkyCoord shortcut

Signed-off-by: nstarman <nstarkman@protonmail.com>

* cleanup section intro

Signed-off-by: nstarman <nstarkman@protonmail.com>

* dr

--- 2022-03-09 src/ms.tex
@@
 % To allow putting figures in a subdir
-% \graphicspath{{figures/}}
+\graphicspath{{figures/}}
 
@@
 \input{preamble.tex}
+
+% Links to documentation: subpackages
+\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
+\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
+\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
+\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
+\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
+\newcommand{\astropyunits}{\astropysubpkg{units}}
+\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
+\newcommand{\astropyconstants}{\astropysubpkg{constants}}
+\newcommand{\astropytable}{\astropysubpkg{table}}
+\newcommand{\astropytime}{\astropysubpkg{time}}
+\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
+\newcommand{\astropywcs}{\astropysubpkg{wcs}}
+\newcommand{\astropyfits}{\astropyiosubpkg{fits}}
+
+% Links to documentation: classes
+\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
+\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}
+
+\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
+\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
+\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
+\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
+GalacticLSR}{GalacticLSR}}
+\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
+\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
+\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
+\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
+\newcommand{\astropyFitsColumn}{\astropyapidoc{io.fits.Column}{Column}}
+\newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
+\newcommand{\astropyTableColumn}{\astropyapidoc{table.Column}{Column}}
+\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
+\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
+\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
+\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
+\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
+\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
+\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
+\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
+\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}
+
+% suggestion: shortcut reference to Paper II
+\newcommand{\paperii}{\cite{astropy:2018}}
+
+% Maybe have command link to Zenodo reference instead, or include it?
+\newcommand{\astropyAPE}[1]{\href{https://github.com/astropy/astropy-APEs/blob/main/APE#1.rst}{APE #1}\xspace}
 
@@
 
-% \secauthor{Adrian Price-Whelan}
+\secauthor{Adrian Price-Whelan}
 
@@
 computational domains, technological sectors, and fields of research.
-This recent and rapid adoption of \python stems from the fact that it enables
+Despite claims to the contrary \citep{Portegies-Zwart:2020}, \python enables
 scalable, time- and energy-efficient code execution \citep[e.g.,][]{Augier:2021}
@@
 Over the last decade, \python has grown enormously in popularity to become a
-dominant programming language also in the astronomical and broader scientific
+dominant programming language in the astronomical and broader scientific
 communities.
@@
 exponential growth in popularity.
-The adoption of \python by astronomy researchers, students, observatories, and
-technical staff combined with an associated increase in awareness and interest
-about open-source software tools is contributing to a paradigm shift in the way
-research is done, data is analyzed, and results are shared in astronomy and
-beyond.
+The rapid adoption of \python by astronomy researchers, students, observatories,
+and technical staff combined with an associated increase in awareness and
+interest about open-source software tools is contributing to a paradigm shift in
+the way research is done, data is analyzed, and results are shared in astronomy
+and beyond.
 
@@
 An early description of the core functionality in \astropypkg can be found in
-the first \astropy paper \citep{astropy:2013} or in detail in the core package
+the first Astropy paper \citep{astropy:2013} or in detail in the core package
 documentation.\footnote{\url{https://docs.astropy.org/}}
-The codebase in the \astropypkg core package is now largely stable, in that the
-software interface does not change without sufficient and significant
-motivation, and the addition of new features into the core package has slowed
-as compared to the first years of its development.
+The \astropypkg core package is now largely stable in that the software
+interface does not change without sufficient and significant motivation, and the
+addition of new features into the core package has slowed significantly as
+compared to the first years of its development.
 This is largely driven by the fact that the core package now represents just one
@@
 is now happening in more specialized packages that are expanding the
-capabilities of the \astropy ecosystem by building on top of the foundations
-laid by the \astropypkg core package.
-Because of this natural expansion, the name \astropy has grown in scope beyond a
+capabilities of the Astropy ecosystem by building on top of the foundations laid
+by the \astropypkg core package.
+Because of this natural expansion, the name Astropy has grown in scope beyond a
 single \python library to become ``the Astropy Project.''
 
-\begin{figure}[th]
-    \begin{centering}
-      \includegraphics[width=\textwidth]{python-mentions.pdf}
-        \caption{
-            Yearly full-text mentions of programming languages (indicated in the
-            figure legend) in refereed publications in the astronomical
-            literature database in the Astrophysics Data System (ADS).
-            \python has rapidly become the dominant programming language
-            mentioned in refereed articles over the last 10 years.
-        }
-        \label{fig:python-mentions}
-    \end{centering}
-\end{figure}
-
-The \astropy Project is a community effort that represents the union of the
+The Astropy Project is a community effort that represents the union of the
 \astropypkg core package, the ecosystem of astronomy-specific software tools
@@
 
+\begin{figure}
+    \begin{centering}
+      % \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
+        \caption{
+            Yearly full-text mentions of programming languages (indicated in the
+            figure legend) in refereed publications in the astronomical
+            literature database in the Astrophysics Data System (ADS;
+            \citealt{ads}).
+            \python has rapidly become the dominant programming language
+            mentioned in refereed articles over the last 10 years.
+        }
+        \label{fig:python-mentions}
+    \end{centering}
+\end{figure}
+
 \section{Major Updates to the Astropy Core Package} \label{sec:core-updates}
@@
 
-% \secauthor{Tom Robitaille}
-
-Major versions of the core package --- that is, versions that add and/or modify
-functionality --- are released approximately every six months, and are then
-maintained with releases that fix issues until the next major version is
-released. However, every two years a major release is designated as a long-term
-support (LTS) release, which continues to be maintained for up to two years
-\citep{ape2}. The motivation for LTS releases is to provide longer-term stable
-versions of \astropypkg that users that require a high level of stability can
-make use of if they do not always need the latest features (this could include,
-for example, telescope operation or data reduction pipelines, etc.). The v5.0
-release of the core package was designated as LTS, and since it was released at
-the end of 2021, it will be maintained until the end of 2023.
+\secauthor{Tom Robitaille}
+
+Major versions of the core package -- that is to say versions which add and/or
+modify functionality -- are released approximately every six months, and are
+then maintained with releases that fix issues, until the next major version
+is released. However, every two years a major release is designated as a
+long-term support (LTS) release which continues to be maintained for up to two
+years \citep{ape2}. The motivation for LTS releases is to provide longer-term stable
+versions of astropy that users requiring a high level of stability can make
+use of if they do not always need the latest features (this could include
+for example telescope pipelines and so on). The 5.0 release of the core package was designated as LTS,
+and since it was released at the end of 2021, it will be maintained until the
+end of 2023.
 
@@
 
-% \secauthor{Nathaniel Starkman, Marten van Kerkwijk}
+\secauthor{Nathaniel Starkman, Marten van Kerkwijk}
 
@@
 Within \astropypkg, development may be roughly split into a few
-categories: new features added, intra- and inter-package interoperability,
+categories: new features added, intra and inter-package interoperability,
 improvements to precision, accuracy, and reproducibility, and performance
-enhancements. We discuss these in the subsections below.
+enhancements. We discuss these in turn.
 
@@
 was too specialized was a Virtual Observatory sub-package; the Simple
-Application Message Protocol was kept in the core package
-(\astropysubpkg{samp}), but other parts more properly belonged in
-\package{astroquery} (see Section~\ref{sec:astroquery}).
-
-\subsubsection{New and Planned Features} \label{sec:core-features-new}
+Application Message Protocol made sense to keep (\astropysubpkg{samp}), but
+other parts more properly belonged in \package{astroquery} (see
+Sect.~\ref{sec:astroquery}).
+
+\subsubsection*{New and Planned Features} \label{sec:core-features-new}
 
@@
 require the manipulation of time series.  The new \astropysubpkg{timeseries}
-sub-package extends the \astropyQTable\ class to support tables of data as a
-function of time, where the data can either represent samples or averages over
-particular time bins. The new classes offer a number of special methods to
-manipulate time series (folding, resampling, etc.) and to read different data
-formats (such as \mission{Kepler} light curves, etc.).
+sub-package extends the \astropyQTable\ class to tables of data as a function of
+time, where the data can either represent samples or averages over particular
+time bins. The new classes offer a number of special methods to manipulate time
+series (folding, resampling, etc.) and to read different data formats (such as
+{\em Kepler} lightcurves, etc.).
 
@@
 \href{https://docs.astropy.org/en/stable/timeseries/bls.html}{box-least-squares}
-periodograms.
+periodogrammes.
 
@@
 
-\subsubsection{Interoperability} \label{sec:core-features-interoperability}
+\subsubsection*{Interoperability} \label{sec:core-features-interoperability}
 
@@
 objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord --
-within a \astropyTable\ and write these to various file formats, such as FITS.
+within a \astropyTable and write these to various file formats, such as FITs.
 For objects not already covered by the mixin protocol, functions can be
@@
 \href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
-Summer of Code} 2017 program,\footnote{ To learn more about this project, please
-see the final report
+Summer of Code} 2017 program\footnote{ To learn more about this project, please
+see the final report,
 \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
-mixin protocol for seamless interoperability}.}, the \astropyfits package now
+mixin protocol for seamless interoperability}}, the \astropyfits package now
 interprets time data correctly, using \astropyTime\ as a `mixin' column.  For
@@
 
+% TODO! better transition
 \astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
@@
 
-\subsection{Precision, Accuracy, \& Reproducibility}
+\subsection*{Precision, Accuracy, \& Reproducibility}
 \label{sec:core-features-precision_accuracy_repreducibility}
@@
 Whether for planning observations, crunching pulsar or VLBI data, missing
-seconds meaningfully impact results. In past versions of \astropypkg, when using
-\astropyTime\ in \astropytime leap seconds had to be manually applied; now, leap
-seconds are applied automatically. Moreover, \astropypkg updates internal time
-data-files to ensure that the correct leap second adjustment is always used. For
-reproducibility, e.g., with an old code, \astropyLeapSeconds\ may also be
+seconds meaningfully impact results. Before, when using \astropyTime\ in
+\astropytime leap seconds had to be manually applied; now, leap seconds are
+applied automatically. Moreover, \astropypkg updates internal time data-files to
+ensure that the correct leap second adjustment is always used. For
+reproducibility, e.g. with an old code, \astropyLeapSeconds\ may also be
 manually applied.
@@
 For reproducibility, \astropypkg allows the constants' (and therefore units')
-definitions to be rolled back to prior values, e.g., to the \texttt{SI/CODATA
+definitions to be rolled back to prior values, e.g. to the \texttt{SI/CODATA
 2014} values. For work sensitive to the values of the fundamental constants we
-recommend including an \astropy configuration file with the work, which
+recommend including with the work an \astropy configuration file, which
 specifies the set of constants used.
@@
 \paragraph{Cosmology}
+% TODO! At least one good sentence about "what is the benefit of
+% astropy.cosmology"
 
@@
 
-Since \citet{astropy:2018}, the following cosmology realizations have been
+Since \citealt{astropy:2018} the following cosmology realizations have been
 added:
@@
 central feature of \astropycoordinates is the ability to transform data between
-reference frames, e.g., \astropyICRS\ \citep{ICRS:1997} to \astropyGalacticLSR\
+reference frames, e.g. \astropyICRS\ \citep{ICRS:1997} to \astropyGalacticLSR\
 \citep{GalacticLSR:2010}. However, there have been some historical limitations
 for spatially proximate transformations, impacting the usefulness of
-\astropycoordinates for ground-based telescopes and astrometry within the
+\astropycoordinates for ground-based telescopes and astronometry within the
 solar system. Transformations in the \astropyAltAz\ frame were reasonably
@@
 
-\subsection{Performance} \label{sec:core-features-performance}
+\subsection*{Performance} \label{sec:core-features-performance}
 
@@
 
-Performance in \astropypkg is addressed based on feedback and requests, and
-through contributions from community members: If you find that parts of
-\astropypkg are not performant, we encourage you to open a \github issue to
-start a discussion with the \astropypkg core maintainers.
-
 \section{Major Updates in the Astropy Project} \label{sec:project-updates}
@@
 
-% \secauthor{Erik Tollerud}
+\secauthor{Erik Tollerud}
 
@@
 
-As part of the process of developing \astropypkg into a long-term sustainable
+As part of the process of developing Astropy into a long-term sustainable
 product, and to improve transparency and accountability, the Project agreed to
 write down and formalize our governance structure (partly supported by explicit
-funding for this purpose; see Section~\ref{sec:project-funding}). At the 2019
+funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
 Astropy Coordination Meeting, input was gathered from participants on what governance
 structures existed in the associated Open Source Software communities, and what
-would fit well with the needs of \astropy. This led into a ``retreat'' planned
+would fit well with the needs of Astropy. This led into a ``retreat'' planned
 for March 2020, but due to the COVID-19 pandemic, this became a series of
-virtual meetings of the ``Astropy Governance Working Group.'' This group
+virtual meetings of the ``Astropy Governance Working Group''.  This group
 drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
@@
 structure, so we refer the reader to that document for a more thorough
-description, and here we only highlight some key elements. While many of these
+description.  However here we highlight some key elements. While many of these
 principles were already de facto true or have been discussed organically (and
 have been discussed in earlier papers in this series), the APE0-based governance
-aims to provide a single place where the community can agree as a starting
+aimed to provide a single place where the community can agree as a starting
 point. With this in mind, it highlights the developer and user community of the
@@
 principle of ``do-ocracy'' that those who do work for the Project (be it coding,
-training, or other harder-to-quantify contributions) gain more influence on the
-outputs of the Project by virtue of their effort. However, APE0 adds the
-concept of ``voting members'': a self-governed part of that community who are
-entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
+training, or other less concrete contributions) gain more influence on the
+outputs of the Project by virtue of their effort.  APE0 adds, however, the
+concept of ``voting members'' - a self-governed part of that community who are
+entrusted to elect the Coordination Committee. While this committee has existed
 from the inception of the project, APE0 establishes a formal voting process for
-CoCo members, and explicitly outlines the rights and responsibilities of the
-CoCo. This role is mainly to facilitate consensus and act as the decision maker
-when other mechanisms have failed, and includes powers that either
-require central authority or secrets (e.g., passwords) However, APE0 also
-charges the Committee to devolve responsibilities and seek community input on
-these items as often as possible.
+this committee, and explicitly outlines the rights and responsibilities of the
+Coordination Committee. This role is mainly to facilitate consensus and act as
+the decision maker when other mechanisms have failed.  However, it also includes
+powers that either require central authority or secrets (e.g., passwords), but
+APE0 also charges the Committee to devolve responsibilities and seek community
+input on these items as often as possible.
 
 The first Coordination Committee election under these rules took place in Fall
-2021, electing a mix of prior and new coordination committee members,
+2021, electing a mix of prior existing and new coordination committee members,
 and was contested in the sense of more candidates than available slots. This
@@
     \begin{centering}
-      \includegraphics[width=0.5\textwidth]{commits_committers_per_month.pdf}
-        \caption{
-            todo
-        }
-        \label{fig:contributor-summary:per-month}
+      % \includegraphics{figures/contributor-summary.pdf}
+        \caption{Placeholder figure!}
+        \label{fig:contributor-summary}
     \end{centering}
@@
 
-\begin{figure}
-    \begin{centering}
-      \includegraphics[width=\textwidth]{Ncommitters_vs_frac_commits.pdf}
-        \caption{
-            todo
-        }
-        \label{fig:contributor-summary:Ncommitters}
-    \end{centering}
-\end{figure}
-
-\begin{figure}
-    \begin{centering}
-      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
-        \caption{
-            todo
-        }
-        \label{fig:contributor-summary:top}
-    \end{centering}
-\end{figure}
-
 % Transition of needs: From new features, to community development, sustainability, ...
@@
 
-% \secauthor{Lía Corrales}
+\secauthor{Lía Corrales}
 
@@
 Github. This process was deemed the most ``open'' because it allowed
-for community-wide feedback to focus and improve proposal
-initiatives. Two highlighted programs that were selected and implemented via
-this process are described below.
-
-% \textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}
+for community wide feedback to focus and improve proposal
+initiatives. Two programs were selected via this process, described
+below.
+
+\textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}
 
@@
 \textbf{\astropy representation at national diversity conferences:}
-Several members of the \astropy community attended virtual conferences of the
-National Society of Black Physicists (NSBP) and SACNAS (focusing on Hispanic and
-Native American scientists across all STEM fields). \astropy representatives
-noted that one underlying topic came up multiple times. A major barrier for
-persons who come from underrepresented communities is the lack of resources and
-expertise that are necessary to train students. Writing PEP 8 compliant software
-and understanding the Git and \github workflow is not part of the standard
-Physics and Astrophysics curriculum. Additionally, the larger astronomical
-community is still in the process of transitioning towards \python-based tools,
-making it difficult for students not currently under advisement by some one with
-extensive \python expertise to get involved with the open source \python
-community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
-have a profound impact by empowering underrepresented groups because it provides
-a free, searchable, and accessible introduction to \python tools for
-astrophysical research. Representatives also noted that offering training and
-teaching materials to PIs of Research Experience for Undergraduate (REU)
-programs throughout the United States could be helpful for encouraging advisors
-to teach students how to use Astropy, create their own libraries, and use
-version control on open source platforms like \github. Such materials could also
-be offered as a workshop at national diversity conferences themselves, as SACNAS
-solicits special session proposals each year.
-
-% \textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
-%   initiative:} \todo{No author assigned}{Describe more about this.}
-
-\subsection{Project funding} \label{sec:project-funding}
-
-% \secauthor{Aarya Patil}
-
-% Summarize funding sources (Moore, NASA) and amounts and what this has been
-% used for.
-
-With the goal of establishing a sustainable financial model for the Astropy
-Project, the Coordination Committee, under the new governance charter (see
-Section~\ref{sec:project-governance}), established a standing Finance Committee.
-This committee oversees the planning and allocation of finances on behalf
-of the Project. The Astropy Project accepts funds from institutions as well as
-individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
-supports and promotes world-class, innovative, open source scientific computing.}
-Previously, no direct financial support was available for the project development,
-and NumFOCUS covered most of the incurred operational costs. However, transitioning
-to long-term financial stability meant having an influx of funds in the form of
-institutional support, and hence the Astropy Project applied for and successfully
-acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
-Project, providing a ``neutral'' space not tied to any particular astronomy
-institution. The two major grants are briefly summarized below.
-
-\begin{description}
-    \item[Moore] The Gordon and Betty Moore Foundation awarded the Astropy
-    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
-    Growing the Astropy Project.'' Over a three-year term, this grant supported
-    sustenance of the Project by providing funds for (1) the development and
-    maintenance of the \astropypkg and its infrastructure, provided to existing
-    project members and targeted hires, (2) the transition of long-term users to
-    contributors and maintainers through mentorship efforts, (3) the
-    formalization of a governance structure (see
-    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
-    diversity, and inclusion efforts in the project (see
-    Section~\ref{sec:project-ide}). Some of this money supported travel for
-    meetings, conferences, and workshops. This generous grant helped reduce
-    reliance on volunteer-driven maintenance of the openly-developed package and
-    paved the way to acquire funding from federal agencies, representing a major
-    milestone for the success of the project.
-
-    \item[NASA] A successful proposal to the NASA ROSES-2020
-    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
-    (Support for Open Source Tools, Frameworks, and Libraries), granted the
-    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
-    term of three years includes (1) project-wide infrastructure maintenance and
-    improvements, (2) targeted work on areas of the \astropy core and
-    coordinated packages, (3) enhancements to the Learn
-    Astropy ecosystem and educational materials (see Section~\ref{sec:learn}),
-    and (4) support for the \astropy affiliated packages. These goals
-    support the sustenance of the project by encouraging further
-    development of the package ecosystem and community engagement.
-\end{description}
-
-In addition to the two grants described above, the project benefited from
-a Gemini Observatory contract under its Science User Support Department,
-which awarded funds for development work that supports both the Astropy and
-DRAGONS projects. The project also received financial support from the Dunlap
-Seed Funding program at the University of Toronto to develop educational
-resources for scientific software packaging within the Learn Astropy
-framework. All of these successful proposals show that the Astropy Project
-is vital for the astronomical infrastructure and community at large, and
-strengthen its promising and sustainable future.
-
-% ADD text for the new finance APE once it has a doi
+Several members of the \astropy community attended virtual conferences
+of the National Society of Black Physicists (NSBP) and SACNAS
+(focusing on Hispanic and Native American scientists across all STEM
+fields). \astropy representatives noted that one underlying topic came
+up multiple times. A major barrier for persons who come from
+underrepresented communities is the lack of resources and expertise
+that are necessary to train students. Writing PEP 8 compliant software
+and understanding the Git and Github workflow is not part of the
+standard Physics and Astrophysics curriculum. Additionally, the larger
+Astronomnical community is still in the process of transitioning
+towards Python based tools, making it difficult for students not
+currently under advisement by some one with extensive Python expertise
+to get involved with the open source Python community. Projects such
+as Learn Astropy could have a profound impact by empowering
+underrepresented groups because it provides a free, searchable, and
+accessible introduction to Python tools for astrophysical
+research. Representatives also noted that offering training and
+teaching materials to PIs of Research Experience for Undergraduate
+(REU) programs throughout the United States could be helpful for
+encouraging advisors to teach students how to use Astropy, create
+their own libraries, and use version control on open source platforms
+like Github. Such materials could also be offered as a workshop at
+national diversity conferences themselves, as SACNAS solicits special
+session proposals each year.
+
+\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
+  initiative:} \todo{No author assigned}{Describe more about this.}
+
+\subsection{Funding} \label{sec:project-funding}
+
+\secauthor{Aarya Patil}
+
+Summarize funding sources (Moore, NASA) and amounts and what this has been used
+for.
 
@@
 
-% \secauthor{Nicholas Earl}
+\secauthor{Nicholas Earl}
 
 The Astropy project supports the broader ecosystem by providing pre-configured
-infrastructure packages that the community can use to support and maintain their
-own software package infrastructure. These include tools and extensions that
-enable easily generating documentation and setting up automated testing, as well
-as providing package scaffolding for new projects.
+infrastructure packages that the community can use to support and maintain
+their own software package infrastructure. These include tools to easily
+generate documentation and setup automated testing, as well as provide
+package scaffolding for new projects.
 
@@
 packages. The Astropy project maintains a default Sphinx configuration along
-with Astropy-specific extensions that can be easily added to community
+with Astropy-specific extensions which can be easily added to community
 projects via the
@@
 (\href{https://github.com/pytest-dev/pytest-cov}{\texttt{pytest-cov}}), and
-configuring packages for property-based testing
+configuring package for property-based testing
 (\href{https://github.com/HypothesisWorks/hypothesis}{\texttt{hypothesis}}).
@@
 package template generation process includes the ability to setup
-interoperability with \github, allowing for easy repository access from
-documentation, as well as an example \github Actions workflow to demonstrate
-the use of \github's continuous integration tooling.
-
-\subsection{Astropy affiliated packages}
-
-% \secauthor{Matt Craig, Brett Morris}
-
-% Highlight a few new affiliated packages and major updates to existing ones.
-% Include a big table of all affiliated packages and references (as in v2.0
-% paper).
-
-As defined in \paperii, Astropy affiliated packages are astronomy-related
-\python packages that provide functionality that builds upon and extends the
-\astropypkg core package that have requested to be included as part of the
-Astropy Project community. These packages support the goals and vision of
-Astropy of improving code re-use, interoperability, and embracing good coding
-practices such as testing and thorough documentation.
-All Astropy-coordinated and other affiliated packages are listed with detailed
-information on the \href{{https://www.astropy.org/affiliated}}{\astropy
-website}; A table summarizing key aspects of the affiliated packages is included
-in the Appendix below (Table~\ref{tab:affiliated-registry}).
-
-Since \paperii, there has been an expansion of affiliated
-packages for gravitational astrophysics, including:
-\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
-modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
-gravitational wave probability maps, and \texttt{EinsteinPy} for general
-relativity and gravitational astronomy. There have also been several packages
-added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
-for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
-Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
-and \texttt{python-cpl} has been added for ESO pipelines and VLT data products.
-For ground-based astronomy, \texttt{baseband} has added IO capabilities for
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
-Other new affiliated packages include \texttt{agnpy} for AGN jets,
-\texttt{statmorph} for fitting galactic morphological diagnostics,
-\texttt{dust\_extinction} for modeling interstellar dust extinction,
-\texttt{feets} for extracting features from time series data, and
-\texttt{corral} for managing data intensive parallel pipelines.
-\texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
-\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
-and \texttt{sbpy} enables calculations for asteroid and
-cometary astrophysics. Finally, \texttt{synphot} provides an interface for
-synthetic photometry.
-% \citep{Mommert2019}
-
-Several of the coordinated and affiliated packages described in
-\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
-\citep{astroquery} has added access to roughly a dozen new missions and data
-services, including the James Webb Space Telescope (JWST) archive, and the
-project has switched to a continuous release model: Every time a change is
-committed to the main development branch it is published on the Python Package
-Index (PyPI) and available for installation. (However, formal releases are still
-done a few times per year.) \texttt{photutils} \cite {photutils} released its
-first stable version, indicating that the API will change less frequently, and
-there have been several significant performance improvements. \texttt{ccdproc}
-\cite{ccdproc} also released a new major version, bringing better performance to
-some image combination operations. The \texttt{regions} package
-\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
-added new ways to manipulate regions and introduced new region types.
-\texttt{reproject}'s \cite{reproject} major new feature is a function to align
-and co-add images to create a mosaic; better support for parallelization was
-also added. \texttt{specutils}, the package that defines containers for 1D and
-2D spectra \cite{specutils}, also had its first stable release and the addition
-of classes to read JWST data. \texttt{stingray} has also had a major performance
-overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
-improvements and has unified its API in preparation for its first stable
-release.
+interoperability with GitHub, allowing for easy repository access from
+documentation, as well as an example GitHub Actions workflow to demonstrate
+the use of GitHub's continuous integration tooling.
+
+\subsection{Affiliated packages}
+
+\secauthor{Matt Craig, Brett Morris}
+
+Highlight a few new affiliated packages and major updates to existing ones.
+Include a big table of all affiliated packages and references (as in v2.0
+paper).
 
 \subsection{Connections with data archives}
+
 \label{sec:astroquery}
-
-% \secauthor{Adam Ginsburg}
-
-\texttt{astroquery} \citep{Ginsburg2019} is the Astropy-coordinated package for
-interacting with online archives of astronomical and related data. It contains
-over 50 modules for querying astronomical databases, large and small. In
-particular, since \paperii, significant contributions to \texttt{astroquery}
-have come from several of the major archives, including the European Space
-Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space
-Telescope Science Institute (STScI), the Infrared Science Archive at the NASA
-Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian
-Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter
-Array archive (ALMA). These contributions represent a formal acknowledgement of
-the utility of a centralized tool suite for archive interaction from \python.
-The widespread usage of \texttt{astroquery} is apparent from the range of
-keywords represented in the journal articles that cite the \texttt{astroquery}
-paper \citep{Ginsburg2019}; everything from asteroids to galaxies is
-represented. Additionally \texttt{astroquery} functionality has been built into
-several special-purpose \python packages, such as \texttt{LightKurve}
-\citep{LightKurve} and SORA \citep{SORA}]. More than 4,000 repositories on
-\github make use of \texttt{astroquery} in some way, and \texttt{astroquery}
-features in a large number of tutorials on various facets of astronomical
-analysis.
-
-Many of the existing and newly-contributed tools rely on Virtual Observatory
-(VO) tools. These use the underlying package \texttt{pyvo}, which has also
-recently become an Astropy-coordinated package. While many or all of the
-functions provided in \texttt{astroquery} can be achieved through direct use of
-VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more
-closely resemble the web interfaces that are more familiar to most users.
+% MHvK: feel free to move above label to where it makes most sense.
+
+\secauthor{Adam Ginsburg}
+
+Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
+It contains over 50 modules for querying astronomical databases, large and small.
+In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
+These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
+
+The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented.
+Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}].
+More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.
+
+Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
+These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
+While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
+% note to coauthors: we need a section on pyvo
 
@@
 
-In this section, we highlight a few efforts in observatory- or mission-driven
-development that have contributed to Astropy, and vice versa.
+Brief summary of efforts in observatory- or mission-driven development that have
+contributed to Astropy, and vice versa.
 
 \subsubsection{James Webb Space Telescope}
-% \secauthor{Larry Bradley}
+\secauthor{Larry Bradley}
 
@@
 infrared telescope that will provide unprecedented resolution and
-sensitivity from 0.6--28 microns. JWST will enable a broad range
+sensitivity from 0.6 -- 28 microns. JWST will enable a broad range
 of scientific investigations from exoplanets and their atmospheres
@@
 Software developers at the Space Telescope Science Institute (STScI),
-the operations center of JWST, have been developing \python-based tools
+the operations center of JWST, have been developing Python-based tools
 for JWST since 2010 (starting with the JWST Calibration Reference Data
@@
 
-% Gemini
-% \secunfilled
-
-\subsubsection{Cherenkov Telescope Array}
-% \secauthor{Axel Donath, Maximilian Nöthe}
+Gemini
+\secunfilled
+
+Cherenkov Telescope Array
+\secauthor{Axel Donath, Maximilian Nöthe}
 
@@
 of Cherenkov telescopes to the reconstructed event lists.
-The high-level analysis (or CTA science tools) will be based on the Astropy affiliated package
+
+The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
 Gammapy~\citep{gammapy}.
-Both Gammapy and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
-tables and FITS I/O.
+
+Both Gammapy and \texttt{ctapipe} are using astropy heavily, mainly for units, times, coordinate transformations,
+tables and FITS IO.
+
 As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
@@
 based on earlier work by B.~Winkel.
+
 Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
@@
 
-% Rubin Observatory
-% \secunfilled
-
-% LIGO/Virgo/KAGRA
-% \secauthor{Leo Singer}
-
-\subsection{Learn Astropy} \label{sec:learn}
+Rubin Observatory
+\secunfilled
+
+LIGO/Virgo/KAGRA
+\secauthor{Leo Singer}
+\section{Future Plans for the Astropy Project} \label{sec:future}
+
+\subsection{Roadmap}
+
+\secauthor{Clara Brasseur}
+
+The Astropy Roadmap\footnote{\url{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}} is a document in the astropy-project repository that captures high level actionable items that the Astropy project aims to undertake to improve the health and stability of the  project.
+It is a static document that is revisited regularly at the Astropy coordination meetings, to keep track of progress and write new versions as needed.
+There is a related project board linked to the Roadmap document, that holds specific issues and efforts related to Roadmap items.
+The project board is a living document that is continually updated as work is planned, assigned, and completed.
+
+All items in the Roadmap have been agreed to be priorities for the Astropy Project, and are color-coded based on resources (both time/effort and developers/expertise) needed to complete the item.
+Green items are well underway, have sufficient resources/support and a plan in place for completion.
+Orange items are well defined, and work for aquiring sufficient resources underway.
+Red items do not yet have a plan for implementation and need more resources.
+\todo{addition}{Add some examples of high priority items from the roadmap in a new paragraph here}
+The Astropy Roadmap originates from the March 2021 Astropy Coordination meeting where it was first drafted before being handed off to a newly formed Astropy Roadmap working group for completion.
+The Roadmap in its currant form was adopted via pull request in December 2021.
+
+\section{Learn Astropy} \label{sec:learn}
 
@@
 
-% \subsubsection{Current status and scope}
+\subsubsection{Current status and scope}
 
@@
 
-% \subsubsection{Learn vision for the future}
+\subsubsection{Learn vision for the future}
 
@@
 
-\section{Future Plans for the Astropy Project} \label{sec:future}
-
-\subsection{A roadmap for future priorities}
-% \secauthor{Clara Brasseur}
-
-The
-\href{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}{Astropy Roadmap} is a document in the \texttt{astropy-project} \github
-repository within the \astropy organization that captures high level actionable
-items that the Astropy project aims to undertake to improve the health and
-stability of the project. It is a static document that is revisited regularly at
-the Astropy coordination meetings, intended to keep track of progress and write
-new versions as needed. There is a related project board linked to the Roadmap
-document that holds specific issues and efforts related to Roadmap items. The
-project board is a living document that is continually updated as work is
-planned, assigned, and completed.
-
-All items in the Roadmap have been agreed to be priorities for the Astropy
-Project, and are color-coded based on resources (both time/effort and
-developers/expertise) needed to complete the item. Green items are well
-underway, have sufficient resources/support and a plan in place for completion;
-For example, efforts to improve the discoverability of documentation and
-educational materials by overhauling the Learn website, which is underway (see
-Section~\ref{sec:learn}).
-Orange items are well defined, and work for acquiring sufficient resources
-underway; For example, the goal of providing next-generation spectroscopic
-reduction, analysis, and visualization tools usable by individual researchers
-and larger surveys.
-Red items do not yet have a plan for implementation and need more resources; For
-example, the goal of improving interoperability with performant I/O file formats
-and libraries such as HDF5 and Dask.
-The Astropy Roadmap originates from the March 2021 Astropy Coordination meeting
-where it was first drafted before being handed off to a newly formed Astropy
-Roadmap working group for completion. The Roadmap in its current form was
-adopted via pull request in December 2021.
-
 \subsection{Current and Future Challenges}
 
-Community Engagement
+\section{Community Engagement}
 
@@
 
-We would like to thank the members of the community who have contributed to
-\astropy, who have opened issues and provided feedback, and have supported the
+We would like to thank the members of the community that have contributed to
+\astropy, that have opened issues and provided feedback, and have supported the
 project in a number of different ways.
 
-The \astropy community is supported by and makes use of a number of
-organizations and services outside the traditional academic community.
-We thank Google for financing and organizing the
+The \astropy community is supported by and makes use
+of a number of organizations and services outside the traditional
+academic community. We thank Google for financing and organizing the
 Google Summer of Code (GSoC) program, that has funded severals
 students per year to work on \astropy related projects over the
-summer. These students often turn into long-term contributors.
-We also thank NumFOCUS and the Python Software Foundation for financial
-support.
-Within the academic community, we thank institutions that make it possible that
-astronomers and other developers on their staff can contribute their time to the
-development of \astropy projects.
-We would like acknowledge the support of the Space Telescope Science Institute,
-Harvard--Smithsonian Center for Astrophysics, and the South African Astronomical
-Observatory.
-
-Supported by the international Gemini Observatory, a program of NSF's NOIRLab,
-which is managed by the Association of Universities for Research in Astronomy
-(AURA) under a cooperative agreement with the National Science Foundation, on
-behalf of the Gemini partnership of Argentina, Brazil, Canada, Chile, the
-Republic of Korea, and the United States of America.
+summer. These students often turn into long-term contributors. We also
+thank NumFOCUS and the Python Software Foundation for financial
+support. Within the academic community, we thank
+institutions that make it possible that astronomers and other developers on
+their staff can contribute their time to the development of
+\astropy projects.  We would like acknowledge the support of the
+Space Telescope Science Institute, Harvard--Smithsonian Center for Astrophysics,
+and the South African Astronomical Observatory.
 
@@
 a number of web services for code hosting, continuous integration, and
-documentation; in particular, \astropy heavily relies on GitHub, Azure
-Pipelines, CircleCI, and Read the Docs.
-
-This research has made use of NASA's Astrophysics Data System.
+documentation; in particular, \astropy heavily relies on GitHub, Travis CI,
+Appveyor, CircleCI, and Read the Docs.
 
@@
 \bibliographystyle{aasjournal}
-\bibliography{refs, static/affiliated-refs}
-
-\appendix
-
-\section{List of Affiliated Packages}
-
-\begin{longrotatetable}
-    \begin{deluxetable*}{ccp{2in}p{2.5in}}
-    \tablecaption{Registry of affiliated packages.}
-    \label{tab:affiliated-registry}
-    \tablehead{
-        \colhead{Package Name} &
+\bibliography{refs}
+
+% \appendix
+
+% \section{List of Affiliated Packages}
+
+% \begin{longrotatetable}
+%     \begin{deluxetable*}{cccp{3in}c}
+%     \tablecaption{Registry of affiliated packages.}
+%     \label{tab:registry}
+%     \tablehead{
+%         \colhead{Package Name} &
 %         \colhead{Stable} &
-        \colhead{PyPI Name} &
-        \colhead{Maintainer(s)} &
-        \colhead{Citation(s)}
-      }
-      \startdata
-        \input{generated/affiliated-table.tex}
-      \enddata
-  \end{deluxetable*}
-\end{longrotatetable}
+%         \colhead{PyPI Name} &
+%         \colhead{Maintainer} &
+%         \colhead{Citation}
+%       }
+%       \startdata
+%         \input{registry.tex}
+%       \enddata
+%   \end{deluxetable*}
+% \end{longrotatetable}
 

### 2022-03-11 src/ms.tex

@@
 
+All coordinated and affiliated packages are listed on the astropy
+website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
+\cite{astropy:2018}, there has been an expansion of affiliated
+packages for gravitational astrophysics including:
+\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
+modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
+gravitational wave probability maps, and \texttt{EinsteinPy} for general
+relativity and gravitational astronomy. There have been several packages
+added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
+for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
+Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
+and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
+For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
+VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
+We have \texttt{agnpy} for AGN jets, and \texttt{statmorph} for fitting galactic
+morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
+routines. For modeling interstellar dust extinction we have 
+\texttt{dust\_extinction}, and for extracting features from time series we
+have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
+pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
+inference, and \texttt{sbpy} \cite{Mommert2019} does calculations for asteroid and cometary
+astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
+
+Several of the coordinated and affiliated packages described in \cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
+\cite{astroquery} has added
+access to roughly a dozen new missions and data services, including JWST, and
+the project has switched to a continuous release model. Every time a change
+is committed to the main development branch it is published on the Python
+Package Index (PyPI, ref) and available for installation. Formal releases are
+still done a couple of times per year. \texttt{photutils} \cite
+{photutils} released its first stable version, indicating that the API will
+change less frequently, and there have been several significant performance
+improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
+version, bringing better performance to some image combination operations.
+The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
+region definitions \cite{ds9}, added new ways to manipulate regions and
+introduced new region types. \texttt{reproject}’s \cite{reproject} major new
+feature is a function to align and co-add images to create a mosaic; better
+support for parallelization was also added. \texttt{specutils}, the package
+that defines containers for 1D and 2D spectra \cite{specutils}, also had its
+first stable release and the addition of classes to read JWST data.
+\texttt{stingray} has also had a major performance overhaul. The package
+\texttt{gammapy} \cite{gammapy} also has major performance improvements,
+has unified its API in preparation for its first stable release.
+
 \subsection{Connections with data archives}

### 2022-03-11 src/ms.tex
words about top contributors

--- 2022-03-11 src/ms.tex
@@
 
-\begin{figure}[ht]
+\begin{figure}[t!]
     \begin{centering}
@@
 
-Since its inception, the most active contributors to the \astropypkg package
-have spanned a range of career stages and career paths.
-
-\begin{figure}[ht]
+\begin{figure}[t!]
     \begin{centering}
@@
 
-\begin{figure}[ht]
+Since its inception, the most active contributors to the \astropypkg package
+have spanned a range of career stages and career paths.
+For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
+career stage/path at the midpoint of the three time eras defined in the previous
+paragraph for all contributors with more than 50 commits to the \astropypkg
+package within each era.
+Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
+contributors moved between these categories (i.e., have graduated as students to
+become research staff / faculty or software engineers), but also a number of
+early contributors moved on to other positions and stopped contributing to the
+core \astropypkg package.
+Given this, it is encouraging that the numbers in each career stage/path have
+stayed relatively constant over the history of \astropy (with some showing
+growth), as new, active contributors have generally joined and stepped in to
+fill top contributor roles within the project.
+On the other hand, the total number of top contributors has not grown
+substantially, reflecting one of our principal challenges: Sustaining a pipeline
+from users to contributors to maintainers of the package.
+One other thing to note in this context is that in all eras, 50--60\% of the
+software engineers who contributed to \astropypkg were (or are) employed by
+STScI, whose leadership have consistently provided resources (financial and
+personnel) towards sustaining \astropypkg.
+
+\begin{figure}[t!]
     \begin{centering}
-      \includegraphics[width=0.5\textwidth]{commits_committers_per_month.pdf}
+      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
         \caption{
@@
 \end{figure}
+
+A final metric we consider ... rate of contributions and contributor participation ...
 

### 2022-03-11 src/ms.tex
caption and figure title

@@
         \caption{
-            todo
+            The cumulative fraction of Git commits to the \astropypkg core
+            package as a function of number of contributors (committers) broken
+            into three equal-length time periods between the start of \astropy
+            and the feature freeze date of v5.0.
+            The time eras early, mid, and recent correspond to the approximate
+            year ranges 2011--2015, 2015--2018, 2018--2021 (defined more
+            precisely in Section~\ref{sec:project-contributors}).
+            In all eras, the majority of commits to the \astropypkg core package
+            have come from a small subset of the total number of contributors.
+            For example, as annotated in the figure, 90\% of the Git commits
+            have come from 18, 34, and 44 committers in the early, mid, and
+            recent eras, respectively.
         }

### 2022-03-11 src/figures/contributor-summary.py

@@ -353,13 +353,17 @@ width = 0.25
 for i, (name, _data) in enumerate(top_contr_data.items()):
     ax.bar(ind + i*width, _data, width, label=name, color=era_colors[i], alpha=0.8)
 
-ax.set_title('Number of contributors with >50 commits per era')
+ax.set_title('Career stage/position of top contributors\nto the Astropy core package', pad=10)
+ax.set_ylabel('Number of contributors\nwith >50 commits per era')
 
 ax.set_xticks(ind + width)
-ax.set_xticklabels(labels)
+ax.set_xticklabels(labels, fontsize=18)
 ax.legend(loc='upper right', fontsize=18,
           title='Era:', title_fontsize=19)
 
 ax.set_ylim(0, 15)
 
-fig.savefig('top_contributors_career.pdf')
+fig.savefig('top_contributors_career.pdf', bbox_inches='tight')
+# -
+
+

### 2022-03-11 src/figures/contributor-summary.py
titles

@@ -187,6 +187,10 @@ ax.set_ylabel('committers per month')
 fig.savefig('commits_committers_per_month.pdf')
 # -
 
+np.mean(commits_per_month[-20:])
+
+np.mean(committers_per_month[-20:])
+
 # ## Number of commits x number of committers
 
 full_shortlog = repo.git.shortlog(
@@ -316,6 +320,8 @@ ax.axhline(0.9, zorder=-10, color='#aaaaaa', ls='--')
 
 ax.set_xlabel('Number of committers')
 ax.set_ylabel('Cumulative fraction of commits\nin each time period')
+ax.set_title('Cumulative fraction of commits\nto the Astropy core package', 
+             pad=10)
 
 fig.savefig('Ncommitters_vs_frac_commits.pdf', bbox_inches='tight')
 # -

### 2022-03-11 src/ms.tex
captions

@@
     \begin{centering}
-      \includegraphics[width=0.8\textwidth]{Ncommitters_vs_frac_commits.pdf}
+      \includegraphics[width=0.9\textwidth]{Ncommitters_vs_frac_commits.pdf}
         \caption{
@@
         \caption{
-            todo
+            The career stages and/or positions of the top contributors within
+            each time era (defined as having contributed $>50$ commits to the
+            \astropypkg core package in each era).
+            The time eras early, mid, and recent correspond to the approximate
+            year ranges 2011--2015, 2015--2018, 2018--2021 (defined more
+            precisely in Section~\ref{sec:project-contributors}).
         }
@@
         \caption{
-            todo
+            \textbf{Top panel:} The number of Git commits per month (defined as
+            consecutive 30 day windows) merged into to the \astropypkg core
+            package over the whole history of the project (up to end of 2021,
+            v5.0).
+            The number clearly peaks just before feature freezes (vertical
+            lines) as a flurry of activity leading up to each release, but in
+            recent years the mean rate has remained steady at about $\sim$300
+            commits per month.
+            \textbf{Bottom panel:} The number of unique Git committers who
+            contribute to the \astropypkg core package each month (again defined
+            as consecutive 30 day windows).
+            While this also shows some variation depending on proximity to
+            feature freezes, in recent years the mean rate of unique
+            contributors has remained consistent at about $\sim$22 committers
+            per month.
         }

### 2022-03-11 src/figures/contributor-summary.py
another title

@@ -184,7 +184,10 @@ ax.set_xlim(month_bins[0].datetime, month_bins[-1].datetime)
 ax.set_xlabel('date')
 ax.set_ylabel('committers per month')
 
-fig.savefig('commits_committers_per_month.pdf')
+axes[0].set_title("Monthly rate of contributions and contributors\n"
+                  "to the Astropy core package", pad=10)
+
+fig.savefig('commits_committers_per_month.pdf', bbox_inches='tight')
 # -
 
 np.mean(commits_per_month[-20:])

### 2022-03-11 src/figures/contributor-summary.py
windowed

@@ -24,6 +24,7 @@ import astropy.units as u
 from git import Repo
 import matplotlib.pyplot as plt
 import numpy as np
+from scipy.ndimage import generic_filter
 # -
 
 astropy_repo_path = pathlib.Path(
@@ -146,12 +147,28 @@ fig, axes = plt.subplots(2, 1, figsize=(8, 8),
                          constrained_layout=True)
 
 ax = axes[0]
+# ax.plot(
+#     month_bins[:-1].datetime,
+#     commits_per_month,
+#     drawstyle='steps',
+#     marker='',
+#     **style
+# )
 ax.plot(
     month_bins[:-1].datetime,
     commits_per_month,
     drawstyle='steps',
     marker='',
-    **style
+    zorder=-10, color='#aaaaaa', lw=1,
+    label='raw monthly'
+)
+ax.plot(
+    month_bins[:-1].datetime,
+    generic_filter(commits_per_month, function=np.mean, size=12),
+    drawstyle='steps',
+    marker='',
+    label='monthly rate, 1 year window',
+    color='k', lw=2
 )
 
 _label = True
@@ -161,25 +178,42 @@ for _date in freeze_dates.datetime:
         _label = False
     else:
         kw = dict()
-    ax.axvline(_date, color='tab:blue', alpha=0.2, zorder=-10, marker='',
+    ax.axvline(_date, color='tab:blue', 
+               alpha=0.2, zorder=-10, marker='', ls='--', lw=1,
                **kw)
 
 ax.set_xlim(month_bins[0].datetime, month_bins[-1].datetime)
 # ax.set_xlabel('date')
 ax.set_ylabel('commits per month')
 
-ax.legend(loc='upper left', fontsize=18)
+ax.legend(loc='upper left', fontsize=15)
 
 # ---
 
 ax = axes[1]
+# ax.plot(
+#     month_bins[:-1].datetime,
+#     committers_per_month,
+#     drawstyle='steps',
+#     marker='',
+#     **style
+# )
 ax.plot(
     month_bins[:-1].datetime,
     committers_per_month,
     drawstyle='steps',
     marker='',
-    **style
+    zorder=-10, color='#aaaaaa', lw=1
 )
+ax.plot(
+    month_bins[:-1].datetime,
+    generic_filter(committers_per_month, function=np.mean, size=12),
+    drawstyle='steps',
+    marker='',
+    color='k', lw=2
+)
+
+
[…]

### 2022-03-11 src/ms.tex
final paragraph on rates

@@
             v5.0).
-            The number clearly peaks just before feature freezes (vertical
+            The dark line shows the mean rate using a 1 year rolling window, but
+            the under-plotted gray line shows the raw monthly rate.
+            The raw number clearly peaks just before feature freezes (vertical
             lines) as a flurry of activity leading up to each release, but in
@@
             commits per month.
-            \textbf{Bottom panel:} The number of unique Git committers who
-            contribute to the \astropypkg core package each month (again defined
-            as consecutive 30 day windows).
+            \textbf{Bottom panel:} Similar to above, but now showing the number
+            of unique Git committers who contribute to the \astropypkg core
+            package each month (again defined as consecutive 30 day windows).
             While this also shows some variation depending on proximity to
@@
 
-A final metric we consider ... rate of contributions and contributor participation ...
+A final metric we consider as a diagnostic of the contributor base and
+contributions to the \astropypkg core package is the rate of contributions and
+contributor participation, defined again using Git commits to quantify
+contributions.
+Figure~\ref{fig:contributor-summary:per-month} shows the monthly rates of
+commits and unique committers to the \astropypkg core package over the history
+of the package.
+After an initial ramp-up in the commit rate and number of committers (lasting
+roughly from 2011 to 2014), the commit rate and number of unique committers have
+both remained fairly constant.
 

### 2022-03-11 src/figures/contributor-summary.py
oops - get rid of APW-specific path
oops also clone astropy

parameter name

paths

paths

fetch depth of astropy

@@ -17,6 +17,7 @@
 # +
 from collections import defaultdict
 import datetime
+import os
 import pathlib
 
 from astropy.time import Time
@@ -27,8 +28,8 @@ import numpy as np
 from scipy.ndimage import generic_filter
 # -
 
-astropy_repo_path = pathlib.Path(
-    '/Users/apricewhelan/projects/astropy-all/astropy')
+_astropy_path = os.environ.get('ASTROPY_REPO_PATH', '../../../astropy')
+astropy_repo_path = pathlib.Path(_astropy_path)
 
 repo = Repo(astropy_repo_path)
 

### 2022-03-12 src/ms.tex
add scipy as dependency
First draft of the Finance Section (#31)
* First draft of funding section

* Fixed minor issue with a reference

* Newline added

* Rview comments addresssed

* added reference

--- 2022-03-11 src/ms.tex
@@
 
-\subsection{Core package contributor base} \label{sec:project-contributors}
+\subsection{Contributor base} \label{sec:project-contributors}
 
@@
 
-% TODO: move this subsection to "updates to core package" section above?
-
-\begin{figure}[t!]
+Overview and statistics of contributors. Highlight changes since v2.0.
+
+\begin{figure}
     \begin{centering}
-      \includegraphics[width=0.8\textwidth]{Ncommitters_vs_frac_commits.pdf}
-        \caption{
-            The cumulative fraction of Git commits to the \astropypkg core
-            package as a function of number of contributors (committers) broken
-            into three equal-length time periods between the start of \astropy
-            and the feature freeze date of v5.0.
-            The time eras early, mid, and recent correspond to the approximate
-            year ranges 2011--2015, 2015--2018, 2018--2021 (defined more
-            precisely in Section~\ref{sec:project-contributors}).
-            In all eras, the majority of commits to the \astropypkg core package
-            have come from a small subset of the total number of contributors.
-            For example, as annotated in the figure, 90\% of the Git commits
-            have come from 18, 34, and 44 committers in the early, mid, and
-            recent eras, respectively.
-        }
-        \label{fig:contributor-summary:Ncommitters}
+      % \includegraphics{figures/contributor-summary.pdf}
+        \caption{Placeholder figure!}
+        \label{fig:contributor-summary}
     \end{centering}
@@
 
-As described in the introduction above, the \astropy Project is broader than the
-core \astropypkg package as it also represents a number of core infrastructure
-packages, affiliated packages, educational initiatives, and user communities.
-In total, well over 1500 people have contributed in some way to the Project,
-whether, for example, through contributing code or documentation, opening issues
-on \github, providing infrastructure support, writing tutorials, or helping to
-coordinate workshops that relate to Astropy.
-The vast majority of these contributors and participants have interacted as
-volunteers choosing to engage with the open-source and open-development software
-community.
-However, while the total number of contributors is large, the number of
-contributions and bulk of the maintenance of the \astropy Project components has
-primarily been sustained by a small number of heavily-engaged or
-explicitly-employed contributors \citep[see, e.g., ][]{astropy:2018}, as is the
-case in many large, open-source software communities \citep[e.g.,][]{numpy}.
-
-As an example of this within the \astropy Project, here we focus on
-contributions made to the \astropypkg core package.
-Figure~\ref{fig:contributor-summary:Ncommitters} shows the number of
-contributors who have contributed a given fraction of the total number of Git
-commits, separated into three separate time spans (``eras'') in the history of
-the Project.\footnote{Note, however, that measuring these statistics in terms of
-Git commits is not a perfect way of encapsulating ``effort,'' but we use this
-metric for its simplicity and clear definition.}
-The annotations show that, within the early (July 2011--January 2015), mid
-(January 2015--June 2018), and recent (June 2018--November 2021) eras, 90\% of
-the commits to the \astropypkg core package have come from 18, 34, and 44
-contributors.
-While the trend is encouraging --- a larger pool of contributors are
-contributing more, fractionally --- this emphasizes that, still, the vast
-majority of the work in \astropypkg is done my a small subset ($<20\%$) of the
-115, 216, and 234 contributors, respectively, within each era.
-
-\begin{figure}[t!]
-    \begin{centering}
-      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
-        \caption{
-            todo
-        }
-        \label{fig:contributor-summary:top}
-    \end{centering}
-\end{figure}
-
-Since its inception, the most active contributors to the \astropypkg package
-have spanned a range of career stages and career paths.
-For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
-career stage/path at the midpoint of the three time eras defined in the previous
-paragraph for all contributors with more than 50 commits to the \astropypkg
-package within each era.
-Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
-contributors moved between these categories (i.e., have graduated as students to
-become research staff / faculty or software engineers), but also a number of
-early contributors moved on to other positions and stopped contributing to the
-core \astropypkg package.
-Given this, it is encouraging that the numbers in each career stage/path have
-stayed relatively constant over the history of \astropy (with some showing
-growth), as new, active contributors have generally joined and stepped in to
-fill top contributor roles within the project.
-On the other hand, the total number of top contributors has not grown
-substantially, reflecting one of our principal challenges: Sustaining a pipeline
-from users to contributors to maintainers of the package.
-One other thing to note in this context is that in all eras, 50--60\% of the
-software engineers who contributed to \astropypkg were (or are) employed by
-STScI, whose leadership have consistently provided resources (financial and
-personnel) towards sustaining \astropypkg.
-
-\begin{figure}[t!]
-    \begin{centering}
-      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
-        \caption{
-            todo
-        }
-        \label{fig:contributor-summary:per-month}
-    \end{centering}
-\end{figure}
-
-A final metric we consider ... rate of contributions and contributor participation ...
-
 % Transition of needs: From new features, to community development, sustainability, ...
@@
 \textbf{\astropy representation at national diversity conferences:}
-Several members of the \astropy community attended virtual conferences of the
-National Society of Black Physicists (NSBP) and SACNAS (focusing on Hispanic and
-Native American scientists across all STEM fields). \astropy representatives
-noted that one underlying topic came up multiple times. A major barrier for
-persons who come from underrepresented communities is the lack of resources and
-expertise that are necessary to train students. Writing PEP 8 compliant software
-and understanding the Git and \github workflow is not part of the standard
-Physics and Astrophysics curriculum. Additionally, the larger astronomical
-community is still in the process of transitioning towards \python-based tools,
-making it difficult for students not currently under advisement by some one with
-extensive \python expertise to get involved with the open source \python
-community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
-have a profound impact by empowering underrepresented groups because it provides
-a free, searchable, and accessible introduction to \python tools for
-astrophysical research. Representatives also noted that offering training and
-teaching materials to PIs of Research Experience for Undergraduate (REU)
-programs throughout the United States could be helpful for encouraging advisors
-to teach students how to use Astropy, create their own libraries, and use
-version control on open source platforms like \github. Such materials could also
-be offered as a workshop at national diversity conferences themselves, as SACNAS
-solicits special session proposals each year.
-
-% \textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
-%   initiative:} \todo{No author assigned}{Describe more about this.}
-
-\subsection{Project funding} \label{sec:project-funding}
-
-% \secauthor{Aarya Patil}
+Several members of the \astropy community attended virtual conferences
+of the National Society of Black Physicists (NSBP) and SACNAS
+(focusing on Hispanic and Native American scientists across all STEM
+fields). \astropy representatives noted that one underlying topic came
+up multiple times. A major barrier for persons who come from
+underrepresented communities is the lack of resources and expertise
+that are necessary to train students. Writing PEP 8 compliant software
+and understanding the Git and Github workflow is not part of the
+standard Physics and Astrophysics curriculum. Additionally, the larger
+Astronomnical community is still in the process of transitioning
+towards Python based tools, making it difficult for students not
+currently under advisement by some one with extensive Python expertise
+to get involved with the open source Python community. Projects such
+as Learn Astropy could have a profound impact by empowering
+underrepresented groups because it provides a free, searchable, and
+accessible introduction to Python tools for astrophysical
+research. Representatives also noted that offering training and
+teaching materials to PIs of Research Experience for Undergraduate
+(REU) programs throughout the United States could be helpful for
+encouraging advisors to teach students how to use Astropy, create
+their own libraries, and use version control on open source platforms
+like Github. Such materials could also be offered as a workshop at
+national diversity conferences themselves, as SACNAS solicits special
+session proposals each year.
+
+\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
+  initiative:} \todo{No author assigned}{Describe more about this.}
+
+\subsection{Funding} \label{sec:project-funding}
+
+\secauthor{Aarya Patil}
 
@@
 With the goal of establishing a sustainable financial model for the Astropy
-Project, the Coordination Committee, under the new governance charter (see
-Section~\ref{sec:project-governance}), established a standing Finance Committee.
-This committee oversees the planning and allocation of finances on behalf
-of the Project. The Astropy Project accepts funds from institutions as well as
-individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
-supports and promotes world-class, innovative, open source scientific computing.}
+Project, the Coordination Committee under the new governance charter (see \S
+\ref{sec:project-governance}) established a standing Finance committee.
+This committee oversee the planning and allocation of finances on behalf
+of the project. The Astropy Project accepts funds from institutions as well as
+individuals through NUMFOCUS\footnote{NUMFOCUS is a 501(c)(3) nonprofit that
+supports and promotes world-class, innovative, open source scientific computing.}.
 Previously, no direct financial support was available for the project development,
-and NumFOCUS covered most of the incurred operational costs. However, transitioning
+and NUMFOCUS covered most of the incurred operational costs. However, transitioning
 to long-term financial stability meant having an influx of funds in the form of
 institutional support, and hence the Astropy Project applied for and successfully
-acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
-Project, providing a ``neutral'' space not tied to any particular astronomy
-institution. The two major grants are briefly summarized below.
-
-\begin{description}
-    \item[Moore] The Gordon and Betty Moore Foundation awarded the Astropy
+acquired four grants. These grants are administered by NumFOCUS on behalf of the
+project, providing a ``neutral'' space not tied to any particular astronomy
+institution. The two major ones are briefly summarized below.
+
+\begin{itemize}
+    \item Moore: The Gordon and Betty Moore Foundation awarded the Astropy
     project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
-    Growing the Astropy Project.'' Over a three-year term, this grant supported
-    sustenance of the Project by providing funds for (1) the development and
-    maintenance of the \astropypkg and its infrastructure, provided to existing
-    project members and targeted hires, (2) the transition of long-term users to
-    contributors and maintainers through mentorship efforts, (3) the
-    formalization of a governance structure (see
-    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
-    diversity, and inclusion efforts in the project (see
-    Section~\ref{sec:project-ide}). Some of this money supported travel for
-    meetings, conferences, and workshops. This generous grant helped reduce
-    reliance on volunteer-driven maintenance of the openly-developed package and
-    paved the way to acquire funding from federal agencies, representing a major
-    milestone for the success of the project.
-
-    \item[NASA] A successful proposal to the NASA ROSES-2020
+    Growing the Astropy Project". Over a three-year term, this grant supported
+    sustenance of the project by providing funds for (1) the development and
+    maintenance of the \astropypkg and its infrastructure to existing project
+    members and targeted hires, (2) the transition of long-term users to
+    contributors and maintainers through mentorship efforts, (3) the formalization
+    of a governance structure (see \S \ref{sec:project-governance}), and (4) the
+    improvement of equity, diversity, and inclusion efforts in the project (see \S
+    \ref{sec:project-ide}). Some of the money supported travel for meetings,
+    conferences, and workshops. This generous grant helped reduce reliance on
+    volunteer-driven maintenance of the openly-developed package and paved
+    the way to acquire funding from federal agencies: a big milestone for the
+    success of the project.
+
+    \item NASA: A successful proposal to the NASA ROSES-2020
     call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
@@
     improvements, (2) targeted work on areas of the \astropy core and
-    coordinated packages, (3) enhancements to the Learn
-    Astropy ecosystem and educational materials (see Section~\ref{sec:learn}),
+    coordinated packages that need support, (3) enhancements to the Learn
+    Astropy ecosystem and educational materials (see \S \ref{sec:learn}),
     and (4) support for the \astropy affiliated packages. These goals
@@
     development of the package ecosystem and community engagement.
-\end{description}
+\end{itemize}
 
@@
 resources for scientific software packaging within the Learn Astropy
-framework. All of these successful proposals show that the Astropy Project
+framework. All these successful proposals show that the Astropy project
 is vital for the astronomical infrastructure and community at large, and
@@
 package template generation process includes the ability to setup
-interoperability with \github, allowing for easy repository access from
-documentation, as well as an example \github Actions workflow to demonstrate
-the use of \github's continuous integration tooling.
-
-\subsection{Astropy affiliated packages}
-
-% \secauthor{Matt Craig, Brett Morris}
-
-% Highlight a few new affiliated packages and major updates to existing ones.
-% Include a big table of all affiliated packages and references (as in v2.0
-% paper).
-
-As defined in \paperii, Astropy affiliated packages are astronomy-related
-\python packages that provide functionality that builds upon and extends the
-\astropypkg core package that have requested to be included as part of the
-Astropy Project community. These packages support the goals and vision of
-Astropy of improving code re-use, interoperability, and embracing good coding
-practices such as testing and thorough documentation.
-All Astropy-coordinated and other affiliated packages are listed with detailed
-information on the \href{{https://www.astropy.org/affiliated}}{\astropy
-website}; A table summarizing key aspects of the affiliated packages is included
-in the Appendix below (Table~\ref{tab:affiliated-registry}).
-
-Since \paperii, there has been an expansion of affiliated
-packages for gravitational astrophysics, including:
+interoperability with GitHub, allowing for easy repository access from
+documentation, as well as an example GitHub Actions workflow to demonstrate
+the use of GitHub's continuous integration tooling.
+
+\subsection{Affiliated packages}
+
+\secauthor{Matt Craig, Brett Morris}
+
+Highlight a few new affiliated packages and major updates to existing ones.
+Include a big table of all affiliated packages and references (as in v2.0
+paper).
+
+All coordinated and affiliated packages are listed on the astropy
+website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
+\cite{astropy:2018}, there has been an expansion of affiliated
+packages for gravitational astrophysics including:
 \texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
@@
 gravitational wave probability maps, and \texttt{EinsteinPy} for general
-relativity and gravitational astronomy. There have also been several packages
+relativity and gravitational astronomy. There have been several packages
 added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
@@
 For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
-Other new affiliated packages include \texttt{agnpy} for AGN jets,
-\texttt{statmorph} for fitting galactic morphological diagnostics,
-\texttt{dust\_extinction} for modeling interstellar dust extinction,
-\texttt{feets} for extracting features from time series data, and
-\texttt{corral} for managing data intensive parallel pipelines.
-\texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
-\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
-and \texttt{sbpy} enables calculations for asteroid and
-cometary astrophysics. Finally, \texttt{synphot} provides an interface for
-synthetic photometry.
-% \citep{Mommert2019}
-
-Several of the coordinated and affiliated packages described in
-\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
-\citep{astroquery} has added access to roughly a dozen new missions and data
-services, including the James Webb Space Telescope (JWST) archive, and the
-project has switched to a continuous release model: Every time a change is
-committed to the main development branch it is published on the Python Package
-Index (PyPI) and available for installation. (However, formal releases are still
-done a few times per year.) \texttt{photutils} \cite {photutils} released its
-first stable version, indicating that the API will change less frequently, and
-there have been several significant performance improvements. \texttt{ccdproc}
-\cite{ccdproc} also released a new major version, bringing better performance to
-some image combination operations. The \texttt{regions} package
-\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
-added new ways to manipulate regions and introduced new region types.
-\texttt{reproject}'s \cite{reproject} major new feature is a function to align
-and co-add images to create a mosaic; better support for parallelization was
-also added. \texttt{specutils}, the package that defines containers for 1D and
-2D spectra \cite{specutils}, also had its first stable release and the addition
-of classes to read JWST data. \texttt{stingray} has also had a major performance
-overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
-improvements and has unified its API in preparation for its first stable
-release.
+VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
+We have \texttt{agnpy} for AGN jets, and \texttt{statmorph} for fitting galactic
+morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
+routines. For modeling interstellar dust extinction we have 
+\texttt{dust\_extinction}, and for extracting features from time series we
+have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
+pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
+inference, and \texttt{sbpy} \cite{Mommert2019} does calculations for asteroid and cometary
+astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
+
+Several of the coordinated and affiliated packages described in \cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
+\cite{astroquery} has added
+access to roughly a dozen new missions and data services, including JWST, and
+the project has switched to a continuous release model. Every time a change
+is committed to the main development branch it is published on the Python
+Package Index (PyPI, ref) and available for installation. Formal releases are
+still done a couple of times per year. \texttt{photutils} \cite
+{photutils} released its first stable version, indicating that the API will
+change less frequently, and there have been several significant performance
+improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
+version, bringing better performance to some image combination operations.
+The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
+region definitions \cite{ds9}, added new ways to manipulate regions and
+introduced new region types. \texttt{reproject}’s \cite{reproject} major new
+feature is a function to align and co-add images to create a mosaic; better
+support for parallelization was also added. \texttt{specutils}, the package
+that defines containers for 1D and 2D spectra \cite{specutils}, also had its
+first stable release and the addition of classes to read JWST data.
+\texttt{stingray} has also had a major performance overhaul. The package
+\texttt{gammapy} \cite{gammapy} also has major performance improvements,
+has unified its API in preparation for its first stable release.
 
 \subsection{Connections with data archives}
+
 \label{sec:astroquery}
-
-% \secauthor{Adam Ginsburg}
-
-\texttt{astroquery} \citep{Ginsburg2019} is the Astropy-coordinated package for
-interacting with online archives of astronomical and related data. It contains
-over 50 modules for querying astronomical databases, large and small. In
-particular, since \paperii, significant contributions to \texttt{astroquery}
-have come from several of the major archives, including the European Space
-Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space
-Telescope Science Institute (STScI), the Infrared Science Archive at the NASA
-Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian
-Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter
-Array archive (ALMA). These contributions represent a formal acknowledgement of
-the utility of a centralized tool suite for archive interaction from \python.
-The widespread usage of \texttt{astroquery} is apparent from the range of
-keywords represented in the journal articles that cite the \texttt{astroquery}
-paper \citep{Ginsburg2019}; everything from asteroids to galaxies is
-represented. Additionally \texttt{astroquery} functionality has been built into
-several special-purpose \python packages, such as \texttt{LightKurve}
-\citep{LightKurve} and SORA \citep{SORA}]. More than 4,000 repositories on
-\github make use of \texttt{astroquery} in some way, and \texttt{astroquery}
-features in a large number of tutorials on various facets of astronomical
-analysis.
-
-Many of the existing and newly-contributed tools rely on Virtual Observatory
-(VO) tools. These use the underlying package \texttt{pyvo}, which has also
-recently become an Astropy-coordinated package. While many or all of the
-functions provided in \texttt{astroquery} can be achieved through direct use of
-VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more
-closely resemble the web interfaces that are more familiar to most users.
+% MHvK: feel free to move above label to where it makes most sense.
+
+\secauthor{Adam Ginsburg}
+
+Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
+It contains over 50 modules for querying astronomical databases, large and small.
+In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
+These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
+
+The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented.
+Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}].
+More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.
+
+Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
+These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
+While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
+% note to coauthors: we need a section on pyvo
 

### 2022-03-12 src/ms.tex
add Gemini ack

@@
 a number of web services for code hosting, continuous integration, and
-documentation; in particular, \astropy heavily relies on GitHub, Travis CI,
-Appveyor, CircleCI, and Read the Docs.
+documentation; in particular, \astropy heavily relies on GitHub, Azure
+Pipelines, CircleCI, and Read the Docs.
 

### 2022-03-12 src/ms.tex
move defs to preamble and intro tweaks

@@
 is now happening in more specialized packages that are expanding the
-capabilities of the Astropy ecosystem by building on top of the foundations laid
-by the \astropypkg core package.
-Because of this natural expansion, the name Astropy has grown in scope beyond a
+capabilities of the \astropy ecosystem by building on top of the foundations
+laid by the \astropypkg core package.
+Because of this natural expansion, the name \astropy has grown in scope beyond a
 single \python library to become ``the Astropy Project.''
 
-The Astropy Project is a community effort that represents the union of the
+The \astropy Project is a community effort that represents the union of the
 \astropypkg core package, the ecosystem of astronomy-specific software tools

### 2022-03-12 src/ms.tex
re-add python mentions figure, now with total astro yearly paper counts

@@
     \begin{centering}
-      % \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
+      \includegraphics[width=\textwidth]{python-mentions.pdf}
         \caption{
@@
             figure legend) in refereed publications in the astronomical
-            literature database in the Astrophysics Data System (ADS;
-            \citealt{ads}).
+            literature database in the Astrophysics Data System (ADS).
             \python has rapidly become the dominant programming language
@@
 
+This research has made use of NASA's Astrophysics Data System.
+
 \astropypkg interfaces with the SIMBAD database, operated at CDS, Strasbourg,

### 2022-03-12 src/figures/python-mentions.py

@@ -1,10 +1,7 @@
-"""
-TODO: describe
-"""
-
 # Standard library
 import datetime
 import pathlib
+import pickle
 
 # Third-party
 import ads
@@ -12,6 +9,7 @@ import astropy.table as at
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
+from tqdm.auto import tqdm
 
 
 def get_papers(query, rows_per_page=100, max_pages=100):
@@ -70,6 +68,25 @@ def get_papers(query, rows_per_page=100, max_pages=100):
     return q, df
 
 
+def get_paper_count(query, rows_per_page=500, max_pages=100):
+    """
+    Parameters
+    ----------
+    query : str
+    rows_per_page : int (optional)
+    max_pages : int (optional)
+
+    Returns
+    -------
+    search_query : `ads.SearchQuery`
+    count : int
+    """
+    q = ads.SearchQuery(q=query)
+    for paper in q:
+        break
+    return q, q.response.numFound
+
+
 def get_dfs(cache_path):
     cache_path = pathlib.Path(cache_path)
     queries = {
@@ -99,7 +116,7 @@ def get_dfs(cache_path):
     return dfs
 
 
-def plot_yearly_mentions(paper_dfs):
+def plot_yearly_mentions(paper_dfs, total_counts, min_year=1991):
     fig, ax = plt.subplots(figsize=(8, 6), constrained_layout=True)
 
     for i, (name, df) in enumerate(paper_dfs.items()):
@@ -116,7 +133,12 @@ def plot_yearly_mentions(paper_dfs):
                 mentions,
                 marker='', drawstyle='steps-mid',
                 lw=2, label=name, zorder=-i)
-
+    
+    ax.plot(total_counts['date'], total_counts['counts'],
+            marker='', lw=2, label='all astronomy',
+            drawstyle='steps-mid', zorder=-10, ls='-', 
+            color='#aaaaaa')
+    
     if len(paper_dfs) == 0:
         # ADS key not found - pull request?
         ax.set_title(
@@ -125,24 +147,52 @@ def plot_yearly_mentions(paper_dfs):
             fontsize=14
         )
 
-    ax.set_xlim(datetime.date(1991, 8, 1),
+    ax.set_xlim(datetime.date(min_year, 8, 1),
                 datetime.date(2021, 6, 1))  # 2021.5
 
-    ax.legend(loc='upper left', fontsize=16)
[…]

### 2022-03-12 src/figures/python-mentions.py
remove force to run without cache

@@ -174,7 +174,7 @@ except ads.exceptions.APIResponseError:
 # +
 this_cache_file = cache_path / 'total_counts.pkl'
 
-if not this_cache_file.exists() or True:
+if not this_cache_file.exists():
     total_astro_counts = []
     years = np.arange(min_year, 2022+1)
     for year in tqdm(years):

### 2022-03-12 src/ms.tex
edits up to section 3

@@
 
+\begin{figure}[th]
+    \begin{centering}
+      \includegraphics[width=\textwidth]{python-mentions.pdf}
+        \caption{
+            Yearly full-text mentions of programming languages (indicated in the
+            figure legend) in refereed publications in the astronomical
+            literature database in the Astrophysics Data System (ADS).
+            \python has rapidly become the dominant programming language
+            mentioned in refereed articles over the last 10 years.
+        }
+        \label{fig:python-mentions}
+    \end{centering}
+\end{figure}
+
 The \astropy Project is a community effort that represents the union of the
@@
 
-\begin{figure}
-    \begin{centering}
-      \includegraphics[width=\textwidth]{python-mentions.pdf}
-        \caption{
-            Yearly full-text mentions of programming languages (indicated in the
-            figure legend) in refereed publications in the astronomical
-            literature database in the Astrophysics Data System (ADS).
-            \python has rapidly become the dominant programming language
-            mentioned in refereed articles over the last 10 years.
-        }
-        \label{fig:python-mentions}
-    \end{centering}
-\end{figure}
-
 \section{Major Updates to the Astropy Core Package} \label{sec:core-updates}

### 2022-03-12 src/ms.tex
edits through section 3

@@
 
-As part of the process of developing Astropy into a long-term sustainable
+As part of the process of developing \astropypkg into a long-term sustainable
 product, and to improve transparency and accountability, the Project agreed to
 write down and formalize our governance structure (partly supported by explicit
-funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
+funding for this purpose; see Section~\ref{sec:project-funding}). At the 2019
 Astropy Coordination Meeting, input was gathered from participants on what governance
 structures existed in the associated Open Source Software communities, and what
-would fit well with the needs of Astropy. This led into a ``retreat'' planned
+would fit well with the needs of \astropy. This led into a ``retreat'' planned
 for March 2020, but due to the COVID-19 pandemic, this became a series of
-virtual meetings of the ``Astropy Governance Working Group''. This group
+virtual meetings of the ``Astropy Governance Working Group.'' This group
 drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
@@
 structure, so we refer the reader to that document for a more thorough
-description. However here we highlight some key elements. While many of these
+description, and here we only highlight some key elements. While many of these
 principles were already de facto true or have been discussed organically (and
 have been discussed in earlier papers in this series), the APE0-based governance
-aimed to provide a single place where the community can agree as a starting
+aims to provide a single place where the community can agree as a starting
 point. With this in mind, it highlights the developer and user community of the
@@
 principle of ``do-ocracy'' that those who do work for the Project (be it coding,
-training, or other less concrete contributions) gain more influence on the
-outputs of the Project by virtue of their effort. APE0 adds, however, the
-concept of ``voting members'' - a self-governed part of that community who are
-entrusted to elect the Coordination Committee. While this committee has existed
+training, or other harder-to-quantify contributions) gain more influence on the
+outputs of the Project by virtue of their effort. However, APE0 adds the
+concept of ``voting members'': a self-governed part of that community who are
+entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
 from the inception of the project, APE0 establishes a formal voting process for
-this committee, and explicitly outlines the rights and responsibilities of the
-Coordination Committee. This role is mainly to facilitate consensus and act as
-the decision maker when other mechanisms have failed. However, it also includes
-powers that either require central authority or secrets (e.g., passwords), but
-APE0 also charges the Committee to devolve responsibilities and seek community
-input on these items as often as possible.
+CoCo members, and explicitly outlines the rights and responsibilities of the
+CoCo. This role is mainly to facilitate consensus and act as the decision maker
+when other mechanisms have failed, and includes powers that either
+require central authority or secrets (e.g., passwords) However, APE0 also
+charges the Committee to devolve responsibilities and seek community input on
+these items as often as possible.
 
 The first Coordination Committee election under these rules took place in Fall
-2021, electing a mix of prior existing and new coordination committee members,
+2021, electing a mix of prior and new coordination committee members,
 and was contested in the sense of more candidates than available slots. This
@@
 \textbf{\astropy representation at national diversity conferences:}
-Several members of the \astropy community attended virtual conferences
-of the National Society of Black Physicists (NSBP) and SACNAS
-(focusing on Hispanic and Native American scientists across all STEM
-fields). \astropy representatives noted that one underlying topic came
-up multiple times. A major barrier for persons who come from
-underrepresented communities is the lack of resources and expertise
-that are necessary to train students. Writing PEP 8 compliant software
-and understanding the Git and Github workflow is not part of the
-standard Physics and Astrophysics curriculum. Additionally, the larger
-Astronomnical community is still in the process of transitioning
-towards Python based tools, making it difficult for students not
-currently under advisement by some one with extensive Python expertise
-to get involved with the open source Python community. Projects such
-as Learn Astropy could have a profound impact by empowering
-underrepresented groups because it provides a free, searchable, and
-accessible introduction to Python tools for astrophysical
-research. Representatives also noted that offering training and
-teaching materials to PIs of Research Experience for Undergraduate
-(REU) programs throughout the United States could be helpful for
-encouraging advisors to teach students how to use Astropy, create
-their own libraries, and use version control on open source platforms
-like Github. Such materials could also be offered as a workshop at
-national diversity conferences themselves, as SACNAS solicits special
-session proposals each year.
-
-\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
-  initiative:} \todo{No author assigned}{Describe more about this.}
-
-\subsection{Funding} \label{sec:project-funding}
-
-\secauthor{Aarya Patil}
+Several members of the \astropy community attended virtual conferences of the
+National Society of Black Physicists (NSBP) and SACNAS (focusing on Hispanic and
+Native American scientists across all STEM fields). \astropy representatives
+noted that one underlying topic came up multiple times. A major barrier for
+persons who come from underrepresented communities is the lack of resources and
+expertise that are necessary to train students. Writing PEP 8 compliant software
+and understanding the Git and \github workflow is not part of the standard
+Physics and Astrophysics curriculum. Additionally, the larger astronomical
+community is still in the process of transitioning towards \python-based tools,
+making it difficult for students not currently under advisement by some one with
+extensive \python expertise to get involved with the open source \python
+community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
+have a profound impact by empowering underrepresented groups because it provides
+a free, searchable, and accessible introduction to \python tools for
+astrophysical research. Representatives also noted that offering training and
+teaching materials to PIs of Research Experience for Undergraduate (REU)
+programs throughout the United States could be helpful for encouraging advisors
+to teach students how to use Astropy, create their own libraries, and use
+version control on open source platforms like \github. Such materials could also
+be offered as a workshop at national diversity conferences themselves, as SACNAS
+solicits special session proposals each year.
+
+% \textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
+%   initiative:} \todo{No author assigned}{Describe more about this.}
+
+\subsection{Project funding} \label{sec:project-funding}
+
+% \secauthor{Aarya Patil}
 
@@
 With the goal of establishing a sustainable financial model for the Astropy
-Project, the Coordination Committee under the new governance charter (see \S
-\ref{sec:project-governance}) established a standing Finance committee.
-This committee oversee the planning and allocation of finances on behalf
-of the project. The Astropy Project accepts funds from institutions as well as
-individuals through NUMFOCUS\footnote{NUMFOCUS is a 501(c)(3) nonprofit that
-supports and promotes world-class, innovative, open source scientific computing.}.
+Project, the Coordination Committee, under the new governance charter (see
+Section~\ref{sec:project-governance}), established a standing Finance Committee.
+This committee oversees the planning and allocation of finances on behalf
+of the Project. The Astropy Project accepts funds from institutions as well as
+individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
+supports and promotes world-class, innovative, open source scientific computing.}
 Previously, no direct financial support was available for the project development,
-and NUMFOCUS covered most of the incurred operational costs. However, transitioning
+and NumFOCUS covered most of the incurred operational costs. However, transitioning
 to long-term financial stability meant having an influx of funds in the form of
 institutional support, and hence the Astropy Project applied for and successfully
-acquired four grants. These grants are administered by NumFOCUS on behalf of the
-project, providing a ``neutral'' space not tied to any particular astronomy
-institution. The two major ones are briefly summarized below.
-
-\begin{itemize}
-    \item Moore: The Gordon and Betty Moore Foundation awarded the Astropy
+acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
+Project, providing a ``neutral'' space not tied to any particular astronomy
+institution. The two major grants are briefly summarized below.
+
+\begin{description}
+    \item[Moore] The Gordon and Betty Moore Foundation awarded the Astropy
     project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
-    Growing the Astropy Project". Over a three-year term, this grant supported
-    sustenance of the project by providing funds for (1) the development and
-    maintenance of the \astropypkg and its infrastructure to existing project
-    members and targeted hires, (2) the transition of long-term users to
-    contributors and maintainers through mentorship efforts, (3) the formalization
-    of a governance structure (see \S \ref{sec:project-governance}), and (4) the
-    improvement of equity, diversity, and inclusion efforts in the project (see \S
-    \ref{sec:project-ide}). Some of the money supported travel for meetings,
-    conferences, and workshops. This generous grant helped reduce reliance on
-    volunteer-driven maintenance of the openly-developed package and paved
-    the way to acquire funding from federal agencies: a big milestone for the
-    success of the project.
-
-    \item NASA: A successful proposal to the NASA ROSES-2020
+    Growing the Astropy Project.'' Over a three-year term, this grant supported
+    sustenance of the Project by providing funds for (1) the development and
+    maintenance of the \astropypkg and its infrastructure, provided to existing
+    project members and targeted hires, (2) the transition of long-term users to
+    contributors and maintainers through mentorship efforts, (3) the
+    formalization of a governance structure (see
+    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
+    diversity, and inclusion efforts in the project (see
+    Section~\ref{sec:project-ide}). Some of this money supported travel for
+    meetings, conferences, and workshops. This generous grant helped reduce
+    reliance on volunteer-driven maintenance of the openly-developed package and
+    paved the way to acquire funding from federal agencies, representing a major
+    milestone for the success of the project.
+
+    \item[NASA] A successful proposal to the NASA ROSES-2020
     call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
@@
     improvements, (2) targeted work on areas of the \astropy core and
-    coordinated packages that need support, (3) enhancements to the Learn
-    Astropy ecosystem and educational materials (see \S \ref{sec:learn}),
+    coordinated packages, (3) enhancements to the Learn
+    Astropy ecosystem and educational materials (see Section~\ref{sec:learn}),
     and (4) support for the \astropy affiliated packages. These goals
@@
     development of the package ecosystem and community engagement.
-\end{itemize}
+\end{description}
 
@@
 resources for scientific software packaging within the Learn Astropy
-framework. All these successful proposals show that the Astropy project
+framework. All of these successful proposals show that the Astropy Project
 is vital for the astronomical infrastructure and community at large, and

### 2022-03-12 src/ms.tex
edits through section 4

@@
 package template generation process includes the ability to setup
-interoperability with GitHub, allowing for easy repository access from
-documentation, as well as an example GitHub Actions workflow to demonstrate
-the use of GitHub's continuous integration tooling.
-
-\subsection{Affiliated packages}
-
-\secauthor{Matt Craig, Brett Morris}
-
-Highlight a few new affiliated packages and major updates to existing ones.
-Include a big table of all affiliated packages and references (as in v2.0
-paper).
-
-All coordinated and affiliated packages are listed on the astropy
-website\footnote{\url{https://www.astropy.org/affiliated}}. Since
-\cite{astropy:2018}, there has been an expansion of affiliated
-packages for gravitational astrophysics including:
+interoperability with \github, allowing for easy repository access from
+documentation, as well as an example \github Actions workflow to demonstrate
+the use of \github's continuous integration tooling.
+
+\subsection{Astropy affiliated packages}
+
+% \secauthor{Matt Craig, Brett Morris}
+
+% Highlight a few new affiliated packages and major updates to existing ones.
+% Include a big table of all affiliated packages and references (as in v2.0
+% paper).
+
+As defined in \paperii, Astropy affiliated packages are astronomy-related
+\python packages that provide functionality that builds upon and extends the
+\astropypkg core package that have requested to be included as part of the
+Astropy Project community. These packages support the goals and vision of
+Astropy of improving code re-use, interoperability, and embracing good coding
+practices such as testing and thorough documentation.
+All Astropy-coordinated and other affiliated packages are listed on the astropy
+website.\footnote{\url{https://www.astropy.org/affiliated}}
+
+Since \paperii, there has been an expansion of affiliated
+packages for gravitational astrophysics, including:
 \texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
@@
 gravitational wave probability maps, and \texttt{EinsteinPy} for general
-relativity and gravitational astronomy. There have been several packages
+relativity and gravitational astronomy. There have also been several packages
 added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
@@
 For ground-based astronomy, \texttt{baseband} has added IO capabilities for
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem.
-We have \texttt{agnpy} for AGN jets, and \texttt{statmorph} for fitting galactic
-morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting
-routines. For modeling interstellar dust extinction we have
-\texttt{dust\_extinction}, and for extracting features from time series we
-have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel
-pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian
-inference, and \texttt{sbpy} \cite{Mommert2019} does calculations for asteroid and cometary
-astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.
-
-Several of the coordinated and affiliated packages described in \cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
-\cite{astroquery} has added
-access to roughly a dozen new missions and data services, including JWST, and
-the project has switched to a continuous release model. Every time a change
-is committed to the main development branch it is published on the Python
-Package Index (PyPI, ref) and available for installation. Formal releases are
-still done a couple of times per year. \texttt{photutils} \cite
-{photutils} released its first stable version, indicating that the API will
-change less frequently, and there have been several significant performance
-improvements. \texttt{ccdproc} \cite{ccdproc} also released a new major
-version, bringing better performance to some image combination operations.
-The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
-region definitions \cite{ds9}, added new ways to manipulate regions and
-introduced new region types. \texttt{reproject}’s \cite{reproject} major new
-feature is a function to align and co-add images to create a mosaic; better
-support for parallelization was also added. \texttt{specutils}, the package
-that defines containers for 1D and 2D spectra \cite{specutils}, also had its
-first stable release and the addition of classes to read JWST data.
-\texttt{stingray} has also had a major performance overhaul. The package
-\texttt{gammapy} \cite{gammapy} also has major performance improvements,
-has unified its API in preparation for its first stable release.
+VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
+Other new affiliated packages include \texttt{agnpy} for AGN jets,
+\texttt{statmorph} for fitting galactic morphological diagnostics,
+\texttt{dust\_extinction} for modeling interstellar dust extinction,
+\texttt{feets} for extracting features from time series data, and
+\texttt{corral} for managing data intensive parallel pipelines.
+\texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
+\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
+and \texttt{sbpy} enables calculations for asteroid and
+cometary astrophysics. Finally, \texttt{synphot} provides an interface for
+synthetic photometry.
+% \citep{Mommert2019}
+
+Several of the coordinated and affiliated packages described in
+\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
+\citep{astroquery} has added access to roughly a dozen new missions and data
+services, including the James Webb Space Telescope (JWST) archive, and the
+project has switched to a continuous release model: Every time a change is
+committed to the main development branch it is published on the Python Package
+Index (PyPI) and available for installation. (However, formal releases are still
+done a few times per year.) \texttt{photutils} \cite {photutils} released its
+first stable version, indicating that the API will change less frequently, and
+there have been several significant performance improvements. \texttt{ccdproc}
+\cite{ccdproc} also released a new major version, bringing better performance to
+some image combination operations. The \texttt{regions} package
+\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
+added new ways to manipulate regions and introduced new region types.
+\texttt{reproject}'s \cite{reproject} major new feature is a function to align
+and co-add images to create a mosaic; better support for parallelization was
+also added. \texttt{specutils}, the package that defines containers for 1D and
+2D spectra \cite{specutils}, also had its first stable release and the addition
+of classes to read JWST data. \texttt{stingray} has also had a major performance
+overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
+improvements and has unified its API in preparation for its first stable
+release.
+
+\todo{APW}{Add a table of all affiliated packages, like in the previous paper}
 
 \subsection{Connections with data archives}
-
 \label{sec:astroquery}
-% MHvK: feel free to move above label to where it makes most sense.
-
-\secauthor{Adam Ginsburg}
-
-Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data.
-It contains over 50 modules for querying astronomical databases, large and small.
-In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA).
-These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.
-
-The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented.
-Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}].
-More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.
-
-Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
-These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package.
-While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
-% note to coauthors: we need a section on pyvo
+
+% \secauthor{Adam Ginsburg}
+
+\texttt{astroquery} \citep{Ginsburg2019} is the Astropy-coordinated package for
+interacting with online archives of astronomical and related data. It contains
+over 50 modules for querying astronomical databases, large and small. In
+particular, since \paperii, significant contributions to \texttt{astroquery}
+have come from several of the major archives, including the European Space
+Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space
+Telescope Science Institute (STScI), the Infrared Science Archive at the NASA
+Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian
+Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter
+Array archive (ALMA). These contributions represent a formal acknowledgement of
+the utility of a centralized tool suite for archive interaction from \python.
+The widespread usage of \texttt{astroquery} is apparent from the range of
+keywords represented in the journal articles that cite the \texttt{astroquery}
+paper \citep{Ginsburg2019}; everything from asteroids to galaxies is
+represented. Additionally \texttt{astroquery} functionality has been built into
+several special-purpose \python packages, such as \texttt{LightKurve}
+\citep{LightKurve} and SORA \citep{SORA}]. More than 4,000 repositories on
+\github make use of \texttt{astroquery} in some way, and \texttt{astroquery}
+features in a large number of tutorials on various facets of astronomical
+analysis.
+
+Many of the existing and newly-contributed tools rely on Virtual Observatory
+(VO) tools. These use the underlying package \texttt{pyvo}, which has also
+recently become an Astropy-coordinated package. While many or all of the
+functions provided in \texttt{astroquery} can be achieved through direct use of
+VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more
+closely resemble the web interfaces that are more familiar to most users.
 
@@
 
-Rubin Observatory
-\secunfilled
-
-LIGO/Virgo/KAGRA
-\secauthor{Leo Singer}
+% Rubin Observatory
+% \secunfilled
+
+% LIGO/Virgo/KAGRA
+% \secauthor{Leo Singer}
+
 \section{Future Plans for the Astropy Project} \label{sec:future}

### 2022-03-12 src/ms.tex
slightly reorganize end sections

@@
 
-\section{Future Plans for the Astropy Project} \label{sec:future}
-
-\subsection{Roadmap}
-
-\secauthor{Clara Brasseur}
-
-The Astropy Roadmap\footnote{\url{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}} is a document in the astropy-project repository that captures high level actionable items that the Astropy project aims to undertake to improve the health and stability of the  project.
-It is a static document that is revisited regularly at the Astropy coordination meetings, to keep track of progress and write new versions as needed.
-There is a related project board linked to the Roadmap document, that holds specific issues and efforts related to Roadmap items.
-The project board is a living document that is continually updated as work is planned, assigned, and completed.
-
-All items in the Roadmap have been agreed to be priorities for the Astropy Project, and are color-coded based on resources (both time/effort and developers/expertise) needed to complete the item.
-Green items are well underway, have sufficient resources/support and a plan in place for completion.
-Orange items are well defined, and work for aquiring sufficient resources underway.
-Red items do not yet have a plan for implementation and need more resources.
-\todo{addition}{Add some examples of high priority items from the roadmap in a new paragraph here}
-The Astropy Roadmap originates from the March 2021 Astropy Coordination meeting where it was first drafted before being handed off to a newly formed Astropy Roadmap working group for completion.
-The Roadmap in its currant form was adopted via pull request in December 2021.
-
-\section{Learn Astropy} \label{sec:learn}
+\subsection{Learn Astropy} \label{sec:learn}
 

### 2022-03-12 src/scripts/affiliated-registry.py
add new snakemake rule to build affiliated package table

@@ -1,0 +1,44 @@
+"""
+Generate the affiliated package table items file.
+"""
+import json
+import re
+
+
+# Downloaded registry into static directory as of 2022-03-12
+
+with open('../static/registry.json') as data_file:
+    data = json.load(data_file)
+
+# with open('bib_mapping.json') as data_file:
+#     bib_map = json.load(data_file)
+
+package_list = sorted(data['packages'], key=lambda k: k['name'].lower())
+print(package_list)
+
+row = (
+    '\\href{{{url}}}{{{name}}} & '
+    '\\href{{https://pypi.python.org/pypi/{pypi_name}}}{{{3}}} & '
+    '{maintainer} & {cite_key} \\\\\n'
+)
+
+with open("../generated/affiliated-table.tex", "w") as f1:
+    for item in package_list:
+
+        # citealt = ("\\citealt{{{0}}}".format(bib_map[item["name"]])
+        #            if item["name"] in bib_map
+        #            else "")
+
+        maintainer = re.sub(' [<(].*?[>)]', '', item["maintainer"])
+
+        name = re.sub('_', '\_', item["name"])
+
+        f1.write(
+            row.format(
+                url=item["repo_url"],
+                name=name,
+                pypi_name=item['pypi_name'],
+                maintainer=maintainer,
+                cite_key='test'
+            )
+        )

### 2022-03-13 src/figures/python-mentions.py
working affiliated package table, but missing citations
pass through dev key / token

assert and remove except

undo dev key overengineering - it should be fine with the env var

upload ms pdf file artifact

ok now i remember - PRs done have access to repository secrets

@@ -133,14 +133,16 @@ def plot_yearly_mentions(paper_dfs, total_counts, min_year=1991):
                 mentions,
                 marker='', drawstyle='steps-mid',
                 lw=2, label=name, zorder=-i)
-    
-    ax.plot(total_counts['date'], total_counts['counts'],
-            marker='', lw=2, label='all astronomy',
-            drawstyle='steps-mid', zorder=-10, ls='-', 
-            color='#aaaaaa')
-    
+
+    if len(total_counts) != 0:
+        ax.plot(total_counts['date'], total_counts['counts'],
+                marker='', lw=2, label='all astronomy',
+                drawstyle='steps-mid', zorder=-10, ls='-',
+                color='#aaaaaa')
+
     if len(paper_dfs) == 0:
-        # ADS key not found - pull request?
+        # ADS key not found - pull request? secrets are only exposed when run
+        # from the main repo (not forks)
         ax.set_title(
             "ADS key not found! If this is a pull request, that is normal: "
             "ADS queries are not performed.",
@@ -154,7 +156,7 @@ def plot_yearly_mentions(paper_dfs, total_counts, min_year=1991):
 
     ax.set_xlabel('Time [year]')
     ax.set_ylabel('Full-text mentions per year')
-    
+
     ax.set_yscale('log')
 
     fig.savefig("python-mentions.pdf", bbox_inches="tight")
@@ -168,13 +170,15 @@ cache_path.mkdir(exist_ok=True, parents=True)
 
 try:
     paper_dfs = get_dfs(cache_path)
+    api_success = True
 except ads.exceptions.APIResponseError:
     paper_dfs = {}
+    api_success = False
 
 # +
 this_cache_file = cache_path / 'total_counts.pkl'
 
-if not this_cache_file.exists():
+if not this_cache_file.exists() and api_success:
     total_astro_counts = []
     years = np.arange(min_year, 2022+1)
     for year in tqdm(years):
@@ -189,10 +193,15 @@ if not this_cache_file.exists():
     }
     with open(this_cache_file, 'wb') as f:
         pickle.dump(total_counts, f)
-        
-else:
+
+elif this_cache_file.exists():
     with open(this_cache_file, 'rb') as f:
         total_counts = pickle.load(f)
+
+else:
+    total_counts = {}
+
+
 # -
 
 plot_yearly_mentions(paper_dfs, total_counts)

### 2022-03-13 src/scripts/affiliated-registry.py

@@ -2,27 +2,28 @@
 Generate the affiliated package table items file.
 """
 import json
+import pathlib
 import re
 
+this_path = pathlib.Path(__file__).parent.resolve()
 
 # Downloaded registry into static directory as of 2022-03-12
 
-with open('../static/registry.json') as data_file:
+with open(this_path / '../static/registry.json') as data_file:
     data = json.load(data_file)
 
 # with open('bib_mapping.json') as data_file:
 #     bib_map = json.load(data_file)
 
 package_list = sorted(data['packages'], key=lambda k: k['name'].lower())
-print(package_list)
 
 row = (
     '\\href{{{url}}}{{{name}}} & '
-    '\\href{{https://pypi.python.org/pypi/{pypi_name}}}{{{3}}} & '
+    '\\href{{https://pypi.python.org/pypi/{pypi_name}}}{{{pypi_name}}} & '
     '{maintainer} & {cite_key} \\\\\n'
 )
 
-with open("../generated/affiliated-table.tex", "w") as f1:
+with open(this_path / "../generated/affiliated-table.tex", "w") as f1:
     for item in package_list:
 
         # citealt = ("\\citealt{{{0}}}".format(bib_map[item["name"]])
@@ -32,12 +33,13 @@ with open("../generated/affiliated-table.tex", "w") as f1:
         maintainer = re.sub(' [<(].*?[>)]', '', item["maintainer"])
 
         name = re.sub('_', '\_', item["name"])
+        pypi_name = re.sub('_', '\_', item["pypi_name"])
 
         f1.write(
             row.format(
                 url=item["repo_url"],
                 name=name,
-                pypi_name=item['pypi_name'],
+                pypi_name=pypi_name,
                 maintainer=maintainer,
                 cite_key='test'
             )

### 2022-03-14 src/ms.tex
remove subsubsections for learn
working affiliated package table, but missing citations
pass through dev key / token

assert and remove except

undo dev key overengineering - it should be fine with the env var

upload ms pdf file artifact

ok now i remember - PRs done have access to repository secrets
Add section for Data Central/AAT Archive

@@
 practices such as testing and thorough documentation.
-All Astropy-coordinated and other affiliated packages are listed on the astropy
-website.\footnote{\url{https://www.astropy.org/affiliated}}
+All Astropy-coordinated and other affiliated packages are listed with detailed
+information on the \href{{https://www.astropy.org/affiliated}}{\astropy
+website}; A table summarizing key aspects of the affiliated packages is included
+in the Appendix below (Table~\ref{tab:affiliated-registry}).
 
@@
 
-\todo{APW}{Add a table of all affiliated packages, like in the previous paper}
-
 \subsection{Connections with data archives}
@@
 
+\subsubsection{Data Central and the Anglo-Australian Telescope Archive}
+\secauthor{James Tocknell}
+Data Central is one of the five Australian Virtual Observatory\footnote{known as
+  the All-Sky Virtual Observatory, or ASVO} nodes, providing both Australian and
+international astronomers access to and services for surveys of the southern
+sky. Whilst primarily hosting optical data, including the Anglo-Australian
+Telescope (AAT) Archive\footnote{The AAT Archive is currently jointly managed by
+  Data Central, Astronomy Australia Limited (AAL), and National Computational
+  Infrastructure (NCI)}, Data Central also assists theoretical and radio
+astronomers host and manage their data.
+
+Python is widely used by Data Central, and \astropy and its wider ecosystem play
+a vital role in the initial ingestion, reduction, access and visualisation of
+the hosted data and services. For example, the AAT Archive hosts data back to
+1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest
+data flowing from the AAT to ensure that a consistent time and coordinate
+system is provided to users for simpler querying.
+
+As many of the surveys hosted by Data Central are spectroscopic surveys from the
+AAT, specutils is heavily used to standardise the variety of encodings of
+spectroscopic data into a form easily readable and visualisable by many standard
+tools (such as ds9 or splat). Data Central has upstreamed into specutils loaders
+for all hosted spectra, including 2dfdr-reduced spectra from the AAT Archive,
+and strongly encourages other archives to provide loaders for their data to
+further reuse of existing archival data.
+
 % Rubin Observatory

### 2022-03-15 src/scripts/affiliated-registry.py
fix script to generated affiliated packages table
fix style and unicode encoding

redundant

write differently

ah, caching was confusing me

@@ -4,6 +4,8 @@ Generate the affiliated package table items file.
 import json
 import pathlib
 import re
+from pylatexenc.latexencode import unicode_to_latex
+
 
 this_path = pathlib.Path(__file__).parent.resolve()
 
@@ -12,35 +14,70 @@ this_path = pathlib.Path(__file__).parent.resolve()
 with open(this_path / '../static/registry.json') as data_file:
     data = json.load(data_file)
 
-# with open('bib_mapping.json') as data_file:
-#     bib_map = json.load(data_file)
+with open(this_path / '../static/affiliated-refs.bib') as bib_file:
+    bib_text = bib_file.read()
+
+pattr = re.compile(r"@[a-zA-Z]+\{([a-zA-Z:0-9\-]+),[\s\S]+?[\}]{0,1}\n\}")
+bib_map = {}
+for m in pattr.finditer(bib_text):
+    cite_name = m.groups()[0]
+    bib_record = m.string[m.start():m.end()]
+
+    pkg_name = cite_name
+    if ":" in cite_name:
+        pkg_name = cite_name.split(':')[0]
+
+    if pkg_name not in bib_map:
+        bib_map[pkg_name] = {'bib': [bib_record], 'cite': [cite_name]}
+    else:
+        bib_map[pkg_name]['bib'].append(bib_record)
+        bib_map[pkg_name]['cite'].append(cite_name)
 
 package_list = sorted(data['packages'], key=lambda k: k['name'].lower())
 
 row = (
     '\\href{{{url}}}{{{name}}} & '
     '\\href{{https://pypi.python.org/pypi/{pypi_name}}}{{{pypi_name}}} & '
-    '{maintainer} & {cite_key} \\\\\n'
+    '{maintainer} & {cite_command} \\\\\n'
 )
 
-with open(this_path / "../generated/affiliated-table.tex", "w") as f1:
-    for item in package_list:
+lines = []
+for pkg in package_list:
+    if pkg['name'] == 'astropy core package':
+        continue
 
-        # citealt = ("\\citealt{{{0}}}".format(bib_map[item["name"]])
-        #            if item["name"] in bib_map
-        #            else "")
+    if pkg['name'] not in bib_map:
+        print(f"{pkg['name']} not in bib file")
 
-        maintainer = re.sub(' [<(].*?[>)]', '', item["maintainer"])
+    cite_cmd = ""
+    if pkg['name'] in bib_map:
+        cite_cmd = r",\newline ".join([
+            f"\\citet{{{cite_key}}}"
[…]

### 2022-03-15 src/scripts/affiliated-registry.py
minor style things

@@ -60,12 +60,12 @@ for pkg in package_list:
     maintainer_block = re.sub(' [<(].*?[>)]', '', pkg["maintainer"])
     maintainers = [x.split(',') for x in maintainer_block.split(' and ')]
     maintainers = [x.strip() for x in sum(maintainers, [])
-                    if len(x.strip()) > 0]
+                   if len(x.strip()) > 0]
     maintainers = [unicode_to_latex(x).replace(" ", "~")
-                    for x in maintainers]
+                   for x in maintainers]
 
-    name = re.sub('_', '\_', pkg["name"])
-    pypi_name = re.sub('_', '\_', pkg["pypi_name"])
+    name = re.sub('_', r'\_', pkg["name"])
+    pypi_name = re.sub('_', r'\_', pkg["pypi_name"])
 
     lines.append(
         row.format(

### 2022-03-15 src/ms.tex
First round of full paper edits (#40)
* move defs to preamble and intro tweaks

* re-add python mentions figure, now with total astro yearly paper counts

* remove force to run without cache

* edits up to section 3

* edits through section 3

* edits through section 4

* slightly reorganize end sections

* remove subsubsections for learn

* add new snakemake rule to build affiliated package table

* working affiliated package table, but missing citations

pass through dev key / token

assert and remove except

undo dev key overengineering - it should be fine with the env var

upload ms pdf fi

@@
 practices such as testing and thorough documentation.
-All Astropy-coordinated and other affiliated packages are listed with detailed
-information on the \href{{https://www.astropy.org/affiliated}}{\astropy
-website}; A table summarizing key aspects of the affiliated packages is included
-in the Appendix below (Table~\ref{tab:affiliated-registry}).
+All Astropy-coordinated and other affiliated packages are listed on the astropy
+website.\footnote{\url{https://www.astropy.org/affiliated}}
 
@@
 
+\todo{APW}{Add a table of all affiliated packages, like in the previous paper}
+
 \subsection{Connections with data archives}
@@
     \tablecaption{Registry of affiliated packages.}
-    \label{tab:affiliated-registry}
+    \label{tab:registry}
     \tablehead{

### 2022-03-15 src/figures/python-mentions.py

@@ -1,10 +1,7 @@
-"""
-TODO: describe
-"""
-
 # Standard library
 import datetime
 import pathlib
+import pickle
 
 # Third-party
 import ads
@@ -12,6 +9,7 @@ import astropy.table as at
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
+from tqdm.auto import tqdm
 
 
 def get_papers(query, rows_per_page=100, max_pages=100):
@@ -70,6 +68,25 @@ def get_papers(query, rows_per_page=100, max_pages=100):
     return q, df
 
 
+def get_paper_count(query, rows_per_page=500, max_pages=100):
+    """
+    Parameters
+    ----------
+    query : str
+    rows_per_page : int (optional)
+    max_pages : int (optional)
+
+    Returns
+    -------
+    search_query : `ads.SearchQuery`
+    count : int
+    """
+    q = ads.SearchQuery(q=query)
+    for paper in q:
+        break
+    return q, q.response.numFound
+
+
 def get_dfs(cache_path):
     cache_path = pathlib.Path(cache_path)
     queries = {
@@ -99,7 +116,7 @@ def get_dfs(cache_path):
     return dfs
 
 
-def plot_yearly_mentions(paper_dfs):
+def plot_yearly_mentions(paper_dfs, total_counts, min_year=1991):
     fig, ax = plt.subplots(figsize=(8, 6), constrained_layout=True)
 
     for i, (name, df) in enumerate(paper_dfs.items()):
@@ -117,32 +134,74 @@ def plot_yearly_mentions(paper_dfs):
                 marker='', drawstyle='steps-mid',
                 lw=2, label=name, zorder=-i)
 
+    if len(total_counts) != 0:
+        ax.plot(total_counts['date'], total_counts['counts'],
+                marker='', lw=2, label='all astronomy',
+                drawstyle='steps-mid', zorder=-10, ls='-',
+                color='#aaaaaa')
+
     if len(paper_dfs) == 0:
-        # ADS key not found - pull request?
+        # ADS key not found - pull request? secrets are only exposed when run
+        # from the main repo (not forks)
         ax.set_title(
             "ADS key not found! If this is a pull request, that is normal: "
             "ADS queries are not performed.",
             fontsize=14
         )
 
-    ax.set_xlim(datetime.date(1991, 8, 1),
[…]

### 2022-03-15 src/scripts/affiliated-registry.py

@@ -1,0 +1,83 @@
+"""
+Generate the affiliated package table items file.
+"""
+import json
+import pathlib
+import re
+from pylatexenc.latexencode import unicode_to_latex
+
+
+this_path = pathlib.Path(__file__).parent.resolve()
+
+# Downloaded registry into static directory as of 2022-03-12
+
+with open(this_path / '../static/registry.json') as data_file:
+    data = json.load(data_file)
+
+with open(this_path / '../static/affiliated-refs.bib') as bib_file:
+    bib_text = bib_file.read()
+
+pattr = re.compile(r"@[a-zA-Z]+\{([a-zA-Z:0-9\-]+),[\s\S]+?[\}]{0,1}\n\}")
+bib_map = {}
+for m in pattr.finditer(bib_text):
+    cite_name = m.groups()[0]
+    bib_record = m.string[m.start():m.end()]
+
+    pkg_name = cite_name
+    if ":" in cite_name:
+        pkg_name = cite_name.split(':')[0]
+
+    if pkg_name not in bib_map:
+        bib_map[pkg_name] = {'bib': [bib_record], 'cite': [cite_name]}
+    else:
+        bib_map[pkg_name]['bib'].append(bib_record)
+        bib_map[pkg_name]['cite'].append(cite_name)
+
+package_list = sorted(data['packages'], key=lambda k: k['name'].lower())
+
+row = (
+    '\\href{{{url}}}{{{name}}} & '
+    '\\href{{https://pypi.python.org/pypi/{pypi_name}}}{{{pypi_name}}} & '
+    '{maintainer} & {cite_command} \\\\\n'
+)
+
+lines = []
+for pkg in package_list:
+    if pkg['name'] == 'astropy core package':
+        continue
+
+    if pkg['name'] not in bib_map:
+        print(f"{pkg['name']} not in bib file")
+
+    cite_cmd = ""
+    if pkg['name'] in bib_map:
+        cite_cmd = r",\newline ".join([
+            f"\\citet{{{cite_key}}}"
+            for cite_key in bib_map[pkg['name']]['cite']
+        ])
+
+    # Parse list of maintainers to make line wrapping better:
+    maintainer_block = re.sub(' [<(].*?[>)]', '', pkg["maintainer"])
+    maintainers = [x.split(',') for x in maintainer_block.split(' and ')]
+    maintainers = [x.strip() for x in sum(maintainers, [])
+                   if len(x.strip()) > 0]
[…]

### 2022-03-15 src/ms.tex
update learn section. Closes #42

--- 2022-03-15 src/ms.tex
@@
 
-\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
+% \secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
 
@@
 educational efforts made by the Astropy Project, which are led by the Learn
-Team.
+Astropy Team.
 The efforts focus on developing online content and workshops covering
@@
 that showcase a short task that can be performed with Astropy or an affiliated
-package; and \textit{documentation}, which is contained within the code base.
+package; and \textit{documentation}, which contains more detailed information
+about the code base and user interface.
 This categorization drives content development, infrastructure choices, and the
-appearance of the Learn Astropy website.
+appearance of the \href{https://learn.astropy.org}{Learn Astropy website}.
 The Learn Team meets weekly to work on creating, expanding, improving these
@@
 
-\begin{description}
-
-\item[Learn Astropy website] The Learn Team re-launched the main website and
-search interface for Learn Astropy in 2021 with a new infrastructure platform,
-built around full-text search and interactive filtering functionality, with the
-goal of making content discoverable as the Learn Astropy content catalog
+The Learn Team recently re-launched the main website and search interface for
+Learn Astropy in 2021 with a new infrastructure platform, built around full-text
+search and interactive filtering functionality, with the goal of making content
+more easily searchable and discoverable as the Learn Astropy content catalog
 expands.
@@
 uses the Algolia service to power its search and filtering user interface.
-Our Python-based application, Learn Astropy Librarian, populates data into the
-Algolia service. We tuned the Librarian around specific content formats (such as
-Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
-accurately index content and heuristically extract metadata.
+Our \python-based application,
+\href{https://github.com/astropy/learn-astropy-librarian/}{Learn Astropy
+Librarian}, populates data into the Algolia service. We tuned the Librarian
+around specific content formats (such as Jupyter Notebook-based tutorial pages
+and Jupyter Book-based guides) to more accurately index content and
+heuristically extract metadata.
 A consequence of the new platform is that we now maintain and compile content
@@
 
-\item[Tutorials] Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
-
-\item[Guides] {\it ccdproc} Jupyter book has been completed.
-
-\item[Workshops] The Project has been conducting workshops at winter
-meetings of the American Astronomical Society since AAS 225 in January 2015.
+We currently host 19 tutorials written as interactive Jupyter Notebooks and
+rendered into static HTML pages with the infrastructure described above.
+The tutorials span a range of astronomical topics, from general tasks like
+reading or creating FITS files with different content or working with
+astronomical coordinate systems, to more specific exercises like analyzing
+spectroscopic data from the UVES instrument.
+Now that the backend infrastructure is stable, we are interested in collecting
+new content to serve and share: If you have ideas for contributions, or have
+tutorials hosted on third-party sites that you would like to make searchable
+through the Learn Astropy interface, please reach out by creating a \github
+issue in the \href{https://github.com/astropy/astropy-tutorials}{Astropy
+Tutorials repository}.
+We currently host one Guide --- a longer-form walkthrough of a more complex
+concept or topic --- that is focused on
+\href{http://www.astropy.org/ccd-reduction-and-photometry-guide}{CCD image
+reduction and photometry}, but are also actively seeking new material that would
+be suitable for new guides.
+
+Beyond developing and serving educational content, the \astropy Project has been
+conducting workshops at winter meetings of the American Astronomical Society
+since AAS 225 in January 2015.
 Up to the start of the coronavirus pandemic, these were full-day in-person
@@
 The Learn team finds that the workshop audience is best found as AAS meetings as
-opposed to more general Python meetings.
-
-The Astropy Project provided another mode of community engagement at AAS
-Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
-AAS Exhibit Hall.
+opposed to more general \python meetings, as the content tends to be more
+applicable for students and researchers in astronomy and astrophysics.
+
+The Astropy Project recently provided another mode of community engagement at
+AAS Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in
+the AAS Exhibit Hall.
 Funding for the exhibit hall was provided alternately by NumFOCUS and later by
-the Moore Foundation funding (\textbf{check this}).
+the Moore Foundation funding.
 The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
@@
 tools.
-\end{description}
 
@@
 
-The Learn plan going forward is to continue to improve the tutorial website and
-facilate new content;
-to index third-party tutorials; and to look for supportable opportunities to
-expand the reach of Astropy workshops beyond the American Astronomical Society
-meetings.
+The focus of Learn Astropy over the coming months and years is to further
+improve the interface for sharing educational content on the main Learn Astropy
+website, and to facilitate the development of new content that highlights
+\astropy and \astropy-affiliated package functionality.
+However, we simultaneously also plan to solicit indexing and ingestion of
+third-party tutorial series, to provide a unified interface for identifying
+astronomy-specific educational content that demonstrates the functionality
+available in the vast ecosystem of open-source software packages.
+We also plan to look for opportunities to expand the reach of \astropy workshops
+beyond the AAS meetings.
 

### 2022-03-15 src/ms.tex
acknowledgements. Closes #41
shuffle Gemini ack

@@
 
+We acknowledge the Gordon and Betty Moore foundation for their continued
+financial support.
+This work is partially supported by NASA under Grant No. 20-OSTFL20-0057 issued
+through the NASA ROSES program.
+This work is partially supported by the international Gemini Observatory, a
+program of NSF's NOIRLab, which is managed by the Association of Universities
+for Research in Astronomy (AURA) under a cooperative agreement with the National
+Science Foundation, on behalf of the Gemini partnership of Argentina, Brazil,
+Canada, Chile, the Republic of Korea, and the United States of America.
+We also thank NumFOCUS and the Python Software Foundation for financial
+support.
+
 The \astropy community is supported by and makes use of a number of
@@
 summer. These students often turn into long-term contributors.
-We also thank NumFOCUS and the Python Software Foundation for financial
-support.
 Within the academic community, we thank institutions that make it possible that
@@
 Observatory.
-
-Supported by the international Gemini Observatory, a program of NSF's NOIRLab,
-which is managed by the Association of Universities for Research in Astronomy
-(AURA) under a cooperative agreement with the National Science Foundation, on
-behalf of the Gemini partnership of Argentina, Brazil, Canada, Chile, the
-Republic of Korea, and the United States of America.
 

### 2022-03-15 src/ms.tex
remove duplicate numpy ref

--- 2022-03-15 src/ms.tex
@@
 
-Overview and statistics of contributors. Highlight changes since v2.0.
-
-\begin{figure}
+% TODO: move this subsection to "updates to core package" section above?
+
+\begin{figure}[t!]
     \begin{centering}
-      % \includegraphics{figures/contributor-summary.pdf}
-        \caption{Placeholder figure!}
-        \label{fig:contributor-summary}
+      \includegraphics[width=0.9\textwidth]{Ncommitters_vs_frac_commits.pdf}
+        \caption{
+            The cumulative fraction of Git commits to the \astropypkg core
+            package as a function of number of contributors (committers) broken
+            into three equal-length time periods between the start of \astropy
+            and the feature freeze date of v5.0.
+            The time eras early, mid, and recent correspond to the approximate
+            year ranges 2011--2015, 2015--2018, 2018--2021 (defined more
+            precisely in Section~\ref{sec:project-contributors}).
+            In all eras, the majority of commits to the \astropypkg core package
+            have come from a small subset of the total number of contributors.
+            For example, as annotated in the figure, 90\% of the Git commits
+            have come from 18, 34, and 44 committers in the early, mid, and
+            recent eras, respectively.
+        }
+        \label{fig:contributor-summary:Ncommitters}
     \end{centering}
 \end{figure}
+
+As described in the introduction above, the \astropy Project is broader than the
+core \astropypkg package as it also represents a number of core infrastructure
+packages, affiliated packages, educational initiatives, and user communities.
+In total, well over 1500 people have contributed in some way to the Project,
+whether, for example, through contributing code or documentation, opening issues
+on \github, providing infrastructure support, writing tutorials, or helping to
+coordinate workshops that relate to Astropy.
+The vast majority of these contributors and participants have interacted as
+volunteers choosing to engage with the open-source and open-development software
+community.
+However, while the total number of contributors is large, the number of
+contributions and bulk of the maintenance of the \astropy Project components has
+primarily been sustained by a small number of heavily-engaged or
+explicitly-employed contributors \citep[see, e.g., ][]{astropy:2018}, as is the
+case in many large, open-source software communities
+\citep[e.g.,][]{numpy:nature}.
+
+As an example of this within the \astropy Project, here we focus on
+contributions made to the \astropypkg core package.
+Figure~\ref{fig:contributor-summary:Ncommitters} shows the number of
+contributors who have contributed a given fraction of the total number of Git
+commits, separated into three separate time spans (``eras'') in the history of
+the Project.\footnote{Note, however, that measuring these statistics in terms of
+Git commits is not a perfect way of encapsulating ``effort,'' but we use this
+metric for its simplicity and clear definition.}
+The annotations show that, within the early (July 2011--January 2015), mid
+(January 2015--June 2018), and recent (June 2018--November 2021) eras, 90\% of
+the commits to the \astropypkg core package have come from 18, 34, and 44
+contributors.
+While the trend is encouraging --- a larger pool of contributors are
+contributing more, fractionally --- this emphasizes that, still, the vast
+majority of the work in \astropypkg is done my a small subset ($<20\%$) of the
+115, 216, and 234 contributors, respectively, within each era.
+
+\begin{figure}[t!]
+    \begin{centering}
+      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
+        \caption{
+            The career stages and/or positions of the top contributors within
+            each time era (defined as having contributed $>50$ commits to the
+            \astropypkg core package in each era).
+            The time eras early, mid, and recent correspond to the approximate
+            year ranges 2011--2015, 2015--2018, 2018--2021 (defined more
+            precisely in Section~\ref{sec:project-contributors}).
+        }
+        \label{fig:contributor-summary:top}
+    \end{centering}
+\end{figure}
+
+Since its inception, the most active contributors to the \astropypkg package
+have spanned a range of career stages and career paths.
+For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
+career stage/path at the midpoint of the three time eras defined in the previous
+paragraph for all contributors with more than 50 commits to the \astropypkg
+package within each era.
+Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
+contributors moved between these categories (i.e., have graduated as students to
+become research staff / faculty or software engineers), but also a number of
+early contributors moved on to other positions and stopped contributing to the
+core \astropypkg package.
+Given this, it is encouraging that the numbers in each career stage/path have
+stayed relatively constant over the history of \astropy (with some showing
+growth), as new, active contributors have generally joined and stepped in to
+fill top contributor roles within the project.
+On the other hand, the total number of top contributors has not grown
+substantially, reflecting one of our principal challenges: Sustaining a pipeline
+from users to contributors to maintainers of the package.
+One other thing to note in this context is that in all eras, 50--60\% of the
+software engineers who contributed to \astropypkg were (or are) employed by
+STScI, whose leadership have consistently provided resources (financial and
+personnel) towards sustaining \astropypkg.
+
+\begin{figure}[t!]
+    \begin{centering}
+      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
+        \caption{
+            \textbf{Top panel:} The number of Git commits per month (defined as
+            consecutive 30 day windows) merged into to the \astropypkg core
+            package over the whole history of the project (up to end of 2021,
+            v5.0).
+            The dark line shows the mean rate using a 1 year rolling window, but
+            the under-plotted gray line shows the raw monthly rate.
+            The raw number clearly peaks just before feature freezes (vertical
+            lines) as a flurry of activity leading up to each release, but in
+            recent years the mean rate has remained steady at about $\sim$300
+            commits per month.
+            \textbf{Bottom panel:} Similar to above, but now showing the number
+            of unique Git committers who contribute to the \astropypkg core
+            package each month (again defined as consecutive 30 day windows).
+            While this also shows some variation depending on proximity to
+            feature freezes, in recent years the mean rate of unique
+            contributors has remained consistent at about $\sim$22 committers
+            per month.
+        }
+        \label{fig:contributor-summary:per-month}
+    \end{centering}
+\end{figure}
+
+A final metric we consider as a diagnostic of the contributor base and
+contributions to the \astropypkg core package is the rate of contributions and
+contributor participation, defined again using Git commits to quantify
+contributions.
+Figure~\ref{fig:contributor-summary:per-month} shows the monthly rates of
+commits and unique committers to the \astropypkg core package over the history
+of the package.
+After an initial ramp-up in the commit rate and number of committers (lasting
+roughly from 2011 to 2014), the commit rate and number of unique committers have
+both remained fairly constant.
 
@@
     \tablecaption{Registry of affiliated packages.}
-    \label{tab:registry}
+    \label{tab:affiliated-registry}
     \tablehead{

### 2022-03-15 src/ms.tex
Add draft conclusions section. Closes #44

--- 2022-03-15 src/ms.tex
@@
 Empowerment) initiatives for project-wide consideration on
-Github. This process was deemed the most ``open'' because it allowed
+\github. This process was deemed the most ``open'' because it allowed
 for community-wide feedback to focus and improve proposal
@@
 
-% \secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
+\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}
 
@@
 educational efforts made by the Astropy Project, which are led by the Learn
-Astropy Team.
+Team.
 The efforts focus on developing online content and workshops covering
@@
 that showcase a short task that can be performed with Astropy or an affiliated
-package; and \textit{documentation}, which contains more detailed information
-about the code base and user interface.
+package; and \textit{documentation}, which is contained within the code base.
 This categorization drives content development, infrastructure choices, and the
-appearance of the \href{https://learn.astropy.org}{Learn Astropy website}.
+appearance of the Learn Astropy website.
 The Learn Team meets weekly to work on creating, expanding, improving these
@@
 
-The Learn Team recently re-launched the main website and search interface for
-Learn Astropy in 2021 with a new infrastructure platform, built around full-text
-search and interactive filtering functionality, with the goal of making content
-more easily searchable and discoverable as the Learn Astropy content catalog
+\begin{description}
+
+\item[Learn Astropy website] The Learn Team re-launched the main website and
+search interface for Learn Astropy in 2021 with a new infrastructure platform,
+built around full-text search and interactive filtering functionality, with the
+goal of making content discoverable as the Learn Astropy content catalog
 expands.
@@
 uses the Algolia service to power its search and filtering user interface.
-Our \python-based application,
-\href{https://github.com/astropy/learn-astropy-librarian/}{Learn Astropy
-Librarian}, populates data into the Algolia service. We tuned the Librarian
-around specific content formats (such as Jupyter Notebook-based tutorial pages
-and Jupyter Book-based guides) to more accurately index content and
-heuristically extract metadata.
+Our Python-based application, Learn Astropy Librarian, populates data into the
+Algolia service. We tuned the Librarian around specific content formats (such as
+Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
+accurately index content and heuristically extract metadata.
 A consequence of the new platform is that we now maintain and compile content
@@
 
-We currently host 19 tutorials written as interactive Jupyter Notebooks and
-rendered into static HTML pages with the infrastructure described above.
-The tutorials span a range of astronomical topics, from general tasks like
-reading or creating FITS files with different content or working with
-astronomical coordinate systems, to more specific exercises like analyzing
-spectroscopic data from the UVES instrument.
-Now that the backend infrastructure is stable, we are interested in collecting
-new content to serve and share: If you have ideas for contributions, or have
-tutorials hosted on third-party sites that you would like to make searchable
-through the Learn Astropy interface, please reach out by creating a \github
-issue in the \href{https://github.com/astropy/astropy-tutorials}{Astropy
-Tutorials repository}.
-We currently host one Guide --- a longer-form walkthrough of a more complex
-concept or topic --- that is focused on
-\href{http://www.astropy.org/ccd-reduction-and-photometry-guide}{CCD image
-reduction and photometry}, but are also actively seeking new material that would
-be suitable for new guides.
-
-Beyond developing and serving educational content, the \astropy Project has been
-conducting workshops at winter meetings of the American Astronomical Society
-since AAS 225 in January 2015.
+\item[Tutorials] Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
+
+\item[Guides] {\it ccdproc} Jupyter book has been completed.
+
+\item[Workshops] The Project has been conducting workshops at winter
+meetings of the American Astronomical Society since AAS 225 in January 2015.
 Up to the start of the coronavirus pandemic, these were full-day in-person
@@
 The Learn team finds that the workshop audience is best found as AAS meetings as
-opposed to more general \python meetings, as the content tends to be more
-applicable for students and researchers in astronomy and astrophysics.
-
-The Astropy Project recently provided another mode of community engagement at
-AAS Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in
-the AAS Exhibit Hall.
+opposed to more general Python meetings.
+
+The Astropy Project provided another mode of community engagement at AAS
+Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
+AAS Exhibit Hall.
 Funding for the exhibit hall was provided alternately by NumFOCUS and later by
-the Moore Foundation funding.
+the Moore Foundation funding (\textbf{check this}).
 The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
@@
 tools.
+\end{description}
 
@@
 
-The focus of Learn Astropy over the coming months and years is to further
-improve the interface for sharing educational content on the main Learn Astropy
-website, and to facilitate the development of new content that highlights
-\astropy and \astropy-affiliated package functionality.
-However, we simultaneously also plan to solicit indexing and ingestion of
-third-party tutorial series, to provide a unified interface for identifying
-astronomy-specific educational content that demonstrates the functionality
-available in the vast ecosystem of open-source software packages.
-We also plan to look for opportunities to expand the reach of \astropy workshops
-beyond the AAS meetings.
+The Learn plan going forward is to continue to improve the tutorial website and
+facilate new content;
+to index third-party tutorials; and to look for supportable opportunities to
+expand the reach of Astropy workshops beyond the American Astronomical Society
+meetings.
 
@@
 
+\section{Conclusions}
+\label{sec:conclusions}
+
+The Astropy Project has continued to serve as an active, community-oriented
+software project with a wide reach and growing impact, both within astronomy and
+astrophysics and in broader scientific software communities.
+The Project is in the midst of setting up infrastructure to ensure its long-term
+sustainability and to continue healthy growth of the software ecosystem,
+developer, and user communities that interface with the Project.
+For example, over the last years, we have formalized and implemented a new
+governance model for the Astropy Project (Section~\ref{sec:project-governance}),
+have continued to engage a growing pool of active and top contributors
+(Section~\ref{sec:project-contributors}), and have begun to receive significant
+formal funding for the project itself (Section~\ref{sec:project-funding}).
+The Astropy Project is also expanding its educational and community-focused
+initiatives, by soliciting and funding initiatives in inclusion, diversity, and
+equity (Section~\ref{sec:project-ide}) and through the Learn Astropy initiative,
+which aims to make tutorials and educational content that highlight software
+tools for astronomy more discoverable and accessible (Section~\ref{sec:learn}).
+We expect to continue in all of these threads over the coming years as we aim to
+make the Astropy Project more sustainable and a larger, more welcoming community
+for open-source software users and enthusiasts in astronomy.
+
 \begin{acknowledgments}

### 2022-03-15 src/ms.tex
comment out affil pkgs

@@
 Empowerment) initiatives for project-wide consideration on
-\github. This process was deemed the most ``open'' because it allowed
+Github. This process was deemed the most ``open'' because it allowed
 for community-wide feedback to focus and improve proposal
@@
 
-\section{Conclusions}
-\label{sec:conclusions}
-
-The Astropy Project has continued to serve as an active, community-oriented
-software project with a wide reach and growing impact, both within astronomy and
-astrophysics and in broader scientific software communities.
-The Project is in the midst of setting up infrastructure to ensure its long-term
-sustainability and to continue healthy growth of the software ecosystem,
-developer, and user communities that interface with the Project.
-For example, over the last years, we have formalized and implemented a new
-governance model for the Astropy Project (Section~\ref{sec:project-governance}),
-have continued to engage a growing pool of active and top contributors
-(Section~\ref{sec:project-contributors}), and have begun to receive significant
-formal funding for the project itself (Section~\ref{sec:project-funding}).
-The Astropy Project is also expanding its educational and community-focused
-initiatives, by soliciting and funding initiatives in inclusion, diversity, and
-equity (Section~\ref{sec:project-ide}) and through the Learn Astropy initiative,
-which aims to make tutorials and educational content that highlight software
-tools for astronomy more discoverable and accessible (Section~\ref{sec:learn}).
-We expect to continue in all of these threads over the coming years as we aim to
-make the Astropy Project more sustainable and a larger, more welcoming community
-for open-source software users and enthusiasts in astronomy.
-
 \begin{acknowledgments}
@@
           \package{matplotlib} \citep{matplotlib},
-          \package{Cython} \citep{cython},
-          \package{photutils} \citep{photutils},
-          \package{specutils} \citep{specutils},
-          \package{regions} \citep{regions},
-          \package{gwcs} \citep{gwcs},
-          \package{synphot} \citep{synphot},
-          \package{ASDF} \citep{ASDF}.
+          \package{Cython} \citep{cython}.
+          % \package{photutils} \citep{photutils},
+          % \package{specutils} \citep{specutils},
+          % \package{regions} \citep{regions},
+          % \package{gwcs} \citep{gwcs},
+          % \package{synphot} \citep{synphot},
+          % \package{ASDF} \citep{ASDF}.
           % \package{SOFA} \citep{sofa},

### 2022-03-16 src/ms.tex
working on author list
trying to work out author list issues

first pass at validating author list

reorder using authorship ordering

read emails from file instead

remove emails from public branch

--- 2022-03-15 src/ms.tex
@@
 \input{preamble.tex}
+\input{generated/git-header.tex}
 
@@
 
-\draft{\today}
+\let\oldtoday\today
+\renewcommand{\today}{\oldtoday;\qquad\textnormal{\gitHeader}}
 
@@
 
-\author{Astropy Collaboration}
-\noaffiliation
-{\let\thefootnote\relax\footnote{{The author list has two parts: the authors that made significant contributions to the writing of the paper in order of contribution, followed by contributors to the \astropy Project in alphabetical order.
-\textbf{The position in the author list does not correspond to contributions to the \astropy Project as a whole.} A more complete list of contributors to the core package can be found in the \href{https://github.com/astropy/astropy/graphs/contributors}{package repository}, and at the \href{http://www.astropy.org/team.html}{\astropy team webpage}.}}}
-
-% \input{author.tex}
+\author{Astropy~Collaboration}
+\altaffiliation{The author list has two parts: the authors that made significant
+contributions to the writing of the paper in order of contribution, followed by
+contributors to the Astropy Project in alphabetical order. \textbf{The position
+in the author list does not correspond to contributions to the Astropy Project
+as a whole.} A more complete list of contributors to the core package can be
+found in the
+\href{https://github.com/astropy/astropy/graphs/contributors}{package
+repository}, and at the \href{http://www.astropy.org/team.html}{Astropy team
+webpage}.}
+
+\input{static/authors.tex}
 
 \begin{abstract}
-The \astropy Project supports and fosters the development of open-source and openly-developed
-\python packages that provide commonly-needed functionality to the astronomical
+The \astropy Project supports and fosters the development of open-source and openly developed
+\python packages that provide commonly needed functionality to the astronomical
 community.
@@
 In this article, we summarize key features in the core package as of the recent major
-release, version 5.0, and provide major updates for the project.
-We then discuss supporting a broader ecosystem of inter-operable packages,
+release, version 5.0, and provide major updates for the Project.
+We then discuss supporting a broader ecosystem of interoperable packages,
 including connections with several astronomical observatories and missions.
@@
 status of Learn Astropy.
-We conclude by raising and discussing the current and future challenges facing the project.
+We conclude by raising and discussing the current and future challenges facing the Project.
 \end{abstract}
@@
 Over the last decade, \python has grown enormously in popularity to become a
-dominant programming language also in the astronomical and broader scientific
+dominant programming language, especially in the astronomical and broader scientific
 communities.
 For example, Figure~\ref{fig:python-mentions} shows the number of yearly
-full-text mentions of \python as compared to a few other programming languages
+full-text mentions of \python as compared to a selection of other programming languages
 in refereed articles in the astronomical literature, demonstrating its nearly
@@
 One of the factors that has led to its rapid ascent in popularity in scientific
-contexts has been the significant, volunteer-driven effort behind developing
-community-oriented open-source software tools and fostering communities of users
-and developers that have grown around these efforts.
+contexts is the combination of volunteer-driven and professionally-supported
+effort behind developing community-oriented open-source software tools and
+fostering communities of users and developers that have grown around these
+efforts.
 Today, a broad and feature-diverse ``ecosystem'' of packages exists in the
-\python scientific computing landscape: Roughly ordered from general-use to
+\python scientific computing landscape: roughly ordered from general-use to
 domain-specific, this landscape now includes packages that provide core
@@
 The \astropypkg \citep{astropy:2013, astropy:2018} core package began in this
-vein, as an effort to consolidate the development of commonly-used functionality
+vein, as an effort to consolidate the development of commonly used functionality
 needed to perform astronomical research into a community-developed \python
@@
 An early description of the core functionality in \astropypkg can be found in
-the first \astropy paper \citep{astropy:2013} or in detail in the core package
-documentation.\footnote{\url{https://docs.astropy.org/}}
+the first \astropy paper \citep{astropy:2013} or in detail in the
+\href{https://docs.astropy.org/}{core package documentation}.
 The codebase in the \astropypkg core package is now largely stable, in that the
@@
 This is largely driven by the fact that the core package now represents just one
-piece of the broader astronomy \python context, and much new feature development
+piece of the broader astronomy \python context, and new feature development
 is now happening in more specialized packages that are expanding the
@@
 \astropypkg core package, the ecosystem of astronomy-specific software tools
-that are interoperable with \astropypkg (Astropy Affiliated Packages),
+that are interoperable with \astropypkg (Astropy Coordinated and Affiliated Packages),
 \emph{and} the community of users, developers, and maintainers that participate
 in Astropy efforts.
-However, there is no institution responsible for managing the Astropy Project,
+However, there is no single institution responsible for managing the Astropy Project,
 for funding or maintaining its development, or sustaining it into the future:
-The Project is maintained and coordinated largely by volunteers.
+The Project is maintained and coordinated largely by volunteers and contributors
+who are primarily paid for other reasons (i.e., Astropy work is just a means to an end).
 While new Astropy-affiliated packages are being developed that expand upon the
@@
 Project are evolving.
-In particular, the transition from focusing our energy on development and
-maintenance of a single core package, to instead sustaining the core package and
-fostering the development of the community and its expansion has been a key
-issue faced by the Astropy Project in the last several years.
+In particular, fostering the development of the community and its expansion has
+been an additional key goal of the Astropy Project over the last several years.
 
@@
 released. However, every two years a major release is designated as a long-term
-support (LTS) release, which continues to be maintained for up to two years
+support (LTS) release, which continues to be maintained for at least two years
 \citep{ape2}. The motivation for LTS releases is to provide longer-term stable
-versions of \astropypkg that users that require a high level of stability can
-make use of if they do not always need the latest features (this could include,
-for example, telescope operation or data reduction pipelines, etc.). The v5.0
-release of the core package was designated as LTS, and since it was released at
-the end of 2021, it will be maintained until the end of 2023.
+versions of \astropypkg that users who require a high level of stability can
+make use of if they do not always need the latest features (this includes but
+is not limited to telescope operations and data reduction pipelines). The v5.0
+release of the core package, replacing v4.0, was designated as LTS at the end of 2021;
+it will be maintained until the end of 2023.
 
@@
 \python libraries have come to rely on \astropypkg. The relationship between
-\astropy and the \python-using community is reciprocal, with the evolving needs
+\astropy and the broader \python community is reciprocal, with the evolving needs
 of the community and the maturation of the broader scientific \python ecosystem
@@
 
-Within \astropypkg, development may be roughly split into a few
-categories: new features added, intra- and inter-package interoperability,
-improvements to precision, accuracy, and reproducibility, and performance
-enhancements. We discuss these in the subsections below.
+Within \astropypkg, development may be roughly split into several
+categories: new features, intra- and inter-package interoperability,
+improvements to numerical precision, accuracy, and reproducibility, performance
+and documentation enhancements, and bug fixes. We discuss some of these in the subsections below.
 
@@
 Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
-carried, as well as the \python wrappers for it. Since this is basic
+carried, as well as the associated \python wrappers. Since this is basic
 infrastructure, with a release cycle set by SOFA, it made more sense to create
@@
 was too specialized was a Virtual Observatory sub-package; the Simple
-Application Message Protocol was kept in the core package
-(\astropysubpkg{samp}), but other parts more properly belonged in
+Application Message Protocol (\astropysubpkg{samp}) is kept in the core package
+for now, but other parts more properly belonged in
 \package{astroquery} (see Section~\ref{sec:astroquery}).
@@
 New features present a particular difficulty for a mature package such as
-\astropypkg, because it is difficult to know {\em a priori} what the best
+\astropypkg, because it is challenging to know {\em a priori} what the best
 interface will be, and hard to address all possible use cases in one round of
@@
 is spanned by five main new features that entered the \astropypkg core since the
-previous summary \citepalias{astropy:2018}. In our descriptions of each we
+previous summary \citepalias{astropy:2018}. In our descriptions of each, we
 include how we hope the new feature will evolve. Ultimately, in a user-driven
@@
 
-\paragraph{Uncertainties and Distributions} Early in our careers we were likely
-all taught that measurements without units and uncertainties are meaningless.
+\paragraph{Uncertainties and Distributions} In scientific analysis,
+measurement values with units and uncertainties are essential.
 \astropyunits provides the ability to associate numbers with units, and
@@
 \astropysubpkg{nddata}, there are options to associate errors with data arrays,
-but these are treated as independent, i.e., covariances are not tracked.
+but covariances are not tracked.
 
@@
 
-A plan for future development of \astropysubpkg{uncertainties} is to also
-support error propagation, tracking covariances, for the case that uncertainties
+A plan for future development of \astropysubpkg{uncertainties} is to
+support error propagation: tracking covariances for the case that uncertainties
 are normally distributed rather than approximated by Monte-Carlo means. One
@@
 points implies that all data points are now covariant with each other, i.e., one
-has to carry $N\times N$ covariances. For a large image, that becomes not just
-pointless but also prohibitive in terms of memory and CPU usage.
-
-\paragraph{Masked quantities}  It can be useful to mask bad data. How one
-supports masked data involves a number of choices, such as whether one indicates
-data values are bad by a separate flag or by replacing bad data with a special
-value, such as ``Not a Number'' (NaN). In \astropypkg, both approaches have
-been used: a flag for masked columns in \astropyTable, mask flags and bitmaps
+has to carry $N\times N$ covariances. For a large image, that becomes unnecessary
+and computationally expensive.
+
+\paragraph{Masked quantities}  It can be useful to mask bad data, either by marking
+bad data values with a flag or by replacing them with a special
+value, e.g., ``Not a Number'' (NaN). In \astropypkg, both approaches are
+used: a flag for masked columns in \astropyTable, mask flags and bitmaps
 for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
-NaN in \astropyTime. However masked quantities were poorly supported, making it
-difficult to use masks in a \astropyQTable\ (in which \astropyQuantity\ is used
+NaN in \astropyTime. However, masked \astropyQuantity\ support is currently very limited,
+hindering usage of masks in a \astropyQTable\ (in which \astropyQuantity\ is used
 for all columns with units).
 
-Unfortunately, the \texttt{MaskedArray} class from \package{numpy} only works
+Furthermore, the \texttt{MaskedArray} class from \package{numpy} only works
 well with plain data arrays, hiding other metadata and attributes, for example
@@
 The masked quantities work largely without any changes in higher-level objects
-such as \astropySkyCoord, but work is still underway to expose the mask in those
-classes, as well as to use masked arrays in \astropyTime\ instead of replacing
-data with NaN.
+such as \astropySkyCoord, but more work is needed to better integrate it with
+the rest of \astropypkg, e.g., using masked arrays in \astropyTime\ instead of NaN.
 
@@
 particular time bins. The new classes offer a number of special methods to
-manipulate time series (folding, resampling, etc.) and to read different data
-formats (such as \mission{Kepler} light curves, etc.).
+manipulate time series (e.g., folding and resampling) and to read different data
+formats (e.g., \mission{Kepler} light curves).
 
@@
 correspondence between ``real-world'' and pixel coordinates. This mapping is the
-essence of the ``World Coordinate System (WCS)'' concept. From its inception,
+essence of the World Coordinate System (WCS) concept. From its inception,
 the \astropysubpkg{wcs} sub-package allowed access to WCS information provided
@@
 well with other \astropypkg\ objects such as \astropySkyCoord\ and \astropyTime.
+Future work includes better integration of this interface with the rest of
+the Project.
 
@@
 both within the various \astropypkg sub-packages and with other \python
-libraries. If done right, all a person notices is that code using many different
-python features and libraries ``just works.'' Here, we describe a few particular
+libraries. We aim for a seamless user experience, where different code blocks
+would ``just work'' when put together. Here, we describe a few particular
 efforts towards this goal.
@@
 would strip a \astropyQuantity\ of its units (or fail outright), limiting
-\astropyQuantity's usefulness. Now, advances in \package{numpy} function
+\astropyQuantity's potential. Now, advances in \package{numpy} function
 overloading (e.g.,
@@
 
-Units were also integrated further within \astropypkg. For an already-defined unit-less
-\astropyModel, e.g. those imported from another library, \astropymodeling can
-now coerce units.
-% TODO! transition words
-A new module – \astropycosmologyunits – has been added to the cosmology
-sub-package for defining and collecting cosmological units and equivalencies.
-The unit \texttt{littleh} and equivalency
-\astropyapidoc{cosmology.units.with_H0}{with\_H0} was moved from the main
-\astropyunits sub-package to \astropycosmologyunits. A new unit,
-\texttt{redshift}, has been added for tracking factors of cosmological redshift.
-To correctly use redshift units in dimensional analysis, \texttt{redshift} is
-\astropyapi{cosmology.units.dimensionless_redshift}{treated as dimensionless}
-by default. To convert between redshift and other cosmological distance
-measures, e.g. CMB temperature or comoving distance, the equivalency
-\astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
-has also been added. This equivalency is actually a composite of other equivalencies,
-which may be used separately.
+Units were also integrated further within \astropypkg. For an already-defined
+unit-less \astropyModel, e.g., those imported from another library,
+\astropymodeling can now coerce units. A new module – \astropycosmologyunits –
+has been added to the cosmology sub-package for defining and collecting
+cosmological units and equivalencies. \astropycosmologyunits\ has a unit and
+equivalencies for \texttt{littleh}, and a new unit, \texttt{redshift}, for
+factors of cosmological redshift. \texttt{redshift} has a number of
+equivalencies for converting between cosmological distance measures, e.g., CMB
+temperature or comoving distance.
 
@@
 
-Within \astropypkg and \python there are numerous ways to represent and store
+Within \astropypkg and \python, there are numerous ways to represent and store
 array-valued data. Some of the differences are historical:
@@
 different use cases. Lastly, some differences are inherent: \astropyQuantity,
-\astropyTime\, and \astropySkyCoord\ represent fundamentally different types of
+\astropyTime, and \astropySkyCoord\ represent fundamentally different types of
 objects.
@@
 We aim to make it possible for any array-valued data to be used as a column in a
-table. A well defined protocol for mixin-columns has been developed for
+table. A well-defined protocol for mixin-columns has been developed for
 \astropytable, allowing the original object to be used as a column with a
@@
 attributes. With this protocol, it is now possible to store \astropy native
-objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord --
+objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord\ --
 within a \astropyTable\ and write these to various file formats, such as FITS.
@@
 The FITS standard was extended to rigorously describe time coordinates in the
-World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared to
+WCS framework \citep{FITS-Time:2015}. Compared to
 other types of coordinates in WCS, time requires more metadata: format, scale,
@@
 \href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
-Summer of Code} 2017 program,\footnote{ To learn more about this project, please
+Summer of Code} 2017 program,\footnote{To learn more about this project,
 see the final report
 \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
-mixin protocol for seamless interoperability}.}, the \astropyfits package now
-interprets time data correctly, using \astropyTime\ as a `mixin' column. For
-backward compatibility with manual systems, this feature may be turned off.
+mixin protocol for seamless interoperability}.} the \astropyfits package now
+interprets time data correctly, using \astropyTime\ as a mixin column. For
+backward compatibility, this feature may be turned off.
 
@@
 \astropy now supports reading and writing tables in the American Astronomical
-Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
-long been missing and as added in this year's Google Summer of Code project adds
-\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}
-to \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
-to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
+Society Journals' Machine-Readable Table (\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}) format. This ASCII format has
+long been missing and was added to \astropyTable\ I/O in a Google Summer of Code 2021 project.
+In addition, \astropyTable\ may also read from and write
+to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},\footnote{Starting in v5.1,
+direct \astropypkg support for ASDF has been deprecated in favor of moving it
+to the new \package{asdf-astropy} \citep{asdf-astropy} package to accommodate
+the differing release schedules between ASDF and \astropypkg.}
 \href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
@@
 The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
-standard} has been updated to version 1.0, adding support three additional data
-subtypes. First, it adds multidimensional column data (both masked and unmasked)
-with fixed dimensions in all table cells. Second, it adds multidimensional
+standard} has been updated to v1.0, adding support of three additional data
+subtypes: multidimensional column data (both masked and unmasked)
+with fixed dimensions in all table cells, multidimensional
 column data with variable-dimension arrays similar to FITS variable-length
-arrays. Third, it adds object-type columns with simple \python objects.
+arrays, and object-type columns with simple \python objects.
 
 As mentioned previously, \package{dask} arrays may be used as a mixin column in
-\astropyTable. Now, \package{dask} may also be the data array in
-\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the array
-will be computed while written, avoiding excessive memory use. \astropyTable\ can
-now be appended to an existing FITS file, and \package{dask} mixin columns
-interoperate seamlessly between the two table types.
+\astropyTable. Now, \package{dask} may also be a data array in
+\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only
+computed while written, avoiding excessive memory use. Furthermore, \astropyTable\ can
+be appended to an existing FITS file, where \package{dask} mixin columns can
+interoperate seamlessly between them.
 
@@
 functions for \astropypkg objects, such as reading from or writing to a file,
-following a given standard. \astropy uses this internally for the I/O methods in
-\astropyTable\ and \astropyCosmology, and users can register custom I/O to
+following a given standard. \astropypkg uses this internally for the I/O methods in
+\astropyTable\ and \astropyCosmology. Users can also register custom I/O to
 extend the options on these classes.
 
-As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
-number of new use cases. for instance, the class-based architecture allows for
-the creation of custom registries. One application is in \astropycosmology,
+As of \astropypkg v5.0, the I/O registry submodule has been generalized to enable a
+number of new use cases; for instance, the class-based architecture allows
+streamlining of the creation of custom registries while minimizing code duplication.
+One application is in \astropycosmology,
 which has two different \textit{kinds} of registries for \astropyCosmology: one
-for reading and writing files, and another for converting between python
-objects. Using \astropyiosubpkg{registry} for custom classes means a developer
-does not have to create a whole new I/O system and can offer to the user a
-unified and familiar API.
-
-\subsection{Precision, Accuracy, \& Reproducibility}
+for reading and writing files, and another for converting between \python objects.
+
+\subsection{Numerical Precision, Accuracy, \& Reproducibility}
 \label{sec:core-features-precision_accuracy_repreducibility}
 
-An important focus of \astropy has been to increase the precision and accuracy
+An important focus of \astropypkg has been to increase the numerical precision and accuracy
 of its functionality, while, where appropriate, making sure results obtained
-using older versions of \astropypkg are reproducible. Within each section below
+using older versions of \astropypkg are reproducible. Within each section below,
 we include in our descriptions how old results may be reproduced. The primary
@@
 In astronomy, time accuracy down to the (nano)second is frequently important.
-Whether for planning observations, crunching pulsar or VLBI data, missing
+Whether for planning observations or crunching pulsar or VLBI data, missing
 seconds meaningfully impact results. In past versions of \astropypkg, when using
-\astropyTime\ in \astropytime leap seconds had to be manually applied; now, leap
+\astropyTime, leap seconds had to be manually applied; now, leap
 seconds are applied automatically. Moreover, \astropypkg updates internal time
-data-files to ensure that the correct leap second adjustment is always used. For
-reproducibility, e.g., with an old code, \astropyLeapSeconds\ may also be
+data files to ensure that the correct leap second adjustment is always used. For
+reproducibility with older code, \astropyLeapSeconds\ may also be
 manually applied.
@@
 \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
-2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
+2018}}. \astropyconstants now defaults to using the \texttt{SI/CODATA 2018}
 values, with the units in \astropyunits based on these constants. Most of
-\astropypkg and affiliate packages build upon \astropyunits, so this update
-effects the entire \astropy ecosystem.
+\astropy and affiliated packages build upon \astropyunits, so this update
+affects the entire \astropy ecosystem.
 
 For reproducibility, \astropypkg allows the constants' (and therefore units')
-definitions to be rolled back to prior values, e.g., to the \texttt{SI/CODATA
-2014} values. For work sensitive to the values of the fundamental constants we
-recommend including an \astropy configuration file with the work, which
-specifies the set of constants used.
+definitions to be rolled back to prior values, e.g., to \texttt{SI/CODATA
+2014}. For work sensitive to the values of the fundamental constants, we
+recommend including an \astropypkg configuration file (i.e., \texttt{astropy.cfg})
+with the work, which specifies the set of constants used.
 
@@
 
-Astropy provides a
+\astropypkg provides a
 \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
@@
 
-Being able to describe the position of objects, i.e., their coordinates, is
+Being able to describe the coordinates of objects is
 fundamental across many knowledge domains. \astropycoordinates allows one to
 work with low-level positional and velocity data all the way to high-level
-coordinate objects with reference frames, atmospheric information, and more. A
+coordinate objects with reference frames, atmospheric information, etc. A
 central feature of \astropycoordinates is the ability to transform data between
@@
 solar system. Transformations in the \astropyAltAz\ frame were reasonably
-precise for very distant objects, but wrong by up to several arcseconds for
+precise for very distant objects, but wrong by up to several arcseconds for,
 e.g., the location of the moon. Now these transformations are much more precise,
@@
 \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
-and associated transformations have been updated to correctly reflect the “true”
-and “mean” terminology.
+and associated transformations have been updated to correctly reflect the ``true''
+and ``mean'' terminology.
 
@@
 
-As features are added and corner cases dealt with, performance often suffers.
-Hence, pull requests that improve performance are welcomed, and were even the
-main goal for one release cycle. For most sub-packages, performance was
+In some cases, feature enhancements and corner-case handling were added
+at the cost of performance.
+Hence, \github pull requests that improve performance are welcomed, and were
+even the main goal for v3.1 release. For most sub-packages, performance was
 improved in the \python\ code, but for some, the most time-critical pieces were
-rewritten in~C (e.g., for convolution of images, sigma-clipping, and converting
-time strings to binary). Furthermore, a particular effort was made to ensure
-\astropypkg\ is thread-safe, so that it can be used on supercomputer clusters.
+rewritten in~C and wrapped with \package{Cython} (\citealt{cython}; e.g., for
+convolution of images, sigma-clipping, and converting time strings to binary).
+Furthermore, a particular effort was made to ensure \astropypkg\ is thread-safe,
+so that it can be used on supercomputing clusters.
 
 Performance in \astropypkg is addressed based on feedback and requests, and
-through contributions from community members: If you find that parts of
+through contributions from community members: if you find that parts of
 \astropypkg are not performant, we encourage you to open a \github issue to
@@
 structures existed in the associated Open Source Software communities, and what
-would fit well with the needs of \astropy. This led into a ``retreat'' planned
+would fit well with the needs of \astropy. This led to a ``retreat'' planned
 for March 2020, but due to the COVID-19 pandemic, this became a series of
 virtual meetings of the ``Astropy Governance Working Group.'' This group
-drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
+drafted the APE 0 document \citep{ape0}, which was eventually ratified and
 implemented by the ``Astropy Governance Implementation Working Group'' in Fall
-2021. While the process emphasized flexibility and the ability to adapt to
-changing circumstances, it is expected that this is the framework Astropy's
-governance will operate in for at least the medium-term future.
-
-The APE0 \citep{ape0} document lays out the principles of this governance
+2021. While the process emphasized flexibility and adaptability,
+it is the expectation that \astropy's governance will operate in
+this framework for at least the medium-term future.
+
+The APE 0 \citep{ape0} document lays out the principles of this governance
 structure, so we refer the reader to that document for a more thorough
-description, and here we only highlight some key elements. While many of these
-principles were already de facto true or have been discussed organically (and
-have been discussed in earlier papers in this series), the APE0-based governance
-aims to provide a single place where the community can agree as a starting
-point. With this in mind, it highlights the developer and user community of the
-Astropy Project as the ultimate sources of authority, as well as the core
-principle of ``do-ocracy'' that those who do work for the Project (be it coding,
+description, and here we only highlight some key elements. The APE 0-based
+governance formalizes the building blocks that are already de facto true or
+previously discussed.
+With this in mind, it highlights the developer and user community of the
+\astropy Project as the ultimate sources of authority, as well as the core
+principle of ``do-ocracy'': those who do work for the Project (coding,
 training, or other harder-to-quantify contributions) gain more influence on the
-outputs of the Project by virtue of their effort. However, APE0 adds the
+outputs of the Project by virtue of their effort. However, APE 0 adds the
 concept of ``voting members'': a self-governed part of that community who are
-entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
-from the inception of the project, APE0 establishes a formal voting process for
-CoCo members, and explicitly outlines the rights and responsibilities of the
-CoCo. This role is mainly to facilitate consensus and act as the decision maker
+entrusted to elect the Coordination Committee (CoCo). While the CoCo
+predates APE 0, membership, rights, and responsibilities are now formally outlined.
+This role is mainly to facilitate consensus and make decision
 when other mechanisms have failed, and includes powers that either
-require central authority or secrets (e.g., passwords) However, APE0 also
+require central authority or secrets (e.g., passwords). However, APE 0 also
 charges the Committee to devolve responsibilities and seek community input on
@@
 
-The first Coordination Committee election under these rules took place in Fall
-2021, electing a mix of prior and new coordination committee members,
-and was contested in the sense of more candidates than available slots. This
+The first CoCo election took place in Fall 2021, electing a mix of prior and new members.
+New roles and voting members were also established. This
 suggests the process is already working to serve the long-term interests of the
@@
             package as a function of number of contributors (committers) broken
-            into three equal-length time periods between the start of \astropy
+            into three equal-length time periods between the start of \astropypkg
             and the feature freeze date of v5.0.
@@
 As described in the introduction above, the \astropy Project is broader than the
-core \astropypkg package as it also represents a number of core infrastructure
-packages, affiliated packages, educational initiatives, and user communities.
+core \astropypkg package as it also represents a number of core infrastructure,
+coordinated, and affiliated packages, educational initiatives, and user communities.
 In total, well over 1500 people have contributed in some way to the Project,
-whether, for example, through contributing code or documentation, opening issues
+whether by contributing code or documentation, reporting issues
 on \github, providing infrastructure support, writing tutorials, or helping to
-coordinate workshops that relate to Astropy.
+coordinate \astropy workshops.
 The vast majority of these contributors and participants have interacted as
@@
 primarily been sustained by a small number of heavily-engaged or
-explicitly-employed contributors \citep[see, e.g., ][]{astropy:2018}, as is the
+explicitly-employed contributors \citep[e.g.,][]{astropy:2018}, as is the
 case in many large, open-source software communities
@@
 contributing more, fractionally --- this emphasizes that, still, the vast
-majority of the work in \astropypkg is done my a small subset ($<20\%$) of the
+majority of the work in \astropypkg is done by a small subset ($<20\%$) of the
 115, 216, and 234 contributors, respectively, within each era.
@@
 Since its inception, the most active contributors to the \astropypkg package
-have spanned a range of career stages and career paths.
+have spanned a range of career stages and paths.
 For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
@@
 contributors moved between these categories (i.e., have graduated as students to
-become research staff / faculty or software engineers), but also a number of
+become research staff/faculty or software engineers), but also a number of
 early contributors moved on to other positions and stopped contributing to the
@@
 growth), as new, active contributors have generally joined and stepped in to
-fill top contributor roles within the project.
+fill top-contributor roles within the Project.
 On the other hand, the total number of top contributors has not grown
-substantially, reflecting one of our principal challenges: Sustaining a pipeline
+substantially, reflecting one of our principal challenges: sustaining a pipeline
 from users to contributors to maintainers of the package.
@@
 software engineers who contributed to \astropypkg were (or are) employed by
-STScI, whose leadership have consistently provided resources (financial and
-personnel) towards sustaining \astropypkg.
+STScI, whose leadership has consistently provided resources (financial and
+personnel) towards sustaining \astropy.
 
@@
             \textbf{Top panel:} The number of Git commits per month (defined as
-            consecutive 30 day windows) merged into to the \astropypkg core
-            package over the whole history of the project (up to end of 2021,
+            consecutive 30 day windows) merged into the \astropypkg core
+            package over the whole history of the Project (up to end of 2021,
             v5.0).
@@
 allocated funding to support mentoring programs. In 2020 a call was
-made to submit proposals for IDE (Inclusion, Diversity, and
-Empowerment) initiatives for project-wide consideration on
-Github. This process was deemed the most ``open'' because it allowed
+made to submit proposals for Inclusion, Diversity, and
+Empowerment (IDE) initiatives for project-wide consideration on
+\github. This process was deemed the most ``open'' because it allowed
 for community-wide feedback to focus and improve proposal
@@
 encouraged to attend monthly webinars to share skills in the context
-of open source software libraries. Every other month, a guest speaker
+of open-source software libraries. Every other month, a guest speaker
 was invited to talk about their career path and share a skill. Program
@@
 Africa, and 15\% from remaining continents). We selected participants
-who we identified as having a high potential for contributing to open
-source projects: intermediate to advanced programming skills with a
+who we identified as having a high potential for contributing to
+open-source projects: intermediate to advanced programming skills with a
 vocal interest in contributing. Of the thirty applicants invited to
@@
 In addition to the supported mentoring initiatives, the \astropy
-project as a whole has taken steps to examine representation of
-marginalized groups within the project and search for avenues of
+Project as a whole has taken steps to examine representation of
+marginalized groups within the Project and search for avenues of
 improvement.
@@
 Native American scientists across all STEM fields). \astropy representatives
-noted that one underlying topic came up multiple times. A major barrier for
-persons who come from underrepresented communities is the lack of resources and
-expertise that are necessary to train students. Writing PEP 8 compliant software
-and understanding the Git and \github workflow is not part of the standard
+noted that one underlying topic came up multiple times: a major barrier for
+persons who come from underrepresented communities is the lack of pedagogical
+resources and expertise. For instance, software engineering skills and best
+practices are not part of the standard
 Physics and Astrophysics curriculum. Additionally, the larger astronomical
 community is still in the process of transitioning towards \python-based tools,
-making it difficult for students not currently under advisement by some one with
-extensive \python expertise to get involved with the open source \python
+hindering students not currently under advisement by someone with
+extensive \python expertise from getting involved with the open-source \python
 community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
@@
 to teach students how to use Astropy, create their own libraries, and use
-version control on open source platforms like \github. Such materials could also
+version control on open-source platforms like \github. Such materials could also
 be offered as a workshop at national diversity conferences themselves, as SACNAS
@@
 
-With the goal of establishing a sustainable financial model for the Astropy
-Project, the Coordination Committee, under the new governance charter (see
+With the goal of establishing a sustainable financial model for the \astropy
+Project, the CoCo, under the new governance charter (see
 Section~\ref{sec:project-governance}), established a standing Finance Committee.
 This committee oversees the planning and allocation of finances on behalf
-of the Project. The Astropy Project accepts funds from institutions as well as
-individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
-supports and promotes world-class, innovative, open source scientific computing.}
-Previously, no direct financial support was available for the project development,
+of the Project. The \astropy Project accepts funds from institutions as well as
+individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) non-profit
+organization in the United States of America that can collect donations,
+administer grants, and perform financial management for projects in open source
+scientific computing.}
+Previously, no direct financial support was available for project development,
 and NumFOCUS covered most of the incurred operational costs. However, transitioning
 to long-term financial stability meant having an influx of funds in the form of
-institutional support, and hence the Astropy Project applied for and successfully
+institutional support, and hence the \astropy Project applied for and successfully
 acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
@@
 \begin{description}
-    \item[Moore] The Gordon and Betty Moore Foundation awarded the Astropy
-    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
+    \item[Moore] The Gordon and Betty Moore Foundation awarded the \astropy
+    Project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
     Growing the Astropy Project.'' Over a three-year term, this grant supported
     sustenance of the Project by providing funds for (1) the development and
-    maintenance of the \astropypkg and its infrastructure, provided to existing
+    maintenance of \astropypkg and its infrastructure, provided to existing
     project members and targeted hires, (2) the transition of long-term users to
@@
     Section~\ref{sec:project-governance}), and (4) the improvement of equity,
-    diversity, and inclusion efforts in the project (see
+    diversity, and inclusion efforts in the Project (see
     Section~\ref{sec:project-ide}). Some of this money supported travel for
@@
     reliance on volunteer-driven maintenance of the openly-developed package and
-    paved the way to acquire funding from federal agencies, representing a major
-    milestone for the success of the project.
+    paved the way to acquire funding from US federal agencies, representing a
+    major milestone for the success of the Project.
 
     \item[NASA] A successful proposal to the NASA ROSES-2020
-    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
+    call\footnote{\url{http://solicitation.nasaprs.com/ROSES2020}}, section E.7
     (Support for Open Source Tools, Frameworks, and Libraries), granted the
-    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
+    \astropy Project $\sim$\$600k (US) to support its work. Funded work for a
     term of three years includes (1) project-wide infrastructure maintenance and
@@
     and (4) support for the \astropy affiliated packages. These goals
-    support the sustenance of the project by encouraging further
+    support the sustenance of the Project by encouraging further
     development of the package ecosystem and community engagement.
@@
 
-In addition to the two grants described above, the project benefited from
+In addition to the two grants described above, the Project benefited from
 a Gemini Observatory contract under its Science User Support Department,
 which awarded funds for development work that supports both the Astropy and
-DRAGONS projects. The project also received financial support from the Dunlap
+DRAGONS\footnote{Data Reduction for Astronomy from Gemini Observatory North and
+South (\url{https://github.com/GeminiDRSoftware/DRAGONS}).} projects. The
+Project also received financial support from the Dunlap
 Seed Funding program at the University of Toronto to develop educational
 resources for scientific software packaging within the Learn Astropy
-framework. All of these successful proposals show that the Astropy Project
+framework. Moreover, the Project indirectly benefited from the NRAO ALMA Development
+Study program, which supported the grant ``Linking CASA to the astropy
+ecosystem''\footnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
+All of these successful funding proposals show that the \astropy Project
 is vital for the astronomical infrastructure and community at large, and
@@
 
-% ADD text for the new finance APE once it has a doi
+The Project follows a set of principles to fairly and effectively
+distribute the above described funds. By drawing inspiration from
+Numpy's NEP 48\footnote{https://numpy.org/neps/nep-0048-spending-project-funds.html},
+the APE 19 \citep{ape19} document lays out these principles. We refer the
+reader to that document for more details.
 
@@
 
-The Astropy project supports the broader ecosystem by providing pre-configured
+The Astropy Project supports the broader ecosystem by providing pre-configured
 infrastructure packages that the community can use to support and maintain their
 own software package infrastructure. These include tools and extensions that
-enable easily generating documentation and setting up automated testing, as well
+enable generating documentation and setting up automated testing, as well
 as providing package scaffolding for new projects.
 
-Sphinx is a common and useful tool for generating documentation for Python
-packages. The Astropy project maintains a default Sphinx configuration along
+Sphinx is a common and useful tool for generating documentation for \python
+packages. The Astropy Project maintains a default Sphinx configuration along
 with Astropy-specific extensions that can be easily added to community
 projects via the
-\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}} meta
-package. This tool provides a pre-configured Sphinx setup compatible with
+\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}}
+meta-package. This tool provides a pre-configured Sphinx setup compatible with
 Astropy projects, which includes several extensions useful for generating API
@@
 (\href{https://github.com/astropy/sphinx-automodapi}{\texttt{sphinx-automodapi}}),
-allowing for Numpy docstring parsing
+allowing for \package{numpy} docstring parsing
 (\href{https://github.com/numpy/numpydoc}{\texttt{numpydoc}}), embedded image
@@
 The Astropy Package Template helps facilitate the setup and creation of new
-Python packages leveraging the Astropy ecosystem. This tool utilizes the
-Cookiecutter project to walk users through the process of creating new
-packages complete with documentation and testing support. Additionally, the
-package template generation process includes the ability to setup
-interoperability with \github, allowing for easy repository access from
-documentation, as well as an example \github Actions workflow to demonstrate
-the use of \github's continuous integration tooling.
+\python packages leveraging the Astropy ecosystem. This tool utilizes the
+\href{https://github.com/cookiecutter/cookiecutter}{Cookiecutter} project to
+walk users through the process of creating new packages complete with
+documentation and testing support. Additionally, the package template
+generation process includes the ability to setup interoperability with
+\github, allowing for easy repository access from documentation, as well as an
+example \github Actions workflow to demonstrate the use of \github's
+continuous integration tooling.
 
@@
 Astropy of improving code re-use, interoperability, and embracing good coding
-practices such as testing and thorough documentation.
+practices such as testing and thorough documentation, but are not solely
+developed and maintained by members of the Astropy Project team.
+However, a subset of the affiliated packages --- the ``Astropy Coordinated
+Packages'' --- are maintained by the Project.
 All Astropy-coordinated and other affiliated packages are listed with detailed
 information on the \href{{https://www.astropy.org/affiliated}}{\astropy
-website}; A table summarizing key aspects of the affiliated packages is included
+website}; a table summarizing key aspects of the affiliated packages is included
 in the Appendix below (Table~\ref{tab:affiliated-registry}).
@@
 \texttt{corral} for managing data intensive parallel pipelines.
+\texttt{casaformats-io} provides infrastructure for reading data stored in CASA table formats.
 \texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
@@
 services, including the James Webb Space Telescope (JWST) archive, and the
-project has switched to a continuous release model: Every time a change is
+project has switched to a continuous release model: every time a change is
 committed to the main development branch it is published on the Python Package
 Index (PyPI) and available for installation. (However, formal releases are still
-done a few times per year.) \texttt{photutils} \cite {photutils} released its
+done a few times per year.) \texttt{photutils} \citep{photutils} released its
 first stable version, indicating that the API will change less frequently, and
 there have been several significant performance improvements. \texttt{ccdproc}
-\cite{ccdproc} also released a new major version, bringing better performance to
+\citep{ccdproc}  released a new major version, bringing better performance to
 some image combination operations. The \texttt{regions} package
-\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
-added new ways to manipulate regions and introduced new region types.
-\texttt{reproject}'s \cite{reproject} major new feature is a function to align
+\citep{pyregions}, for manipulating ds9-, CASA-, or FITS-style region definitions \citep{ds9},
+added new ways to manipulate regions, introduced new region types, and obtained
+expanded visualization and interactive region editing capabilities.
+\texttt{reproject}'s \citep{reproject} major new feature is a function to align
 and co-add images to create a mosaic; better support for parallelization was
 also added. \texttt{specutils}, the package that defines containers for 1D and
-2D spectra \cite{specutils}, also had its first stable release and the addition
-of classes to read JWST data. \texttt{stingray} has also had a major performance
-overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
+2D spectra \citep{specutils},  had its first stable release and the addition
+of classes to read JWST data. \texttt{stingray} has had a major performance
+overhaul. The package \texttt{gammapy} \citep{gammapy} has had major performance
 improvements and has unified its API in preparation for its first stable
@@
 several special-purpose \python packages, such as \texttt{LightKurve}
-\citep{LightKurve} and SORA \citep{SORA}]. More than 4,000 repositories on
+\citep{LightKurve} and SORA \citep{SORA}. More than 4,000 repositories on
 \github make use of \texttt{astroquery} in some way, and \texttt{astroquery}
@@
 inception. The JWST instrument calibration pipelines, exposure-time
-calculators, and data analysis tools are all written in Python and
+calculators, and data analysis tools are all written in \python and
 depend on the \astropypkg core package and some coordinated and
@@
 Likewise, JWST developers have provided significant contributions
-to the \package{photutils} \citep{photutils}, \package{specutils}
-\citep{specutils}, and \package{regions} \citep{regions} coordinated
-packages and the \package{gwcs} \citep{gwcs} and \package{synphot}
-\citep{synphot} affiliated packages. For example, development of the
+to the several affiliated packages. For example, development of the
 \package{photutils} coordinated package for source detection and
@@
 was created at STScI and is a dependency of the JWST exposure-time
-calculators. Further, the \package{ASDF} (Advanced Scientific Data
+calculators. Furthermore, the \package{ASDF} (Advanced Scientific Data
 Format) package \citep{ASDF}, a next-generation interchange format for
@@
 CTA will be able to observe gamma rays in a broad energy range from around $20\,\mathrm{GeV}$ to over $300\,\mathrm{TeV}$
-using three different types of telescopes, in total over 100 telescopes are planned at the two sites.
+using three different types of telescopes. In total over 100 telescopes are planned at the two sites.
 CTA will also be the first open gamma-ray observatory.
 
-The data analysis pipeline is developed as open source software and essentially split in two domains:
+The data analysis pipeline is developed as open-source software and essentially split in two domains:
 \begin{enumerate}
@@
     have to be estimated from the raw data.
-    The raw data consists of very short (\textasciitilde $40--100\,\mathrm{ns}$) videos recorded with the fast and
+    The raw data consists of very short ($\sim40$--$100\,\mathrm{ns}$) videos recorded with the fast and
     sensitive cameras of the telescopes.
@@
 A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
-a python package developed to perform all the necessary tasks to from the raw data
+a \python package developed to perform all the necessary tasks to perform data processing, from the raw data
 of Cherenkov telescopes to the reconstructed event lists.
 The high-level analysis (or CTA science tools) will be based on the Astropy affiliated package
-Gammapy~\citep{gammapy}.
-Both Gammapy and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
-tables and FITS I/O.
+\package{gammapy}~\citep{gammapy}.
+Both \package{gammapy} and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
+tables, and FITS I/O.
 As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
@@
 based on earlier work by B.~Winkel.
-Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
+Together with \package{gammapy} maintainer A.~Donath and former maintainer C.~Deil,
 a total of 107 merged pull requests were contributed to astropy.
+
+\subsubsection{Data Central and the Anglo-Australian Telescope Archive}
+% \secauthor{James Tocknell}
+Data Central is one of the five Australian Virtual Observatory\footnote{Known as
+  the All-Sky Virtual Observatory, or ASVO.} nodes, providing both Australian and
+international astronomers access to and services for surveys of the southern
+sky. Whilst primarily hosting optical data, including the Anglo-Australian
+Telescope (AAT) Archive\footnote{The AAT Archive is currently jointly managed by
+  Data Central, Astronomy Australia Limited (AAL), and National Computational
+  Infrastructure (NCI).}, Data Central also assists theoretical and radio
+astronomers with hosting and managing their data.
+
+\python is widely used by Data Central, and \astropy and its wider ecosystem
+play a vital role in the initial ingestion, reduction, access and visualization
+of the hosted data and services. For example, the AAT Archive hosts data from as early as
+1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest
+data flowing from the AAT to ensure that a consistent time and coordinate system
+is provided to users for simpler querying.
+
+As many of the surveys hosted by Data Central are spectroscopic surveys from the
+AAT, \package{specutils} is heavily used to standardize the variety of encodings
+of spectroscopic data into a form easily readable and visualizable by many
+standard tools (such as \package{ds9} or \package{splat}). Data Central has
+upstreamed into \package{specutils} loaders for all hosted spectra, including
+2dfdr-reduced spectra from the AAT Archive, and strongly encourages other
+archives to provide loaders for their data to further the reuse of existing archival
+data.
 
@@
 educational efforts made by the Astropy Project, which are led by the Learn
-Team.
+Astropy Team.
 The efforts focus on developing online content and workshops covering
-astronomy-specific coding tasks in Python.
+astronomy-specific coding tasks in \python.
 As introduced in \citet{astropy:2018}, there are four different types of
@@
 type of astronomical analyses; \textit{examples}, which are snippets of code
-that showcase a short task that can be performed with Astropy or an affiliated
-package; and \textit{documentation}, which is contained within the code base.
+that showcase a short task that can be performed with \astropypkg or an affiliated
+package; and \textit{documentation}, which contains more detailed information
+about the code base and user interface.
 This categorization drives content development, infrastructure choices, and the
-appearance of the Learn Astropy website.
-The Learn Team meets weekly to work on creating, expanding, improving these
+appearance of the \href{https://learn.astropy.org}{Learn Astropy website}.
+The Learn Team meets weekly to work on creating, expanding, and improving these
 educational resources.
 
-\begin{description}
-
-\item[Learn Astropy website] The Learn Team re-launched the main website and
-search interface for Learn Astropy in 2021 with a new infrastructure platform,
-built around full-text search and interactive filtering functionality, with the
-goal of making content discoverable as the Learn Astropy content catalog
+The Learn Team recently re-launched the main website and search interface for
+Learn Astropy in 2021 with a new infrastructure platform, built around full-text
+search and interactive filtering functionality, with the goal of making content
+more easily searchable and discoverable as the Learn Astropy content catalog
 expands.
@@
 Guides, which utilize the \href{https://jupyterbook.org/}{Jupyter Book} build
-infrastructure, are also deployed as separate websites using GitHub pages.
+infrastructure, are also deployed as separate websites using \github pages.
 This architecture opens future possibilities of indexing third-party content,
@@
 
-\item[Tutorials] Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.
-
-\item[Guides] {\it ccdproc} Jupyter book has been completed.
-
-\item[Workshops] The Project has been conducting workshops at winter
-meetings of the American Astronomical Society since AAS 225 in January 2015.
+We currently host 19 tutorials written as interactive Jupyter Notebooks and
+rendered into static HTML pages with the infrastructure described above.
+The tutorials span a range of astronomical topics, from general tasks like
+reading or creating FITS files with different content or working with
+astronomical coordinate systems, to more specific exercises like analyzing
+spectroscopic data from the UVES instrument.
+Now that the backend infrastructure is stable, we are interested in collecting
+new content to serve and share.
+Ideas for contributions, new tutorials, or suggestions for new tutorial series
+to be made searchable through the Learn Astropy interface can be raised by
+creating a \github issue in the
+\href{https://github.com/astropy/astropy-tutorials}{Astropy Tutorials GitHub
+repository}.
+We currently host one Guide --- a longer-form walkthrough of a more complex
+concept or topic --- that is focused on
+\href{http://www.astropy.org/ccd-reduction-and-photometry-guide}{CCD image
+reduction and photometry}, but are also actively seeking new material that would
+be suitable for new guides.
+
+Beyond developing and serving educational content, the \astropy Project has been
+conducting workshops at winter meetings of the American Astronomical Society
+since AAS 225 in January 2015.
 Up to the start of the coronavirus pandemic, these were full-day in-person
 workshops with as many as 90 participants and a dozen facilitators from the
-project.
+Project.
 During the pandemic, these workshops were moved to an online format and split
@@
 expanded to Summer AAS meetings.
-The Learn team finds that the workshop audience is best found as AAS meetings as
-opposed to more general Python meetings.
-
-The Astropy Project provided another mode of community engagement at AAS
-Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
-AAS Exhibit Hall.
+The Learn team finds that the workshop audience is best found at AAS meetings as
+opposed to more general \python meetings, as the content tends to be more
+applicable for students and researchers in astronomy and astrophysics.
+
+The Astropy Project recently provided another mode of community engagement at
+AAS Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in
+the AAS Exhibit Hall.
 Funding for the exhibit hall was provided alternately by NumFOCUS and later by
-the Moore Foundation funding (\textbf{check this}).
+the Moore Foundation funding.
 The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
 during the virtual AAS 237 meetings, to provide the general astronomy community
-information and access to experts on a variety of open source astronomical
+information and access to experts on a variety of open-source astronomical
 tools.
-\end{description}
 
@@
 
-The Learn plan going forward is to continue to improve the tutorial website and
-facilate new content;
-to index third-party tutorials; and to look for supportable opportunities to
-expand the reach of Astropy workshops beyond the American Astronomical Society
-meetings.
+The focus of Learn Astropy over the coming months and years is to further
+improve the interface for sharing educational content on the main Learn Astropy
+website, and to facilitate the development of new content that highlights
+\astropypkg and \astropy-affiliated package functionality.
+However, we simultaneously also plan to solicit indexing and ingestion of
+third-party tutorial series, to provide a unified interface for identifying
+astronomy-specific educational content that demonstrates the functionality
+available in the vast ecosystem of open-source software packages.
+We also plan to look for opportunities to expand the reach of \astropy workshops
+beyond the AAS meetings.
+
+% \begin{itemize}
+% \item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+% be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+% open-source projects.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+% a Discourse site which is more open than the Facebook group and more user-oriented
+% than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+% developers are able to invoice as independent contractors the time they
+% spend helping users on these forums.
+% \end{itemize}
 
@@
 
-\subsection{A roadmap for future priorities}
+\subsection{A Roadmap for Future Priorities}
 % \secauthor{Clara Brasseur}
@@
 \href{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}{Astropy Roadmap} is a document in the \texttt{astropy-project} \github
-repository within the \astropy organization that captures high level actionable
-items that the Astropy project aims to undertake to improve the health and
-stability of the project. It is a static document that is revisited regularly at
+repository within the \astropy organization that captures high-level actionable
+items that the Astropy Project aims to undertake to improve the health and
+stability of the Project. It is a static document that is revisited regularly at
 the Astropy coordination meetings, intended to keep track of progress and write
@@
 underway, have sufficient resources/support and a plan in place for completion;
-For example, efforts to improve the discoverability of documentation and
+for example, efforts to improve the discoverability of documentation and
 educational materials by overhauling the Learn website, which is underway (see
@@
 Orange items are well defined, and work for acquiring sufficient resources
-underway; For example, the goal of providing next-generation spectroscopic
+underway; for example, the goal of providing next-generation spectroscopic
 reduction, analysis, and visualization tools usable by individual researchers
 and larger surveys.
-Red items do not yet have a plan for implementation and need more resources; For
+Red items do not yet have a plan for implementation and need more resources; for
 example, the goal of improving interoperability with performant I/O file formats
@@
 
+\begin{figure}[th!]
+  \begin{centering}
+    \includegraphics[width=\textwidth]{docu_analytics.pdf}
+      \caption{
+          The number of users who accessed the \astropypkg core package
+          documentation in 2021, based on data from Google Analytics.
+          The circle markers show the approximate locations of home institutions
+          for \astropy voting members (green) and coordination committee members
+          (red), demonstrating the over-representation of the United States and,
+          less so, Europe given the global spread of \astropypkg users.
+      }
+      \label{fig:User-analytics-locations}
+  \end{centering}
+\end{figure}
+
 \subsection{Current and Future Challenges}
 
-Community Engagement
-
-\begin{itemize}
-\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
-This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
-The Moore Foundation grant allows this Slack space to
-be on a paid plan; additionally NumFOCUS has negotiated a special rate for
-open-source projects.
-While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
-The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
-a Discourse site which is more open than the Facebook group and more user-oriented
-than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
-developers are able to invoice as independent contractors the time they
-spend helping users on these forums.
-\end{itemize}
-
-\begin{itemize}
-\item Attracting new contributors when the code has become quite complex,
-\item Contributor to maintainer mentoring,
-\item Long-term / sustained funding for maintaining infrastructure,
-\item ...
-\end{itemize}
-
-\begin{acknowledgments}
+Over the history of \astropy, the needs and challenges of the \astropy Project
+have evolved, but many common themes have persisted and are shared by other
+open-source software communities.
+However, \astropy also faces a number of unique challenges stemming from the
+fact that the Project is in a special position in the astronomy software stack:
+the \astropy code and ecosystem are more specialized than core scientific tools
+like \package{numpy} and \package{scipy}, but more general than software
+packages developed to enable research in specific subdomains such as
+\package{exoplanet} \citep{exoplanet:2021}.
+Below we discuss a few challenges faced by the Project in more detail, but we
+stress that this list is not complete or exhaustive.
+\begin{description}
+  \item[Supporting an active group of maintainers] Throughout its history,
+  development and maintenance of code and educational materials in the \astropy
+  Project --- the \astropypkg core package, coordinated packages, Learn Astropy
+  repositories, and Project infrastructure packages --- has been done by a mix
+  of volunteer effort and institutional support.
+  At present, the \href{https://www.astropy.org/team.html}{named roles} and
+  maintainer positions in the \astropy Project are also split roughly evenly
+  between these types of contributors.
+  As is the case with many other open-source software communities, the culture
+  of the Project and overall coordination relies heavily on volunteer effort.
+  We therefore have significant amounts of critical infrastructure or core
+  functionality that is either maintained with no redundancy or unmaintained as
+  contributors naturally move on from and stop contributing to the Project.
+  A prime example within the core package is the \astropyfits subpackage, which
+  is largely in maintenance mode and maintained by just one contributor (despite
+  the significant codebase and user community behind this subpackage).
+  While we now have funding to support the longer-term sustainability of the
+  Project and thus to address these issues, we are finding it difficult to use
+  this funding in ways that address our needs.
+  For one, many of the needs of the Project require long-term and sustained
+  funding rather than short-term contracts.
+  Furthermore, the specialized nature of the software requires being able to hire or
+  contract software engineers with domain knowledge about astronomical software
+  tools.
+  Finally, we want to be able to preserve and sustain the engaged community of
+  volunteers, software-engaged researchers, and existing software engineers that
+  contribute to the Project and that have enabled its success in ways that
+  support their careers.
+  These challenges will require both cultural and larger systemic changes to
+  astronomical research and career paths, and so will take time to influence and
+  implement.
+  \item[Diversity of users, contributors, maintainers, and coordinators]
+  The \astropy Project aims to be made up of members at all levels (users to
+  contributors to coordinators) with a diverse set of skills, backgrounds, and
+  experiences.
+  While \astropy is used globally --- for example,
+  Figure~\ref{fig:User-analytics-locations} shows the number of users per
+  country who accessed the \astropypkg core package documentation in 2021 ---
+  our user community is underrepresented in non-English-speaking or -teaching
+  countries, and our maintainers and coordinators are fairly narrowly
+  concentrated in the United States (US) and Europe (see the circle markers on
+  Figure~\ref{fig:User-analytics-locations}).
+  The gap in users in many countries is understandable given that much of our
+  content and documentation is written in English and therefore only available
+  in other languages through imperfect, automated translation services (like
+  Google translate).
+  The fact that our maintainers and coordinators are generally in the US (even
+  then, mostly on the East coast) and Europe is a reflection of where
+  institutions have explicitly supported \astropy, and where volunteers have had
+  the ability to contribute to the Project.
+  \item[Long-term and sustained funding for maintaining infrastructure] The
+  software and packaging infrastructure within the \astropy Project includes
+  (but is not limited to) Continuous Integration (CI), test plugins,
+  documentation builds, and package delivery.
+  Keeping the infrastructure working is essential in software development and
+  operations.
+  This kind of work has low visibility and is not attractive to scientists or
+  volunteer contributors; therefore, it is most effective to delegate such tasks
+  to paid roles.
+  \astropy secured a three-year grant for this but the situation remains
+  uncertain once the grant runs out, as infrastructure evolves and requires
+  constant maintenance and upgrades.
+  Some of the engineers currently in these roles are also employees of research
+  institutions that may leave the Project without much notice, adding to the
+  uncertainty factor.
+  \item[Understanding, communicating with, and engaging our user communities]
+  The \astropy Project is used by a number of intersecting and overlapping
+  communities, such as telescope operations and data management, data archives,
+  researchers, students, educators, and the general public.
+  These communities have different needs from and uses for \astropy, many of
+  which we are only aware of through individual engagement from members of these
+  respective communities.
+  This kind of communication may provide biased perspectives on these
+  communities.
+  We therefore need to develop more formal and democratic ways of engaging these
+  communities, such as through running and analyzing user surveys, or
+  creating and engaging with user groups that can better represent these
+  communities.
+  Additionally, as the \astropy ecosystem has and continues to grow, we need to
+  ensure that users in these different communities have access to educational
+  materials to train new users to use \astropy and \python software tools.
+  This may involve tailoring workshops, tutorials, and educational materials to
+  these communities, however this represents a significant maintenance and
+  content-development burden on the Project.
+\end{description}
+
+\section{Conclusions}
+\label{sec:conclusions}
+
+The Astropy Project has continued to serve as an active, community-oriented
+software project with a wide reach and growing impact, both within astronomy and
+astrophysics and in broader scientific software communities.
+The Project is in the midst of setting up infrastructure to ensure its long-term
+sustainability and to continue healthy growth of the software ecosystem,
+developer, and user communities that interface with the Project.
+For example, over the last years, we have formalized and implemented a new
+governance model for the Astropy Project (Section~\ref{sec:project-governance}),
+have continued to engage a growing pool of active and top contributors
+(Section~\ref{sec:project-contributors}), and have begun to receive significant
+formal funding for the Project itself (Section~\ref{sec:project-funding}).
+The Astropy Project is also expanding its educational and community-focused
+initiatives by soliciting and funding initiatives in inclusion, diversity, and
+equity (Section~\ref{sec:project-ide}) as well as through the Learn Astropy initiative,
+which aims to make tutorials and educational content that highlight software
+tools for astronomy more discoverable and accessible (Section~\ref{sec:learn}).
+We expect to continue in all of these threads over the coming years as we aim to
+make the Astropy Project more sustainable and a larger, more welcoming community
+for open-source software users and enthusiasts in astronomy.
+
+% Due to a bug in AAStex 6.3.1, using this with amsmath leads to line numbers in
+% the acknowledgements...
+% https://github.com/AASJournals/AASTeX60/issues/130
+% \begin{acknowledgments}
+\vspace{2em}
 
 We would like to thank the members of the community who have contributed to
-\astropy, who have opened issues and provided feedback, and have supported the
-project in a number of different ways.
+\astropy, opened issues and provided feedback, and supported the
+Project in a number of different ways.
+
+We acknowledge the Gordon and Betty Moore foundation for their continued
+financial support.
+This work is partially supported by NASA under Grant No. 80NSSC22K0347 issued
+through the NASA ROSES program.
+This work is partially supported by the international Gemini Observatory, a
+program of NSF's NOIRLab, which is managed by the Association of Universities
+for Research in Astronomy (AURA) under a cooperative agreement with the National
+Science Foundation, on behalf of the Gemini partnership of Argentina, Brazil,
+Canada, Chile, the Republic of Korea, and the United States of America.
+We also thank NumFOCUS and the Python Software Foundation for financial
+support.
 
@@
 summer. These students often turn into long-term contributors.
-We also thank NumFOCUS and the Python Software Foundation for financial
-support.
-Within the academic community, we thank institutions that make it possible that
-astronomers and other developers on their staff can contribute their time to the
+Within the academic community, we thank institutions that make it possible for
+astronomers and other developers on their staff to contribute their time to the
 development of \astropy projects.
@@
 
-Supported by the international Gemini Observatory, a program of NSF's NOIRLab,
-which is managed by the Association of Universities for Research in Astronomy
-(AURA) under a cooperative agreement with the National Science Foundation, on
-behalf of the Gemini partnership of Argentina, Brazil, Canada, Chile, the
-Republic of Korea, and the United States of America.
-
 Furthermore, the \astropy packages would not exist in their current form without
 a number of web services for code hosting, continuous integration, and
-documentation; in particular, \astropy heavily relies on GitHub, Azure
+documentation; in particular, \astropy heavily relies on \github, Azure
 Pipelines, CircleCI, and Read the Docs.
@@
 \astropypkg interfaces with the SIMBAD database, operated at CDS, Strasbourg,
-France. It also makes use of the ERFA library \citep{erfa}, which in turn
+France. It also makes use of the \package{ERFA} library \citep{erfa}, which in turn
 derives from the IAU SOFA Collection\footnote{\url{http://www.iausofa.org}}
@@
 
-\end{acknowledgments}
+% \end{acknowledgments}
 

### 2022-03-16 src/static/authors.tex
working on author list
trying to work out author list issues

first pass at validating author list

reorder using authorship ordering

read emails from file instead

remove emails from public branch

\newcommand{\affstsci}{
    Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD 21218, USA
}
\newcommand{\affcca}{
    Center for Computational Astrophysics, Flatiron Institute, 162 5th Ave, New York, NY 10010, USA
}
\newcommand{\affcfa}{
    Center for Astrophysics | Harvard \& Smithsonian, 60 Garden Street, Cambridge, MA, 02138, USA
}
\newcommand{\affbaeri}{
    Bay Area Environmental Research Institute, P.O. Box 25, Moffett Field, CA 94035, USA
}
\newcommand{\affprinceton}{
    Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA
}
\newcommand{\affaperio}{
    Aperio Software Ltd., Insight House, Riverside Business Park, Stoney Common Road, Stansted, Essex, CM24 8PL, UK
}
\newcommand{\afftorontoastro}{
    David A. Dunlap Department of Astronomy \& Astrophysics, University of Toronto, 50 St. George Street, Toronto ON M5S 3H4, Canada
}
\newcommand{\affipac}{
    IPAC, MC 100-22, Caltech, 1200E. California Blvd. Pasadena, CA 91125
}

% Paper coordinators, ordered by contributions:
\author[0000-0003-0872-7098]{Adrian~M.~Price-Whelan}
\affiliation{\affcca}

\author[0000-0003-0079-4114]{Pey~Lian~Lim}
\affiliation{\affstsci}

%FLAG
\author[0000-0003-1714-7415]{Nicholas~Earl}
\affiliation{Department of Astronomy, University of Illinois at Urbana-Champaign, 1002 W. Green St., Urbana, IL 61801 USA}

% ----------------------------------------------------------------------------
% Paper contributors, ordered by contributions:
\author[0000-0003-3954-3291]{Nathaniel~Starkman}
\affiliation{\afftorontoastro}

\author[0000-0002-7908-9284]{Larry~Bradley}
\affiliation{\affstsci}

\author[0000-0003-4401-0430]{David~L.~Shupe}
\affiliation{\affipac}

\author[0000-0002-7626-506X]{A.~A.~Patil}
\affiliation{\afftorontoastro}
\affiliation{Dunlap Institute for Astronomy \& Astrophysics, University of Toronto, 50 St George Street, Toronto, ON M5S 3H4, Canada}

\author[0000-0002-5466-3817]{Lia~Corrales}
\affiliation{University of Michigan Astronomy, 1085 S. University, Ann Arbor MI 48109, United States}

\author[0000-0002-9314-960X]{C.~E.~Brasseur}
\affiliation{School of Physics and Astronomy, University of St Andrews, North Haugh, St Andrews KY16 9SS, Scotland}

\author[0000-0001-7993-8189]{Maximilian~N\"{o}the}
\affiliation{Astroparticle Physics, TU Dortmund, Otto-Hahn-Str. 4a, 44227 Dortmund, Germany}

\author[0000-0002-9599-310X]{Erik~Tollerud}
\affiliation{\affstsci}

\author[0000-0003-2528-3409]{Brett~M.~Morris}
\affiliation{Center for Space and Habitability, University of Bern, Gesellschaftsstrasse 6, 3012 Bern, Switzerland}

\author[0000-0001-6431-9633]{Adam~Ginsburg}
\affiliation{Department of Astronomy, University of Florida, P.O. Box 112055, Gainesville, FL, USA}

\author{Eero~Vaher}
\affiliation{Lund Observatory, Department of Astronomy and Theoretical Physics, Lund University, Box 43, 22100 Lund, Sweden}

\author{Benjamin~A.~Weaver}
\affiliation{NSF's National Optical-Infrared Astronomy Research Laboratory, 950 North Cherry Avenue, Tucson, AZ 85719, United States}

\author[0000-0001-6637-6922]{James~Tocknell}
\affiliation{Department of Physics and Astronomy, Macquarie University, Sydney, NSW 2109, Australia}
\affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
\affiliation{Australian Astronomical Optics - Macquarie, Macquarie University, Sydney, NSW 2109, Australia}

\author[0000-0001-5976-4492]{William~Jamieson}
\affiliation{\affstsci}

\author[0000-0002-5830-8505]{Marten~H.~van~Kerkwijk}
\affiliation{\afftorontoastro}

\author[0000-0002-8642-1329]{Thomas~P.~Robitaille}
\affiliation{\affaperio}

%FLAG
\author{Bruce~Merry}
\affiliation{South African Radio Astronomy Observatory}

\author{Matteo~Bachetti}
\affiliation{Istituto Nazionale di Astrofisica/Osservatorio Astronomico di Cagliari, via della Scienza 5, I-09047 Selargius (CA), Italy}

\author[0000-0003-4243-2840]{H.~Moritz~G\"{u}nther}
\affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}

% ----------------------------------------------------------------------------
% All others, alphabetical:
\author{Thomas~L.~Aldcroft}
\affiliation{}

\author[0000-0003-0353-9741]{Jaime~A.~Alvarado-Montes}
\affiliation{School of Mathematical and Physical Sciences, Macquarie University -- Sydney, NSW 2109, Australia.}
\affiliation{Centre for Astronomy, Astrophysics and Astrophotonics, Macquarie University -- Sydney, NSW 2109, Australia.}

\author[0000-0003-0638-3340]{Anne~M.~Archibald}
\affiliation{Newcastle University}

\author[0000-0002-8585-4544]{Attila~B\'odi}
\affiliation{Konkoly Observatory, Research Centre for Astronomy and Earth Sciences (ELKH), H-1121 Budapest, Konkoly Thege Mikl\'os \'ut 15-17, Hungary}
\affiliation{MTA CSFK Lend\"ulet Near-Field Cosmology Research Group}

\author[0000-0002-0870-4665]{Shreyas~Bapat}
\affiliation{School of Computing and Electrical Engineering, Indian Institute of Technology Mandi, Kamand, Mandi (H.P.), 175005, India}

\author[0000-0002-3306-3484]{Geert~Barentsen}
\affiliation{\affbaeri}

%FLAG
\author[0000-0001-7699-3983]{Juanjo~Baz{\'a}n}
\affiliation{CIEMAT, Astroparticle physics, Madrid, Spain.}

%FLAG
\author{Manish~Biswas}
\affiliation{Savitribai phule pune university}

\author[0000-0003-0946-6176]{M\'ed\'eric~Boquien}
\affiliation{Centro de Astronomía (CITEVA), Universidad de Antofagasta, Avenida Angamos 601, Antofagasta, Chile}

\author[0000-0003-4428-7835]{D.~J.~Burke}
\affiliation{\affcfa}

\author[0000-0002-9294-6551]{Mihai~Cara}
\affiliation{\affstsci}

\author[0000-0003-3833-1668]{Daria~Cara}
\affiliation{Wake Forest University, 1834 Wake Forest Rd, Winston-Salem, NC, 27109, USA}

\author[0000-0001-5410-5551]{Jos\'e~Vin\'icius~de~Miranda~Cardoso}
\affiliation{The Hong Kong University of Science and Technology, Department of Electronic and Computer Engineering, Hong Kong SAR China}

\author[0000-0002-9583-263X]{ZHANG~Chen}
\affiliation{Purple Mountain Observatory, CAS}

\author[0000-0002-5442-8550]{Kyle~E~Conroy}
\affiliation{\affstsci}

\author{Simon~Conseil}
\affiliation{Univ Lyon, Univ Claude Bernard Lyon 1, CNRS/IN2P3, IP2I Lyon, UMR 5822, F-69622, Villeurbanne, France}

%FLAG
\author[0000-0001-7988-8919]{Matthew~W.~Craig}
\affiliation{Minnesota State University Moorhead}

\author[0000-0003-0081-8024]{Robert~M.~Cross}
\noaffiliation

\author[0000-0002-1821-0650]{Kelle~L.~Cruz}
\affiliation{Hunter College, Physics and Astronomy, 695 Park Avenue, New York, NY 10065}
\affiliation{American Museum of Natural History, Astrophysics, Central Park West at 79th St, New York, NY 10024}
\affiliation{Flatiron Institute, 162 Fifth Avenue, New York, NY 10010}

\author[0000-0003-2388-8172]{Francesco~D'Eugenio}
\affiliation{Cavendish Laboratory and Kavli Institute for Cosmology, University of Cambridge, Madingley Rise, Cambridge, CB3 0HA, United Kingdom}

\author[0000-0002-5686-9632]{Nadia~Dencheva}
\affiliation{\affstsci}

\author[0000-0001-9226-1870]{Hadrien~A.~R.~Devillepoix}
\affiliation{School of Earth and Planetary Sciences, Curtin University, Perth WA 6845, Australia}

\author[0000-0002-8134-9591]{J\"org~P.~Dietrich}
\affiliation{Arndtstr. 10, 55124 Mainz, Germany}

\author[0000-0003-4568-7005]{Axel~Donath}
\affiliation{\affcfa}

\author[0000-0003-0810-4368]{Arthur~Davis~Eigenbrot}
\affiliation{National Solar Observatory, 3665 Discovery Drive, Boulder CO. 80303}

\author{Thomas~Erben}
\affiliation{Argelander-Institut f\"ur Astronomie (University of Bonn), Auf dem H\"ugel 71, 53121 Bonn, Germany}

%FLAG
\author[0000-0002-8919-079X]{Leonardo~Ferreira}
\affiliation{University of Nottingham}

\author[0000-0002-9328-5652]{Daniel~Foreman-Mackey}
\affiliation{\affcca}

\author{Ryan~Fox}
\noaffiliation

\author[0000-0002-6253-082X]{Nabil~Freij}
\affiliation{\affbaeri}
\affiliation{Lockheed Martin Solar \& Astrophysics Laboratory, Org. A021S, Bldg. 252, 3251 Hanover St., Palo Alto, CA 94304, USA}

\author[0000-0002-2309-9731]{Suyog~Garg}
\affiliation{Department of Physics, The University of Tokyo, Bunkyo-ku, Tokyo 113-0033, Japan}
\affiliation{Institute of Cosmic Ray Research, The University of Tokyo, Kashiwa City, Chiba 277-8582, Japan}

\author[0000-0003-1509-9966]{Robel~Geda}
\affiliation{\affprinceton}

\author{Lauren~Glattly}
\affiliation{ScienceBetter Consulting, LLC, 4 West 101st Street, No.29, New York, NY 10025}

\author[0000-0002-6646-4225]{Yash~Gondhalekar}
\affiliation{BITS Pilani/Computer Science, Sancoale, Goa 403726, India}

\author[0000-0001-5340-6774]{Karl~D.~Gordon}
\affiliation{\affstsci}

\author[0000-0001-5878-618X]{David~Grant}
\affiliation{University of Oxford, Department of Physics, Keble Road, Oxford, OX1 3RH, UK}
\affiliation{University of Bristol, HH Wills Physics Laboratory, Tyndall Avenue, Bristol, BS8 1TL, UK}

\author[0000-0003-2269-0551]{Perry~Greenfield}
\affiliation{\affstsci}

\author[0000-0002-6508-2938]{Austen~M.~Groener}
\affiliation{Drexel University Department of Physics, 816 Disque Hall 32 S. 32nd Street Philadelphia, PA 19104}

%FLAG
\author{Steve~Guest}
\affiliation{RAL Space}

%FLAG
\author[0000-0001-5848-4077]{Sebastian~Gurovich}
\affiliation{IATE-OAC-UNC}

\author[0000-0001-8725-4502]{Rasmus~Handberg}
\affiliation{Stellar Astrophysics Centre, Department of Physics and Astronomy, Aarhus University, Ny Munkegade 120, DK-8000 Aarhus C, Denmark.}

\author[0000-0002-8138-1479]{Akeem~Hart}
\noaffiliation

%FLAG
\author[0000-0002-8646-8362]{Zac~Hatfield-Dodds}
\affiliation{Australian National University, Canberra, Australia}

\author[0000-0002-8546-9128]{Derek~Homeier}
\affiliation{\affaperio}
\affiliation{F{\"o}rderkreis Planetarium G{\"o}ttingen, G{\"o}ttingen, Germany}

\author[0000-0002-0832-2974]{Griffin~Hosseinzadeh}
\affiliation{Steward Observatory, University of Arizona, 933 North Cherry Avenue, Tucson, AZ 85721-0065, USA}

\author{Anany~Shrey~Jain}
\noaffiliation

\author[0000-0001-5982-167X]{Tim~Jenness}
\affiliation{Vera C. Rubin Observatory, 950 N. Cherry Ave, Tucson, AZ 85719}

%FLAG
\author[0000-0002-0629-3006]{Craig~K~Jones}
\affiliation{The Malone Center for Healthcare in Engineering, Johns Hopkins University}

\author[0000-0003-1409-1903]{Prajwel~Joseph}
\affiliation{Indian Institute of Astrophysics, Bangalore 560034, India}
\affiliation{Department of Physics, CHRIST (Deemed to be University), Bangalore 560029, India}

\author[0000-0002-6825-5283]{J.~Bryce~Kalmbach}
\affiliation{DIRAC Institute, Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195}

\author[0000-0001-6209-838X]{Emir~Karamehmetoglu}
\affiliation{Aarhus University, Department of Physics and Astronomy, Ny Munkegade 120, DK-8000 Aarhus C, Denmark}

\author[0000-0003-0109-5833]{Miko{\l}aj~Ka{\l}uszy{\'{n}}ski}
\affiliation{Nicolaus Copernicus Astronomical Center, Polish Academy of Sciences, Bartycka 18, 00-716 Warsaw, Poland}
\affiliation{AkondLab, Walbrzyska 6-8, Wroclaw, Poland}

\author[0000-0002-6702-7676]{Michael~S.~P.~Kelley}
\affiliation{Department of Astronomy, University of Maryland, 4296 Stadium Drive, College Park, MD 20742-0001}

\author[0000-0002-8211-1892]{Nicholas~Kern}
\affiliation{Massachusetts Institute of Technology / Dept. of Physics, 182 Memorial Dr, Cambridge, USA}

\author[0000-0002-0479-7235]{Wolfgang~E.~Kerzendorf}
\affiliation{Department of Physics \& Astronomy/Computational Mathematics, Science, and Engineering, Michigan State University, East Lansing, MI 48824, USA}

\author[0000-0001-9605-780X]{Eric~W.~Koch}
\affiliation{\affcfa}

\author[0000-0002-4472-8517]{Samuel~J.~Van~Kooten}
\affiliation{Southwest Research Institute, 1050 Walnut St. Ste 300, Boulder CO 80302, USA}

\author[0000-0002-7822-0471]{Shankar~Kulumani}
\affiliation{Johns Hopkins University Applied Physics Laboratory, Laurel, MD USA}

\author[0000-0003-2193-5369]{Antony~Lee}
\affiliation{Laboratoire Photonique, Numérique et Nanosciences UMR 5298, Université de Bordeaux, Institut d'Optique, CNRS, 33400 Talence, France}

%FLAG
\author[0000-0002-4245-2318]{Chun~Ly}
\affiliation{University of Arizona}

\author[0000-0002-3713-6337]{Brigitta~M.~Sip\H{o}cz}
\affiliation{\affipac}

%FLAG
\author[0000-0003-3270-6844]{Zhiyuan~Ma}
\affiliation{University of Massachusetts Amherst}

\author[0000-0002-9901-8723]{Conor~MacBride}
\affiliation{Astrophysics Research Centre, School of Mathematics and Physics, Queen's University Belfast, Belfast, BT7 1NN, UK}

%FLAG
\author[0000-0001-8100-9514]{Jakob~~M.~Maljaars}
\affiliation{Science \& Technology Corporation}

\author[0000-0001-6628-8033]{N.~A.~Murphy}
\affiliation{\affcfa}

%FLAG
\author[0000-0003-4189-3450]{Henrik~Norman}
\affiliation{Winter Way, Lerum, Sweden}

\author[0000-0002-2432-8946]{Richard~O'Steen}
\affiliation{\affstsci}

\author[0000-0001-9857-7788]{Kyle~A.~Oman}
\affiliation{Institute for Computational Cosmology, Durham University, South Road, Durham DH1 3LE, United Kingdom}
\affiliation{Department of Physics, Durham University, South Road, Durham DH1 3LE, United Kingdom}

\author{Camilla~Pacifici}
\affiliation{\affstsci}

\author[0000-0002-9351-6051]{Sergio~Pascual}
\affiliation{Departamento de F\'{i}sica de la Tierra y Astrof\'{i}sica, Fac. CC. F\'{i}sicas, Universidad Complutense de Madrid, E28040, Madrid, Spain}
\affiliation{Instituto de F\'{i}sica de Part\'{i}culas y del Cosmos, IPARCOS, Fac. CC. F\'{i}sicas, Universidad Complutense de Madrid, E28040, Madrid, Spain}

\author[0000-0003-0139-6951]{J.~Pascual-Granado}
\affiliation{Instituto de Astrof\'{\i}sica de Andaluc\'{\i}a (CSIC). Glorieta de la Astronom\'{\i}a s/n. 18008, Granada, Spain}

%FLAG
\author[0000-0002-5923-4966]{Rohit~R~Patil}
\affiliation{Syracuse University, Syracuse, NY}

\author[0000-0002-1852-9653]{Gabriel~I~Perren}
\affiliation{Instituto de Astrof\'isica de La Plata (IALP-CONICET), La Plata, Argentina}

%FLAG
\author[0000-0002-9427-5448]{Timothy~E~Pickering}
\affiliation{MMT Observatory}

\author{Tanuj~Rastogi}
\noaffiliation

\author[0000-0002-9453-7735]{Benjamin~R.~Roulston}
\affiliation{\affcfa}
\affiliation{Department of Astronomy, Boston University, 725 Commonwealth Avenue, Boston, MA 02215, USA}

\author[0000-0001-8661-3825]{Daniel~F~Ryan}
\affiliation{University of Applied Sciences and Arts Northwest Switzerland, Bahnhofstrasse 6, 5210 Windisch, Switzerland}
\affiliation{American University, 4400 Massachusetts Avenue, NW Washington, DC 20016, USA}

\author[0000-0001-9376-3135]{Eli~S.~Rykoff}
\affiliation{SLAC National Accelerator Laboratory, Menlo Park, CA 94025, USA}

\author[0000-0003-1149-6294]{Jose~Sabater}
\affiliation{STFC UK Astronomy Technology Centre, Royal Observatory, Blackford Hill, Edinburgh, EH9 3HJ, UK}

%FLAG
\author[0000-0002-9523-5640]{Parikshit~Sakurikar}
\affiliation{International Institute of Information Technology - Hyderabad}

%FLAG
\author[0000-0002-3680-4364]{Jes\'us~Salgado}
\affiliation{SKAO Observatory, Jodrell Bank, Cheshire, UK}

\author[0000-0002-1838-4757]{Aniket~Sanghi}
\affiliation{Department of Astronomy, The University of Texas at Austin, 2515 Speedway, C1400, Austin, TX 78712, USA}

\author[0000-0003-2657-3889]{Nicholas~Saunders}
\affiliation{Institute for Astronomy, University of Hawaiʻi at M\=anoa, 2680 Woodlawn Drive, Honolulu, HI 96822, USA}
\affiliation{American Museum of Natural History, 200 Central Park West, Manhattan, NY 10024, USA}
\altaffiliation{NSF Graduate Research Fellow}

%FLAG
\author[0000-0001-6353-0808]{Volodymyr~Savchenko}
\affiliation{University of Geneva, Department of Astronomy}

\author{Ludwig~Schwardt}
\affiliation{South African Radio Astronomy Observatory, Black River Park, 2 Fir Street, Observatory 7925, South Africa}

\author{Michael~Seifert-Eckert}
\noaffiliation

\author[0000-0001-6874-2594]{Albert~Y.~Shih}
\affiliation{NASA Goddard Space Flight Center, 8800 Greenbelt Road, Greenbelt, MD 20771, USA}

%FLAG
\author[0000-0001-9268-3803]{Gyanendra~Shukla}
\affiliation{Maulana Azad National Institute Of Technology, Bhopal}

\author[0000-0003-3001-676X]{Jonathan~Sick}
\affiliation{J.Sick Codes Inc., Penetanguishene ON Canada}

\author[0000-0001-8589-4055]{Chris~Simpson}
\affiliation{Gemini Observatory/NSF's NOIRLab, 670 N. A`ohoku Place, Hilo HI 96720}

\author[0000-0003-2462-7273]{Sudheesh~Singanamalla}
\affiliation{Paul G. Allen School of Computer Science and Engineering, University of Washington, Seattle, WA, USA}

\author[0000-0001-9898-5597]{Leo~P.~Singer}
\affiliation{Astroparticle Physics Laboratory, NASA Goddard Space Flight Center, Code 661, Greenbelt, MD 20771, USA}

%FLAG
\author[0000-0002-8310-0829]{Jaladh~Singhal}
\affiliation{TARDIS collaboration}

\author[0000-0002-4845-1228]{Manodeep~Sinha}
\affiliation{Centre for Astrophysics \& Supercomputing, Swinburne University of Technology, Hawthorn, VIC 3122, Australia}
\affiliation{ARC Centre of Excellence for All Sky Astrophysics in 3 Dimensions (ASTRO 3D)}

\author[0000-0001-5185-9876]{Lee~R.~Spitler}
\affiliation{Department of Physics and Astronomy, Macquarie University, Sydney, NSW 2109, Australia}

\author[0000-0002-1365-1908]{David~Stansby}
\affiliation{Mullard Space Science Laboratory, University College London, Holmbury St. Mary, Surrey, RH5 6NT, UK}

\author[0000-0001-7751-1843]{Ole~Streicher}
\affiliation{Leibniz Institute for Astrophysics, An der Sternwarte 16, 14482 Potsdam, Germany}

\author{Jani~\v{S}umak}
\noaffiliation

\author[0000-0001-9445-1846]{John~D.~Swinbank}
\affiliation{ASTRON, Oude Hoogeveensedijk 4, 7991 PD Dwingeloo, The Netherlands}

\author[0000-0001-6268-1882]{Dan~S.~Taranu}
\affiliation{\affprinceton}

\author{Nikita~Tewary}
\affiliation{Bharati Vidyapeeth's College of Engineering/ Electronics and Communication Department, A-4 Block, Baba Ramdev Marg, Shiva Enclave, Paschim Vihar, New Delhi, 110063}

\author[0000-0002-5445-5401]{Grant~R.~Tremblay}
\affiliation{\affcfa}

\author[0000-0002-0455-9384]{Miguel~de~Val-Borro}
\affiliation{Planetary Science Institute, 1700 East Fort Lowell, Tucson, AZ 85719}

\author{Zlatan~Vasovi\'{c}}
\affiliation{Faculty of Physics, University of Belgrade, Studentski trg 12-16, Belgrade, Serbia}

\author[0000-0003-0370-5471]{Shresth~Verma}
\affiliation{Atal Bihari Vajpayee Indian institute of Information Technology and Management Gwalior, MP, India}

\author[0000-0003-3734-3587]{Peter~K.~G.~Williams}
\affiliation{\affcfa}
\affiliation{American Astroomical Society, 1667 K St. NW Suite 800, Washington, DC 20006, USA}

\author[0000-0001-6352-9735]{Tom~J.~Wilson}
\affiliation{School of Physics, University of Exeter, Stocker Road, Exeter, EX4 4QL, UK}
\affiliation{\affstsci}

\author[0000-0001-6999-3635]{Benjamin~Winkel}
\affiliation{Max-Planck-Institut f{\"u}r Radioastronomie, Auf dem H{\"u}gel 69, 53121 Bonn, Germany}

\author[0000-0001-7113-1233]{W.~M.~Wood-Vasey}
\affiliation{Pittsburgh Particle Physics, Astrophysics, and Cosmology Center (PITT PACC). Physics and Astronomy Department, University of Pittsburgh, Pittsburgh, PA 15260, USA}

\author[0000-0001-7689-9305]{Rui~Xue}
\affiliation{National Radio Astronomy Observatory, 520 Edgemont Road, Charlottesville, VA 22903, USA}

\author[0000-0003-2874-6464]{Peter~Yoachim}
\affiliation{Department of Astronomy, University of Washington, Seattle, WA 98195, USA}

%FLAG
\author[0000-0002-9583-263X]{Chen~ZHANG}
\affiliation{Purple Mountain Observatory, CAS}

\author[0000-0001-6841-1058]{Andrea~Zonca}
\affiliation{San Diego Supercomputer Center, University of California San Diego, 9500 Gilman Dr, La Jolla, CA 92093, USA}

### 2022-03-16 src/figures/python-mentions.py
apply suggestions from ADS help desk

@@ -81,10 +81,9 @@ def get_paper_count(query, rows_per_page=500, max_pages=100):
     search_query : `ads.SearchQuery`
     count : int
     """
-    q = ads.SearchQuery(q=query)
-    for paper in q:
-        break
-    return q, q.response.numFound
+    q = ads.SearchQuery(q=query, rows=1)
+    q.execute()
+    return q.response.numFound()
 
 
 def get_dfs(cache_path):
@@ -182,7 +181,7 @@ if not this_cache_file.exists() and api_success:
     total_astro_counts = []
     years = np.arange(min_year, 2022+1)
     for year in tqdm(years):
-        q, count = get_paper_count(
+        count = get_paper_count(
             f'collection:astronomy year:{year} property:refereed'
         )
         total_astro_counts.append(count)

### 2022-03-16 src/ms.tex
minor edits to new AAT section

--- 2022-03-15 src/ms.tex
@@
 
+\subsubsection{Data Central and the Anglo-Australian Telescope Archive}
+% \secauthor{James Tocknell}
+Data Central is one of the five Australian Virtual Observatory\footnote{known as
+  the All-Sky Virtual Observatory, or ASVO} nodes, providing both Australian and
+international astronomers access to and services for surveys of the southern
+sky. Whilst primarily hosting optical data, including the Anglo-Australian
+Telescope (AAT) Archive\footnote{The AAT Archive is currently jointly managed by
+  Data Central, Astronomy Australia Limited (AAL), and National Computational
+  Infrastructure (NCI)}, Data Central also assists theoretical and radio
+astronomers host and manage their data.
+
+\python is widely used by Data Central, and \astropy and its wider ecosystem
+play a vital role in the initial ingestion, reduction, access and visualization
+of the hosted data and services. For example, the AAT Archive hosts data back to
+1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest
+data flowing from the AAT to ensure that a consistent time and coordinate system
+is provided to users for simpler querying.
+
+As many of the surveys hosted by Data Central are spectroscopic surveys from the
+AAT, \package{specutils} is heavily used to standardize the variety of encodings
+of spectroscopic data into a form easily readable and visualizable by many
+standard tools (such as \package{ds9} or \package{splat}). Data Central has
+upstreamed into \package{specutils} loaders for all hosted spectra, including
+2dfdr-reduced spectra from the AAT Archive, and strongly encourages other
+archives to provide loaders for their data to further reuse of existing archival
+data.
+
 % Rubin Observatory

### 2022-03-16 src/figures/python-mentions.py
typo

@@ -83,7 +83,7 @@ def get_paper_count(query, rows_per_page=500, max_pages=100):
     """
     q = ads.SearchQuery(q=query, rows=1)
     q.execute()
-    return q.response.numFound()
+    return q.response.numFound
 
 
 def get_dfs(cache_path):

### 2022-03-16 src/ms.tex
fix ROSES grant number

--- 2022-03-16 src/ms.tex
@@
 
-\subsubsection{Data Central and the Anglo-Australian Telescope Archive}
-\secauthor{James Tocknell}
-Data Central is one of the five Australian Virtual Observatory\footnote{known as
-  the All-Sky Virtual Observatory, or ASVO} nodes, providing both Australian and
-international astronomers access to and services for surveys of the southern
-sky. Whilst primarily hosting optical data, including the Anglo-Australian
-Telescope (AAT) Archive\footnote{The AAT Archive is currently jointly managed by
-  Data Central, Astronomy Australia Limited (AAL), and National Computational
-  Infrastructure (NCI)}, Data Central also assists theoretical and radio
-astronomers host and manage their data.
-
-Python is widely used by Data Central, and \astropy and its wider ecosystem play
-a vital role in the initial ingestion, reduction, access and visualisation of
-the hosted data and services. For example, the AAT Archive hosts data back to
-1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest
-data flowing from the AAT to ensure that a consistent time and coordinate
-system is provided to users for simpler querying.
-
-As many of the surveys hosted by Data Central are spectroscopic surveys from the
-AAT, specutils is heavily used to standardise the variety of encodings of
-spectroscopic data into a form easily readable and visualisable by many standard
-tools (such as ds9 or splat). Data Central has upstreamed into specutils loaders
-for all hosted spectra, including 2dfdr-reduced spectra from the AAT Archive,
-and strongly encourages other archives to provide loaders for their data to
-further reuse of existing archival data.
-
 % Rubin Observatory
@@
 
+We acknowledge the Gordon and Betty Moore foundation for their continued
+financial support.
+This work is partially supported by NASA under Grant No. 80NSSC22K0347 issued
+through the NASA ROSES program.
+This work is partially supported by the international Gemini Observatory, a
+program of NSF's NOIRLab, which is managed by the Association of Universities
+for Research in Astronomy (AURA) under a cooperative agreement with the National
+Science Foundation, on behalf of the Gemini partnership of Argentina, Brazil,
+Canada, Chile, the Republic of Korea, and the United States of America.
+We also thank NumFOCUS and the Python Software Foundation for financial
+support.
+
 The \astropy community is supported by and makes use of a number of
@@
 
-Supported by the international Gemini Observatory, a program of NSF's NOIRLab,
-which is managed by the Association of Universities for Research in Astronomy
-(AURA) under a cooperative agreement with the National Science Foundation, on
-behalf of the Gemini partnership of Argentina, Brazil, Canada, Chile, the
-Republic of Korea, and the United States of America.
-
 Furthermore, the \astropy packages would not exist in their current form without
@@
           \package{matplotlib} \citep{matplotlib},
-          \package{Cython} \citep{cython}.
-          % \package{photutils} \citep{photutils},
-          % \package{specutils} \citep{specutils},
-          % \package{regions} \citep{regions},
-          % \package{gwcs} \citep{gwcs},
-          % \package{synphot} \citep{synphot},
-          % \package{ASDF} \citep{ASDF}.
+          \package{Cython} \citep{cython},
+          \package{photutils} \citep{photutils},
+          \package{specutils} \citep{specutils},
+          \package{regions} \citep{regions},
+          \package{gwcs} \citep{gwcs},
+          \package{synphot} \citep{synphot},
+          \package{ASDF} \citep{ASDF}.
         %   \package{SOFA} \citep{sofa},

### 2022-03-17 src/ms.tex

@@
 a total of 107 merged pull requests were contributed to astropy.
+
+\subsubsection{Data Central and the Anglo-Australian Telescope Archive}
+% \secauthor{James Tocknell}
+Data Central is one of the five Australian Virtual Observatory\footnote{known as
+  the All-Sky Virtual Observatory, or ASVO} nodes, providing both Australian and
+international astronomers access to and services for surveys of the southern
+sky. Whilst primarily hosting optical data, including the Anglo-Australian
+Telescope (AAT) Archive\footnote{The AAT Archive is currently jointly managed by
+  Data Central, Astronomy Australia Limited (AAL), and National Computational
+  Infrastructure (NCI)}, Data Central also assists theoretical and radio
+astronomers host and manage their data.
+
+\python is widely used by Data Central, and \astropy and its wider ecosystem
+play a vital role in the initial ingestion, reduction, access and visualization
+of the hosted data and services. For example, the AAT Archive hosts data back to
+1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest
+data flowing from the AAT to ensure that a consistent time and coordinate system
+is provided to users for simpler querying.
+
+As many of the surveys hosted by Data Central are spectroscopic surveys from the
+AAT, \package{specutils} is heavily used to standardize the variety of encodings
+of spectroscopic data into a form easily readable and visualizable by many
+standard tools (such as \package{ds9} or \package{splat}). Data Central has
+upstreamed into \package{specutils} loaders for all hosted spectra, including
+2dfdr-reduced spectra from the AAT Archive, and strongly encourages other
+archives to provide loaders for their data to further reuse of existing archival
+data.
 

### 2022-03-17 src/figures/contributor-summary.py
Add section and plots describing contributor statistics (#39)
* add contributor section figure script

* add figures to tex file

* fix figures build

* new figure and words

* numbers

* words about top contributors

* caption and figure title

* titles

* captions

* another title

* windowed

* final paragraph on rates

* add scipy as dependency

* oops - get rid of APW-specific path

oops also clone astropy

parameter name

paths

paths

fetch depth of astropy

* merge

* remove duplicate numpy ref

@@ -1,13 +1,413 @@
-"""
-TODO: This currently just produces a placeholder figure!
-"""
-import numpy as np
+# -*- coding: utf-8 -*-
+# Ideas for plots:
+# - Commits per month
+# - Commiters per month
+# - Lines of code added/removed per month (harder)
+# - Contributors x commits in bins of astropy eras
+# - Breakdown of career stage of top N committers in different eras (early, mid, present)
+#     - Eras: 2012–2015, 2015–2018, 2018–present
+#     - Or maybe only include committers that contribute a certain number of line changes?
+#
+#
+# Stretch goals:
+# - For committers, did they start by opening an issue or go right to PR?
+
+# ---
+
+# +
+from collections import defaultdict
+import datetime
+import os
+import pathlib
+
+from astropy.time import Time
+import astropy.units as u
+from git import Repo
 import matplotlib.pyplot as plt
+import numpy as np
+from scipy.ndimage import generic_filter
+# -
 
-rng = np.random.default_rng(42)
+_astropy_path = os.environ.get('ASTROPY_REPO_PATH', '../../../astropy')
+astropy_repo_path = pathlib.Path(_astropy_path)
 
-fig, ax = plt.subplots(1, 1, figsize=(6, 6))
-xy = rng.uniform(size=(2, 128))
-ax.plot(*xy, linestyle='none')
+repo = Repo(astropy_repo_path)
 
-fig.savefig("contributor-summary.pdf", bbox_inches="tight")
+first_commit_hash = repo.git.log('--pretty=format:%h', '--reverse').split('\n')[0]
+first_commit = list(repo.iter_commits(max_count=1, rev=first_commit_hash))[0]
+first_date = Time(first_commit.committed_date, format='unix')
+first_date.datetime
+
+# Date of last commit in v5.0
+v50_date = Time(datetime.datetime(2021, 11, 19))
+total_days = int((v50_date - first_date).jd)
+total_days
+
+# ### Time ranges:
+
+month_bins = (
+    first_date + np.arange(0, total_days, 30).astype(int) * u.day
+)
+
+# +
+# 4 bin edges for 3 eras
+era_edges = (
+    first_date + np.linspace(0, total_days, 4).astype(int) * u.day
+)
+
+eras = {
[…]

### 2022-03-18 src/ms.tex
add git commit or tag name to latex header
starting to sketch challenges

@@
 \begin{abstract}
-The \astropy Project supports and fosters the development of open-source and openly-developed
-\python packages that provide commonly-needed functionality to the astronomical
+The \astropy Project supports and fosters the development of open-source and openly developed
+\python packages that provide commonly needed functionality to the astronomical
 community.
@@
 release, version 5.0, and provide major updates for the project.
-We then discuss supporting a broader ecosystem of inter-operable packages,
+We then discuss supporting a broader ecosystem of interoperable packages,
 including connections with several astronomical observatories and missions.
@@
 Today, a broad and feature-diverse ``ecosystem'' of packages exists in the
-\python scientific computing landscape: Roughly ordered from general-use to
+\python scientific computing landscape: roughly ordered from general-use to
 domain-specific, this landscape now includes packages that provide core
@@
 \citep{ape2}. The motivation for LTS releases is to provide longer-term stable
-versions of \astropypkg that users that require a high level of stability can
-make use of if they do not always need the latest features (this could include,
-for example, telescope operation or data reduction pipelines, etc.). The v5.0
-release of the core package was designated as LTS, and since it was released at
-the end of 2021, it will be maintained until the end of 2023.
+versions of \astropypkg that users who require a high level of stability can
+make use of if they do not always need the latest features (this includes but
+not limited to telescope operations and data reduction pipelines). The v5.0
+release of the core package, replacing v4.0, was designated as LTS at the end of 2021;
+it will be maintained until the end of 2023.
 
@@
 points implies that all data points are now covariant with each other, i.e., one
-has to carry $N\times N$ covariances. For a large image, that becomes not just
-pointless but also prohibitive in terms of memory and CPU usage.
-
-\paragraph{Masked quantities}  It can be useful to mask bad data. How one
-supports masked data involves a number of choices, such as whether one indicates
-data values are bad by a separate flag or by replacing bad data with a special
-value, such as ``Not a Number'' (NaN). In \astropypkg, both approaches have
-been used: a flag for masked columns in \astropyTable, mask flags and bitmaps
+has to carry $N\times N$ covariances. For a large image, that becomes unnecessary
+and computationally expensive.
+
+\paragraph{Masked quantities}  It can be useful to mask bad data, either by marking
+bad data values with a flag or by replacing them with a special
+value, e.g., ``Not a Number'' (NaN). In \astropypkg, both approaches are
+used: a flag for masked columns in \astropyTable, mask flags and bitmaps
 for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
-NaN in \astropyTime. However masked quantities were poorly supported, making it
-difficult to use masks in a \astropyQTable\ (in which \astropyQuantity\ is used
+NaN in \astropyTime. However masked \astropyQuantity support is currently very limited,
+hindering usage of masks in a \astropyQTable\ (in which \astropyQuantity\ is used
 for all columns with units).
 
-Unfortunately, the \texttt{MaskedArray} class from \package{numpy} only works
+Furthermore, the \texttt{MaskedArray} class from \package{numpy} only works
 well with plain data arrays, hiding other metadata and attributes, for example
@@
 
-Units were also integrated further within \astropypkg. For an already-defined unit-less
-\astropyModel, e.g. those imported from another library, \astropymodeling can
-now coerce units.
-% TODO! transition words
-A new module – \astropycosmologyunits – has been added to the cosmology
-sub-package for defining and collecting cosmological units and equivalencies.
-The unit \texttt{littleh} and equivalency
-\astropyapidoc{cosmology.units.with_H0}{with\_H0} was moved from the main
-\astropyunits sub-package to \astropycosmologyunits. A new unit,
-\texttt{redshift}, has been added for tracking factors of cosmological redshift.
-To correctly use redshift units in dimensional analysis, \texttt{redshift} is
-\astropyapi{cosmology.units.dimensionless_redshift}{treated as dimensionless}
-by default. To convert between redshift and other cosmological distance
-measures, e.g. CMB temperature or comoving distance, the equivalency
-\astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
-has also been added. This equivalency is actually a composite of other equivalencies,
-which may be used separately.
+Units were also integrated further within \astropypkg. For an already-defined
+unit-less \astropyModel, e.g., those imported from another library,
+\astropymodeling can now coerce units. A new module – \astropycosmologyunits –
+has been added to the cosmology sub-package for defining and collecting
+cosmological units and equivalencies. \astropycosmologyunits has a unit and
+equivalencies for \texttt{littleh}, and a new unit, \texttt{redshift}, for
+factors of cosmological redshift. \texttt{redshift} has a number of
+equivalencies for converting between cosmological distance measures, e.g., CMB
+temperature or comoving distance.
 
@@
 \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
-mixin protocol for seamless interoperability}.}, the \astropyfits package now
-interprets time data correctly, using \astropyTime\ as a `mixin' column. For
-backward compatibility with manual systems, this feature may be turned off.
+mixin protocol for seamless interoperability}.} the \astropyfits package now
+interprets time data correctly, using \astropyTime\ as a mixin column. For
+backward compatibility, this feature may be turned off.
 
@@
 functions for \astropypkg objects, such as reading from or writing to a file,
-following a given standard. \astropy uses this internally for the I/O methods in
-\astropyTable\ and \astropyCosmology, and users can register custom I/O to
+following a given standard. \astropypkg uses this internally for the I/O methods in
+\astropyTable\ and \astropyCosmology. Users can also register custom I/O to
 extend the options on these classes.
 
-As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
-number of new use cases. for instance, the class-based architecture allows for
-the creation of custom registries. One application is in \astropycosmology,
+As of \astropypkg v5.0, the I/O registry submodule has been generalized to enable a
+number of new use cases; for instance, the class-based architecture allows
+streamlining of the creation of custom registries while minimizing code duplication.
+One application is in \astropycosmology,
 which has two different \textit{kinds} of registries for \astropyCosmology: one
-for reading and writing files, and another for converting between python
-objects. Using \astropyiosubpkg{registry} for custom classes means a developer
-does not have to create a whole new I/O system and can offer to the user a
-unified and familiar API.
-
-\subsection{Precision, Accuracy, \& Reproducibility}
+for reading and writing files, and another for converting between \python objects.
+
+\subsection{Numerical Precision, Accuracy, \& Reproducibility}
 \label{sec:core-features-precision_accuracy_repreducibility}
@@
 In astronomy, time accuracy down to the (nano)second is frequently important.
-Whether for planning observations, crunching pulsar or VLBI data, missing
+Whether for planning observations or crunching pulsar or VLBI data, missing
 seconds meaningfully impact results. In past versions of \astropypkg, when using
-\astropyTime\ in \astropytime leap seconds had to be manually applied; now, leap
+\astropyTime\, leap seconds had to be manually applied; now, leap
 seconds are applied automatically. Moreover, \astropypkg updates internal time
-data-files to ensure that the correct leap second adjustment is always used. For
-reproducibility, e.g., with an old code, \astropyLeapSeconds\ may also be
+data files to ensure that the correct leap second adjustment is always used. For
+reproducibility with older code, \astropyLeapSeconds\ may also be
 manually applied.
@@
 \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
-2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
+2018}}. \astropyconstants now defaults to using the \texttt{SI/CODATA 2018}
 values, with the units in \astropyunits based on these constants. Most of
 \astropypkg and affiliate packages build upon \astropyunits, so this update
-effects the entire \astropy ecosystem.
+affects the entire \astropy ecosystem.
 
@@
 \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
-and associated transformations have been updated to correctly reflect the “true”
-and “mean” terminology.
+and associated transformations have been updated to correctly reflect the ``true''
+and ``mean" terminology.
 
@@
 Performance in \astropypkg is addressed based on feedback and requests, and
-through contributions from community members: If you find that parts of
+through contributions from community members: if you find that parts of
 \astropypkg are not performant, we encourage you to open a \github issue to
@@
 Astropy Project as the ultimate sources of authority, as well as the core
-principle of ``do-ocracy'' that those who do work for the Project (be it coding,
+principle of ``do-ocracy'': those who do work for the Project (be it coding,
 training, or other harder-to-quantify contributions) gain more influence on the
@@
 when other mechanisms have failed, and includes powers that either
-require central authority or secrets (e.g., passwords) However, APE0 also
+require central authority or secrets (e.g., passwords). However, APE0 also
 charges the Committee to devolve responsibilities and seek community input on
@@
             \textbf{Top panel:} The number of Git commits per month (defined as
-            consecutive 30 day windows) merged into to the \astropypkg core
+            consecutive 30 day windows) merged into the \astropypkg core
             package over the whole history of the project (up to end of 2021,
@@
 own software package infrastructure. These include tools and extensions that
-enable easily generating documentation and setting up automated testing, as well
+enable generating documentation and setting up automated testing, as well
 as providing package scaffolding for new projects.
@@
 Python packages leveraging the Astropy ecosystem. This tool utilizes the
-Cookiecutter project to walk users through the process of creating new
-packages complete with documentation and testing support. Additionally, the
-package template generation process includes the ability to setup
-interoperability with \github, allowing for easy repository access from
-documentation, as well as an example \github Actions workflow to demonstrate
-the use of \github's continuous integration tooling.
+\href{https://github.com/cookiecutter/cookiecutter}{Cookiecutter} project to
+walk users through the process of creating new packages complete with
+documentation and testing support. Additionally, the package template
+generation process includes the ability to setup interoperability with
+\github, allowing for easy repository access from documentation, as well as an
+example \github Actions workflow to demonstrate the use of \github's
+continuous integration tooling.
 
@@
 services, including the James Webb Space Telescope (JWST) archive, and the
-project has switched to a continuous release model: Every time a change is
+project has switched to a continuous release model: every time a change is
 committed to the main development branch it is published on the Python Package
 Index (PyPI) and available for installation. (However, formal releases are still
-done a few times per year.) \texttt{photutils} \cite {photutils} released its
+done a few times per year.) \texttt{photutils} \cite{photutils} released its
 first stable version, indicating that the API will change less frequently, and
@@
 A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
-a python package developed to perform all the necessary tasks to from the raw data
+a python package developed to perform all the necessary tasks to perform data processing, from the raw data
 of Cherenkov telescopes to the reconstructed event lists.
@@
 Both Gammapy and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
-tables and FITS I/O.
+tables, and FITS I/O.
 As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
@@
 % \secauthor{James Tocknell}
-Data Central is one of the five Australian Virtual Observatory\footnote{known as
-  the All-Sky Virtual Observatory, or ASVO} nodes, providing both Australian and
+Data Central is one of the five Australian Virtual Observatory\footnote{Known as
+  the All-Sky Virtual Observatory, or ASVO.} nodes, providing both Australian and
 international astronomers access to and services for surveys of the southern
@@
   Data Central, Astronomy Australia Limited (AAL), and National Computational
-  Infrastructure (NCI)}, Data Central also assists theoretical and radio
-astronomers host and manage their data.
+  Infrastructure (NCI).}, Data Central also assists theoretical and radio
+astronomers with hosting and managing their data.
 
@@
 play a vital role in the initial ingestion, reduction, access and visualization
-of the hosted data and services. For example, the AAT Archive hosts data back to
+of the hosted data and services. For example, the AAT Archive hosts data from as early as
 1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest
@@
 2dfdr-reduced spectra from the AAT Archive, and strongly encourages other
-archives to provide loaders for their data to further reuse of existing archival
+archives to provide loaders for their data to further the reuse of existing archival
 data.
@@
 appearance of the \href{https://learn.astropy.org}{Learn Astropy website}.
-The Learn Team meets weekly to work on creating, expanding, improving these
+The Learn Team meets weekly to work on creating, expanding, and improving these
 educational resources.
@@
 Now that the backend infrastructure is stable, we are interested in collecting
-new content to serve and share: If you have ideas for contributions, or have
+new content to serve and share: if you have ideas for contributions, or have
 tutorials hosted on third-party sites that you would like to make searchable
@@
 expanded to Summer AAS meetings.
-The Learn team finds that the workshop audience is best found as AAS meetings as
+The Learn team finds that the workshop audience is best found at AAS meetings as
 opposed to more general \python meetings, as the content tends to be more
@@
 
+% \begin{itemize}
+% \item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+% be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+% open-source projects.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+% a Discourse site which is more open than the Facebook group and more user-oriented
+% than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+% developers are able to invoice as independent contractors the time they
+% spend helping users on these forums.
+% \end{itemize}
+
 \section{Future Plans for the Astropy Project} \label{sec:future}
@@
 
-Community Engagement
+- Astropy is established; challenges faced have evolved
+-
 
 \begin{itemize}
-\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
-This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
-The Moore Foundation grant allows this Slack space to
-be on a paid plan; additionally NumFOCUS has negotiated a special rate for
-open-source projects.
-While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
-The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
-a Discourse site which is more open than the Facebook group and more user-oriented
-than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
-developers are able to invoice as independent contractors the time they
-spend helping users on these forums.
-\end{itemize}
-
-\begin{itemize}
-\item Attracting new contributors when the code has become quite complex,
-\item Contributor to maintainer mentoring,
-\item Long-term / sustained funding for maintaining infrastructure,
-\item ...
+\item Attracting new contributors when the code has become quite complex, project established (astropy as an "institution", walled-off group). (how to attract? Fast feedback and code reviews, ...)
+\item Support for existing contributors: Contributor and maintainer mentoring (...)
+\item Infrastructure: Long-term / sustained funding for maintaining infrastructure (grants last 3 years; infrastructure evolves, nature of open source ecosystem)
+\item As ecosystem grows, making sure users know about all of the functionality (workshops, tutorials, educational materials, discoverability, accessibility) (Continued engagement with user community)
 \end{itemize}
@@
 The Astropy Project is also expanding its educational and community-focused
-initiatives, by soliciting and funding initiatives in inclusion, diversity, and
-equity (Section~\ref{sec:project-ide}) and through the Learn Astropy initiative,
+initiatives by soliciting and funding initiatives in inclusion, diversity, and
+equity (Section~\ref{sec:project-ide}) as well as through the Learn Astropy initiative,
 which aims to make tutorials and educational content that highlight software
@@
 We would like to thank the members of the community who have contributed to
-\astropy, who have opened issues and provided feedback, and have supported the
+\astropy, opened issues and provided feedback, and supported the
 project in a number of different ways.
@@
 summer. These students often turn into long-term contributors.
-Within the academic community, we thank institutions that make it possible that
-astronomers and other developers on their staff can contribute their time to the
+Within the academic community, we thank institutions that make it possible for
+astronomers and other developers on their staff to contribute their time to the
 development of \astropy projects.

### 2022-03-18 src/scripts/make-repo-info.py
make repo info script for git header in latex

@@ -1,0 +1,21 @@
+import pathlib
+from git import Repo
+
+this_path = pathlib.Path(__file__).parent.resolve()
+
+repo = Repo(this_path / '../..')
+try:
+    tag = repo.git.describe("HEAD", exact_match=True, tags=True)
+except Exception:
+    tag = None
+    short_sha = repo.head.commit.hexsha[:7]
+
+if tag is None:
+    header = f"Git commit: {short_sha}"
+else:
+    header = f"Git tag: {tag}"
+
+command = r"\newcommand{\gitHeader}" + f"{{{header}}}" + "\n"
+
+with open(this_path / '../generated/git-header.tex', "w") as f:
+    f.write(command)

### 2022-03-18 src/ms.tex
intro tweaks and title case

@@
 In particular, the transition from focusing our energy on development and
-maintenance of a single core package, to instead sustaining the core package and
-fostering the development of the community and its expansion has been a key
-issue faced by the Astropy Project in the last several years.
+maintenance of a single core package, to instead maintaining the core package
+and fostering the development of the community and its expansion has been a key
+goal of the Astropy Project over the last several years.
 

### 2022-03-18 src/ms.tex
Lim edit: Beginning - end of Sec 2.2

--- 2022-03-18 src/ms.tex
@@
 categories: new features, intra- and inter-package interoperability,
-improvements to numerical precision, accuracy, and reproducibility, performance
-and documentation enhancements, and bug fixes. We discuss some of these in the subsections below.
+improvements to precision and reproducibility, performance and documentation
+enhancements, and bug fixes. We discuss some of these in the subsections below.
 
@@
 In addition, \astropyTable\ may also read from and write
-to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},\footnote{Starting in v5.1,
-direct \astropypkg support for ASDF has been deprecated in favor of moving it
-to the new \package{asdf-astropy} \citep{asdf-astropy} package to accommodate
-the differing release schedules between ASDF and \astropypkg.}
+to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
 \href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
@@
 
-\subsection{Numerical Precision, Accuracy, \& Reproducibility}
+\subsection{Precision \& Reproducibility}
 \label{sec:core-features-precision_accuracy_repreducibility}
@@
 
-% \begin{itemize}
-% \item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
-This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
-The Moore Foundation grant allows this Slack space to
-% be on a paid plan; additionally NumFOCUS has negotiated a special rate for
-% open-source projects.
-While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
-The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
-% a Discourse site which is more open than the Facebook group and more user-oriented
-% than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
-% developers are able to invoice as independent contractors the time they
-% spend helping users on these forums.
-% \end{itemize}
-
 \section{Future Plans for the Astropy Project} \label{sec:future}
@@
 
-- Astropy is established; challenges faced have evolved
--
+Community Engagement
 
 \begin{itemize}
-\item Attracting new contributors when the code has become quite complex, project established (astropy as an "institution", walled-off group). (how to attract? Fast feedback and code reviews, ...)
-\item Support for existing contributors: Contributor and maintainer mentoring (...)
-\item Infrastructure: Long-term / sustained funding for maintaining infrastructure (grants last 3 years; infrastructure evolves, nature of open source ecosystem)
-\item As ecosystem grows, making sure users know about all of the functionality (workshops, tutorials, educational materials, discoverability, accessibility) (Continued engagement with user community)
+\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+open-source projects.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easy-to-navigate interface for users to ask for help.
+Accordingly, we have commissioned
+a Discourse site which is more open than the Facebook group and more user-oriented
+than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+developers are able to invoice as independent contractors the time they
+spend helping users on these forums.
+\end{itemize}
+
+\begin{itemize}
+\item Attracting new contributors when the code has become quite complex,
+\item Contributor to maintainer mentoring,
+\item Long-term / sustained funding for maintaining infrastructure,
+\item ...
 \end{itemize}

### 2022-03-19 src/ms.tex
Address #53 on paragraph apy on units
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
 \begin{abstract}
-The \astropy Project supports and fosters the development of open-source and openly developed
-\python packages that provide commonly needed functionality to the astronomical
+The \astropy Project supports and fosters the development of open-source and openly-developed
+\python packages that provide commonly-needed functionality to the astronomical
 community.
@@
 release, version 5.0, and provide major updates for the project.
-We then discuss supporting a broader ecosystem of interoperable packages,
+We then discuss supporting a broader ecosystem of inter-operable packages,
 including connections with several astronomical observatories and missions.
@@
 Over the last decade, \python has grown enormously in popularity to become a
-dominant programming language, especially in the astronomical and broader scientific
+dominant programming language also in the astronomical and broader scientific
 communities.
 For example, Figure~\ref{fig:python-mentions} shows the number of yearly
-full-text mentions of \python as compared to a selection of other programming languages
+full-text mentions of \python as compared to a few other programming languages
 in refereed articles in the astronomical literature, demonstrating its nearly
@@
 Today, a broad and feature-diverse ``ecosystem'' of packages exists in the
-\python scientific computing landscape: roughly ordered from general-use to
+\python scientific computing landscape: Roughly ordered from general-use to
 domain-specific, this landscape now includes packages that provide core
@@
 The \astropypkg \citep{astropy:2013, astropy:2018} core package began in this
-vein, as an effort to consolidate the development of commonly used functionality
+vein, as an effort to consolidate the development of commonly-used functionality
 needed to perform astronomical research into a community-developed \python
@@
 This is largely driven by the fact that the core package now represents just one
-piece of the broader astronomy \python context, and new feature development
+piece of the broader astronomy \python context, and much new feature development
 is now happening in more specialized packages that are expanding the
@@
 \astropypkg core package, the ecosystem of astronomy-specific software tools
-that are interoperable with \astropypkg (Astropy Coordinated and Affiliated Packages),
+that are interoperable with \astropypkg (Astropy Affiliated Packages),
 \emph{and} the community of users, developers, and maintainers that participate
 in Astropy efforts.
-However, there is no single institution responsible for managing the Astropy Project,
+However, there is no institution responsible for managing the Astropy Project,
 for funding or maintaining its development, or sustaining it into the future:
-The Project is maintained and coordinated largely by volunteers and contributors
-who are primarily paid for other reasons (i.e., Astropy work is just a means to an end).
+The Project is maintained and coordinated largely by volunteers.
 While new Astropy-affiliated packages are being developed that expand upon the
@@
 \citep{ape2}. The motivation for LTS releases is to provide longer-term stable
-versions of \astropypkg that users who require a high level of stability can
-make use of if they do not always need the latest features (this includes but
-not limited to telescope operations and data reduction pipelines). The v5.0
-release of the core package, replacing v4.0, was designated as LTS at the end of 2021;
-it will be maintained until the end of 2023.
+versions of \astropypkg that users that require a high level of stability can
+make use of if they do not always need the latest features (this could include,
+for example, telescope operation or data reduction pipelines, etc.). The v5.0
+release of the core package was designated as LTS, and since it was released at
+the end of 2021, it will be maintained until the end of 2023.
 
@@
 \python libraries have come to rely on \astropypkg. The relationship between
-\astropy and the broader \python community is reciprocal, with the evolving needs
+\astropy and the \python-using community is reciprocal, with the evolving needs
 of the community and the maturation of the broader scientific \python ecosystem
@@
 
-Within \astropypkg, development may be roughly split into several
-categories: new features, intra- and inter-package interoperability,
-improvements to precision and reproducibility, performance and documentation
-enhancements, and bug fixes. We discuss some of these in the subsections below.
+Within \astropypkg, development may be roughly split into a few
+categories: new features added, intra- and inter-package interoperability,
+improvements to precision, accuracy, and reproducibility, and performance
+enhancements. We discuss these in the subsections below.
 
@@
 Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
-carried, as well as the associated \python wrappers. Since this is basic
+carried, as well as the \python wrappers for it. Since this is basic
 infrastructure, with a release cycle set by SOFA, it made more sense to create
@@
 was too specialized was a Virtual Observatory sub-package; the Simple
-Application Message Protocol (\astropysubpkg{samp}) is kept in the core package
-for now, but other parts more properly belonged in
+Application Message Protocol was kept in the core package
+(\astropysubpkg{samp}), but other parts more properly belonged in
 \package{astroquery} (see Section~\ref{sec:astroquery}).
@@
 New features present a particular difficulty for a mature package such as
-\astropypkg, because it is challenging to know {\em a priori} what the best
+\astropypkg, because it is difficult to know {\em a priori} what the best
 interface will be, and hard to address all possible use cases in one round of
@@
 is spanned by five main new features that entered the \astropypkg core since the
-previous summary \citepalias{astropy:2018}. In our descriptions of each, we
+previous summary \citepalias{astropy:2018}. In our descriptions of each we
 include how we hope the new feature will evolve. Ultimately, in a user-driven
@@
 
-\paragraph{Uncertainties and Distributions} In scientific analysis,
-measurement values with units and uncertainties are essential.
+\paragraph{Uncertainties and Distributions} Early in our careers we were likely
+all taught that measurements without units and uncertainties are meaningless.
 \astropyunits provides the ability to associate numbers with units, and
@@
 \astropysubpkg{nddata}, there are options to associate errors with data arrays,
-but covariances are not tracked.
+but these are treated as independent, i.e., covariances are not tracked.
 
@@
 
-A plan for future development of \astropysubpkg{uncertainties} is to
-support error propagation: tracking covariances for the case that uncertainties
+A plan for future development of \astropysubpkg{uncertainties} is to also
+support error propagation, tracking covariances, for the case that uncertainties
 are normally distributed rather than approximated by Monte-Carlo means. One
@@
 points implies that all data points are now covariant with each other, i.e., one
-has to carry $N\times N$ covariances. For a large image, that becomes unnecessary
-and computationally expensive.
-
-\paragraph{Masked quantities}  It can be useful to mask bad data, either by marking
-bad data values with a flag or by replacing them with a special
-value, e.g., ``Not a Number'' (NaN). In \astropypkg, both approaches are
-used: a flag for masked columns in \astropyTable, mask flags and bitmaps
+has to carry $N\times N$ covariances. For a large image, that becomes not just
+pointless but also prohibitive in terms of memory and CPU usage.
+
+\paragraph{Masked quantities}  It can be useful to mask bad data. How one
+supports masked data involves a number of choices, such as whether one indicates
+data values are bad by a separate flag or by replacing bad data with a special
+value, such as ``Not a Number'' (NaN). In \astropypkg, both approaches have
+been used: a flag for masked columns in \astropyTable, mask flags and bitmaps
 for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
-NaN in \astropyTime. However masked \astropyQuantity support is currently very limited,
-hindering usage of masks in a \astropyQTable\ (in which \astropyQuantity\ is used
+NaN in \astropyTime. However masked quantities were poorly supported, making it
+difficult to use masks in a \astropyQTable\ (in which \astropyQuantity\ is used
 for all columns with units).
 
-Furthermore, the \texttt{MaskedArray} class from \package{numpy} only works
+Unfortunately, the \texttt{MaskedArray} class from \package{numpy} only works
 well with plain data arrays, hiding other metadata and attributes, for example
@@
 The masked quantities work largely without any changes in higher-level objects
-such as \astropySkyCoord, but more work is needed to better integrate it with
-the rest of \astropypkg, e.g., using masked arrays in \astropyTime\ instead of NaN.
+such as \astropySkyCoord, but work is still underway to expose the mask in those
+classes, as well as to use masked arrays in \astropyTime\ instead of replacing
+data with NaN.
 
@@
 particular time bins. The new classes offer a number of special methods to
-manipulate time series (e.g., folding and resampling) and to read different data
-formats (e.g., \mission{Kepler} light curves).
+manipulate time series (folding, resampling, etc.) and to read different data
+formats (such as \mission{Kepler} light curves, etc.).
 
@@
 correspondence between ``real-world'' and pixel coordinates. This mapping is the
-essence of the World Coordinate System (WCS) concept. From its inception,
+essence of the ``World Coordinate System (WCS)'' concept. From its inception,
 the \astropysubpkg{wcs} sub-package allowed access to WCS information provided
@@
 well with other \astropypkg\ objects such as \astropySkyCoord\ and \astropyTime.
-Future work includes better integration of this interface with the rest of
-the Project.
 
@@
 both within the various \astropypkg sub-packages and with other \python
-libraries. We aim for a seamless user experience, where different code blocks
-would ``just work'' when put together. Here, we describe a few particular
+libraries. If done right, all a person notices is that code using many different
+python features and libraries ``just works.'' Here, we describe a few particular
 efforts towards this goal.
@@
 would strip a \astropyQuantity\ of its units (or fail outright), limiting
-\astropyQuantity's potential. Now, advances in \package{numpy} function
+\astropyQuantity's usefulness. Now, advances in \package{numpy} function
 overloading (e.g.,
@@
 Units were also integrated further within \astropypkg. For an already-defined
-unit-less \astropyModel, e.g., those imported from another library,
+unit-less \astropyModel, e.g. those imported from another library,
 \astropymodeling can now coerce units. A new module – \astropycosmologyunits –
@@
 factors of cosmological redshift. \texttt{redshift} has a number of
-equivalencies for converting between cosmological distance measures, e.g., CMB
+equivalencies for converting between cosmological distance measures, e.g. CMB
 temperature or comoving distance.
@@
 
-Within \astropypkg and \python, there are numerous ways to represent and store
+Within \astropypkg and \python there are numerous ways to represent and store
 array-valued data. Some of the differences are historical:
@@
 different use cases. Lastly, some differences are inherent: \astropyQuantity,
-\astropyTime, and \astropySkyCoord\ represent fundamentally different types of
+\astropyTime\, and \astropySkyCoord\ represent fundamentally different types of
 objects.
@@
 attributes. With this protocol, it is now possible to store \astropy native
-objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord\ --
+objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord --
 within a \astropyTable\ and write these to various file formats, such as FITS.
@@
 The FITS standard was extended to rigorously describe time coordinates in the
-WCS framework \citep{FITS-Time:2015}. Compared to
+World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared to
 other types of coordinates in WCS, time requires more metadata: format, scale,
@@
 \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
-mixin protocol for seamless interoperability}.} the \astropyfits package now
-interprets time data correctly, using \astropyTime\ as a mixin column. For
-backward compatibility, this feature may be turned off.
+mixin protocol for seamless interoperability}.}, the \astropyfits package now
+interprets time data correctly, using \astropyTime\ as a `mixin' column. For
+backward compatibility with manual systems, this feature may be turned off.
 
@@
 \astropy now supports reading and writing tables in the American Astronomical
-Society Journals' Machine-Readable Table (\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}) format. This ASCII format has
-long been missing and was added to \astropyTable\ I/O in a Google Summer of Code 2021 project.
-In addition, \astropyTable\ may also read from and write
+Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
+long been missing and as added in this year's Google Summer of Code project adds
+\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}
+to \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
 to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
@@
 The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
-standard} has been updated to v1.0, adding support of three additional data
-subtypes: multidimensional column data (both masked and unmasked)
-with fixed dimensions in all table cells, multidimensional
+standard} has been updated to version 1.0, adding support three additional data
+subtypes. First, it adds multidimensional column data (both masked and unmasked)
+with fixed dimensions in all table cells. Second, it adds multidimensional
 column data with variable-dimension arrays similar to FITS variable-length
-arrays, and object-type columns with simple \python objects.
+arrays. Third, it adds object-type columns with simple \python objects.
 
 As mentioned previously, \package{dask} arrays may be used as a mixin column in
-\astropyTable. Now, \package{dask} may also be a data array in
-\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only
-computed while written, avoiding excessive memory use. Furthermore, \astropyTable\ can
-be appended to an existing FITS file, where \package{dask} mixin columns can
-interoperate seamlessly between them.
+\astropyTable. Now, \package{dask} may also be the data array in
+\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the array
+will be computed while written, avoiding excessive memory use. \astropyTable\ can
+now be appended to an existing FITS file, and \package{dask} mixin columns
+interoperate seamlessly between the two table types.
 
@@
 functions for \astropypkg objects, such as reading from or writing to a file,
-following a given standard. \astropypkg uses this internally for the I/O methods in
-\astropyTable\ and \astropyCosmology. Users can also register custom I/O to
+following a given standard. \astropy uses this internally for the I/O methods in
+\astropyTable\ and \astropyCosmology, and users can register custom I/O to
 extend the options on these classes.
 
-As of \astropypkg v5.0, the I/O registry submodule has been generalized to enable a
-number of new use cases; for instance, the class-based architecture allows
-streamlining of the creation of custom registries while minimizing code duplication.
-One application is in \astropycosmology,
+As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
+number of new use cases. for instance, the class-based architecture allows for
+the creation of custom registries. One application is in \astropycosmology,
 which has two different \textit{kinds} of registries for \astropyCosmology: one
-for reading and writing files, and another for converting between \python objects.
-
-\subsection{Precision \& Reproducibility}
+for reading and writing files, and another for converting between python
+objects. Using \astropyiosubpkg{registry} for custom classes means a developer
+does not have to create a whole new I/O system and can offer to the user a
+unified and familiar API.
+
+\subsection{Precision, Accuracy, \& Reproducibility}
 \label{sec:core-features-precision_accuracy_repreducibility}
@@
 In astronomy, time accuracy down to the (nano)second is frequently important.
-Whether for planning observations or crunching pulsar or VLBI data, missing
+Whether for planning observations, crunching pulsar or VLBI data, missing
 seconds meaningfully impact results. In past versions of \astropypkg, when using
-\astropyTime\, leap seconds had to be manually applied; now, leap
+\astropyTime\ in \astropytime leap seconds had to be manually applied; now, leap
 seconds are applied automatically. Moreover, \astropypkg updates internal time
-data files to ensure that the correct leap second adjustment is always used. For
-reproducibility with older code, \astropyLeapSeconds\ may also be
+data-files to ensure that the correct leap second adjustment is always used. For
+reproducibility, e.g., with an old code, \astropyLeapSeconds\ may also be
 manually applied.
@@
 \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
-2018}}. \astropyconstants now defaults to using the \texttt{SI/CODATA 2018}
+2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
 values, with the units in \astropyunits based on these constants. Most of
 \astropypkg and affiliate packages build upon \astropyunits, so this update
-affects the entire \astropy ecosystem.
+effects the entire \astropy ecosystem.
 
@@
 \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
-and associated transformations have been updated to correctly reflect the ``true''
-and ``mean" terminology.
+and associated transformations have been updated to correctly reflect the “true”
+and “mean” terminology.
 
@@
 Performance in \astropypkg is addressed based on feedback and requests, and
-through contributions from community members: if you find that parts of
+through contributions from community members: If you find that parts of
 \astropypkg are not performant, we encourage you to open a \github issue to
@@
 Astropy Project as the ultimate sources of authority, as well as the core
-principle of ``do-ocracy'': those who do work for the Project (be it coding,
+principle of ``do-ocracy'' that those who do work for the Project (be it coding,
 training, or other harder-to-quantify contributions) gain more influence on the
@@
 when other mechanisms have failed, and includes powers that either
-require central authority or secrets (e.g., passwords). However, APE0 also
+require central authority or secrets (e.g., passwords) However, APE0 also
 charges the Committee to devolve responsibilities and seek community input on
@@
             \textbf{Top panel:} The number of Git commits per month (defined as
-            consecutive 30 day windows) merged into the \astropypkg core
+            consecutive 30 day windows) merged into to the \astropypkg core
             package over the whole history of the project (up to end of 2021,
@@
 own software package infrastructure. These include tools and extensions that
-enable generating documentation and setting up automated testing, as well
+enable easily generating documentation and setting up automated testing, as well
 as providing package scaffolding for new projects.
@@
 Python packages leveraging the Astropy ecosystem. This tool utilizes the
-\href{https://github.com/cookiecutter/cookiecutter}{Cookiecutter} project to
-walk users through the process of creating new packages complete with
-documentation and testing support. Additionally, the package template
-generation process includes the ability to setup interoperability with
-\github, allowing for easy repository access from documentation, as well as an
-example \github Actions workflow to demonstrate the use of \github's
-continuous integration tooling.
+Cookiecutter project to walk users through the process of creating new
+packages complete with documentation and testing support. Additionally, the
+package template generation process includes the ability to setup
+interoperability with \github, allowing for easy repository access from
+documentation, as well as an example \github Actions workflow to demonstrate
+the use of \github's continuous integration tooling.
 
@@
 services, including the James Webb Space Telescope (JWST) archive, and the
-project has switched to a continuous release model: every time a change is
+project has switched to a continuous release model: Every time a change is
 committed to the main development branch it is published on the Python Package
 Index (PyPI) and available for installation. (However, formal releases are still
-done a few times per year.) \texttt{photutils} \cite{photutils} released its
+done a few times per year.) \texttt{photutils} \cite {photutils} released its
 first stable version, indicating that the API will change less frequently, and
@@
 several special-purpose \python packages, such as \texttt{LightKurve}
-\citep{LightKurve} and SORA \citep{SORA}. More than 4,000 repositories on
+\citep{LightKurve} and SORA \citep{SORA}]. More than 4,000 repositories on
 \github make use of \texttt{astroquery} in some way, and \texttt{astroquery}
@@
 A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
-a python package developed to perform all the necessary tasks to perform data processing, from the raw data
+a python package developed to perform all the necessary tasks to from the raw data
 of Cherenkov telescopes to the reconstructed event lists.
@@
 Both Gammapy and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
-tables, and FITS I/O.
+tables and FITS I/O.
 As CTA will record gamma-ray events with a rate of up to $10\,000$ events per second,
@@
 % \secauthor{James Tocknell}
-Data Central is one of the five Australian Virtual Observatory\footnote{Known as
-  the All-Sky Virtual Observatory, or ASVO.} nodes, providing both Australian and
+Data Central is one of the five Australian Virtual Observatory\footnote{known as
+  the All-Sky Virtual Observatory, or ASVO} nodes, providing both Australian and
 international astronomers access to and services for surveys of the southern
@@
   Data Central, Astronomy Australia Limited (AAL), and National Computational
-  Infrastructure (NCI).}, Data Central also assists theoretical and radio
-astronomers with hosting and managing their data.
+  Infrastructure (NCI)}, Data Central also assists theoretical and radio
+astronomers host and manage their data.
 
@@
 play a vital role in the initial ingestion, reduction, access and visualization
-of the hosted data and services. For example, the AAT Archive hosts data from as early as
+of the hosted data and services. For example, the AAT Archive hosts data back to
 1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest
@@
 2dfdr-reduced spectra from the AAT Archive, and strongly encourages other
-archives to provide loaders for their data to further the reuse of existing archival
+archives to provide loaders for their data to further reuse of existing archival
 data.
@@
 appearance of the \href{https://learn.astropy.org}{Learn Astropy website}.
-The Learn Team meets weekly to work on creating, expanding, and improving these
+The Learn Team meets weekly to work on creating, expanding, improving these
 educational resources.
@@
 Now that the backend infrastructure is stable, we are interested in collecting
-new content to serve and share: if you have ideas for contributions, or have
+new content to serve and share: If you have ideas for contributions, or have
 tutorials hosted on third-party sites that you would like to make searchable
@@
 expanded to Summer AAS meetings.
-The Learn team finds that the workshop audience is best found at AAS meetings as
+The Learn team finds that the workshop audience is best found as AAS meetings as
 opposed to more general \python meetings, as the content tends to be more
@@
 underway, have sufficient resources/support and a plan in place for completion;
-for example, efforts to improve the discoverability of documentation and
+For example, efforts to improve the discoverability of documentation and
 educational materials by overhauling the Learn website, which is underway (see
@@
 Orange items are well defined, and work for acquiring sufficient resources
-underway; for example, the goal of providing next-generation spectroscopic
+underway; For example, the goal of providing next-generation spectroscopic
 reduction, analysis, and visualization tools usable by individual researchers
 and larger surveys.
-Red items do not yet have a plan for implementation and need more resources; for
+Red items do not yet have a plan for implementation and need more resources; For
 example, the goal of improving interoperability with performant I/O file formats
@@
 While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
-The Astropy Project identified a need for a public, archived, searchable, and easy-to-navigate interface for users to ask for help.
-Accordingly, we have commissioned
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
 a Discourse site which is more open than the Facebook group and more user-oriented
@@
 The Astropy Project is also expanding its educational and community-focused
-initiatives by soliciting and funding initiatives in inclusion, diversity, and
-equity (Section~\ref{sec:project-ide}) as well as through the Learn Astropy initiative,
+initiatives, by soliciting and funding initiatives in inclusion, diversity, and
+equity (Section~\ref{sec:project-ide}) and through the Learn Astropy initiative,
 which aims to make tutorials and educational content that highlight software
@@
 We would like to thank the members of the community who have contributed to
-\astropy, opened issues and provided feedback, and supported the
+\astropy, who have opened issues and provided feedback, and have supported the
 project in a number of different ways.
@@
 summer. These students often turn into long-term contributors.
-Within the academic community, we thank institutions that make it possible for
-astronomers and other developers on their staff to contribute their time to the
+Within the academic community, we thank institutions that make it possible that
+astronomers and other developers on their staff can contribute their time to the
 development of \astropy projects.

### 2022-03-19 src/ms.tex
Typo and format fixes

@@
 \citep{ape2}. The motivation for LTS releases is to provide longer-term stable
-versions of \astropypkg that users that require a high level of stability can
+versions of \astropypkg that users who require a high level of stability can
 make use of if they do not always need the latest features (this could include,
@@
 Units were also integrated further within \astropypkg. For an already-defined
-unit-less \astropyModel, e.g. those imported from another library,
+unit-less \astropyModel, e.g., those imported from another library,
 \astropymodeling can now coerce units. A new module – \astropycosmologyunits –
@@
 factors of cosmological redshift. \texttt{redshift} has a number of
-equivalencies for converting between cosmological distance measures, e.g. CMB
+equivalencies for converting between cosmological distance measures, e.g., CMB
 temperature or comoving distance.
@@
 Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
-long been missing and as added in this year's Google Summer of Code project adds
+long been missing and as part of this year's Google Summer of Code project
 \href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}
-to \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
+has been added to the \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
 to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
@@
 The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
-standard} has been updated to version 1.0, adding support three additional data
+standard} has been updated to version 1.0, adding support for three additional data
 subtypes. First, it adds multidimensional column data (both masked and unmasked)
@@
 As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
-number of new use cases. for instance, the class-based architecture allows for
+number of new use cases. For instance, the class-based architecture allows for
 the creation of custom registries. One application is in \astropycosmology,

### 2022-03-22 src/ms.tex
Formatting and verbiage improvements

--- 2022-03-19 src/ms.tex
@@
 Python packages leveraging the Astropy ecosystem. This tool utilizes the
-Cookiecutter project to walk users through the process of creating new
-packages complete with documentation and testing support. Additionally, the
-package template generation process includes the ability to setup
-interoperability with \github, allowing for easy repository access from
-documentation, as well as an example \github Actions workflow to demonstrate
-the use of \github's continuous integration tooling.
+\href{https://github.com/cookiecutter/cookiecutter}{Cookiecutter} project to 
+walk users through the process of creating new packages complete with 
+documentation and testing support. Additionally, the package template 
+generation process includes the ability to setup interoperability with 
+\github, allowing for easy repository access from documentation, as well as an 
+example \github Actions workflow to demonstrate the use of \github's 
+continuous integration tooling.
 
@@
 play a vital role in the initial ingestion, reduction, access and visualization
-of the hosted data and services. For example, the AAT Archive hosts data back to
+of the hosted data and services. For example, the AAT Archive hosts date back to
 1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest
@@
 While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
-The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+The Astropy Project identified a need for a public, archived, searchable, and easy-to-navigate interface for users to ask for help.
+Accordingly, we have commissioned
 a Discourse site which is more open than the Facebook group and more user-oriented

### 2022-03-22 src/ms.tex

@@
 play a vital role in the initial ingestion, reduction, access and visualization
-of the hosted data and services. For example, the AAT Archive hosts date back to
+of the hosted data and services. For example, the AAT Archive hosts data from as early as
 1974, so \astropycoordinates, \astropytime and \astropyfits are used to ingest

### 2022-03-22 src/ms.tex
Add back accuracy

@@
 versions of \astropypkg that users who require a high level of stability can
-make use of if they do not always need the latest features (this could include,
-for example, telescope operation or data reduction pipelines, etc.). The v5.0
-release of the core package was designated as LTS, and since it was released at
-the end of 2021, it will be maintained until the end of 2023.
+make use of if they do not always need the latest features (this includes but
+not limited to telescope operations and data reduction pipelines). The v5.0
+release of the core package, replacing v4.0, was designated as LTS at the end of 2021;
+it will be maintained until the end of 2023.
 
@@
 \astropy now supports reading and writing tables in the American Astronomical
-Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
-long been missing and as part of this year's Google Summer of Code project
-\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}
-has been added to the \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
+Society Journals' Machine-Readable Table (\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}) format. This ASCII format has
+long been missing and was added to \astropyTable\ I/O in a Google Summer of Code 2021 project.
+In addition, \astropyTable\ may also read from and write
 to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
@@
 The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
-standard} has been updated to version 1.0, adding support for three additional data
-subtypes. First, it adds multidimensional column data (both masked and unmasked)
-with fixed dimensions in all table cells. Second, it adds multidimensional
+standard} has been updated to v1.0, adding support of three additional data
+subtypes: multidimensional column data (both masked and unmasked)
+with fixed dimensions in all table cells, multidimensional
 column data with variable-dimension arrays similar to FITS variable-length
-arrays. Third, it adds object-type columns with simple \python objects.
+arrays, and object-type columns with simple \python objects.
 
 As mentioned previously, \package{dask} arrays may be used as a mixin column in
-\astropyTable. Now, \package{dask} may also be the data array in
-\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the array
-will be computed while written, avoiding excessive memory use. \astropyTable\ can
-now be appended to an existing FITS file, and \package{dask} mixin columns
-interoperate seamlessly between the two table types.
+\astropyTable. Now, \package{dask} may also be a data array in
+\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only
+computed while written, avoiding excessive memory use. Furthermore, \astropyTable\ can
+be appended to an existing FITS file, where \package{dask} mixin columns can
+interoperate seamlessly between them.
 
@@
 functions for \astropypkg objects, such as reading from or writing to a file,
-following a given standard. \astropy uses this internally for the I/O methods in
-\astropyTable\ and \astropyCosmology, and users can register custom I/O to
+following a given standard. \astropypkg uses this internally for the I/O methods in
+\astropyTable\ and \astropyCosmology. Users can also register custom I/O to
 extend the options on these classes.
 
-As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
-number of new use cases. For instance, the class-based architecture allows for
-the creation of custom registries. One application is in \astropycosmology,
+As of \astropypkg v5.0, the I/O registry submodule has been generalized to enable a
+number of new use cases; for instance, the class-based architecture allows
+streamlining of the creation of custom registries while minimizing code duplication.
+One application is in \astropycosmology,
 which has two different \textit{kinds} of registries for \astropyCosmology: one
-for reading and writing files, and another for converting between python
-objects. Using \astropyiosubpkg{registry} for custom classes means a developer
-does not have to create a whole new I/O system and can offer to the user a
-unified and familiar API.
-
-\subsection{Precision, Accuracy, \& Reproducibility}
+for reading and writing files, and another for converting between \python objects.
+
+\subsection{Numerical Precision, Accuracy, \& Reproducibility}
 \label{sec:core-features-precision_accuracy_repreducibility}

### 2022-03-23 src/ms.tex
Clarify future asdf support
Co-authored-by: William Jamieson <wjamieson@stsci.edu>

@@
 In addition, \astropyTable\ may also read from and write
-to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
+to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},\footnote{Starting in v5.1,
+direct \astropypkg support for ASDF has been deprecated in favor of moving it
+to the new \package{asdf-astropy} \citep{asdf-astropy} package to accommodate
+the differing release schedules between ASDF and \astropypkg.}
 \href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},

### 2022-03-23 src/ms.tex
Lim edits for Sections 2.3 and 2.4

--- 2022-03-22 src/ms.tex
@@
 
-An important focus of \astropy has been to increase the precision and accuracy
+An important focus of \astropypkg has been to increase the numerical precision and accuracy
 of its functionality, while, where appropriate, making sure results obtained
-using older versions of \astropypkg are reproducible. Within each section below
+using older versions of \astropypkg are reproducible. Within each section below,
 we include in our descriptions how old results may be reproduced. The primary
@@
 seconds meaningfully impact results. In past versions of \astropypkg, when using
-\astropyTime\, leap seconds had to be manually applied; now, leap
+\astropyTime, leap seconds had to be manually applied; now, leap
 seconds are applied automatically. Moreover, \astropypkg updates internal time
@@
 values, with the units in \astropyunits based on these constants. Most of
-\astropypkg and affiliate packages build upon \astropyunits, so this update
+\astropy and affiliated packages build upon \astropyunits, so this update
 affects the entire \astropy ecosystem.
@@
 For reproducibility, \astropypkg allows the constants' (and therefore units')
-definitions to be rolled back to prior values, e.g., to the \texttt{SI/CODATA
-2014} values. For work sensitive to the values of the fundamental constants we
-recommend including an \astropy configuration file with the work, which
-specifies the set of constants used.
+definitions to be rolled back to prior values, e.g., to \texttt{SI/CODATA
+2014}. For work sensitive to the values of the fundamental constants, we
+recommend including an \astropypkg configuration file (i.e., \texttt{astropy.cfg})
+with the work, which specifies the set of constants used.
 
@@
 
-As features are added and corner cases dealt with, performance often suffers.
-Hence, pull requests that improve performance are welcomed, and were even the
-main goal for one release cycle. For most sub-packages, performance was
+In some cases, feature enhancements and corner-case handling were added
+at the cost of performance.
+Hence, \github pull requests that improve performance are welcomed, and were even the
+main goal for v3.1 release. For most sub-packages, performance was
 improved in the \python\ code, but for some, the most time-critical pieces were
@@
 time strings to binary). Furthermore, a particular effort was made to ensure
-\astropypkg\ is thread-safe, so that it can be used on supercomputer clusters.
+\astropypkg\ is thread-safe, so that it can be used on supercomputing clusters.
 

### 2022-03-24 src/ms.tex
challenges outline

--- 2022-03-23 src/ms.tex
@@
 
+% \begin{itemize}
+% \item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
+This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
+The Moore Foundation grant allows this Slack space to
+% be on a paid plan; additionally NumFOCUS has negotiated a special rate for
+% open-source projects.
+While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
+The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
+% a Discourse site which is more open than the Facebook group and more user-oriented
+% than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
+% developers are able to invoice as independent contractors the time they
+% spend helping users on these forums.
+% \end{itemize}
+
 \section{Future Plans for the Astropy Project} \label{sec:future}
 
-\subsection{A roadmap for future priorities}
+\subsection{A Roadmap for Future Priorities}
 % \secauthor{Clara Brasseur}
@@
 
-Community Engagement
-
-\begin{itemize}
-\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community.
-This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace.
-The Moore Foundation grant allows this Slack space to
-be on a paid plan; additionally NumFOCUS has negotiated a special rate for
-open-source projects.
-While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source.
-The Astropy Project identified a need for a public, archived, searchable, and easy-to-navigate interface for users to ask for help.
-Accordingly, we have commissioned
-a Discourse site which is more open than the Facebook group and more user-oriented
-than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
-developers are able to invoice as independent contractors the time they
-spend helping users on these forums.
-\end{itemize}
-
-\begin{itemize}
-\item Attracting new contributors when the code has become quite complex,
-\item Contributor to maintainer mentoring,
-\item Long-term / sustained funding for maintaining infrastructure,
-\item ...
-\end{itemize}
+Over the last decade, the needs of (and challenges faced by) the \astropy
+Project have evolved, but many common themes persist.
+For example, ...list...
+In many cases, these themes are not unique to \astropy: Many Open-Source
+Software communities are facing and addressing similar issues.
+Below we discuss a few of these challenges and ...
+
+\begin{description}
+  \item[Supporting an active group of maintainers] With funding, can pay maintainers.
+But: time commitment, some positions don't allow.
+Need for new career paths on the spectrum from software engineer to astronomy researcher.
+Also need to attract new non-hired contributors to sustain the project, but challenging when the code has become quite complex (astropy as an "institution", don't want to be a walled-off group).
+(Need for fast feedback and code reviews to empower contributors, but puts a burden on maintainers)
+  \item[Long-term / sustained funding for maintaining infrastructure] Further from "science," still critical to astropy, but who will provide updates and maintenance?
+More than just updating versions and releasing.
+(grants last 3 years; infrastructure evolves, nature of open source ecosystem)
+  \item[Understanding, communicating with, and engaging our user communities] Different overlapping and intersecting communities of users: telescopes, archives, researchers, students, educators, general public.
+Communities have different needs from Astropy.
+Also, as ecosystem grows, making sure users know about all of the functionality (workshops, tutorials, educational materials, discoverability, accessibility) (Continued engagement with user community).
+\end{description}
 

### 2022-04-02 src/ms.tex
Update src/ms.tex
Co-authored-by: P. L. Lim <2090236+pllim@users.noreply.github.com>

--- 2022-03-24 src/ms.tex
@@
 (Need for fast feedback and code reviews to empower contributors, but puts a burden on maintainers)
-  \item[Long-term / sustained funding for maintaining infrastructure] Further from "science," still critical to astropy, but who will provide updates and maintenance?
-More than just updating versions and releasing.
-(grants last 3 years; infrastructure evolves, nature of open source ecosystem)
+  \item[Diversity of contributor, maintainer, coordinator communities] Practical issues: translation of content and documentation, github is not accessible in some countries.
+  \item[Long-term / sustained funding for maintaining infrastructure] Software and packaging infrastructure
+  includes but not limited to Continuous Integration (CI), test plugins, documentation builds, and package
+  delivery. Keeping the infrastructure working is essential in software development and operations.
+  This kind of work has low visibility and is not attractive to scientists or volunteer contributors;
+  therefore, it is most effective to delegate such tasks to paid roles. \astropy secured a three-year grant
+  for this but the situation remains uncertain once the grant runs out, as infrastructure evolves and
+  requires constant maintenance and upgrades. Some of the engineers currently in these roles are
+  also employees of research institutions that may leave the Project without much notice, adding to
+  the uncertainty factor.
   \item[Understanding, communicating with, and engaging our user communities] Different overlapping and intersecting communities of users: telescopes, archives, researchers, students, educators, general public.

### 2022-04-02 src/ms.tex
supporting maintainers

--- 2022-03-24 src/ms.tex
@@
 
-Over the last decade, the needs of (and challenges faced by) the \astropy
-Project have evolved, but many common themes persist.
-For example, ...list...
-In many cases, these themes are not unique to \astropy: Many Open-Source
-Software communities are facing and addressing similar issues.
-Below we discuss a few of these challenges and ...
-
+Over the history of \astropy, the needs and challenges of the \astropy Project
+have evolved, but many common themes have persisted and are shared by other
+open source software communities.
+However, \astropy also faces a number of unique challenges stemming from the
+fact that the Project is in a special position in the astronomy software stack:
+The \astropy code and ecosystem are more specialized than core scientific tools
+like \package{numpy} and \package{scipy}, but more general than software
+packages developed to enable specific research subdomains like
+\package{exoplanet} \citep{exoplanet:2021}.
+Below we discuss a few challenges faced by the Project in more detail, but we
+stress that this list is not complete or exhaustive.
 \begin{description}
-  \item[Supporting an active group of maintainers] With funding, can pay maintainers.
-But: time commitment, some positions don't allow.
-Need for new career paths on the spectrum from software engineer to astronomy researcher.
-Also need to attract new non-hired contributors to sustain the project, but challenging when the code has become quite complex (astropy as an "institution", don't want to be a walled-off group).
-(Need for fast feedback and code reviews to empower contributors, but puts a burden on maintainers)
-  \item[Long-term / sustained funding for maintaining infrastructure] Further from "science," still critical to astropy, but who will provide updates and maintenance?
-More than just updating versions and releasing.
-(grants last 3 years; infrastructure evolves, nature of open source ecosystem)
+  \item[Supporting an active group of maintainers] Throughout its history,
+  development and maintenance of code and educational materials in the \astropy
+  Project --- the \astropypkg core package, coordinated packages, Learn Astropy
+  repositories, and Project infrastructure packages --- has been done by a mix
+  of volunteer effort and institutional support.
+  At present, the \href{https://www.astropy.org/team.html}{named roles} and
+  maintainer positions in the \astropy Project are also split roughly evenly
+  between these types of contributors.
+  As is the case with many other open-source software communities, the culture
+  of the Project and overall coordination relies heavily on volunteer effort,
+  but we therefore have significant amounts of critical infrastructure or core
+  functionality that is either maintained with no redundancy or unmaintained as
+  contributors naturally move on from and stop contributing to the Project.
+  Within the core package, a prime example is the \astropyfits subpackage, which
+  is largely in maintenance mode and maintained by just one contributor (despite
+  the significant codebase and user community behind this subpackage).
+  While we now have funding to support the longer-term sustainability of the
+  Project and thus to address these issues, we are finding it difficult to use
+  this funding in ways that address our needs.
+  For one, many of the needs of the Project require long-term and sustained
+  funding rather than short-term contracts.
+  But also the specialized nature of the software requires being able to hire or
+  contract software engineers with domain knowledge about astronomical software
+  tools.
+  Finally, we want to be able to preserve and sustain the engaged community of
+  volunteers, software-engaged researchers, and existing software engineers that
+  contribute to the Project and that have enabled its success in ways that
+  support their careers.
+  These challenges will require cultural and larger systemic changes to
+  astronomical research and career paths, and so will take time to influence and
+  implement.
+  \item[Diversity of users, contributors, maintainers, and coordinators]
+  Astropy is
+  Practical issues: translation of content and documentation, github is not accessible in some countries.
+  \item[Long-term / sustained funding for maintaining infrastructure] Software
+  and packaging infrastructure includes but not limited to Continuous
+  Integration (CI), test plugins, documentation builds, and package delivery.
+  Keeping the infrastructure working is essential in software development and
+  operations. This kind of work has low visibility and is not attractive to
+  scientists or volunteer contributors; therefore, it is most effective to
+  delegate such tasks to paid roles. \astropy secured a three-year grant for
+  this but the situation remains uncertain once the grant runs out, as
+  infrastructure evolves and requires constant maintenance and upgrades. Some of
+  the engineers currently in these roles are also employees of research
+  institutions that may leave the Project without much notice, adding to the
+  uncertainty factor.
   \item[Understanding, communicating with, and engaging our user communities] Different overlapping and intersecting communities of users: telescopes, archives, researchers, students, educators, general public.

### 2022-04-02 src/figures/python-mentions.py
adjust julia

@@ -124,9 +124,11 @@ def plot_yearly_mentions(paper_dfs, total_counts, min_year=1991):
         mentions = g['pubdate.1'].values
 
         if name.lower() == 'julia':
+            # HACK: Correct julia counts for people named "Julia"
             jmask = group_dates > datetime.date(2012, 1, 1)
+            prev_jmask = (group_dates > datetime.date(2000, 1, 1)) & ~jmask
             group_dates = group_dates[jmask]
-            mentions = mentions[jmask]
+            mentions = mentions[jmask] - np.mean(mentions[prev_jmask])
 
         ax.plot(group_dates,
                 mentions,
@@ -151,7 +153,7 @@ def plot_yearly_mentions(paper_dfs, total_counts, min_year=1991):
     ax.set_xlim(datetime.date(min_year, 8, 1),
                 datetime.date(2021, 6, 1))  # 2021.5
 
-    ax.legend(loc='lower right', fontsize=16, ncol=2)
+    ax.legend(loc='lower center', fontsize=15, ncol=2)
 
     ax.set_xlabel('Time [year]')
     ax.set_ylabel('Full-text mentions per year')

### 2022-04-02 src/ms.tex
some words about diversity

@@
   \item[Diversity of users, contributors, maintainers, and coordinators]
-  Astropy is
-  Practical issues: translation of content and documentation, github is not accessible in some countries.
-  \item[Long-term / sustained funding for maintaining infrastructure] Software
-  and packaging infrastructure includes but not limited to Continuous
-  Integration (CI), test plugins, documentation builds, and package delivery.
+  The \astropy project aims to be a project that is made up of members at all
+  levels (users to contributors to coordinators) with a diverse set of skills,
+  backgrounds, and experiences.
+  While \astropy is used globally --- TODO figure with analytics --- our user
+  community is underrepresented in non-English-speaking or -teaching countries
+  and our maintainers and coordinators are fairly narrowly concentrated in the
+  United States (US) and Europe.
+  The gap in users in many countries is understandable given that much of our
+  content and documentation is written in English and therefore only available
+  in other languages through imperfect, automated translation services (like
+  Google translate).
+  The fact that our maintainers and coordinators are generally in the US (even
+  then, mostly on the East coast) and Europe is a reflection of where
+  institutions have explicitly supported \astropy, and where volunteers have had
+  the ability to contribute to the project.
+  \item[Long-term and sustained funding for maintaining infrastructure] The
+  software and packaging infrastructure within the \astropy Project includes
+  (but is not limited to) Continuous Integration (CI), test plugins,
+  documentation builds, and package delivery.
   Keeping the infrastructure working is essential in software development and
-  operations. This kind of work has low visibility and is not attractive to
-  scientists or volunteer contributors; therefore, it is most effective to
-  delegate such tasks to paid roles. \astropy secured a three-year grant for
-  this but the situation remains uncertain once the grant runs out, as
-  infrastructure evolves and requires constant maintenance and upgrades. Some of
-  the engineers currently in these roles are also employees of research
+  operations.
+  This kind of work has low visibility and is not attractive to scientists or
+  volunteer contributors; therefore, it is most effective to delegate such tasks
+  to paid roles.
+  \astropy secured a three-year grant for this but the situation remains
+  uncertain once the grant runs out, as infrastructure evolves and requires
+  constant maintenance and upgrades.
+  Some of the engineers currently in these roles are also employees of research
   institutions that may leave the Project without much notice, adding to the

### 2022-04-02 src/ms.tex
words about user communities

@@
   uncertainty factor.
-  \item[Understanding, communicating with, and engaging our user communities] Different overlapping and intersecting communities of users: telescopes, archives, researchers, students, educators, general public.
-Communities have different needs from Astropy.
-Also, as ecosystem grows, making sure users know about all of the functionality (workshops, tutorials, educational materials, discoverability, accessibility) (Continued engagement with user community).
+  \item[Understanding, communicating with, and engaging our user communities]
+  The \astropy Project is used by a number of intersecting and overlapping
+  communities, such as telescope operations and data management, data archives,
+  researchers, students, educators, and the general public.
+  These communities have different needs from and uses for \astropy, many of
+  which we are only aware of through individual engagement from members of these
+  respective communities.
+  This kind of communication may provide bias pespectives on these communities.
+  We therefore need to develop more formal ways of engaging these communities in
+  more democratic ways, such as through running and analyzing user surveys, or
+  creating and engaging with user groups that can better represent these
+  communities.
+  Additionally, as the \astropy ecosystem has and continues to grow, we need to
+  ensure that users in these different communities have access to educational
+  materials to train new users to use \astropy and \python software tools.
+  This may involve tailoring workshops, tutorials, and educational materials to
+  these communities, however this represents a significant maintenance and
+  content-development burden on the Project.
 \end{description}

### 2022-04-02 src/ms.tex
add new figure and text about new figure

@@
 adopted via pull request in December 2021.
+
+\begin{figure}[th!]
+  \begin{centering}
+    \includegraphics[width=\textwidth]{docu_analytics.pdf}
+      \caption{
+          The number of users who accessed the \astropypkg core package
+          documentation in 2021, based on data from Google Analytics.
+          The circle markers show the approximate locations of home institutions
+          for \astropy voting members (green) and coordination committee members
+          (red), demonstrating the over-representation of the United States and,
+          less so, Europe given the global spread of \astropypkg users.
+      }
+      \label{fig:User-analytics-locations}
+  \end{centering}
+\end{figure}
 
@@
   backgrounds, and experiences.
-  While \astropy is used globally --- TODO figure with analytics --- our user
-  community is underrepresented in non-English-speaking or -teaching countries
-  and our maintainers and coordinators are fairly narrowly concentrated in the
-  United States (US) and Europe.
+  While \astropy is used globally --- for example,
+  Figure~\ref{fig:User-analytics-locations} shows the numnber of users per
+  country who accessed the \astropypkg core package documentation in 2021 ---
+  our user community is underrepresented in non-English-speaking or -teaching
+  countries and our maintainers and coordinators are fairly narrowly
+  concentrated in the United States (US) and Europe (see the circle markers on
+  Figure~\ref{fig:User-analytics-locations}).
   The gap in users in many countries is understandable given that much of our

### 2022-04-02 src/figures/User-analytics-locations.py

@@ -1,0 +1,199 @@
+# -*- coding: utf-8 -*-
+# +
+import copy
+
+import astropy.coordinates as coord
+import astropy.table as at
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+# %matplotlib inline
+import numpy as np
+
+import pandas as pd
+import geopandas as gp
+# -
+
+# Voting members:
+
+# +
+voting_member_locations = [
+    "Boston, MA USA", 
+    "Baltimore, MD USA", 
+    "St. Andrews, Scotland", 
+    "Amherst, MA USA", 
+    "Paris, France", 
+    "Madrid, Spain", 
+    "Baltimore, MD USA", 
+    "Marseille, France", 
+    "Ann Arbor, MI USA", 
+    "Moorhead, MN USA", 
+    "New York, NY USA", 
+    "Baltimore, MD USA", 
+    "Baltimore, MD USA", 
+    "Urbana, IL USA", 
+    "Baltimore, MD USA", 
+    "Hilo, Hawaii USA", 
+    "Gainesville, FL USA", 
+    "New York, NY USA", 
+    "Baltimore, MD USA", 
+    "Boston, MA USA", 
+    "Göttingen, Germany", 
+    "Baltimore, MD USA", 
+    "Sheffield, England", 
+    "Bern, Switzerland", 
+    "Holmfirth, England", 
+    "Toronto, Canada", 
+    "New York, NY USA", 
+    "Holmfirth, England", 
+    "Greenbelt, MD USA", 
+    "Pasadena, CA USA", 
+    "Penticton, Ontario, Canada", 
+    "Greenbelt, MD USA", 
+    "Pasadena, CA USA", 
+    "Baltimore, MD USA", 
+    "Toronto, Canada", 
+    "Potsdam, Germany", 
+    "dwingeloo, netherlands", 
+    "Baltimore, MD USA", 
+    "La Serena, Chile", 
+    "Toronto, Canada", 
+    "Hong Kong"
+]
+
+coco_locations = [
+    "Boston, MA USA", 
+    "Baltimore, MD USA", 
+    "Moorhead, MN USA",
+    "New York, NY USA",
+    "New York, NY USA"
+]
+
+# +
+_loc_cache = {}
+for name in voting_member_locations:
+    if name not in _loc_cache:
+        _loc_cache[name] = coord.EarthLocation.of_address(name)
+        
+for name in coco_locations:
+    if name not in _loc_cache:
[…]

### 2022-04-02 src/figures/User-analytics-locations.py
figure style tweaks

@@ -177,7 +177,7 @@ ax.legend(loc='lower left', fontsize=16)
 
 ax.annotate('16/41 voting members\n4/5 CoCo members', 
             xy=(-72, 39), xytext=(-65, 23), ha='left', 
-            fontsize=12, color='#444444',
+            fontsize=11, color='#444444',
             arrowprops=dict(facecolor='#555555', 
                             width=2, headwidth=8, headlength=10))
 

### 2022-04-02 src/figures/contributor-summary.py

@@ -70,7 +70,7 @@ for name, era in eras.items():
 
 # ### Plot style things:
 
-figsize = (8.5, 6)
+figsize = (9.5, 6)
 style = dict(color='k')
 era_colors = ['#1b9e77', '#d95f02', '#7570b3']
 

### 2022-04-04 src/ms.tex
acknowledgements hack

@@
 
-\begin{acknowledgments}
+% Due to a bug in AAStex 6.3.1, using this with amsmath leads to line numbers in
+% the acknowledgements...
+% https://github.com/AASJournals/AASTeX60/issues/130
+% \begin{acknowledgments}
+\vspace{2em}
 

### 2022-04-04 src/ms.tex
Word flow alterations

@@
 fact that the Project is in a special position in the astronomy software stack:
-The \astropy code and ecosystem are more specialized than core scientific tools
+the \astropy code and ecosystem are more specialized than core scientific tools
 like \package{numpy} and \package{scipy}, but more general than software
-packages developed to enable specific research subdomains like
+packages developed to enable research in specific subdomains such as
 \package{exoplanet} \citep{exoplanet:2021}.
@@
   As is the case with many other open-source software communities, the culture
-  of the Project and overall coordination relies heavily on volunteer effort,
-  but we therefore have significant amounts of critical infrastructure or core
+  of the Project and overall coordination relies heavily on volunteer effort.
+  We therefore have significant amounts of critical infrastructure or core
   functionality that is either maintained with no redundancy or unmaintained as
   contributors naturally move on from and stop contributing to the Project.
-  Within the core package, a prime example is the \astropyfits subpackage, which
+  A prime example within the core package is the \astropyfits subpackage, which
   is largely in maintenance mode and maintained by just one contributor (despite
@@
   funding rather than short-term contracts.
-  But also the specialized nature of the software requires being able to hire or
+  Furthermore, the specialized nature of the software requires being able to hire or
   contract software engineers with domain knowledge about astronomical software
@@
   support their careers.
-  These challenges will require cultural and larger systemic changes to
+  These challenges will require both cultural and larger systemic changes to
   astronomical research and career paths, and so will take time to influence and
@@
   While \astropy is used globally --- for example,
-  Figure~\ref{fig:User-analytics-locations} shows the numnber of users per
+  Figure~\ref{fig:User-analytics-locations} shows the number of users per
   country who accessed the \astropypkg core package documentation in 2021 ---
   our user community is underrepresented in non-English-speaking or -teaching
-  countries and our maintainers and coordinators are fairly narrowly
+  countries, and our maintainers and coordinators are fairly narrowly
   concentrated in the United States (US) and Europe (see the circle markers on
@@
   This kind of communication may provide bias pespectives on these communities.
-  We therefore need to develop more formal ways of engaging these communities in
-  more democratic ways, such as through running and analyzing user surveys, or
+  We therefore need to develop more formal and democratic ways of engaging these 
+  communities, such as through running and analyzing user surveys, or
   creating and engaging with user groups that can better represent these

### 2022-04-04 src/figures/User-analytics-locations.py

@@ -139,7 +139,7 @@ plot_world = gp.GeoDataFrame(joined.to_pandas())
 joined[joined['Users'] < 0]['name', 'Country', 'Users']
 
 water_color = '#d6f3ff'
-cmap = copy.copy(plt.get_cmap('magma_r'))
+cmap = copy.copy(plt.get_cmap('magma'))
 cmap.set_bad(color='w')
 
 # +

### 2022-04-04 src/ms.tex
missing "is"

--- 2022-04-04 src/ms.tex
@@
 make use of if they do not always need the latest features (this includes but
-not limited to telescope operations and data reduction pipelines). The v5.0
+is not limited to telescope operations and data reduction pipelines). The v5.0
 release of the core package, replacing v4.0, was designated as LTS at the end of 2021;

### 2022-04-04 src/ms.tex
Lim edits for Section 3.1

--- 2022-04-04 src/ms.tex
@@
 for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
-NaN in \astropyTime. However masked \astropyQuantity support is currently very limited,
+NaN in \astropyTime. However, masked \astropyQuantity\ support is currently very limited,
 hindering usage of masks in a \astropyQTable\ (in which \astropyQuantity\ is used
@@
 has been added to the cosmology sub-package for defining and collecting
-cosmological units and equivalencies. \astropycosmologyunits has a unit and
+cosmological units and equivalencies. \astropycosmologyunits\ has a unit and
 equivalencies for \texttt{littleh}, and a new unit, \texttt{redshift}, for
@@
 and associated transformations have been updated to correctly reflect the ``true''
-and ``mean" terminology.
+and ``mean'' terminology.
 
@@
 
-As part of the process of developing \astropypkg into a long-term sustainable
+As part of the process of developing \astropy into a long-term sustainable
 product, and to improve transparency and accountability, the Project agreed to
@@
 structures existed in the associated Open Source Software communities, and what
-would fit well with the needs of \astropy. This led into a ``retreat'' planned
+would fit well with the needs of \astropy. This led to a ``retreat'' planned
 for March 2020, but due to the COVID-19 pandemic, this became a series of
 virtual meetings of the ``Astropy Governance Working Group.'' This group
-drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
+drafted the APE 0 document \citep{ape0}, which was eventually ratified and
 implemented by the ``Astropy Governance Implementation Working Group'' in Fall
-2021. While the process emphasized flexibility and the ability to adapt to
-changing circumstances, it is expected that this is the framework Astropy's
-governance will operate in for at least the medium-term future.
-
-The APE0 \citep{ape0} document lays out the principles of this governance
+2021. While the process emphasized flexibility and adaptability,
+it is the expectation that \astropy's governance will operate in
+this framework for at least the medium-term future.
+
+The APE 0 \citep{ape0} document lays out the principles of this governance
 structure, so we refer the reader to that document for a more thorough
-description, and here we only highlight some key elements. While many of these
-principles were already de facto true or have been discussed organically (and
-have been discussed in earlier papers in this series), the APE0-based governance
-aims to provide a single place where the community can agree as a starting
-point. With this in mind, it highlights the developer and user community of the
-Astropy Project as the ultimate sources of authority, as well as the core
-principle of ``do-ocracy'': those who do work for the Project (be it coding,
+description, and here we only highlight some key elements. The APE 0-based
+governance formalizes the building blocks that are already de facto true or
+previously discussed.
+With this in mind, it highlights the developer and user community of the
+\astropy Project as the ultimate sources of authority, as well as the core
+principle of ``do-ocracy'': those who do work for the Project (coding,
 training, or other harder-to-quantify contributions) gain more influence on the
-outputs of the Project by virtue of their effort. However, APE0 adds the
+outputs of the Project by virtue of their effort. However, APE 0 adds the
 concept of ``voting members'': a self-governed part of that community who are
-entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
-from the inception of the project, APE0 establishes a formal voting process for
-CoCo members, and explicitly outlines the rights and responsibilities of the
-CoCo. This role is mainly to facilitate consensus and act as the decision maker
+entrusted to elect the Coordination Committee (CoCo). While the CoCo
+predates APE 0, membership, rights, and responsibilities are now formally outlined.
+This role is mainly to facilitate consensus and make decision
 when other mechanisms have failed, and includes powers that either
-require central authority or secrets (e.g., passwords). However, APE0 also
+require central authority or secrets (e.g., passwords). However, APE 0 also
 charges the Committee to devolve responsibilities and seek community input on
@@
 
-The first Coordination Committee election under these rules took place in Fall
-2021, electing a mix of prior and new coordination committee members,
-and was contested in the sense of more candidates than available slots. This
+The first CoCo election took place in Fall 2021, electing a mix of prior and new members.
+New roles and voting members were also establish. This
 suggests the process is already working to serve the long-term interests of the
@@
 persons who come from underrepresented communities is the lack of resources and
-expertise that are necessary to train students. Writing PEP 8 compliant software
+expertise that are necessary to train students. Writing PEP 8-compliant software
 and understanding the Git and \github workflow is not part of the standard
@@
 community is still in the process of transitioning towards \python-based tools,
-making it difficult for students not currently under advisement by some one with
-extensive \python expertise to get involved with the open source \python
+making it difficult for students not currently under advisement by someone with
+extensive \python expertise to get involved with the open-source \python
 community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
@@
 individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
-supports and promotes world-class, innovative, open source scientific computing.}
+supports and promotes world-class, innovative, open-source scientific computing.}
 Previously, no direct financial support was available for the project development,
@@
     reliance on volunteer-driven maintenance of the openly-developed package and
-    paved the way to acquire funding from federal agencies, representing a major
-    milestone for the success of the project.
+    paved the way to acquire funding from US federal agencies, representing a
+    major milestone for the success of the project.
 
@@
 
-Sphinx is a common and useful tool for generating documentation for Python
+Sphinx is a common and useful tool for generating documentation for \python
 packages. The Astropy project maintains a default Sphinx configuration along
@@
 projects via the
-\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}} meta
-package. This tool provides a pre-configured Sphinx setup compatible with
+\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}}
+meta-package. This tool provides a pre-configured Sphinx setup compatible with
 Astropy projects, which includes several extensions useful for generating API
@@
 The Astropy Package Template helps facilitate the setup and creation of new
-Python packages leveraging the Astropy ecosystem. This tool utilizes the
+\python packages leveraging the Astropy ecosystem. This tool utilizes the
 \href{https://github.com/cookiecutter/cookiecutter}{Cookiecutter} project to
@@
 Index (PyPI) and available for installation. (However, formal releases are still
-done a few times per year.) \texttt{photutils} \cite{photutils} released its
+done a few times per year.) \texttt{photutils} \citep{photutils} released its
 first stable version, indicating that the API will change less frequently, and
 there have been several significant performance improvements. \texttt{ccdproc}
-\cite{ccdproc} also released a new major version, bringing better performance to
+\citep{ccdproc} also released a new major version, bringing better performance to
 some image combination operations. The \texttt{regions} package
-\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
+\citep{pyregions}, for manipulating ds9-style region definitions \citep{ds9},
 added new ways to manipulate regions and introduced new region types.
-\texttt{reproject}'s \cite{reproject} major new feature is a function to align
+\texttt{reproject}'s \citep{reproject} major new feature is a function to align
 and co-add images to create a mosaic; better support for parallelization was
 also added. \texttt{specutils}, the package that defines containers for 1D and
-2D spectra \cite{specutils}, also had its first stable release and the addition
+2D spectra \citep{specutils}, also had its first stable release and the addition
 of classes to read JWST data. \texttt{stingray} has also had a major performance
-overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
+overhaul. The package \texttt{gammapy} \citep{gammapy} also has major performance
 improvements and has unified its API in preparation for its first stable
@@
 A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
-a python package developed to perform all the necessary tasks to perform data processing, from the raw data
+a \python package developed to perform all the necessary tasks to perform data processing, from the raw data
 of Cherenkov telescopes to the reconstructed event lists.
 The high-level analysis (or CTA science tools) will be based on the Astropy affiliated package
-Gammapy~\citep{gammapy}.
-Both Gammapy and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
+\package{gammapy}~\citep{gammapy}.
+Both \package{gammapy} and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
 tables, and FITS I/O.
@@
 based on earlier work by B.~Winkel.
-Together with Gammapy maintainer A.~Donath and former maintainer C.~Deil,
+Together with \package{gammapy} maintainer A.~Donath and former maintainer C.~Deil,
 a total of 107 merged pull requests were contributed to astropy.
@@
 The efforts focus on developing online content and workshops covering
-astronomy-specific coding tasks in Python.
+astronomy-specific coding tasks in \python.
 As introduced in \citet{astropy:2018}, there are four different types of
@@
 type of astronomical analyses; \textit{examples}, which are snippets of code
-that showcase a short task that can be performed with Astropy or an affiliated
+that showcase a short task that can be performed with \astropypkg or an affiliated
 package; and \textit{documentation}, which contains more detailed information
@@
 during the virtual AAS 237 meetings, to provide the general astronomy community
-information and access to experts on a variety of open source astronomical
+information and access to experts on a variety of open-source astronomical
 tools.
@@
 website, and to facilitate the development of new content that highlights
-\astropy and \astropy-affiliated package functionality.
+\astropypkg and \astropy-affiliated package functionality.
 However, we simultaneously also plan to solicit indexing and ingestion of
@@
 \href{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}{Astropy Roadmap} is a document in the \texttt{astropy-project} \github
-repository within the \astropy organization that captures high level actionable
+repository within the \astropy organization that captures high-level actionable
 items that the Astropy project aims to undertake to improve the health and
@@
 have evolved, but many common themes have persisted and are shared by other
-open source software communities.
+open-source software communities.
 However, \astropy also faces a number of unique challenges stemming from the
@@
   respective communities.
-  This kind of communication may provide bias pespectives on these communities.
+  This kind of communication may provide biased perspectives on these
+  communities.
   We therefore need to develop more formal and democratic ways of engaging these
@@
 a number of web services for code hosting, continuous integration, and
-documentation; in particular, \astropy heavily relies on GitHub, Azure
+documentation; in particular, \astropy heavily relies on \github, Azure
 Pipelines, CircleCI, and Read the Docs.

### 2022-04-05 src/ms.tex
Add a missing comma and space
- Comma after "However": generally good style, and seems to be used
  everywhere else "however" is used.
- Escape a space after a \command.

--- 2022-04-04 src/ms.tex
@@
 make use of if they do not always need the latest features (this includes but
-is not limited to telescope operations and data reduction pipelines). The v5.0
+not limited to telescope operations and data reduction pipelines). The v5.0
 release of the core package, replacing v4.0, was designated as LTS at the end of 2021;
@@
   While \astropy is used globally --- for example,
-  Figure~\ref{fig:User-analytics-locations} shows the numnber of users per
+  Figure~\ref{fig:User-analytics-locations} shows the number of users per
   country who accessed the \astropypkg core package documentation in 2021 ---

### 2022-04-05 src/ms.tex
Escape another space
Hyphenate "well-defined" when a compound adjective

@@
 for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
-NaN in \astropyTime. However, masked \astropyQuantity\ support is currently very limited,
+NaN in \astropyTime. However masked \astropyQuantity support is currently very limited,
 hindering usage of masks in a \astropyQTable\ (in which \astropyQuantity\ is used

### 2022-04-05 src/ms.tex
Typo fix: "done my" -> "done by"

@@
 We aim to make it possible for any array-valued data to be used as a column in a
-table. A well-defined protocol for mixin-columns has been developed for
+table. A well defined protocol for mixin-columns has been developed for
 \astropytable, allowing the original object to be used as a column with a

### 2022-04-05 src/ms.tex
Fix a comma splice
I split it into two sentences, but a semicolon or mdash would also work.

--- 2022-04-05 src/ms.tex
@@
 persons who come from underrepresented communities is the lack of resources and
-expertise that are necessary to train students. Writing PEP 8-compliant software
+expertise that are necessary to train students. Writing PEP 8 compliant software
 and understanding the Git and \github workflow is not part of the standard
@@
 CTA will be able to observe gamma rays in a broad energy range from around $20\,\mathrm{GeV}$ to over $300\,\mathrm{TeV}$
-using three different types of telescopes, in total over 100 telescopes are planned at the two sites.
+using three different types of telescopes. In total over 100 telescopes are planned at the two sites.
 CTA will also be the first open gamma-ray observatory.

### 2022-04-05 src/ms.tex
Fix missing backslash before "python"

--- 2022-04-05 src/ms.tex
@@
 A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
-a python package developed to perform all the necessary tasks to perform data processing, from the raw data
+a \python package developed to perform all the necessary tasks to perform data processing, from the raw data
 of Cherenkov telescopes to the reconstructed event lists.

### 2022-04-05 src/ms.tex
Add another missing hyphen

--- 2022-04-05 src/ms.tex
@@
 CTA will be able to observe gamma rays in a broad energy range from around $20\,\mathrm{GeV}$ to over $300\,\mathrm{TeV}$
-using three different types of telescopes. In total over 100 telescopes are planned at the two sites.
+using three different types of telescopes, in total over 100 telescopes are planned at the two sites.
 CTA will also be the first open gamma-ray observatory.

### 2022-04-05 src/figures/User-analytics-locations.py
Revert colormap changes

@@ -139,7 +139,7 @@ plot_world = gp.GeoDataFrame(joined.to_pandas())
 joined[joined['Users'] < 0]['name', 'Country', 'Users']
 
 water_color = '#d6f3ff'
-cmap = copy.copy(plt.get_cmap('magma'))
+cmap = copy.copy(plt.get_cmap('magma_r'))
 cmap.set_bad(color='w')
 
 # +

### 2022-04-05 src/ms.tex
Use "open-source" everywhere, not "open source"
Previously the text contained both "open-source" and "open source". The
instances of "open source" have now been replaced with "open-source" for
the sake of consistency, except the one instance that is commented out.

@@
 making it difficult for students not currently under advisement by some one with
-extensive \python expertise to get involved with the open source \python
+extensive \python expertise to get involved with the open-source \python
 community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
@@
 
-The data analysis pipeline is developed as open source software and essentially split in two domains:
+The data analysis pipeline is developed as open-source software and essentially split in two domains:
 \begin{enumerate}

### 2022-04-05 src/ms.tex
Minor updates to Future Plans section (#64)
* Word flow alterations

* Web analytics colormap update

* Revert colormap changes

--- 2022-04-05 src/ms.tex
@@
   our user community is underrepresented in non-English-speaking or -teaching
-  countries and our maintainers and coordinators are fairly narrowly
+  countries, and our maintainers and coordinators are fairly narrowly
   concentrated in the United States (US) and Europe (see the circle markers on

### 2022-04-05 src/ms.tex
Fix a handful of typos

@@
 community is still in the process of transitioning towards \python-based tools,
-making it difficult for students not currently under advisement by some one with
+making it difficult for students not currently under advisement by someone with
 extensive \python expertise to get involved with the open-source \python

### 2022-04-05 src/ms.tex
Improve consistency of formatting software names

@@
 The high-level analysis (or CTA science tools) will be based on the Astropy affiliated package
-Gammapy~\citep{gammapy}.
-Both Gammapy and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
+\package{gammapy}~\citep{gammapy}.
+Both \package{gammapy} and \texttt{ctapipe} make heavy use of the \astropy core package, mainly for units, times, coordinate transformations,
 tables, and FITS I/O.

### 2022-04-06 src/ms.tex

--- 2022-04-05 src/ms.tex
@@
 
-The APE0 \citep{ape0} document lays out the principles of this governance
+The APE 0 \citep{ape0} document lays out the principles of this governance
 structure, so we refer the reader to that document for a more thorough
@@
 principles were already de facto true or have been discussed organically (and
-have been discussed in earlier papers in this series), the APE0-based governance
+have been discussed in earlier papers in this series), the APE 0-based governance
 aims to provide a single place where the community can agree as a starting
@@
 training, or other harder-to-quantify contributions) gain more influence on the
-outputs of the Project by virtue of their effort. However, APE0 adds the
+outputs of the Project by virtue of their effort. However, APE 0 adds the
 concept of ``voting members'': a self-governed part of that community who are
 entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
-from the inception of the project, APE0 establishes a formal voting process for
+from the inception of the project, APE 0 establishes a formal voting process for
 CoCo members, and explicitly outlines the rights and responsibilities of the
@@
 when other mechanisms have failed, and includes powers that either
-require central authority or secrets (e.g., passwords). However, APE0 also
+require central authority or secrets (e.g., passwords). However, APE 0 also
 charges the Committee to devolve responsibilities and seek community input on

### 2022-04-08 src/ms.tex
Minor text corrections

@@
 On the other hand, the total number of top contributors has not grown
-substantially, reflecting one of our principal challenges: Sustaining a pipeline
+substantially, reflecting one of our principal challenges: sustaining a pipeline
 from users to contributors to maintainers of the package.
@@
 software engineers who contributed to \astropypkg were (or are) employed by
-STScI, whose leadership have consistently provided resources (financial and
+STScI, whose leadership has consistently provided resources (financial and
 personnel) towards sustaining \astropypkg.

### 2022-04-08 src/ms.tex
Lim edits for Sections 3.2-3.4

@@
 structures existed in the associated Open Source Software communities, and what
-would fit well with the needs of \astropy. This led into a ``retreat'' planned
+would fit well with the needs of \astropy. This led to a ``retreat'' planned
 for March 2020, but due to the COVID-19 pandemic, this became a series of
 virtual meetings of the ``Astropy Governance Working Group.'' This group
-drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
+drafted the APE 0 document \citep{ape0}, which was eventually ratified and
 implemented by the ``Astropy Governance Implementation Working Group'' in Fall
-2021. While the process emphasized flexibility and the ability to adapt to
-changing circumstances, it is expected that this is the framework Astropy's
-governance will operate in for at least the medium-term future.
+2021. While the process emphasized flexibility and adaptability,
+it is the expectation that \astropy's governance will operate in
+this framework for at least the medium-term future.
 
@@
 structure, so we refer the reader to that document for a more thorough
-description, and here we only highlight some key elements. While many of these
-principles were already de facto true or have been discussed organically (and
-have been discussed in earlier papers in this series), the APE 0-based governance
-aims to provide a single place where the community can agree as a starting
-point. With this in mind, it highlights the developer and user community of the
-Astropy Project as the ultimate sources of authority, as well as the core
-principle of ``do-ocracy'': those who do work for the Project (be it coding,
+description, and here we only highlight some key elements. The APE 0-based
+governance formalizes the building blocks that are already de facto true or
+previously discussed.
+With this in mind, it highlights the developer and user community of the
+\astropy Project as the ultimate sources of authority, as well as the core
+principle of ``do-ocracy'': those who do work for the Project (coding,
 training, or other harder-to-quantify contributions) gain more influence on the
@@
 concept of ``voting members'': a self-governed part of that community who are
-entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
-from the inception of the project, APE 0 establishes a formal voting process for
-CoCo members, and explicitly outlines the rights and responsibilities of the
-CoCo. This role is mainly to facilitate consensus and act as the decision maker
+entrusted to elect the Coordination Committee (CoCo). While the CoCo
+predates APE 0, membership, rights, and responsibilities are now formally outlined.
+This role is mainly to facilitate consensus and make decision
 when other mechanisms have failed, and includes powers that either
@@
 As described in the introduction above, the \astropy Project is broader than the
-core \astropypkg package as it also represents a number of core infrastructure
-packages, affiliated packages, educational initiatives, and user communities.
+core \astropypkg package as it also represents a number of core infrastructure,
+coordinated, and affiliated packages, educational initiatives, and user communities.
 In total, well over 1500 people have contributed in some way to the Project,
-whether, for example, through contributing code or documentation, opening issues
+e.g., through contributing code or documentation, reporting issues
 on \github, providing infrastructure support, writing tutorials, or helping to
-coordinate workshops that relate to Astropy.
+coordinate \astropy workshops.
 The vast majority of these contributors and participants have interacted as
@@
 growth), as new, active contributors have generally joined and stepped in to
-fill top contributor roles within the project.
+fill top-contributor roles within the project.
 On the other hand, the total number of top contributors has not grown
@@
 STScI, whose leadership has consistently provided resources (financial and
-personnel) towards sustaining \astropypkg.
+personnel) towards sustaining \astropy.
 
@@
 Native American scientists across all STEM fields). \astropy representatives
-noted that one underlying topic came up multiple times. A major barrier for
-persons who come from underrepresented communities is the lack of resources and
-expertise that are necessary to train students. Writing PEP 8-compliant software
-and understanding the Git and \github workflow is not part of the standard
+noted that one underlying topic came up multiple times: A major barrier for
+persons who come from underrepresented communities is the lack of pedagogical
+resources and expertise. For instance, software engineering skills and best
+practices are not part of the standard
 Physics and Astrophysics curriculum. Additionally, the larger astronomical
 community is still in the process of transitioning towards \python-based tools,
-making it difficult for students not currently under advisement by someone with
-extensive \python expertise to get involved with the open-source \python
+hindering students not currently under advisement by someone with
+extensive \python expertise from getting involved with the open-source \python
 community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
@@
 
-With the goal of establishing a sustainable financial model for the Astropy
-Project, the Coordination Committee, under the new governance charter (see
+With the goal of establishing a sustainable financial model for the \astropy
+Project, the CoCo, under the new governance charter (see
 Section~\ref{sec:project-governance}), established a standing Finance Committee.
 This committee oversees the planning and allocation of finances on behalf
-of the Project. The Astropy Project accepts funds from institutions as well as
+of the Project. The \astropy Project accepts funds from institutions as well as
 individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
 supports and promotes world-class, innovative, open-source scientific computing.}
-Previously, no direct financial support was available for the project development,
+Previously, no direct financial support was available for project development,
 and NumFOCUS covered most of the incurred operational costs. However, transitioning
 to long-term financial stability meant having an influx of funds in the form of
-institutional support, and hence the Astropy Project applied for and successfully
+institutional support, and hence the \astropy Project applied for and successfully
 acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
-Project, providing a ``neutral'' space not tied to any particular astronomy
+Project, providing a ``neutral'' space not tied to any particular astronomical
 institution. The two major grants are briefly summarized below.
@@
     formalization of a governance structure (see
-    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
-    diversity, and inclusion efforts in the project (see
+    Section~\ref{sec:project-governance}), and (4) the improvement of
+    IDE efforts in the project (see
     Section~\ref{sec:project-ide}). Some of this money supported travel for
@@
     (Support for Open Source Tools, Frameworks, and Libraries), granted the
-    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
+    \astropy Project $\sim$\$600k (US) to support its work. Funded work for a
     term of three years includes (1) project-wide infrastructure maintenance and
@@
 
-In addition to the two grants described above, the project benefited from
+In addition to the two grants described above, the Project benefited from
 a Gemini Observatory contract under its Science User Support Department,
 which awarded funds for development work that supports both the Astropy and
-DRAGONS projects. The project also received financial support from the Dunlap
+DRAGONS projects. The Project also received financial support from the Dunlap
 Seed Funding program at the University of Toronto to develop educational
 resources for scientific software packaging within the Learn Astropy
-framework. All of these successful proposals show that the Astropy Project
+framework. The project also indirectly benefited from the NRAO ALMA Development
+Study program, which supported the grant ``Linking CASA to the astropy
+ecosystem''\footnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
+All of these successful proposals show that the \astropy Project
 is vital for the astronomical infrastructure and community at large, and
@@
 information on the \href{{https://www.astropy.org/affiliated}}{\astropy
-website}; A table summarizing key aspects of the affiliated packages is included
+website}; a table summarizing key aspects of the affiliated packages is included
 in the Appendix below (Table~\ref{tab:affiliated-registry}).
@@
 there have been several significant performance improvements. \texttt{ccdproc}
-\citep{ccdproc} also released a new major version, bringing better performance to
+\citep{ccdproc}  released a new major version, bringing better performance to
 some image combination operations. The \texttt{regions} package
-\citep{pyregions}, for manipulating ds9-style region definitions \citep{ds9},
-added new ways to manipulate regions and introduced new region types.
+\citep{pyregions}, for manipulating ds9-, CASA-, or FITS-style region definitions \citep{ds9},
+added new ways to manipulate regions, introduced new region types, and obtained
+expanded visualization and interactive region editing capabilities.
 \texttt{reproject}'s \citep{reproject} major new feature is a function to align
@@
 also added. \texttt{specutils}, the package that defines containers for 1D and
-2D spectra \citep{specutils}, also had its first stable release and the addition
-of classes to read JWST data. \texttt{stingray} has also had a major performance
-overhaul. The package \texttt{gammapy} \citep{gammapy} also has major performance
+2D spectra \citep{specutils},  had its first stable release and the addition
+of classes to read JWST data. \texttt{stingray} has had a major performance
+overhaul. The package \texttt{gammapy} \citep{gammapy} has had major performance
 improvements and has unified its API in preparation for its first stable

### 2022-04-08 src/ms.tex
Add an indirect support link
I'm open to discussion on whether/how this should be included.  I can expand on what components were relevant to astropy.
capitalization

--- 2022-04-08 src/ms.tex
@@
 resources for scientific software packaging within the Learn Astropy
-framework. All of these successful proposals show that the Astropy Project
+framework.  The project also indirectly benefited from the NRAO ALMA Development
+Study program, which supported the grant ``Linking CASA to the astropy ecosystem''\foontnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
+All of these successful proposals show that the Astropy Project
 is vital for the astronomical infrastructure and community at large, and

### 2022-04-08 src/ms.tex
add casaformats-io reference
https://github.com/radio-astro-tools/casa-formats-io
expand on astropy-regions (new interactive editing tools!)

@@
 some image combination operations. The \texttt{regions} package
-\citep{pyregions}, for manipulating ds9-style region definitions \citep{ds9},
-added new ways to manipulate regions and introduced new region types.
+\citep{pyregions}, for manipulating ds9-, CASA-, or FITS style region definitions \citep{ds9},
+added new ways to manipulate regions, introduced new region types, and obtained
+expanded visualization and interactive region editing capabilities.
 \texttt{reproject}'s \citep{reproject} major new feature is a function to align

### 2022-04-08 src/ms.tex
cut down on 'also'
I'd suggest doing a global search for 'also's and remove them when unnecessary

@@
 there have been several significant performance improvements. \texttt{ccdproc}
-\citep{ccdproc} also released a new major version, bringing better performance to
+\citep{ccdproc}  released a new major version, bringing better performance to
 some image combination operations. The \texttt{regions} package
@@
 also added. \texttt{specutils}, the package that defines containers for 1D and
-2D spectra \citep{specutils}, also had its first stable release and the addition
-of classes to read JWST data. \texttt{stingray} has also had a major performance
-overhaul. The package \texttt{gammapy} \citep{gammapy} also has major performance
+2D spectra \citep{specutils},  had its first stable release and the addition
+of classes to read JWST data. \texttt{stingray} has had a major performance
+overhaul. The package \texttt{gammapy} \citep{gammapy} has major performance
 improvements and has unified its API in preparation for its first stable

### 2022-04-12 src/ms.tex
Update src/ms.tex
Co-authored-by: Nicholas Earl <contact@nicholasearl.me>

@@
 some image combination operations. The \texttt{regions} package
-\citep{pyregions}, for manipulating ds9-, CASA-, or FITS style region definitions \citep{ds9},
+\citep{pyregions}, for manipulating ds9-, CASA-, or FITS-style region definitions \citep{ds9},
 added new ways to manipulate regions, introduced new region types, and obtained

### 2022-04-12 src/ms.tex
Update src/ms.tex
Co-authored-by: Nicholas Earl <contact@nicholasearl.me>

@@
 of classes to read JWST data. \texttt{stingray} has had a major performance
-overhaul. The package \texttt{gammapy} \citep{gammapy} has major performance
+overhaul. The package \texttt{gammapy} \citep{gammapy} has had major performance
 improvements and has unified its API in preparation for its first stable

### 2022-04-15 src/ms.tex

@@
 framework.  The project also indirectly benefited from the NRAO ALMA Development
-Study program, which supported the grant ``Linking CASA to the astropy ecosystem''\foontnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
+Study program, which supported the grant ``Linking CASA to the astropy ecosystem''\footnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
 All of these successful proposals show that the Astropy Project

### 2022-04-15 src/ms.tex
Apply suggestions from code review

--- 2022-04-12 src/ms.tex
@@
 On the other hand, the total number of top contributors has not grown
-substantially, reflecting one of our principal challenges: sustaining a pipeline
+substantially, reflecting one of our principal challenges: Sustaining a pipeline
 from users to contributors to maintainers of the package.
@@
 software engineers who contributed to \astropypkg were (or are) employed by
-STScI, whose leadership has consistently provided resources (financial and
+STScI, whose leadership have consistently provided resources (financial and
 personnel) towards sustaining \astropypkg.
@@
 resources for scientific software packaging within the Learn Astropy
-framework.  The project also indirectly benefited from the NRAO ALMA Development
-Study program, which supported the grant ``Linking CASA to the astropy ecosystem''\foontnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
-All of these successful proposals show that the Astropy Project
+framework. All of these successful proposals show that the Astropy Project
 is vital for the astronomical infrastructure and community at large, and
@@
 information on the \href{{https://www.astropy.org/affiliated}}{\astropy
-website}; a table summarizing key aspects of the affiliated packages is included
+website}; A table summarizing key aspects of the affiliated packages is included
 in the Appendix below (Table~\ref{tab:affiliated-registry}).
@@
 \texttt{corral} for managing data intensive parallel pipelines.
-\texttt{casaformats-io} provides infrastructure for reading data stored in CASA table formats.
 \texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
@@
 there have been several significant performance improvements. \texttt{ccdproc}
-\citep{ccdproc}  released a new major version, bringing better performance to
+\citep{ccdproc} also released a new major version, bringing better performance to
 some image combination operations. The \texttt{regions} package
-\citep{pyregions}, for manipulating ds9-, CASA-, or FITS-style region definitions \citep{ds9},
-added new ways to manipulate regions, introduced new region types, and obtained
-expanded visualization and interactive region editing capabilities.
+\citep{pyregions}, for manipulating ds9-style region definitions \citep{ds9},
+added new ways to manipulate regions and introduced new region types.
 \texttt{reproject}'s \citep{reproject} major new feature is a function to align
@@
 also added. \texttt{specutils}, the package that defines containers for 1D and
-2D spectra \citep{specutils},  had its first stable release and the addition
-of classes to read JWST data. \texttt{stingray} has had a major performance
-overhaul. The package \texttt{gammapy} \citep{gammapy} has major performance
+2D spectra \citep{specutils}, also had its first stable release and the addition
+of classes to read JWST data. \texttt{stingray} has also had a major performance
+overhaul. The package \texttt{gammapy} \citep{gammapy} also has major performance
 improvements and has unified its API in preparation for its first stable

### 2022-04-15 src/ms.tex

@@
 Native American scientists across all STEM fields). \astropy representatives
-noted that one underlying topic came up multiple times. A major barrier for
-persons who come from underrepresented communities is the lack of resources and
-expertise that are necessary to train students. Writing PEP 8-compliant software
-and understanding the Git and \github workflow is not part of the standard
+noted that one underlying topic came up multiple times: a major barrier for
+persons who come from underrepresented communities is the lack of pedagogical
+resources and expertise. For instance, software engineering skills and best
+practices are not part of the standard
 Physics and Astrophysics curriculum. Additionally, the larger astronomical
 community is still in the process of transitioning towards \python-based tools,
-making it difficult for students not currently under advisement by someone with
-extensive \python expertise to get involved with the open-source \python
+hindering students not currently under advisement by someone with
+extensive \python expertise from getting involved with the open-source \python
 community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
@@
 
-With the goal of establishing a sustainable financial model for the Astropy
-Project, the Coordination Committee, under the new governance charter (see
+With the goal of establishing a sustainable financial model for the \astropy
+Project, the CoCo, under the new governance charter (see
 Section~\ref{sec:project-governance}), established a standing Finance Committee.
 This committee oversees the planning and allocation of finances on behalf
-of the Project. The Astropy Project accepts funds from institutions as well as
+of the Project. The \astropy Project accepts funds from institutions as well as
 individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
 supports and promotes world-class, innovative, open-source scientific computing.}
-Previously, no direct financial support was available for the project development,
+Previously, no direct financial support was available for project development,
 and NumFOCUS covered most of the incurred operational costs. However, transitioning
 to long-term financial stability meant having an influx of funds in the form of
-institutional support, and hence the Astropy Project applied for and successfully
+institutional support, and hence the \astropy Project applied for and successfully
 acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
@@
 
-In addition to the two grants described above, the project benefited from
+In addition to the two grants described above, the Project benefited from
 a Gemini Observatory contract under its Science User Support Department,
 which awarded funds for development work that supports both the Astropy and
-DRAGONS projects. The project also received financial support from the Dunlap
+DRAGONS projects. The Project also received financial support from the Dunlap
 Seed Funding program at the University of Toronto to develop educational
@@
 framework. The project also indirectly benefited from the NRAO ALMA Development
-Study program, which supported the grant ``Linking CASA to the astropy ecosystem''\footnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
-All of these successful proposals show that the Astropy Project
+Study program, which supported the grant ``Linking CASA to the astropy
+ecosystem''\footnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
+All of these successful proposals show that the \astropy Project
 is vital for the astronomical infrastructure and community at large, and

### 2022-04-15 src/ms.tex
fix language around LTS. fixes #98
fix en dash issue

@@
 released. However, every two years a major release is designated as a long-term
-support (LTS) release, which continues to be maintained for up to two years
+support (LTS) release, which continues to be maintained for at least two years
 \citep{ape2}. The motivation for LTS releases is to provide longer-term stable

### 2022-04-15 src/ms.tex
fix capitalization in project vs Project

@@
 In this article, we summarize key features in the core package as of the recent major
-release, version 5.0, and provide major updates for the project.
+release, version 5.0, and provide major updates for the Project.
 We then discuss supporting a broader ecosystem of interoperable packages,
@@
 \href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
-Summer of Code} 2017 program,\footnote{ To learn more about this project, please
+Summer of Code} 2017 program,\footnote{To learn more about this project, please
 see the final report
@@
 growth), as new, active contributors have generally joined and stepped in to
-fill top-contributor roles within the project.
+fill top-contributor roles within the Project.
 On the other hand, the total number of top contributors has not grown
@@
             consecutive 30 day windows) merged into the \astropypkg core
-            package over the whole history of the project (up to end of 2021,
+            package over the whole history of the Project (up to end of 2021,
             v5.0).
@@
 allocated funding to support mentoring programs. In 2020 a call was
-made to submit proposals for IDE (Inclusion, Diversity, and
-Empowerment) initiatives for project-wide consideration on
+made to submit proposals for Inclusion, Diversity, and
+Empowerment (IDE) initiatives for project-wide consideration on
 \github. This process was deemed the most ``open'' because it allowed
@@
     paved the way to acquire funding from US federal agencies, representing a
-    major milestone for the success of the project.
+    major milestone for the success of the Project.
 
@@
 
-The Astropy project supports the broader ecosystem by providing pre-configured
+The Astropy Project supports the broader ecosystem by providing pre-configured
 infrastructure packages that the community can use to support and maintain their
@@
 Sphinx is a common and useful tool for generating documentation for \python
-packages. The Astropy project maintains a default Sphinx configuration along
+packages. The Astropy Project maintains a default Sphinx configuration along
 with Astropy-specific extensions that can be easily added to community
@@
 workshops with as many as 90 participants and a dozen facilitators from the
-project.
+Project.
 During the pandemic, these workshops were moved to an online format and split
@@
 repository within the \astropy organization that captures high-level actionable
-items that the Astropy project aims to undertake to improve the health and
-stability of the project. It is a static document that is revisited regularly at
+items that the Astropy Project aims to undertake to improve the health and
+stability of the Project. It is a static document that is revisited regularly at
 the Astropy coordination meetings, intended to keep track of progress and write
@@
   \item[Diversity of users, contributors, maintainers, and coordinators]
-  The \astropy project aims to be a project that is made up of members at all
-  levels (users to contributors to coordinators) with a diverse set of skills,
-  backgrounds, and experiences.
+  The \astropy Project aims to be made up of members at all levels (users to
+  contributors to coordinators) with a diverse set of skills, backgrounds, and
+  experiences.
   While \astropy is used globally --- for example,
@@
   institutions have explicitly supported \astropy, and where volunteers have had
-  the ability to contribute to the project.
+  the ability to contribute to the Project.
   \item[Long-term and sustained funding for maintaining infrastructure] The
@@
 (Section~\ref{sec:project-contributors}), and have begun to receive significant
-formal funding for the project itself (Section~\ref{sec:project-funding}).
+formal funding for the Project itself (Section~\ref{sec:project-funding}).
 The Astropy Project is also expanding its educational and community-focused
@@
 \astropy, opened issues and provided feedback, and supported the
-project in a number of different ways.
+Project in a number of different ways.
 

### 2022-04-15 src/ms.tex
update numfocus footnote

@@
 of the Project. The \astropy Project accepts funds from institutions as well as
-individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
-supports and promotes world-class, innovative, open-source scientific computing.}
+individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) non-profit
+organization in the United States of America that can collect donations,
+administer grants, and perform financial management for projects in open source
+scientific computing.}
 Previously, no direct financial support was available for project development,

### 2022-04-15 src/ms.tex
add footnote defining and linking to DRAGONS

@@
 which awarded funds for development work that supports both the Astropy and
-DRAGONS projects. The Project also received financial support from the Dunlap
+DRAGONS\footnote{Data Reduction for Astronomy from Gemini Observatory North and
+South (\url{https://github.com/GeminiDRSoftware/DRAGONS}).} projects. The
+Project also received financial support from the Dunlap
 Seed Funding program at the University of Toronto to develop educational

### 2022-04-15 src/ms.tex
address comment about transition of project focus

@@
 Project are evolving.
-In particular, the transition from focusing our energy on development and
-maintenance of a single core package, to instead maintaining the core package
-and fostering the development of the community and its expansion has been a key
-goal of the Astropy Project over the last several years.
+In particular, fostering the development of the community and its expansion has
+been an additional key goal of the Astropy Project over the last several years.
 

### 2022-04-15 src/ms.tex
rephrase to remove please request

@@
 \href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
-Summer of Code} 2017 program,\footnote{To learn more about this project, please
+Summer of Code} 2017 program,\footnote{To learn more about this project,
 see the final report
@@
 Now that the backend infrastructure is stable, we are interested in collecting
-new content to serve and share: if you have ideas for contributions, or have
-tutorials hosted on third-party sites that you would like to make searchable
-through the Learn Astropy interface, please reach out by creating a \github
-issue in the \href{https://github.com/astropy/astropy-tutorials}{Astropy
-Tutorials repository}.
+new content to serve and share.
+Ideas for contributions, new tutorials, or suggestions for new tutorial series
+to be made searchable through the Learn Astropy interface can be raised by
+creating a \github issue in the
+\href{https://github.com/astropy/astropy-tutorials}{Astropy Tutorials GitHub
+repository}.
 We currently host one Guide --- a longer-form walkthrough of a more complex

### 2022-04-15 src/ms.tex
added description of the coordinated packages

@@
 Astropy of improving code re-use, interoperability, and embracing good coding
-practices such as testing and thorough documentation.
+practices such as testing and thorough documentation, but are not solely
+developed and maintained by members of the Astropy Project team.
+However, a subset of the affiliated packages --- the ``Astropy Coordinated
+Packages'' --- are maintained by the Project.
 All Astropy-coordinated and other affiliated packages are listed with detailed

### 2022-04-15 src/ms.tex
rephrase volunteer effort claim

@@
 One of the factors that has led to its rapid ascent in popularity in scientific
-contexts has been the significant, volunteer-driven effort behind developing
-community-oriented open-source software tools and fostering communities of users
-and developers that have grown around these efforts.
+contexts is the combination of volunteer-driven and professionally-supported
+effort behind developing community-oriented open-source software tools and
+fostering communities of users and developers that have grown around these
+efforts.
 Today, a broad and feature-diverse ``ecosystem'' of packages exists in the

### 2022-04-15 src/ms.tex
remove redundant list of contributed packages

@@
     \item[NASA] A successful proposal to the NASA ROSES-2020
-    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
+    call\footnote{\url{http://solicitation.nasaprs.com/ROSES2020}}, section E.7
     (Support for Open Source Tools, Frameworks, and Libraries), granted the

### 2022-04-16 src/ms.tex
comment out section author

@@
 resources for scientific software packaging within the Learn Astropy
-framework. The project also indirectly benefited from the NRAO ALMA Development
+framework. Moreover, the Project indirectly benefited from the NRAO ALMA Development
 Study program, which supported the grant ``Linking CASA to the astropy
 ecosystem''\footnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
-All of these successful proposals show that the \astropy Project
+All of these successful funding proposals show that the \astropy Project
 is vital for the astronomical infrastructure and community at large, and

### 2022-04-16 src/ms.tex
add cython reference

@@
 at the cost of performance.
-Hence, \github pull requests that improve performance are welcomed, and were even the
-main goal for v3.1 release. For most sub-packages, performance was
+Hence, \github pull requests that improve performance are welcomed, and were
+even the main goal for v3.1 release. For most sub-packages, performance was
 improved in the \python\ code, but for some, the most time-critical pieces were
-rewritten in~C (e.g., for convolution of images, sigma-clipping, and converting
-time strings to binary). Furthermore, a particular effort was made to ensure
-\astropypkg\ is thread-safe, so that it can be used on supercomputing clusters.
+rewritten in~C and wrapped with \package{Cython} (\citealt{cython}; e.g., for
+convolution of images, sigma-clipping, and converting time strings to binary).
+Furthermore, a particular effort was made to ensure \astropypkg\ is thread-safe,
+so that it can be used on supercomputing clusters.
 

### 2022-04-17 src/static/repo-emails.py
add script to collect all emails for people who have committed to astropy org projects

@@ -1,0 +1,58 @@
+from collections import defaultdict
+import pathlib
+import git
+
+# Get all emails that have contributed to astropy repositories:
+repos_path = pathlib.Path(
+    '~/projects/astropy-all/astropy-sustainability/cache/repos'
+).expanduser().resolve()
+
+repos = {}
+for path in repos_path.glob("*"):
+    if path.parts[-1].startswith('.'):
+        continue
+
+    repo_name = path.parts[-1]
+    repos[repo_name] = git.Repo(path)
+
+all_emails = []
+for repo_name, repo in repos.items():
+    shortlog_raw = repo.git.shortlog(
+        'HEAD', s=True, n=True, e=True, no_merges=True,
+    )
+    lines = shortlog_raw.split('\n')
+
+    shortlog = defaultdict(lambda *_: 0)
+    for line in lines:
+        ncommits, name_email = line.strip().split('\t')
+        _, email = name_email.split('<')
+        email = email[:-1]
+        all_emails.append(email)
+all_emails = set(all_emails)
+
+_emails = [f'{x}\n' for x in list(set(all_emails))
+           if 'users.noreply' not in x
+           and 'github' not in x
+           and '.local' not in x
+           and 'HP-Pavilion' not in x
+           and '@' in x
+           and 'gitter.im' not in x
+           and '0101010101.com' not in x
+           and x != 'anonymous@overleaf.com']
+
+with open('repo-emails.txt', 'w') as f:
+    f.writelines(_emails)
+repo_emails = [x.strip() for x in _emails]
+
+# Load the form emails:
+with open("form-emails.txt", "r") as f:
+    form_emails = [x.strip() for x in f.readlines()]
+
+all_emails = sorted(set(repo_emails + form_emails))
+print(len(all_emails))
+print()
+print("--- BATCH 1: ---")
+print(", ".join(all_emails[:400]))
+
+print("\n\n\n--- BATCH 2: ---")
+print(", ".join(all_emails[400:]))

### 2022-04-18 src/ms.tex
more tweaks to author list formatting

@@
 \documentclass[modern]{aastex631}
+\NewPageAfterKeywords
 \usepackage{xspace}
@@
 
-\submitjournal{ApJ}
+% \submitjournal{ApJ}
 
@@
 
-\author{Astropy Collaboration}
+\author{Astropy~Collaboration}
 \noaffiliation
-{\let\thefootnote\relax\footnote{{The author list has two parts: the authors that made significant contributions to the writing of the paper in order of contribution, followed by contributors to the \astropy Project in alphabetical order.
-\textbf{The position in the author list does not correspond to contributions to the \astropy Project as a whole.} A more complete list of contributors to the core package can be found in the \href{https://github.com/astropy/astropy/graphs/contributors}{package repository}, and at the \href{http://www.astropy.org/team.html}{\astropy team webpage}.}}}
-
-% \input{author.tex}
+{\let\thefootnote\relax\footnote{{
+  The author list has two parts: the authors that made significant contributions
+  to the writing and/or coordination of the paper (in order of contribution),
+  followed by maintainers of and contributors to to the Astropy Project (in
+  alphabetical order). \textbf{The position in the author list does not
+  correspond to contributions to the Astropy Project as a whole.} A more
+  complete list of contributors to the core package can be found in the
+  \href{https://github.com/astropy/astropy/graphs/contributors}{package
+  repository}, and at the \href{http://www.astropy.org/team.html}{Astropy team
+  webpage}.
+}}}
+
+\input{static/authors.tex}
 
@@
 
-\keywords{%
-    Astrophysics - Instrumentation and Methods for Astrophysics
-    ---
-    methods: data analysis
-    ---
-    methods: miscellaneous
-}
+% \keywords{%
+%     Astrophysics - Instrumentation and Methods for Astrophysics
+%     ---
+%     methods: data analysis
+%     ---
+%     methods: miscellaneous
+% }
 
@@
 
-\begin{figure}[th]
+\begin{figure}[t!]
     \begin{centering}

### 2022-04-18 src/static/authors.tex
altaffil coco
fix tom a affil
Apply suggestions from code review
Co-authored-by: Volodymyr <volodymyrss@users.noreply.github.com>
authorlist minor fixes
more tweaks to author list formatting
updated flagged affiliations

@@
 }
+\newcommand{\affcoco}{
+    Astropy Coordination Committee
+}
+\newcommand{\affsteward}{
+    Steward Observatory, University of Arizona, 933 North Cherry Avenue, Tucson,
+    AZ 85721-0065, USA
+}
 
@@
 \author[0000-0003-0872-7098]{Adrian~M.~Price-Whelan}
+\altaffiliation{\affcoco}
 \affiliation{\affcca}
@@
 
-%FLAG
 \author[0000-0003-1714-7415]{Nicholas~Earl}
@@
 \author[0000-0002-9599-310X]{Erik~Tollerud}
+\altaffiliation{\affcoco}
 \affiliation{\affstsci}
@@
 
-%FLAG
 \author{Bruce~Merry}
-\affiliation{South African Radio Astronomy Observatory}
+\affiliation{South African Radio Astronomy Observatory, 2 Fir Street, Black River Park, Observatory 7925, South Africa}
 
@@
 \author[0000-0003-4243-2840]{H.~Moritz~G\"{u}nther}
+\altaffiliation{\affcoco}
 \affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}
+
+\collaboration{23}{Paper Authors}
 
@@
 \author{Thomas~L.~Aldcroft}
-\affiliation{}
+\affiliation{\affcfa}
 
@@
 
-%FLAG
 \author[0000-0001-7699-3983]{Juanjo~Baz{\'a}n}
-\affiliation{CIEMAT, Astroparticle physics, Madrid, Spain.}
+\affiliation{CIEMAT, Astroparticle physics, Madrid, Spain}
 
-%FLAG
 \author{Manish~Biswas}
-\affiliation{Savitribai phule pune university}
+\affiliation{Savitribai Phule Pune University, Pune, Maharashtra, India}
 
@@
 
-\author[0000-0001-5410-5551]{Jos\'e~Vin\'icius~de~Miranda~Cardoso}
-\affiliation{The Hong Kong University of Science and Technology, Department of Electronic and Computer Engineering, Hong Kong SAR China}
-
 \author[0000-0002-9583-263X]{ZHANG~Chen}
@@
 
-%FLAG
 \author[0000-0001-7988-8919]{Matthew~W.~Craig}
-\affiliation{Minnesota State University Moorhead}
+\altaffiliation{\affcoco}
+\affiliation{Department of Physics and Astronomy, Minnesota State University Moorhead, Moorhead, MN 56563}
 
@@
 \author[0000-0002-1821-0650]{Kelle~L.~Cruz}
+\altaffiliation{\affcoco}
 \affiliation{Hunter College, Physics and Astronomy, 695 Park Avenue, New York, NY 10065}
@@
 
-%FLAG
 \author[0000-0002-8919-079X]{Leonardo~Ferreira}
-\affiliation{University of Nottingham}
+\affiliation{Centre for Astronomy and Particle Physics, University of Nottingham, Nottingham, UK}
 
@@
 
-%FLAG
 \author{Steve~Guest}
-\affiliation{RAL Space}
+\affiliation{RAL Space, STFC Rutherford Appleton Laboratory, Harwell, Didcot OX11 0QX, UK}
 
-%FLAG
 \author[0000-0001-5848-4077]{Sebastian~Gurovich}
-\affiliation{IATE-OAC-UNC}
+\affiliation{Instituto De Astronom\'{i}a Te\'{o}rica y Experimental -
+Observatorio Astron\'{o}mico C\'{o}rdoba (IATE–OAC–UNC–CONICET), C\'{o}rdoba,
+Argentina}
 
@@
 
-%FLAG
 \author[0000-0002-8646-8362]{Zac~Hatfield-Dodds}
@@
 \author[0000-0002-0832-2974]{Griffin~Hosseinzadeh}
-\affiliation{Steward Observatory, University of Arizona, 933 North Cherry Avenue, Tucson, AZ 85721-0065, USA}
-
-\author{Anany~Shrey~Jain}
-\noaffiliation
+\affiliation{\affsteward}
 
@@
 
-%FLAG
-\author[0000-0002-0629-3006]{Craig~K~Jones}
+\author[0000-0002-0629-3006]{Craig~K.~Jones}
 \affiliation{The Malone Center for Healthcare in Engineering, Johns Hopkins University}
@@
 
-%FLAG
 \author[0000-0002-4245-2318]{Chun~Ly}
-\affiliation{University of Arizona}
+\affiliation{\affsteward}
 
-\author[0000-0002-3713-6337]{Brigitta~M.~Sip\H{o}cz}
-\affiliation{\affipac}
-
-%FLAG
 \author[0000-0003-3270-6844]{Zhiyuan~Ma}
-\affiliation{University of Massachusetts Amherst}
+\affiliation{Department of Astronomy, University of Massachusetts - Amherst, 710 North Pleasant Street, Amherst, MA 01003, USA}
 
@@
 
-%FLAG
-\author[0000-0001-8100-9514]{Jakob~~M.~Maljaars}
-\affiliation{Science \& Technology Corporation}
+\author[0000-0001-8100-9514]{Jakob~M.~Maljaars}
+\affiliation{Science \& Technology Corporation, Delft, The Netherlands}
 
@@
 
-%FLAG
 \author[0000-0003-4189-3450]{Henrik~Norman}
@@
 
-%FLAG
-\author[0000-0002-5923-4966]{Rohit~R~Patil}
-\affiliation{Syracuse University, Syracuse, NY}
+\author[0000-0002-5923-4966]{Rohit~R.~Patil}
+\affiliation{Syracuse University, Syracuse, NY, USA}
 
@@
 
-%FLAG
-\author[0000-0002-9427-5448]{Timothy~E~Pickering}
-\affiliation{MMT Observatory}
+\author[0000-0002-9427-5448]{Timothy~E.~Pickering}
+\affiliation{MMT Observatory, Tucson, AZ, USA}
 
@@
 
-%FLAG
 \author[0000-0002-9523-5640]{Parikshit~Sakurikar}
-\affiliation{International Institute of Information Technology - Hyderabad}
+\affiliation{International Institute of Information Technology, Hyderabad, India}
 
-%FLAG
 \author[0000-0002-3680-4364]{Jes\'us~Salgado}
@@
 \author[0000-0003-2657-3889]{Nicholas~Saunders}
+\altaffiliation{NSF Graduate Research Fellow}
 \affiliation{Institute for Astronomy, University of Hawaiʻi at M\=anoa, 2680 Woodlawn Drive, Honolulu, HI 96822, USA}
 \affiliation{American Museum of Natural History, 200 Central Park West, Manhattan, NY 10024, USA}
-\altaffiliation{NSF Graduate Research Fellow}
 
-%FLAG
 \author[0000-0001-6353-0808]{Volodymyr~Savchenko}
-\affiliation{University of Geneva, Department of Astronomy}
+\affiliation{Department of Astronomy, University of Geneva, Chemin d'\'Ecogia, 16 CH-1290 Versoix, Switzerland}
+\affiliation{EPFL SCITAS, Station 9, CH-1015 Lausanne, Switzerland}
 
@@
 
-%FLAG
+\author{Anany~Shrey~Jain}
+\noaffiliation
+
 \author[0000-0001-9268-3803]{Gyanendra~Shukla}
-\affiliation{Maulana Azad National Institute Of Technology, Bhopal}
+\affiliation{Maulana Azad National Institute Of Technology, Bhopal, India}
 
@@
 \affiliation{ARC Centre of Excellence for All Sky Astrophysics in 3 Dimensions (ASTRO 3D)}
+
+\author[0000-0002-3713-6337]{Brigitta~M.~Sip\H{o}cz}
+\affiliation{\affipac}
 
@@
 
+\author[0000-0001-5410-5551]{Jos\'e~Vin\'icius~de~Miranda~Cardoso}
+\affiliation{The Hong Kong University of Science and Technology, Department of Electronic and Computer Engineering, Hong Kong SAR China}
+
 \author[0000-0003-3734-3587]{Peter~K.~G.~Williams}
@@
 
-%FLAG
 \author[0000-0002-9583-263X]{Chen~ZHANG}
-\affiliation{Purple Mountain Observatory, CAS}
+\affiliation{Purple Mountain Observatory, CAS, Qixia District, Nanjing 210023, China}
 
@@
 \affiliation{San Diego Supercomputer Center, University of California San Diego, 9500 Gilman Dr, La Jolla, CA 92093, USA}
+
+\collaboration{113}{Astropy Project Contributors}

### 2022-04-18 src/ms.tex
fix graphicspath because of syw incompatibility

@@
 \usepackage{ulem}
-
-% To allow putting figures in a subdir
-\graphicspath{{figures/}}
 
@@
     \begin{centering}
-      \includegraphics[width=\textwidth]{python-mentions.pdf}
+      \includegraphics[width=\textwidth]{figures/python-mentions.pdf}
         \caption{
@@
     \begin{centering}
-      \includegraphics[width=0.9\textwidth]{Ncommitters_vs_frac_commits.pdf}
+      \includegraphics[width=0.9\textwidth]{figures/Ncommitters_vs_frac_commits.pdf}
         \caption{
@@
     \begin{centering}
-      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
+      \includegraphics[width=0.75\textwidth]{figures/top_contributors_career.pdf}
         \caption{
@@
     \begin{centering}
-      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
+      \includegraphics[width=0.75\textwidth]{figures/commits_committers_per_month.pdf}
         \caption{
@@
   \begin{centering}
-    \includegraphics[width=\textwidth]{docu_analytics.pdf}
+    \includegraphics[width=\textwidth]{figures/docu_analytics.pdf}
       \caption{

### 2022-04-18 src/ms.tex
JWST

@@
 representations had to be supported for new missions and observatories (e.g.,
-the James Webb Space Telescope and the Rubin Observatory). To harmonize these
+JWST and the Rubin Observatory). To harmonize these
 needs, a new high-level interface was created based on a formal design first
@@
 \citep{astroquery} has added access to roughly a dozen new missions and data
-services, including the James Webb Space Telescope (JWST) archive, and the
+services, including the JWST archive, and the
 project has switched to a continuous release model: every time a change is
@@
 
-\subsubsection{James Webb Space Telescope}
+\subsubsection{JWST}
 % \secauthor{Larry Bradley}
 
-The James Webb Space Telescope (JWST) is a 6.5-meter space-based
+JWST is a 6.5-meter space-based
 infrared telescope that will provide unprecedented resolution and

### 2022-04-19 src/ms.tex
acknowledgements from authorship form

@@
 
+JAA-M acknowledges funding support from Macquarie University through the
+International Macquarie University Research Excellence Scholarship (`iMQRES').
+AB was supported by the Lend\"ulet Program of the Hungarian Academy of Sciences,
+project No. LP2018-7, and the KKP-137523 `SeismoLab' \'Elvonal grant of the
+Hungarian Research, Development and Innovation Office (NKFIH).
+MB gratefully acknowledges support by the ANID BASAL project FB210003 and the
+FONDECYT regular grant 1211000.
+FDE acknowledges funding through the H2020 ERC Consolidator Grant 683184, the
+ERC Advanced grant 695671 ``QUENCH'' and support by the Science and Technology
+Facilities Council (STFC).
+SG made contributions to Astropy a part of the Google Summer of Code 2021.
+SG acknowledges some intellectual support from the CCAD-UNC-NOVA.
+Funding for the Stellar Astrophysics Centre is provided by The Danish National
+Research Foundation (Grant agreement no.: DNRF106).
+NK acknowledges support from the MIT Pappalardo fellowship.
+KAO acknowledges support by the European Research Council (ERC) through Advanced
+Investigator grant to C.S. Frenk, DMIDAS (GA 786910).
+CP is supported by the Canadian Space Agency under a contract with NRC Herzberg
+Astronomy and Astrophysics.
+SP has been supported by Spanish MINECO-FEDER grant RTI2018-096188-B-I00
+JPG acknowledges funding support from Spanish public funds for research from
+project PID2019-107061GB-C63 from the ``Programas Estatales de Generaci\'on de
+Conocimiento y Fortalecimiento Cient\'ifico y Tecnol\'ogico del Sistema de I+D+i
+y de I+D+i Orientada a los Retos de la Sociedad,'' as well as from the State
+Agency for Research of the Spanish MCIU through the ``Center of Excellence
+Severo Ochoa'' award to the Instituto de Astrof\'{\i}sica de Andaluc\'{\i}a
+(SEV-2017-0709).
+NS acknowledges support from the National Science Foundation through the
+Graduate Research Fellowship Program under Grant 1842402. Any opinions,
+findings, and conclusions or recommendations expressed in this material are
+those of the authors and do not necessarily reflect the views of the National
+Science Foundation.
+Supported by the international Gemini Observatory, a program of NSF's NOIRLab,
+which is managed by the Association of Universities for Research in Astronomy
+(AURA) under a cooperative agreement with the National Science Foundation, on
+behalf of the Gemini partnership of Argentina, Brazil, Canada, Chile, the
+Republic of Korea, and the United States of America.
+Parts of this research was supported by the Australian Research Council Centre
+of Excellence for All Sky Astro-physics in 3 Dimensions (ASTRO 3D), through
+project number CE170100013.
+DS is supported by STFC grant ST/S000240/1.
+NS acknowledges the support of the Science and Engineering Research Council of
+Canada (NSERC) Canadian Graduate Scholarship — Doctoral Program, [funding
+reference numbers CGSD — 54721 — 2020].
+
 The \astropy community is supported by and makes use of a number of

### 2022-04-19 src/ms.tex
short title
Fix "to to"

@@
 
-\shorttitle{Astropy Project III}
+\shorttitle{Astropy: Sustaining the Project and v5.0}
 \shortauthors{The Astropy Collaboration}
@@
   to the writing and/or coordination of the paper (in order of contribution),
-  followed by maintainers of and contributors to to the Astropy Project (in
+  followed by maintainers of and contributors to the Astropy Project (in
   alphabetical order). \textbf{The position in the author list does not

### 2022-04-19 src/static/authors.tex
final flag
Unify affiliations for SARAO
They were written with the address line in different order. I've used
the version that matches https://www.sarao.ac.za/contact/

@@
     AZ 85721-0065, USA
+}
+\newcommand{\affsarao}{
+    South African Radio Astronomy Observatory, 2 Fir Street, Black River Park,
+    Observatory 7925, South Africa
 }
@@
 \author{Bruce~Merry}
-\affiliation{South African Radio Astronomy Observatory, 2 Fir Street, Black River Park, Observatory 7925, South Africa}
+\affiliation{\affsarao}
 
@@
 \author{Ludwig~Schwardt}
-\affiliation{South African Radio Astronomy Observatory, Black River Park, 2 Fir Street, Observatory 7925, South Africa}
+\affiliation{\affsarao}
 
@@
 
-%FLAG
 \author[0000-0002-8310-0829]{Jaladh~Singhal}
-\affiliation{TARDIS collaboration}
+% \affiliation{TARDIS collaboration}
+\noaffiliation
 

### 2022-04-19 src/ms.tex
Fix regions package reference

@@
 some image combination operations. The \texttt{regions} package
-\citep{pyregions}, for manipulating ds9-, CASA-, or FITS-style region definitions \citep{ds9},
+\citep{regions}, for manipulating ds9-, CASA-, or FITS-style region definitions \citep{ds9},
 added new ways to manipulate regions, introduced new region types, and obtained

### 2022-04-19 src/static/authors.tex
update Macquarie Univ affiliations
A Department became a School and making other affiliations at Macquarie University consistent.

@@
 \author[0000-0001-6637-6922]{James~Tocknell}
-\affiliation{Department of Physics and Astronomy, Macquarie University, Sydney, NSW 2109, Australia}
+\affiliation{School of Mathematical and Physical Sciences, Macquarie University -- Sydney, NSW 2109, Australia.}
 \affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
@@
 \affiliation{School of Mathematical and Physical Sciences, Macquarie University -- Sydney, NSW 2109, Australia.}
-\affiliation{Centre for Astronomy, Astrophysics and Astrophotonics, Macquarie University -- Sydney, NSW 2109, Australia.}
+\affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
 
@@
 
-\author{Simon~Conseil}
-\affiliation{Univ Lyon, Univ Claude Bernard Lyon 1, CNRS/IN2P3, IP2I Lyon, UMR 5822, F-69622, Villeurbanne, France}
+\author[0000-0002-3657-4191]{Simon~Conseil}
+\affiliation{Aix Marseille Univ, CNRS, CNES, LAM, Marseille, France}
 
@@
 \author[0000-0001-5185-9876]{Lee~R.~Spitler}
-\affiliation{Department of Physics and Astronomy, Macquarie University, Sydney, NSW 2109, Australia}
+\affiliation{School of Mathematical and Physical Sciences, Macquarie University -- Sydney, NSW 2109, Australia.}
+\affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
+\affiliation{Australian Astronomical Optics - Macquarie, Macquarie University, Sydney, NSW 2109, Australia}
 

### 2022-04-20 src/ms.tex

@@
 for funding or maintaining its development, or sustaining it into the future:
-The Project is maintained and coordinated largely by volunteers and contributors
+The Project is maintained and coordinated by both volunteers and contributors
 who are primarily paid for other reasons (i.e., Astropy work is just a means to an end).

### 2022-04-20 src/static/authors.tex
fixes as per discussion

@@
 \author[0000-0001-6637-6922]{James~Tocknell}
-\affiliation{School of Mathematical and Physical Sciences, Macquarie University -- Sydney, NSW 2109, Australia.}
+\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia.}
 \affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
-\affiliation{Australian Astronomical Optics - Macquarie, Macquarie University, Sydney, NSW 2109, Australia}
+\affiliation{Australian Astronomical Optics, Macquarie University, Sydney, NSW 2109, Australia}
 
@@
 \author[0000-0003-0353-9741]{Jaime~A.~Alvarado-Montes}
-\affiliation{School of Mathematical and Physical Sciences, Macquarie University -- Sydney, NSW 2109, Australia.}
+\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia.}
 \affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
@@
 \author[0000-0001-5185-9876]{Lee~R.~Spitler}
-\affiliation{School of Mathematical and Physical Sciences, Macquarie University -- Sydney, NSW 2109, Australia.}
+\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia.}
 \affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
-\affiliation{Australian Astronomical Optics - Macquarie, Macquarie University, Sydney, NSW 2109, Australia}
+\affiliation{Australian Astronomical Optics, Macquarie University, Sydney, NSW 2109, Australia}
 

### 2022-04-20 src/static/authors.tex
Update src/static/authors.tex
Co-authored-by: P. L. Lim <2090236+pllim@users.noreply.github.com>

@@
 \author[0000-0001-6637-6922]{James~Tocknell}
-\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia.}
+\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia}
 \affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
@@
 \author[0000-0003-0353-9741]{Jaime~A.~Alvarado-Montes}
-\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia.}
+\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia}
 \affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}
@@
 \author[0000-0002-1852-9653]{Gabriel~I~Perren}
-\affiliation{Instituto de Astrof\'isica de La Plata (IALP-CONICET), La Plata, Argentina}
+\affiliation{Instituto de Astrof\'isica de La Plata, IALP (CONICET-UNLP), 1900 La Plata, Argentina}
 
@@
 \author[0000-0001-5185-9876]{Lee~R.~Spitler}
-\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia.}
+\affiliation{School of Mathematical and Physical Sciences, Macquarie University, Sydney, NSW 2109, Australia}
 \affiliation{Astronomy, Astrophysics and Astrophotonics Research Centre, Macquarie University, Sydney, NSW 2109, Australia}

### 2022-04-21 src/ms.tex
add pandas and geopandas citations
Font change for 2dfdr

@@
 upstreamed into \package{specutils} loaders for all hosted spectra, including
-2dfdr-reduced spectra from the AAT Archive, and strongly encourages other
+\package{2dfdr}-reduced spectra from the AAT Archive, and strongly encourages other
 archives to provide loaders for their data to further the reuse of existing archival
@@
 \software{\package{astropy} \citep{astropy:2013, astropy:2018},
+          \package{Cython} \citep{cython}.
+          \package{geopandas} \citep{geopandas},
+          \package{matplotlib} \citep{matplotlib},
           \package{numpy} \citep{numpy:nature},
-          \package{scipy} \citep{scipy},
-          \package{matplotlib} \citep{matplotlib},
-          \package{Cython} \citep{cython}.
+          \package{pandas} \citep{pandas, pandas2},
+          \package{scipy} \citep{scipy}.
           % \package{photutils} \citep{photutils},

### 2022-04-22 src/ms.tex
adjust wording about leap seconds

@@
 seconds meaningfully impact results. In past versions of \astropypkg, when using
-\astropyTime, leap seconds had to be manually applied; now, leap
-seconds are applied automatically. Moreover, \astropypkg updates internal time
-data files to ensure that the correct leap second adjustment is always used. For
-reproducibility with older code, \astropyLeapSeconds\ may also be
-manually applied.
+\astropyTime, leap second tables had to be manually updated. Now, \astropypkg
+automatically updates internal time data files to ensure that the correct leap
+second adjustment is always used. For reproducibility with older code,
+\astropyLeapSeconds\ may also be manually applied.
 

### 2022-04-22 src/ms.tex
Add references to body for new affiliated packages

--- 2022-04-21 src/ms.tex
@@
 packages for gravitational astrophysics, including:
-\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
+\texttt{PyCBC} \cite{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} \cite{lenstronomy:1} for
 modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
-gravitational wave probability maps, and \texttt{EinsteinPy} for general
+gravitational wave probability maps, and \texttt{EinsteinPy} \cite{EinsteinPy} for general
 relativity and gravitational astronomy. There have also been several packages
@@
 for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
-Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
-and \texttt{python-cpl} has been added for ESO pipelines and VLT data products.
-For ground-based astronomy, \texttt{baseband} has added IO capabilities for
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
-Other new affiliated packages include \texttt{agnpy} for AGN jets,
-\texttt{statmorph} for fitting galactic morphological diagnostics,
+Coverage maps. \texttt{astroalign}  \cite{astroalign} has been introduced for astrometric registration,
+and \texttt{python-cpl} products \cite{pythoncpl} has been added for ESO pipelines and VLT data.
+For ground-based astronomy, \texttt{baseband} \cite{baseband} has added IO capabilities for
+VLBI, and \texttt{SpectraPy} \cite{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
+Other new affiliated packages include \texttt{agnpy} \cite{agnpy} for AGN jets,
+\texttt{statmorph}  \cite{statmorph}for fitting galactic morphological diagnostics,
 \texttt{dust\_extinction} for modeling interstellar dust extinction,
 \texttt{feets} for extracting features from time series data, and
-\texttt{corral} for managing data intensive parallel pipelines.
+\texttt{corral} \cite{corral} for managing data intensive parallel pipelines.
 \texttt{casaformats-io} provides infrastructure for reading data stored in CASA table formats.
 \texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
-\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
-and \texttt{sbpy} enables calculations for asteroid and
-cometary astrophysics. Finally, \texttt{synphot} provides an interface for
+\texttt{BayesicFitting} \cite{BayesicFitting} provides an interface for generic Bayesian inference,
+and \texttt{sbpy} \cite{sbpy} enables calculations for asteroid and
+cometary astrophysics. Finally, \texttt{synphot} \cite{synphot} provides an interface for
 synthetic photometry.

### 2022-04-22 src/figures/contributor-summary.py
update tom a career stage

@@ -180,7 +180,7 @@ for _date in freeze_dates.datetime:
         _label = False
     else:
         kw = dict()
-    ax.axvline(_date, color='tab:blue', 
+    ax.axvline(_date, color='tab:blue',
                alpha=0.2, zorder=-10, marker='', ls='--', lw=1,
                **kw)
 
@@ -283,7 +283,7 @@ for name, era_lim in eras.items():
     ).split('\n')
     era_shortlogs[name] = _shortlog
     era_n_commits[name] = np.array([int(x.split('\t')[0]) for x in _shortlog])
-    
+
     print(f"era={name} had {len(era_n_commits[name])} committers")
 
 # +
@@ -330,29 +330,29 @@ fig, ax = plt.subplots(figsize=figsize)
 for i, (name, _n_commits) in enumerate(era_n_commits.items()):
     y = np.cumsum(_n_commits) / np.sum(_n_commits)
     x = np.arange(1, len(y)+1)
-    ax.semilogx(x, y, 
+    ax.semilogx(x, y,
                 marker='', color=era_colors[i], lw=2,
                 label=name)
-    
+
     _idx = np.abs(y - 0.9).argmin()
-    
+
     btm = 0.9 - 0.4 - i * 0.15
     ax.plot(
-        [x[_idx]]*2, [btm, 0.9], 
-        marker='', ls='--', 
+        [x[_idx]]*2, [btm, 0.9],
+        marker='', ls='--',
         alpha=0.6, color=era_colors[i]
     )
-    
+
     _val = np.where(y > 0.9)[0][0]
-    ax.text(1.15 * x[_idx], btm, 
+    ax.text(1.15 * x[_idx], btm,
             f'{_val} committers',
             fontsize=16,
             ha='right', va='top',
             color=era_colors[i]
            )
-    
+
     print(name, np.where(y > 0.5)[0][0], np.where(y > 0.9)[0][0])
-    
+
 ax.set_xlim(1, 300)
 ax.set_ylim(0, 1.04)
 ax.legend(
@@ -360,13 +360,13 @@ ax.legend(
     title='Era:', title_fontsize=19
 )
 
-ax.text(1.1, 0.9, '90% of commits', ha='left', va='bottom', 
+ax.text(1.1, 0.9, '90% of commits', ha='left', va='bottom',
         fontsize=16, color='#666666')
 ax.axhline(0.9, zorder=-10, color='#aaaaaa', ls='--')
 
[…]

### 2022-04-22 src/ms.tex
Add missing citations in major updates section

@@
 2D spectra \citep{specutils},  had its first stable release and the addition
-of classes to read JWST data. \texttt{stingray} has had a major performance
+of classes to read JWST data. \texttt{stingray} \citep{stingray:2} has had a major performance
 overhaul. The package \texttt{gammapy} \citep{gammapy} has had major performance

### 2022-04-23 src/figures/python-mentions.py
fortran not all caps

@@ -97,7 +97,7 @@ def get_dfs(cache_path):
             'full:"matlab" property:"refereed" collection:"astronomy"'),
         'Julia': (
             'full:"julia" property:"refereed" collection:"astronomy"'),
-        'FORTRAN': (
+        'Fortran': (
             'full:"fortran" property:"refereed" collection:"astronomy"')
     }
 

### 2022-04-23 src/ms.tex
change cite to citep

--- 2022-04-22 src/ms.tex
@@
 packages for gravitational astrophysics, including:
-\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
+\texttt{PyCBC} \citep{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} \citep{lenstronomy:1} for
 modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
-gravitational wave probability maps, and \texttt{EinsteinPy} for general
+gravitational wave probability maps, and \texttt{EinsteinPy} \citep{EinsteinPy} for general
 relativity and gravitational astronomy. There have also been several packages
@@
 for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
-Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
-and \texttt{python-cpl} has been added for ESO pipelines and VLT data products.
-For ground-based astronomy, \texttt{baseband} has added IO capabilities for
-VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
-Other new affiliated packages include \texttt{agnpy} for AGN jets,
-\texttt{statmorph} for fitting galactic morphological diagnostics,
+Coverage maps. \texttt{astroalign}  \citep{astroalign} has been introduced for astrometric registration,
+and \texttt{python-cpl} products \citep{pythoncpl} has been added for ESO pipelines and VLT data.
+For ground-based astronomy, \texttt{baseband} \citep{baseband} has added IO capabilities for
+VLBI, and \texttt{SpectraPy} \citep{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
+Other new affiliated packages include \texttt{agnpy} \citep{agnpy} for AGN jets,
+\texttt{statmorph}  \citep{statmorph}for fitting galactic morphological diagnostics,
 \texttt{dust\_extinction} for modeling interstellar dust extinction,
 \texttt{feets} for extracting features from time series data, and
-\texttt{corral} for managing data intensive parallel pipelines.
+\texttt{corral} \citep{corral} for managing data intensive parallel pipelines.
 \texttt{casaformats-io} provides infrastructure for reading data stored in CASA table formats.
 \texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
-\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
-and \texttt{sbpy} enables calculations for asteroid and
-cometary astrophysics. Finally, \texttt{synphot} provides an interface for
+\texttt{BayesicFitting} \citep{BayesicFitting} provides an interface for generic Bayesian inference,
+and \texttt{sbpy} \citep{sbpy} enables calculations for asteroid and
+cometary astrophysics. Finally, \texttt{synphot} \citep{synphot} provides an interface for
 synthetic photometry.
@@
 Several of the coordinated and affiliated packages described in
-\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
+\citet{astropy:2018} have had substantial improvements. \texttt{astroquery}
 \citep{astroquery} has added access to roughly a dozen new missions and data

### 2022-04-24 src/ms.tex
rephrase around mixin
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
 
-\paragraph{Table Mixin Columns}
+\paragraph{Table Column Types}
 
 Within \astropypkg and \python, there are numerous ways to represent and store
-array-valued data. Some of the differences are historical:
-\astropyapidoc{table.Column}{table.Column} and
-\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
-differences are computational: \package{numpy}, \package{cupy}, and
-\package{dask} arrays have almost identical APIs, but are optimized for
-different use cases. Lastly, some differences are inherent: \astropyQuantity,
-\astropyTime, and \astropySkyCoord\ represent fundamentally different types of
-objects.
-
-We aim to make it possible for any array-valued data to be used as a column in a
-table. A well-defined protocol for mixin-columns has been developed for
-\astropytable, allowing the original object to be used as a column with a
-familiar API, and to ``round-trip'' through tables with no loss of data or
-attributes. With this protocol, it is now possible to store \astropy native
-objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord\ --
-within a \astropyTable\ and write these to various file formats, such as FITS.
-For objects not already covered by the mixin protocol, functions can be
-registered with \astropyTable\ to convert any array-like object into a mixin
-column. As an example, the mixin functions are used to integrate \package{dask}
-arrays with \astropyTable, allowing cloud-stored or cluster-scale data to be
-used as a column in a \astropyTable.
+array-valued data. Some of the differences are historical: \astropytable and
+\astropyapidoc{io.fits.BinTableHDU}{io.fits.BinTableHDU} are not the same since
+\astropyfits was originally developed in a separate package (\package{PyFITS})
+yet both are for holding tabular data. Some differences are computational:
+\package{numpy}, \package{cupy}, and \package{dask} arrays have almost identical
+APIs, but are optimized for different use cases. Lastly, some differences are
+inherent: \astropyQuantity, \astropyTime, and \astropySkyCoord\ represent
+fundamentally different types of objects.
+
+\astropytable is \astropypkg's most general container for array-valued data.
+Rather than requiring component data be converted to a
+\astropyapidoc{table.Column}{table.Column}, a well-defined protocol has been
+developed for any array-valued data to be used in a \astropytable, without
+having to convert the data. The protocol for ``mixing'' column types provides
+the same familiar table API, yet loses none the object's data or attributes.
+With this protocol, it is now possible to store \astropy native objects --
+including \astropyTime, \astropyQuantity, and \astropySkyCoord\ -- within a
+\astropyTable\ and write these to various file formats, such as FITS. For
+objects not already covered by the ``mixing'' protocol, functions can be
+registered with \astropyTable\ to enable the interoperability. As an example,
+\astropypkg arelady implements the ``mixing'' functions to integrate
+\package{dask} arrays with \astropyTable, allowing cloud-stored or cluster-scale
+data to be used as a column in a \astropyTable.
 
@@
 The FITS standard was extended to rigorously describe time coordinates in the
-WCS framework \citep{FITS-Time:2015}. Compared to
-other types of coordinates in WCS, time requires more metadata: format, scale,
-position, reference, etc. This metadata had to be manually specified and the
-nuances understood by the user. \astropyfits could read this data as a standard
-\astropyFitsColumn\, but would not interpret the time-related metadata and
-attributes. Through the support of the
+WCS framework \citep{FITS-Time:2015}. Compared to other types of coordinates in
+WCS, time requires more metadata: format, scale, position, reference, etc. This
+metadata had to be manually specified and the nuances understood by the user.
+\astropyfits could read this data as a standard \astropyFitsColumn, but would
+not interpret the time-related metadata and attributes. Through the support of
+the
 \href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
-Summer of Code} 2017 program,\footnote{To learn more about this project,
-see the final report
+Summer of Code} 2017 program,\footnote{To learn more about this project, see the
+final report
 \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
 mixin protocol for seamless interoperability}.} the \astropyfits package now
-interprets time data correctly, using \astropyTime\ as a mixin column. For
-backward compatibility, this feature may be turned off.
+interprets time data correctly, using \astropyTime\ as a column. For backward
+compatibility, this feature may be turned off.
 
@@
 
-As mentioned previously, \package{dask} arrays may be used as a mixin column in
+As mentioned previously, \package{dask} arrays may be used as a column in
 \astropyTable. Now, \package{dask} may also be a data array in
-\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only
-computed while written, avoiding excessive memory use. Furthermore, \astropyTable\ can
-be appended to an existing FITS file, where \package{dask} mixin columns can
-interoperate seamlessly between them.
+\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only computed while written,
+avoiding excessive memory use. Furthermore, \astropyTable\ can be appended to an
+existing FITS file, where \package{dask} columns can interoperate seamlessly
+between them.
 

### 2022-04-26 src/ms.tex
explain Astropy’s idiosyncratic mixin definition
Signed-off-by: nstarman <nstarkman@protonmail.com>

@@
 
-Within \astropypkg and \python, there are numerous ways to represent and store
-array-valued data. Some of the differences are historical: \astropytable and
-\astropyapidoc{io.fits.BinTableHDU}{io.fits.BinTableHDU} are not the same since
-\astropyfits was originally developed in a separate package (\package{PyFITS})
-yet both are for holding tabular data. Some differences are computational:
-\package{numpy}, \package{cupy}, and \package{dask} arrays have almost identical
-APIs, but are optimized for different use cases. Lastly, some differences are
-inherent: \astropyQuantity, \astropyTime, and \astropySkyCoord\ represent
-fundamentally different types of objects.
+Within \astropypkg and \python, there are numerous different ways to represent
+and store array-valued data. Some differences are historical: \astropytable and
+\astropyapidoc{io.fits.BinTableHDU}{io.fits.BinTableHDU} are both for holding
+tabular data but the latter was developed for and remains largely specific to
+FITS. Some differences are computational: \package{numpy}, \package{cupy}, and
+\package{dask} arrays have almost identical APIs, but are optimized for
+different use cases. Lastly, some differences are inherent: \astropyQuantity,
+\astropyTime, and \astropySkyCoord\ represent fundamentally different types of
+objects.
 
@@
 developed for any array-valued data to be used in a \astropytable, without
-having to convert the data. The protocol for ``mixing'' column types provides
-the same familiar table API, yet loses none the object's data or attributes.
-With this protocol, it is now possible to store \astropy native objects --
-including \astropyTime, \astropyQuantity, and \astropySkyCoord\ -- within a
-\astropyTable\ and write these to various file formats, such as FITS. For
-objects not already covered by the ``mixing'' protocol, functions can be
-registered with \astropyTable\ to enable the interoperability. As an example,
-\astropypkg arelady implements the ``mixing'' functions to integrate
-\package{dask} arrays with \astropyTable, allowing cloud-stored or cluster-scale
-data to be used as a column in a \astropyTable.
+having to convert the data. The protocol for mixing column types, which astropy
+terms ``mixin''-columns, provides the same familiar table API, yet loses none
+the object's data or attributes. Note ``mixin'' here does not refer to Python's
+inheritance definition. Rather, it is a composition protocol that makes it
+possible to store \astropy native objects -- including \astropyTime,
+\astropyQuantity, and \astropySkyCoord\ -- within a \astropyTable\ and write
+these to various file formats, such as FITS. For objects not already covered by
+the astropy mixin-protocol, functions can be registered with \astropyTable\ to
+enable interoperability. As an example, \astropypkg already implements these
+mixin functions to integrate \package{dask} arrays with \astropyTable, allowing
+cloud-stored or cluster-scale data to be used as a column in a \astropyTable.
 

### 2022-06-23 src/ms.tex
Update src/ms.tex
Co-authored-by: Eero Vaher <eero.vaher@gmail.com>

@@
 developed for any array-valued data to be used in a \astropytable, without
-having to convert the data. The protocol for mixing column types, which astropy
+having to convert the data. The protocol for mixing column types, which \astropypkg
 terms ``mixin''-columns, provides the same familiar table API, yet loses none

### 2022-06-23 src/ms.tex
Update src/ms.tex
Co-authored-by: Derek Homeier <dhomeie@gwdg.de>
Update src/ms.tex
Co-authored-by: Eero Vaher <eero.vaher@gmail.com>

@@
 terms ``mixin''-columns, provides the same familiar table API, yet loses none
-the object's data or attributes. Note ``mixin'' here does not refer to Python's
-inheritance definition. Rather, it is a composition protocol that makes it
+the object's data or attributes. Note that ``mixin'' here is not implemented
+in the commonly used Python scheme of multiple inheritance.
+Rather, it is a composition protocol that makes it
 possible to store \astropy native objects -- including \astropyTime,
@@
 these to various file formats, such as FITS. For objects not already covered by
-the astropy mixin-protocol, functions can be registered with \astropyTable\ to
+the \astropypkg mixin-protocol, functions can be registered with \astropyTable\ to
 enable interoperability. As an example, \astropypkg already implements these

### 2022-06-27 src/figures/contributor-summary.py
nomerges

@@ -80,7 +80,7 @@ era_colors = ['#1b9e77', '#d95f02', '#7570b3']
 commits_per_month = []
 for after_dt, before_dt in zip(month_bins[:-1].datetime,
                                month_bins[1:].datetime):
-    count = repo.git.rev_list(['--count', '--all', '--no-merges',
+    count = repo.git.rev_list(['--count', '--all',
                                f'--after={after_dt:%Y-%m-%d}',
                                f'--before={before_dt:%Y-%m-%d}'])
     commits_per_month.append(int(count))
@@ -125,7 +125,7 @@ committers_per_month = []
 for after_dt, before_dt in zip(month_bins[:-1],
                                month_bins[1:]):
     shortlog = repo.git.shortlog(
-        'HEAD', s=True, n=True, no_merges=True,
+        'HEAD', s=True, n=True,
         after=f"{after_dt.datetime:%Y-%m-%d}",
         before=f"{(before_dt + 1*u.day).datetime:%Y-%m-%d}"
     )
@@ -240,7 +240,7 @@ np.mean(committers_per_month[-20:])
 # ## Number of commits x number of committers
 
 full_shortlog = repo.git.shortlog(
-    'HEAD', s=True, n=True, no_merges=True,
+    'HEAD', s=True, n=True,
     before=f"{(v50_date + 1*u.day).datetime:%Y-%m-%d}"
 ).split('\n')
 full_n_commits = [int(x.split('\t')[0]) for x in full_shortlog]

### 2022-06-27 src/ms.tex
sub- to sub

@@
 \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
-was too specialized was a Virtual Observatory sub-package; the Simple
+was too specialized was a Virtual Observatory subpackage; the Simple
 Application Message Protocol (\astropysubpkg{samp}) is kept in the core package
@@
 Error propagation is difficult, and often is best approximated using Monte Carlo
-methods. The new \astropysubpkg{uncertainties} sub-package is our first step
+methods. The new \astropysubpkg{uncertainties} subpackage is our first step
 towards enabling seamless error propagation. The package allows one to
@@
 require the manipulation of time series. The new \astropysubpkg{timeseries}
-sub-package extends the \astropyQTable\ class to support tables of data as a
+subpackage extends the \astropyQTable\ class to support tables of data as a
 function of time, where the data can either represent samples or averages over
@@
 essence of the World Coordinate System (WCS) concept. From its inception,
-the \astropysubpkg{wcs} sub-package allowed access to WCS information provided
+the \astropysubpkg{wcs} subpackage allowed access to WCS information provided
 in, e.g., FITS files, but it became clear that other WCS standards and
@@
 computing and astronomy, interoperability is an important focus for \astropy,
-both within the various \astropypkg sub-packages and with other \python
+both within the various \astropypkg subpackages and with other \python
 libraries. We aim for a seamless user experience, where different code blocks
@@
 \astropymodeling can now coerce units. A new module – \astropycosmologyunits –
-has been added to the cosmology sub-package for defining and collecting
+has been added to the cosmology subpackage for defining and collecting
 cosmological units and equivalencies. \astropycosmologyunits\ has a unit and
@@
 
-As of \astropypkg v5.0, the I/O registry submodule has been generalized to enable a
+As of \astropypkg v5.0, the I/O registry subpackage has been generalized to enable a
 number of new use cases; for instance, the class-based architecture allows
@@
 Hence, \github pull requests that improve performance are welcomed, and were
-even the main goal for v3.1 release. For most sub-packages, performance was
+even the main goal for v3.1 release. For most subpackages, performance was
 improved in the \python\ code, but for some, the most time-critical pieces were
@@
 (\href{https://github.com/astropy/pytest-arraydiff}{\texttt{pytest-arraydiff}}),
-sub-package command-line testing support
+subpackage command-line testing support
 (\href{https://github.com/astropy/pytest-filter-subpackage}{\texttt{pytest-filter-subpackage}}),
@@
 Tutorials, which are written as Jupyter Notebooks, are now compiled into their
-own Learn Astropy sub-site using
+own Learn Astropy website using
 \href{https://github.com/astropy/nbcollection}{\texttt{nbcollection}}.
